---
title: "Advanced: Customization and tuning"
pkgdown:
  slug: "advanced/customization-and-tuning"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  cache = TRUE,
  cache.path = "legacy-cache/",
  cache.lazy = FALSE
)

if (!require("devtools", quietly = TRUE)) {
  library(devtools)
}
pkg_root <- rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
devtools::load_all(path = pkg_root, quiet = TRUE)

library(DPmixGPD)
library(nimble)
library(ggplot2)
library(kableExtra)

set.seed(1)

use_fast_mcmc <- TRUE
mcmc_long <- list(niter = 2000, nburnin = 500, thin = 3, nchains = 3, seed = 1:3)
mcmc_fast <- list(niter = 600, nburnin = 150, thin = 2, nchains = 1, seed = 1)
mcmc <- if (use_fast_mcmc) mcmc_fast else mcmc_long

# Avoid compiling NIMBLE models during pkgdown builds.
is_pkgdown <- any(grepl("pkgdown|callr", Sys.getenv(), ignore.case = TRUE)) ||
  exists(".__PKGDOWN__", envir = .GlobalEnv, inherits = FALSE) ||
  any(grepl("pkgdown", commandArgs(), ignore.case = TRUE)) ||
  nzchar(Sys.getenv("IN_PKGDOWN", "")) ||
  nzchar(Sys.getenv("PKGDOWN", ""))
if (is_pkgdown) {
  knitr::opts_chunk$set(eval=TRUE)
}

safe_plot <- function(x, ...) {
  obj <- tryCatch(plot(x, ...), error = function(e) e)
  if (inherits(obj, "error")) return(invisible(NULL))
  if (inherits(obj, "ggplot")) {
    try(print(obj), silent = TRUE)
    return(invisible(obj))
  }
  if (is.list(obj)) {
    for (elt in obj) if (inherits(elt, "ggplot")) try(print(elt), silent = TRUE)
  }
  invisible(obj)
}
```

# Overview

This vignette documents how to construct and run NIMBLE models using the low-level building blocks exported by DPmixGPD. The intended audience is advanced users who want full control of the model specification while still reusing the package's NIMBLE-ready distributions (mixture kernels and bulk-tail splices) and the standard NIMBLE compilation and MCMC workflow.

The examples emphasize transparency. Each model is specified in `nimbleCode{}`; constants, data, and initial values are stated explicitly; and the model is compiled and sampled using the canonical NIMBLE sequence.

The vignette covers:

1) A manual stick-breaking (SB) Normal mixture using `dNormMix` as the likelihood.

2) A spliced bulk-tail model (SB + Amoroso + GPD) in which DPmixGPD generates the model code, while the compilation and sampling steps are carried out explicitly to facilitate inspection.

3) A causal (two-arm) model built via `build_causal_bundle()`, followed by downstream summaries via S3 methods such as `predict()`, `ate()/qte()` (marginal), `att()/qtt()` (treated-only), and `cate()/cqte()` (conditional).

Throughout, the focus is on reproducible model construction and on using the exported NIMBLE distributions directly in the likelihood. In this setting, a ones/zeros trick is not required.

# Data

```{r}
data("mtcars", package = "datasets")

df <- mtcars
y <- df$mpg
X <- df[, c("wt", "hp")]
X <- as.data.frame(X)
```

# SB Case: Manual Stick-Breaking Model with `dNormMix`

In this first example we write an SB mixture model by hand and use the exported
`dNormMix()` nimbleFunction as the per-observation log-likelihood. The point is to show
that DPmixGPD kernels are ordinary NIMBLE components: you can call them inside your own
model code, monitor whatever nodes you like, and control the MCMC configuration exactly
as you would in a standalone NIMBLE workflow.

## Note on custom likelihoods

NIMBLE can work with user-defined distributions, but for a vignette it is often easier to
avoid registration and keep everything "plain" model code. DPmixGPD already exposes the mixture and MixGPD building blocks as NIMBLE distributions (e.g., `dNormMix` for a Normal mixture and `dNormMixGpd` for the spliced bulk+GPD version). That means you can write the likelihood directly in `nimbleCode` using `y[i] ~ dNormMix(...)` or `y[i] ~ dNormMixGpd(...)`--no ones/zeros trick is needed here.

## Constants, data, and inits

```{r}
N <- length(y)
K <- 3

constants_sb <- list(N = N, K = K)
data_sb <- list(y = y)

inits_sb <- function() {
  list(
    alpha = 1,
    v = c(rep(0.5, K - 1), 1),
    mu = rnorm(K, mean(y), 1),
    sigma = rep(sd(y), K)
  )
}
```

## NIMBLE code

```{r}
code_sb <- nimble::nimbleCode({
  alpha ~ dgamma(1, 1)

  for (k in 1:(K - 1)) {
    v[k] ~ dbeta(1, alpha)
  }
  v[K] <- 1

  w[1] <- v[1]
  for (k in 2:K) {
    w[k] <- v[k] * prod(1 - v[1:(k - 1)])
  }

  for (k in 1:K) {
    mu[k] ~ dnorm(0, sd = 5)
    sigma[k] ~ dunif(0.1, 5)
  }

  for (i in 1:N) {
    y[i] ~ dNormMix(w = w[1:K], mean = mu[1:K], sd = sigma[1:K])
  }
})
```

## Compile and run (manual)

```{r, cache=TRUE, results='hide'}
Rmodel_sb <- nimble::nimbleModel(
  code = code_sb,
  constants = constants_sb,
  data = data_sb,
  inits = inits_sb()
)

conf_sb <- nimble::configureMCMC(
  Rmodel_sb,
  monitors = c("alpha", "v", "w", "mu", "sigma")
)

Rmcmc_sb <- nimble::buildMCMC(conf_sb)
Cmodel_sb <- nimble::compileNimble(Rmodel_sb, showCompilerOutput = FALSE)
Cmcmc_sb <- nimble::compileNimble(Rmcmc_sb, project = Rmodel_sb, showCompilerOutput = FALSE)

samples_sb <- nimble::runMCMC(
  Cmcmc_sb,
  niter = mcmc$niter,
  nburnin = mcmc$nburnin,
  thin = mcmc$thin,
  nchains = mcmc$nchains,
  setSeed = mcmc$seed
)
```

## Sample extraction

```{r}
head(samples_sb[, 1:min(6, ncol(samples_sb))])
summary(samples_sb[, c("alpha", "mu[1]", "mu[2]", "mu[3]")])
```

## Optional: S3 methods via the DPmixGPD bundle

For comparison, we build a **bundle** for the same SB Normal mixture and use
DPmixGPD's S3 methods for posterior inference.

```{r, cache=TRUE, results='hide'}
bundle_sb <- build_nimble_bundle(
  y = y,
  X = NULL,
  backend = "sb",
  kernel = "normal",
  GPD = FALSE,
  components = K,
  mcmc = mcmc
)

fit_sb <- run_mcmc_bundle_manual(bundle_sb, show_progress = FALSE)
```

```{r, results='asis'}
print(bundle_sb)
summary(bundle_sb)
```

```{r, results='asis'}
print(fit_sb)
summary(fit_sb)
```

```{r}
safe_plot(fit_sb, family = "traceplot")
```

```{r}
pred_sb <- predict(fit_sb, type = "mean", interval = "credible")
head(pred_sb$fit)
```

# SB + GPD Case: Manual Pipeline with Codegen

The second example adds a GPD tail. Rather than writing the full spliced model by hand,
we let `build_nimble_bundle()` generate the NIMBLE code and the supporting objects
(constants, data, inits, monitors). The "customization" part is that we then treat those
outputs as if you had written them yourself: we construct the `nimbleModel`, configure and
compile the MCMC, run it, and inspect the samples. This is a good pattern when you want a
reproducible code generator, but still need low-level access for debugging or extensions.

## Specification

```{r}
param_specs_crp <- list(
  bulk = list(
    loc = list(
      mode = "dist",
      dist = "normal",
      args = list(mean = 0, sd = 2)
    ),
    scale = list(
      mode = "dist",
      dist = "gamma",
      args = list(shape = 2, rate = 1)
    ),
    shape1 = list(mode = "fixed", value = 1)
  ),
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "gamma",
      args = list(shape = 2, rate = 1)
    ),
    tail_scale = list(
      mode = "dist",
      dist = "gamma",
      args = list(shape = 2, rate = 1)
    )
  )
)
```

## Build bundle and extract pieces

```{r}
X_crp <- NULL
bundle_crp <- build_nimble_bundle(
  y = y,
  X = X_crp,
  backend = "sb",
  kernel = "amoroso",
  GPD = TRUE,
  components = 5,
  param_specs = param_specs_crp,
  mcmc = mcmc
)

code_crp <- bundle_crp$code$nimble
constants_crp <- bundle_crp$constants
data_crp <- bundle_crp$data
inits_crp <- if (is.function(bundle_crp$inits_fun)) bundle_crp$inits_fun else bundle_crp$inits
```

## Compile and run (manual)

```{r, cache=TRUE, results='hide'}
Rmodel_crp <- nimble::nimbleModel(
  code = code_crp,
  constants = constants_crp,
  data = data_crp,
  inits = if (is.function(inits_crp)) inits_crp() else inits_crp
)

conf_crp <- nimble::configureMCMC(
  Rmodel_crp,
  monitors = bundle_crp$monitors
)

if (!is.null(conf_crp$samplerConfs) && length(conf_crp$samplerConfs) > 0) {
  for (i in seq_along(conf_crp$samplerConfs)) {
    ctl <- conf_crp$samplerConfs[[i]]$control
    if (is.null(ctl)) ctl <- list()
    if (is.null(ctl$checkConjugacy)) ctl$checkConjugacy <- FALSE
    conf_crp$samplerConfs[[i]]$control <- ctl
  }
}

Rmcmc_crp <- nimble::buildMCMC(conf_crp)
Cmodel_crp <- nimble::compileNimble(Rmodel_crp, showCompilerOutput = FALSE)
Cmcmc_crp <- nimble::compileNimble(Rmcmc_crp, project = Rmodel_crp, showCompilerOutput = FALSE)

samples_crp <- nimble::runMCMC(
  Cmcmc_crp,
  niter = mcmc$niter,
  nburnin = mcmc$nburnin,
  thin = mcmc$thin,
  nchains = mcmc$nchains,
  setSeed = mcmc$seed
)
```

## Sample extraction

```{r}
head(samples_crp[, 1:min(6, ncol(samples_crp))])
```

## S3 methods for posterior inference

```{r, cache=TRUE, results='hide'}
fit_crp <- run_mcmc_bundle_manual(bundle_crp, show_progress = FALSE)
```

```{r, results='asis'}
print(fit_crp)
summary(fit_crp)
```

```{r}
safe_plot(fit_crp, family = "traceplot")
```

```{r}
pred_crp_mean <- predict(fit_crp, type = "mean", interval = "credible")
pred_crp_q90 <- predict(fit_crp, type = "quantile", index = 0.90, interval = "credible")
head(pred_crp_mean$fit)
head(pred_crp_q90$fit)
```

# Causal Model: From Bundle to S3 Inference

Finally, we switch from "one distribution for everyone" to a two-arm causal setup.
DPmixGPD's causal bundle wraps (i) an outcome model for the treated arm, (ii) an outcome
model for the control arm, and (iii) a propensity-score model (optional, but often useful).
Once the sampler has run, the fitted object exposes high-level S3 methods so you can move
from posterior samples to causal estimands with minimal friction--e.g., average treatment
effects via `ate()` and quantile treatment effects via `qte()`, with treated-only counterparts `att()` / `qtt()` and conditional counterparts `cate()` / `cqte()`.

```{r}
X_causal <- df[, c("wt", "hp", "qsec", "cyl")]
X_causal <- as.data.frame(X_causal)
T_ind <- df$am

param_specs_causal <- list(
  trt = list(
    bulk = list(
      mean = list(
        mode = "link",
        link = "identity",
        beta_prior = list(dist = "normal", args = list(mean = 0, sd = 2))
      ),
      sd = list(
        mode = "dist",
        dist = "invgamma",
        args = list(shape = 2, scale = 1)
      )
    )
  ),
  con = list(
    bulk = list(
      location = list(
        mode = "link",
        link = "identity",
        beta_prior = list(dist = "normal", args = list(mean = 0, sd = 2))
      ),
      scale = list(
        mode = "dist",
        dist = "gamma",
        args = list(shape = 2, rate = 1)
      )
    )
  )
)
```

```{r, cache=TRUE, results='hide'}
causal_bundle <- build_causal_bundle(
  y = y,
  X = X_causal,
  T = T_ind,
  backend = "sb",
  kernel = c("normal", "laplace"),
  GPD = FALSE,
  components = 5,
  param_specs = param_specs_causal,
  PS = "logit",
  mcmc_outcome = mcmc,
  mcmc_ps = mcmc
)

causal_fit <- run_mcmc_causal(causal_bundle, show_progress = FALSE)
```

```{r, results='asis'}
print(causal_bundle)
summary(causal_bundle)
```

```{r, results='asis'}
print(causal_fit)
summary(causal_fit)
```

```{r}
safe_plot(causal_fit, family = "traceplot")
```

```{r}
pred_causal_mean <- predict(causal_fit, x = X_causal, type = "mean", interval = "credible")
head(pred_causal_mean)
```

```{r, results='asis'}
ate_result <- ate(causal_fit, interval = "credible", nsim_mean = 50)
qte_result <- qte(causal_fit, probs = c(0.25, 0.5, 0.75), interval = "credible")
att_result <- att(causal_fit, interval = "credible", nsim_mean = 50)
qtt_result <- qtt(causal_fit, probs = c(0.25, 0.5, 0.75), interval = "credible")
cate_result <- cate(causal_fit, newdata = X_causal[1:5, ], interval = "credible", nsim_mean = 50)
cqte_result <- cqte(causal_fit, probs = c(0.25, 0.5, 0.75), newdata = X_causal[1:5, ], interval = "credible")

print(ate_result)
summary(ate_result)

safe_plot(ate_result, type = "effect")
safe_plot(qte_result, type = "effect")
```

```{r sessioninfo, include=FALSE}
sessionInfo()
```


