---
title: "Advanced: GPD-in-DPM architecture (CRP-only)"
description: "Advanced user vignette. CRP-only prototype for a Dirichlet process mixture in which the cluster atom is a scalar kernel parameter and the observation model uses the exported *GPD-spliced* kernel distributions."
keywords: [advanced, CRP, DPM, GPD, nimble]
pkgdown:
  slug: "advanced/gpd-in-dpm-architecture"
format:
  html:
    toc: true
    number-sections: true
execute:
  enabled: true
---

::: callout-warning
## Advanced users only
This vignette is a code-oriented prototype. It assumes familiarity with NIMBLE model code and with validating custom Bayesian nonparametric models.
:::

# Objective

The package currently provides bulk-only mixtures (`d*Mix`) and bulk-tail spliced distributions (`d*Gpd`, `d*MixGpd`) as NIMBLE `nimbleFunction`s. However, it does not yet expose a "scalar-atom DPM with an embedded GPD tail" as a one-line high-level interface.

The goal here is to show how to write that model directly in `nimbleCode{}` for the **CRP backend**, while reusing the **exported** spliced distributions (e.g., `dGammaGpd`, `dLognormalGpd`, `dInvGaussGpd`, `dNormGpd`, etc.). No ones/zeros trick is required because the likelihood is written using these distributions directly.

# Model specification

Let \(y_1,\dots,y_N\) be observations. Under a CRP DPM:

- Cluster allocations follow a Chinese restaurant process:
  \[
    z[1{:}N] \sim \mathrm{CRP}(\alpha),
  \]
  implemented in NIMBLE as `z[1:N] ~ dCRP(conc = alpha, size = N)`.

- Each cluster \(k\) has a *scalar* atom \(\theta_k\). The atom is the only cluster-specific kernel parameter; all other kernel parameters are fixed or global.

- Conditional on allocation, the observation likelihood uses a **GPD-spliced kernel distribution** exported by the package. For a positive-support outcome, a convenient choice is Gamma+GPD:
  \[
    y_i \mid z_i=k \sim \mathrm{GammaGpd}(\text{shape}=a,\ \text{scale}=\theta_k,\ u,\ \sigma_{\text{tail}},\ \xi),
  \]
  where the Gamma is used below the threshold \(u\), and exceedances use the GPD tail via `dGpd` inside `dGammaGpd`.

This construction is "GPD-in-DPM" in the sense that the **tail is part of the per-observation distribution** (not a separate post-processing step), while the mixture is induced by the CRP allocations.

# A concrete working template (Gamma bulk, scalar scale atom)

The template below uses:

- scalar atom: `scale_k = theta[k] > 0`
- fixed bulk shape: `shape0` (constant)
- tail parameters: `threshold`, `tail_scale`, `tail_shape` (constants or global parameters)

You can substitute a different exported spliced distribution by changing the likelihood line (for example, `dLognormalGpd`, `dInvGaussGpd`, `dNormGpd`, etc.).

```{r}
library(CausalMixGPD)
library(nimble)

# Data shipped with the package (positive-support, tail-designed)
data("nc_pos_tail200_k4")
y <- nc_pos_tail200_k4$y
N <- length(y)

# Use the dataset's generating tail settings (keeps the example coherent)
threshold  <- nc_pos_tail200_k4$truth$threshold
tail_scale <- nc_pos_tail200_k4$truth$tail_params$scale
tail_shape <- nc_pos_tail200_k4$truth$tail_params$shape


# Choose a truncation level for the finite representation of clusters
components <- 25

# Fixed bulk shape (example)
shape0 <- 2.0


code <- nimbleCode({
  # Concentration
  alpha ~ dgamma(1, 1)

  # Scalar atoms (cluster-specific). Here: positive scale parameters.
  for (k in 1:components) {
    theta[k] ~ dgamma(a0, b0)
  }

  # CRP memberships (native NIMBLE BNP distribution)
  z[1:N] ~ dCRP(conc = alpha, size = N)

  # Likelihood: exported spliced kernel distribution
  for (i in 1:N) {
    y[i] ~ dGammaGpd(shape = shape0,
                     scale = theta[z[i]],
                     threshold = threshold,
                     tail_scale = tail_scale,
                     tail_shape = tail_shape)
  }
})

constants <- list(
  N = N, components = components,
  a0 = 2, b0 = 2,
  shape0 = shape0,
  threshold = threshold,
  tail_scale = tail_scale,
  tail_shape = tail_shape
)

data <- list(y = y)

inits <- list(
  alpha = 1,
  theta = rgamma(components, 2, 2),
  z = sample.int(5, N, replace = TRUE)
)

m <- nimbleModel(code, constants = constants, data = data, inits = inits, check = TRUE)
cm <- compileNimble(m)

conf <- configureMCMC(m, monitors = c("alpha", "theta", "z"))
mcmc <- buildMCMC(conf)
cmcmc <- compileNimble(mcmc, project = m)

samps <- runMCMC(cmcmc, niter = 600, nburnin = 300, thin = 2, setSeed = 1)
```


# Posterior analysis

This section sketches a minimal posterior workflow after `runMCMC()` returns `samps`. The object returned by `runMCMC()` is typically a matrix with one row per retained iteration and one column per monitored node.

## Basic parameter summaries

```{r}
# Posterior summary for alpha
alpha_draws <- samps[, "alpha"]
summary(alpha_draws)

# Posterior summaries for the scalar atoms theta[k]
theta_cols  <- grep("^theta\\[", colnames(samps), value = TRUE)
theta_draws <- samps[, theta_cols, drop = FALSE]

theta_summary <- t(apply(theta_draws, 2, function(x) {
  c(mean = mean(x), sd = sd(x),
    q025 = unname(quantile(x, 0.025)),
    q50  = unname(quantile(x, 0.50)),
    q975 = unname(quantile(x, 0.975)))
}))

kableExtra::kbl(theta_summary, digits = 3, caption = "Posterior summaries of scalar atoms (theta[k])") |>
  kableExtra::kable_styling(full_width = FALSE)
```

## Posterior clustering diagnostics

A practical summary of the induced partition is the posterior distribution of the number of occupied clusters,
\[
  K^{(m)} = \left|\left\{ z^{(m)}_1,\dots,z^{(m)}_N \right\}\right|,
\]
computed for each retained iteration \(m\).

```{r}
z_cols  <- grep("^z\\[", colnames(samps), value = TRUE)
z_draws <- samps[, z_cols, drop = FALSE]

K_occ <- apply(z_draws, 1, function(zrow) length(unique(as.integer(zrow))))

summary(K_occ)

dfK <- data.frame(K = K_occ)
ggplot2::ggplot(dfK, ggplot2::aes(x = K)) +
  ggplot2::geom_histogram(bins = 30) +
  ggplot2::labs(x = "Number of occupied clusters", y = "Frequency")
```

If you monitor `z`, this can be memory-heavy for large \(N\). For posterior diagnostics only, consider monitoring `z` for a subsample of observations, or derive `K_occ` inside NIMBLE using deterministic nodes (advanced).

## Posterior predictive checks

A simple posterior predictive check is to generate replicated outcomes \(y^{rep}\) from the fitted model and compare distributional features.

For the CRP mixture, one convenient approximation is to sample a cluster label for a new observation using empirical cluster frequencies from the current iteration (ignoring the "new cluster" probability). This yields a practical in-sample predictive check and is typically sufficient for early debugging.

```{r}
# Extract theta and z at each iteration
theta_mat <- theta_draws
z_mat     <- z_draws

# One posterior predictive draw per iteration (increase B if needed)
M <- nrow(samps)
y_rep <- numeric(M)

for (m in 1:M) {
  z_m <- as.integer(z_mat[m, ])
  # empirical cluster weights
  tab <- table(z_m)
  labs <- as.integer(names(tab))
  w <- as.numeric(tab) / sum(tab)

  k_new <- sample(labs, size = 1, prob = w)
  scale_new <- theta_mat[m, sprintf("theta[%d]", k_new)]

  y_rep[m] <- rGammaGpd(
    1,
    shape = shape0,
    scale = scale_new,
    threshold = threshold,
    tail_scale = tail_scale,
    tail_shape = tail_shape
  )
}

# Compare observed vs replicated summaries
quantile(y, probs = c(0.5, 0.9, 0.95, 0.99))
quantile(y_rep, probs = c(0.5, 0.9, 0.95, 0.99))

dfppc <- data.frame(value = c(y, y_rep),
                    which = rep(c("observed", "replicated"), c(length(y), length(y_rep))))

ggplot2::ggplot(dfppc, ggplot2::aes(x = value, colour = which)) +
  ggplot2::geom_density() +
  ggplot2::labs(x = "y", y = "Density")
```

## Tail summaries from the posterior

Two basic tail summaries are:

- exceedance probability \( \Pr(Y > u) \),
- extreme quantiles (e.g., \(q_{0.99}\)).

With posterior predictive draws `y_rep`, you can estimate both directly:

```{r}
mean(y_rep > threshold)                      # posterior predictive exceedance rate
quantile(y_rep, probs = c(0.95, 0.99, 0.995))
```

For more precise quantiles (and to reduce Monte Carlo error), increase the number of predictive draws per iteration or compute the predictive CDF as a mixture of `p*Gpd` terms and solve for the quantile (more expensive; recommended only after the baseline model is stable).

# How to adapt this template

1. Select the appropriate spliced distribution for the support of your outcome:

- Positive support: `dGammaGpd`, `dLognormalGpd`, `dInvGaussGpd`, `dAmorosoGpd`, etc.
- Real line: `dNormGpd` or a suitable real-line kernel if available.

2. Identify a single kernel parameter to serve as the scalar atom \(\theta_k\), and fix or globalize the remaining kernel parameters.

3. Decide whether `threshold`, `tail_scale`, and `tail_shape` are fixed constants, global parameters with priors, or covariate-linked (the last option is more involved and should be introduced only after the baseline model is stable).

# Minimal validation requirements

Before applying the model to real data:

- Perform simulation recovery under the same kernel and tail settings.
- Check posterior predictive exceedance rates above `threshold`.
- Check sensitivity to the chosen `threshold` and priors on `alpha` and tail parameters.












