---
title: "Workflow 0: Start Here"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\\VignetteIndexEntry{Workflow 0: Start Here}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'cookbook-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("COOKBOOK_FAST"))) as.logical(Sys.getenv("COOKBOOK_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(DPmixGPD)

# Precomputed results for fast website builds
source("_cookbook-setup.R")
```


# Theory (brief)

The DP mixture models the bulk density as $f(y)=\\int K(y;\\theta)\\,dG(\\theta)$
with $G \\sim \\mathrm{DP}(\\alpha, G_0)$. This gives a flexible, data-driven
mixture representation while keeping inference in a Bayesian framework.

# Getting Started

This vignette gives a minimal, fully working workflow for an **unconditional** model and a **conditional** model. Everything uses short MCMC runs so the vignette renders quickly.

---

## Unconditional Model (CRP, bulk-only)

```{r start-uncond-data}
data("nc_pos200_k3")
y <- nc_pos200_k3$y
```

```{r start-uncond-bundle}
bundle_uncond <- build_nimble_bundle(
  y = y,
  backend = "crp",
  kernel = "gamma",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)
```

```{r start-uncond-fit}
fit_uncond <- load_or_fit("v00-start-here-fit_uncond", quiet_mcmc(run_mcmc_bundle_manual(bundle_uncond, show_progress = FALSE)))
summary(fit_uncond)
```

```{r start-uncond-predict, cache=FALSE}
pred_q <- predict(fit_uncond, type = "quantile", index = c(0.5, 0.9), interval = "credible")
head(pred_q$fit)
plot(pred_q)
```

---

## Conditional Model (SB, bulk-only)

```{r start-cond-data}
data("nc_posX100_p3_k2")
yc <- nc_posX100_p3_k2$y
X <- as.matrix(nc_posX100_p3_k2$X)
```

```{r start-cond-bundle}
bundle_cond <- build_nimble_bundle(
  y = yc,
  X = X,
  backend = "sb",
  kernel = "lognormal",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)
```

```{r start-cond-fit}
fit_cond <- load_or_fit("v00-start-here-fit_cond", quiet_mcmc(run_mcmc_bundle_manual(bundle_cond, show_progress = FALSE)))
summary(fit_cond)
```

```{r start-cond-predict, cache=FALSE}
x_new <- X[1:20, , drop = FALSE]
pred_mean <- predict(fit_cond, x = x_new, type = "mean", interval = "credible", nsim_mean = 200)
head(pred_mean$fit)
plot(pred_mean)
```

---

## Useful S3 Methods

```{r start-s3}
params(fit_uncond)
plot(fit_uncond, params = "shape", family = "traceplot")
```

---

## Next Steps

- `vignette 1`: package overview and terminology
- `vignette 5`: full three-phase workflow (spec ? bundle ? MCMC)
- `vignette 6-13`: unconditional/conditional models with and without GPD
- `vignette 14-19`: causal workflows



