---
title: "13. Conditional DPmixGPD with Stick-Breaking Backend"
output:
  rmarkdown::html_vignette:
    highlight: null
    code_folding: show
vignette: >
  %\VignetteIndexEntry{13. Conditional DPmixGPD with Stick-Breaking Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

source("_load_pkg.R")

# Precomputed results for fast website builds
source("_legacy-setup.R")
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).


# Conditional DPmixGPD: Stick-Breaking Backend

**Purpose**: Apply fixed-component stick-breaking truncation to covariate-dependent mixtures while keeping the GPD tail. This vignette mirrors `v10` but with the SB backend.

---

## Data Setup

```{r data-setup}
data("nc_posX100_p5_k4")
y <- nc_posX100_p5_k4$y
X <- as.matrix(nc_posX100_p5_k4$X)
if (is.null(colnames(X))) {
  colnames(X) <- paste0("x", seq_len(ncol(X)))
}

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

ggplot(data.frame(y = y, x1 = X[, 1]), aes(x = x1, y = y)) +
  geom_point(alpha = 0.6, color = "purple") +
  geom_smooth(method = "loess", color = "orange", fill = NA) +
  labs(title = "Tail Outcome vs x1 (SB)", x = "x1", y = "y") +
  theme_minimal()
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Conditional Tail Summary (SB)", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

---

## Threshold

```{r threshold}
u_threshold <- quantile(y, 0.85)

ggplot(data.frame(y = y), aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "lightgreen", alpha = 0.6, color = "black") +
  geom_vline(xintercept = u_threshold, linetype = "dashed", color = "black") +
  labs(title = "Threshold for SB tail (85%)", x = "y", y = "Density") +
  theme_minimal()
```

---

## Model Specification

```{r bundle, cache=FALSE}
bundle_sb_cond_gpd_gamma <- build_nimble_bundle(
  y = y,
  X = X,
  kernel = "gamma",
  backend = "sb",
  GPD = TRUE,
  components = 5,
  param_specs = list(
    gpd = list(
      threshold = list(mode = "link", link = "exp")
    )
  ),
  mcmc = mcmc
)

bundle_sb_cond_gpd_laplace <- build_nimble_bundle(
  y = y,
  X = X,
  kernel = "laplace",
  backend = "sb",
  GPD = TRUE,
  components = 5,
  param_specs = list(
    gpd = list(
      threshold = list(mode = "link", link = "exp")
    )
  ),
  mcmc = mcmc
)
```

---

## MCMC Execution

```{r mcmc, cache=FALSE}
fit_sb_cond_gpd_gamma <- load_or_fit("v13-conditional-DPmixGPD-SB-fit_sb_cond_gpd_gamma", run_mcmc_bundle_manual(bundle_sb_cond_gpd_gamma))
fit_sb_cond_gpd_laplace <- load_or_fit("v13-conditional-DPmixGPD-SB-fit_sb_cond_gpd_laplace", run_mcmc_bundle_manual(bundle_sb_cond_gpd_laplace))
summary(fit_sb_cond_gpd_gamma)
summary(fit_sb_cond_gpd_laplace)
```

```{r params-cond-gpd-sb}
params_sb_cond <- params(fit_sb_cond_gpd_gamma)
params_sb_cond
```

---

## Conditional Predictions

```{r cond-predict}
X_new <- rbind(
  c(-1, 0, 0, 0, 0),
  c(0, 0, 0, 0, 0),
  c(1, 1, 0, 0, 0)
)
colnames(X_new) <- colnames(X)
y_grid <- seq(0, max(y) * 1.2, length.out = 200)

df_pred_gamma <- lapply(seq_len(nrow(X_new)), function(i) {
  pred <- predict(fit_sb_cond_gpd_gamma, x = as.matrix(X_new[i, , drop = FALSE]), y = y_grid, type = "density")
  data.frame(
    y = pred$fit$y,
    density = pred$fit$density,
    label = paste("x1=", X_new[i, 1], ", x2=", X_new[i, 2], sep = ""),
    model = "Gamma"
  )
})

df_pred_laplace <- lapply(seq_len(nrow(X_new)), function(i) {
  pred <- predict(fit_sb_cond_gpd_laplace, x = as.matrix(X_new[i, , drop = FALSE]), y = y_grid, type = "density")
  data.frame(
    y = pred$fit$y,
    density = pred$fit$density,
    label = paste("x1=", X_new[i, 1], ", x2=", X_new[i, 2], sep = ""),
    model = "Laplace"
  )
})

bind_rows(df_pred_gamma, df_pred_laplace) %>%
  ggplot(aes(x = y, y = density, color = label)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ model) +
  labs(title = "Conditional Density (SB + GPD)", x = "y", y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Tail Quantiles

```{r quantiles}
X_grid <- cbind(x1 = seq(-1, 1, length.out = 5), x2 = 0, x3 = 0, x4 = 0, x5 = 0)
colnames(X_grid) <- colnames(X)
quant_probs <- c(0.90, 0.95)

pred_q_gamma <- predict(fit_sb_cond_gpd_gamma, x = as.matrix(X_grid), type = "quantile", index = quant_probs)
pred_q_laplace <- predict(fit_sb_cond_gpd_laplace, x = as.matrix(X_grid), type = "quantile", index = quant_probs)

quant_df_gamma <- pred_q_gamma$fit
quant_df_gamma$x1 <- X_grid[quant_df_gamma$id, "x1"]
quant_df_gamma$model <- "Gamma"

quant_df_laplace <- pred_q_laplace$fit
quant_df_laplace$x1 <- X_grid[quant_df_laplace$id, "x1"]
quant_df_laplace$model <- "Laplace"

bind_rows(quant_df_gamma, quant_df_laplace) %>%
  ggplot(aes(x = x1, y = estimate, color = factor(index), group = index)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ model) +
  labs(title = "Tail Quantiles vs x1 (SB)", x = "x1", y = "Quantile", color = "Probability") +
  theme_minimal()
```

---

## Residuals & Diagnostics

```{r residuals, cache=FALSE}
if (interactive()) plot(fitted(fit_sb_cond_gpd_gamma))
```

```{r diagnostics}
if (interactive()) plot(fit_sb_cond_gpd_gamma, family = c("traceplot", "autocorrelation", "geweke"))
if (interactive()) plot(fit_sb_cond_gpd_laplace, family = c("density", "running", "caterpillar"))
```

---

## Takeaways

- Conditional stick-breaking mixtures capture covariate-dependent bulk structure while the GPD handles extremes.
- `predict()` and `if (interactive()) plot()` remain consistent for densities, posterior-mean quantiles, and residuals.
- Expect threshold-selected posterior-mean tail quantiles to shift with `x1` even when `components` is fixed.
- Next: move into causal models starting with same-backend CRP (v12).









