---
title: "19. Causal Inference: Mixed Backends with GPD - Amoroso Kernel"
output:
  rmarkdown::html_vignette:
    highlight: null
    code_folding: show
vignette: >
  %\VignetteIndexEntry{19. Causal Inference: Mixed Backends with GPD - Amoroso Kernel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

pkgload::load_all()
source("_load_pkg.R")

# Precomputed results for fast website builds
source("_legacy-setup.R")
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).


# Causal Inference: Mixed Backends with GPD - Amoroso Kernel {#causal-different-backends-gpd}

This vignette fits two mixed-backend causal models with GPD tails and an Amoroso kernel:

- Model A: Treated = SB, Control = CRP
- Model B: Treated = CRP, Control = SB

We shift y to keep outcomes in a stable positive range.

---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2")
y_raw <- causal_alt_real500_p4_k2$y
T <- causal_alt_real500_p4_k2$T
X <- as.matrix(causal_alt_real500_p4_k2$X)

shift <- abs(min(y_raw)) + 0.1
y <- y_raw + shift

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Shifted Outcome Summary (Amoroso)", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")

x_eval <- X[1:40, , drop = FALSE]
y_eval <- y[1:40]
u_threshold <- as.numeric(stats::quantile(y, 0.8, names = FALSE))
```

```{r data-scatter}
df_causal <- data.frame(y = y, T = as.factor(T), x1 = X[, 1])

p_scatter <- ggplot(df_causal, aes(x = x1, y = y, color = T)) +
  geom_point(alpha = 0.5) +
  labs(title = "Shifted Outcome vs X1", x = "X1", y = "y (shifted)", color = "Treatment") +
  theme_minimal()

p_scatter
```

---

## Model A: Treated SB, Control CRP (Amoroso + GPD)

```{r bundle-sb-crp, cache=FALSE}
param_specs_gpd <- list(
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "lognormal",
      args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
    )
  )
)

bundle_sb_crp <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = c("amoroso", "amoroso"),
  backend = c("sb", "crp"),
  PS = "logit",
  GPD = c(TRUE, TRUE),
  components = c(5, 5),
  param_specs = param_specs_gpd,
  mcmc_outcome = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 1),
  mcmc_ps = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 1)
)

bundle_sb_crp
summary(bundle_sb_crp)
```

```{r fit-sb-crp, cache=FALSE}
fit_sb_crp <- run_mcmc_causal(bundle_sb_crp)
summary(fit_sb_crp)
params(fit_sb_crp)
```

```{r plot-fit-sb-crp}
if (interactive()) plot(fit_sb_crp, family = c("traceplot", "autocorrelation", "running"))
```

```{r predict-mean-sb-crp}
pred_mean_sb_crp <- predict(fit_sb_crp, x = x_eval, type = "mean",
                            interval = "credible", nsim_mean = 100)
if (interactive()) plot(pred_mean_sb_crp)
```

```{r predict-quantile-sb-crp}
pred_q_sb_crp <- predict(fit_sb_crp, x = x_eval, type = "quantile",
                         p = 0.5, interval = "credible")
if (interactive()) plot(pred_q_sb_crp)
```

```{r predict-density-sb-crp}
pred_d_sb_crp <- predict(fit_sb_crp, x = x_eval, y = y_eval,
                         type = "density", interval = "credible")
if (interactive()) plot(pred_d_sb_crp)
```

```{r predict-survival-sb-crp}
pred_surv_sb_crp <- predict(fit_sb_crp, x = x_eval, y = y_eval,
                            type = "survival", interval = "credible")
if (interactive()) plot(pred_surv_sb_crp)
```

```{r ate-sb-crp}
ate_sb_crp <- ate(fit_sb_crp, newdata = x_eval,
                  interval = "credible", nsim_mean = 100)
if (interactive()) plot(ate_sb_crp)
```

```{r qte-sb-crp}
qte_sb_crp <- qte(fit_sb_crp, probs = c(0.25, 0.5, 0.75),
                  newdata = x_eval, interval = "credible")
if (interactive()) plot(qte_sb_crp)
```

---

## Model B: Treated CRP, Control SB (Amoroso + GPD)

```{r bundle-crp-sb, cache=FALSE}
bundle_crp_sb <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = c("amoroso", "amoroso"),
  backend = c("crp", "sb"),
  PS = "logit",
  GPD = c(TRUE, TRUE),
  components = c(5, 5),
  param_specs = param_specs_gpd,
  mcmc_outcome = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 2),
  mcmc_ps = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 2)
)

bundle_crp_sb
summary(bundle_crp_sb)
```

```{r fit-crp-sb, cache=FALSE}
fit_crp_sb <- run_mcmc_causal(bundle_crp_sb)
summary(fit_crp_sb)
params(fit_crp_sb)
```

```{r plot-fit-crp-sb}
if (interactive()) plot(fit_crp_sb, family = c("density", "geweke", "caterpillar"))
```

```{r predict-mean-crp-sb}
pred_mean_crp_sb <- predict(fit_crp_sb, x = x_eval, type = "mean",
                            interval = "credible", nsim_mean = 100)
if (interactive()) plot(pred_mean_crp_sb)
```

```{r predict-quantile-crp-sb}
pred_q_crp_sb <- predict(fit_crp_sb, x = x_eval, type = "quantile",
                         p = 0.5, interval = "credible")
if (interactive()) plot(pred_q_crp_sb)
```

```{r predict-density-crp-sb}
pred_d_crp_sb <- predict(fit_crp_sb, x = x_eval, y = y_eval,
                         type = "density", interval = "credible")
if (interactive()) plot(pred_d_crp_sb)
```

```{r predict-survival-crp-sb}
pred_surv_crp_sb <- predict(fit_crp_sb, x = x_eval, y = y_eval,
                            type = "survival", interval = "credible")
if (interactive()) plot(pred_surv_crp_sb)
```

```{r ate-crp-sb}
ate_crp_sb <- ate(fit_crp_sb, newdata = x_eval,
                  interval = "credible", nsim_mean = 100)
if (interactive()) plot(ate_crp_sb)
```

```{r qte-crp-sb}
qte_crp_sb <- qte(fit_crp_sb, probs = c(0.25, 0.5, 0.75),
                  newdata = x_eval, interval = "credible")
if (interactive()) plot(qte_crp_sb)
```









