---
title: "Introduction with GPD kernel"
pkgdown:
  slug: "kernels/introduction-with-gpd-kernel"
---

```{r setup, include=FALSE}
source("_load_pkg.R")
```

## Introduction with GPD kernel {#introduction-with-gpd-kernel}


### Generalized Pareto distribution (standalone tail)

The standalone GPD functions `dGpd`, `pGpd`, `qGpd`, and `rGpd` implement the distribution of \(X\) **above a threshold** \(u\). For \(x\ge u\), the density is

\[
f(x;u,\sigma,\xi)=\frac{1}{\sigma}\left(1+\xi\frac{x-u}{\sigma}\right)^{-1/\xi-1},
\quad \sigma>0,
\quad 1+\xi\frac{x-u}{\sigma}>0.
\]

The CDF is
\[
F(x)=1-\left(1+\xi\frac{x-u}{\sigma}\right)^{-1/\xi},
\quad x\ge u,
\]
and the quantile is
\[
Q(p)=u+\frac{\sigma}{\xi}\Big[(1-p)^{-\xi}-1\Big],\quad p\in(0,1),\ \xi\ne 0,
\]
with the exponential-limit case obtained as \(\xi\to 0\).

**Parameter mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `scale`, \(\xi\to\) `shape`. Log-densities are returned when `log = TRUE`.


We show the DPmixGPD kernels in pairs: bulk-only and GPD-augmented. A GPD tail above threshold $u$ uses

$$f_{GPD}(x; u, \sigma, \xi) = \frac{1}{\sigma} \left(1 + \xi \frac{x-u}{\sigma}\right)^{-1/\xi - 1} \ \text{for } x \ge u.$$

The standalone GPD tail distribution is available via `dGpd`, `pGpd`, `qGpd`, and `rGpd` functions.


```{r gpd-density}
dGpd(1.8, threshold = 1.5, scale = 0.5, shape = 0.2)
dGpd(1.8, threshold = 1.5, scale = 0.5, shape = 0.2, log = TRUE)
```

```{r gpd-cdf}
pGpd(1.8, threshold = 1.5, scale = 0.5, shape = 0.2)
pGpd(1.8, threshold = 1.5, scale = 0.5, shape = 0.2, lower.tail = FALSE)
pGpd(1.8, threshold = 1.5, scale = 0.5, shape = 0.2, log.p = TRUE)
```

```{r gpd-quantile}
q_vec(qGpd, c(0.25, 0.5, 0.75), threshold = 1.5, scale = 0.5, shape = 0.2)
q_vec(qGpd, c(0.25, 0.5, 0.75), threshold = 1.5, scale = 0.5, shape = 0.2,
      lower.tail = FALSE)
q_vec(qGpd, c(log(0.25), log(0.5), log(0.75)), threshold = 1.5, scale = 0.5,
      shape = 0.2, log.p = TRUE)
```

```{r gpd-random}
draw_many(rGpd, list(threshold = 1.5, scale = 0.5, shape = 0.2))
```

```{r gpd-plot}
grid <- seq(-4, 15, length.out = 500)
gpd_sets <- list(
  list(label = "GPD A", threshold = 1.5, tail_scale = 0.5, tail_shape = 0.2),
  list(label = "GPD B", threshold = 2.0, tail_scale = 0.6, tail_shape = 0.15),
  list(label = "GPD C", threshold = 1.0, tail_scale = 0.4, tail_shape = 0.25)
)

df_gpd <- do.call(rbind, lapply(gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dGpd, list(threshold = ps$threshold, scale = ps$tail_scale, shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Standalone GPD tails", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

Each subsection below defines the available $d/p/q/r$ functions, prints example outputs for fixed parameters, and overlays density curves for **three parameter sets** with clear legends. The same parameter sets are reused within the bulk and GPD variants of a section.





