---
title: "Lognormal"
pkgdown:
  slug: "kernels/kernel-lognormal"
---

```{r setup, include=FALSE, cache=FALSE}
source("_load_pkg.R")
```

## Lognormal {#v02-lognormal}


### Lognormal mixture kernel

A lognormal component is defined by \(\log Y \sim \mathcal{N}(\mu,\sigma^2)\), i.e.,
\[
f(y\mid\mu,\sigma)=\frac{1}{y\,\sigma\sqrt{2\pi}}\exp\!\left(-\frac{(\log y-\mu)^2}{2\sigma^2}\right),\quad y>0.
\]

A finite lognormal mixture has density
\[
f(y)=\sum_{j=1}^J w_j\,\text{Lognormal}(y\mid\mu_j,\sigma_j^2),\quad y>0.
\]

**Parameter mapping (math \(\rightarrow\) code):** \(\mu_j\to\) `meanlog[j]`, \(\sigma_j\to\) `sdlog[j]`, \(w_j\to\) `w[j]`.

### Lognormal mixture with GPD tail

The `*MixGpd` variant uses the same splicing idea: lognormal mixture below \(u\), GPD tail above \(u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD (mixture kernel)

```{r lognormal-mix}
grid <- seq(0, 8, length.out = 400)
logn_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), meanlog = c(0.0, 0.3, 0.6), sdlog = c(0.4, 0.5, 0.6)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), meanlog = c(0.1, 0.4, 0.7), sdlog = c(0.35, 0.45, 0.55)),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), meanlog = c(0.2, 0.5, 2), sdlog = c(0.3, 0.4, 0.5))
)

example <- logn_sets[[1]]
```

```{r lognormal-mix-d}
dLognormalMix(1, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog)
dLognormalMix(1, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, log = TRUE)
```

```{r lognormal-mix-p}
pLognormalMix(1, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog)
pLognormalMix(1, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, lower.tail = FALSE)
pLognormalMix(1, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, log.p = TRUE)
```

```{r lognormal-mix-q}
q_vec(qLognormalMix, c(0.25, 0.5, 0.75), w = example$w,
      meanlog = example$meanlog, sdlog = example$sdlog)
q_vec(qLognormalMix, c(0.25, 0.5, 0.75), w = example$w,
      meanlog = example$meanlog, sdlog = example$sdlog, lower.tail = FALSE)
q_vec(qLognormalMix, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      meanlog = example$meanlog, sdlog = example$sdlog, log.p = TRUE)
```

```{r lognormal-mix-r}
draw_many(rLognormalMix, list(w = example$w, meanlog = example$meanlog, sdlog = example$sdlog))
```

```{r lognormal-mix-plot}
df_logn <- do.call(rbind, lapply(logn_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dLognormalMix, list(w = ps$w, meanlog = ps$meanlog, sdlog = ps$sdlog)), label = ps$label)
}))

ggplot(df_logn, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Lognormal mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

```{r lognormal-gpd}
logn_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.4), meanlog = c(0, 1), sdlog = c(0.3, 0.5), threshold = 2.5, tail_scale = 0.5, tail_shape = 0.2),
  list(label = "Mix B", w = c(0.5, 0.5), meanlog = c(0.5, 1.2), sdlog = c(0.4, 0.4), threshold = 2.0, tail_scale = 0.4, tail_shape = 0.15),
  list(label = "Mix C", w = c(0.7, 0.3), meanlog = c(0.8, 1.5), sdlog = c(0.25, 0.6), threshold = 3.0, tail_scale = 0.6, tail_shape = 0.18)
)

example <- logn_gpd_sets[[1]]
```

```{r lognormal-gpd-d}
dLognormalMixGpd(2.5, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r lognormal-gpd-p}
pLognormalMixGpd(2.5, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r lognormal-gpd-q}
q_vec(qLognormalMixGpd, c(0.5, 0.9), w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r lognormal-gpd-r}
draw_many(rLognormalMixGpd, example)
```

```{r lognormal-gpd-plot}
df_logn_gpd <- do.call(rbind, lapply(logn_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dLognormalMixGpd, list(w = ps$w, meanlog = ps$meanlog, sdlog = ps$sdlog, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_logn_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Lognormal mixtures with GPD tail (different thresholds)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```









