---
title: "Amoroso (bulk kernel and mixture)"
pkgdown:
  slug: "kernels/kernel-amoroso"
---

```{r setup, include=FALSE}
source("_load_pkg.R")
```

## Amoroso (bulk kernel and mixture) {#v02-amoroso-bulk-kernel-and-mixture}


### Amoroso mixture kernel

An Amoroso component with location \(a\in\mathbb{R}\), scale \(\theta\neq 0\), and shapes \(\alpha>0\), \(\beta>0\) has density (for \(y>a\))
\[
f(y\mid a,\theta,\alpha,\beta)
=
\frac{1}{\Gamma(\alpha)}
\left|\frac{\beta}{\theta}\right|
\left(\frac{y-a}{\theta}\right)^{\alpha\beta-1}
\exp\!\left[-\left(\frac{y-a}{\theta}\right)^{\beta}\right].
\]

A finite Amoroso mixture is
\[
f(y)=\sum_{j=1}^J w_j\,f_j(y\mid a_j,\theta_j,\alpha_j,\beta_j).
\]

**Parameter mapping (math \(\rightarrow\) code):** \(a_j\to\) `loc[j]`, \(\theta_j\to\) `scale[j]`, \(\alpha_j\to\) `shape1[j]`, \(\beta_j\to\) `shape2[j]`, \(w_j\to\) `w[j]`.

### Amoroso mixture with GPD tail

The spliced kernel attaches a GPD tail above \(u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD

```{r amoroso-mix}
grid <- seq(0.01, 6, length.out = 400)
amor_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), loc = c(0.5, 0.5, 0.5), scale = c(1.0, 1.3, 1.6), shape1 = c(2.5, 3.0, 4.0), shape2 = c(1.2, 1.2, 1.2)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), loc = c(0.4, 0.6, 0.6), scale = c(1.1, 1.2, 1.5), shape1 = c(2.2, 2.8, 3.8), shape2 = c(1.1, 1.2, 1.3)),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), loc = c(0.6, 0.6, 0.6), scale = c(0.9, 1.1, 1.4), shape1 = c(2.0, 2.6, 3.5), shape2 = c(1.0, 1.1, 1.2))
)

example <- amor_sets[[1]]
```

```{r amoroso-mix-d}
dAmorosoMix(2, w = example$w, loc = example$loc, scale = example$scale,
            shape1 = example$shape1, shape2 = example$shape2)
dAmorosoMix(2, w = example$w, loc = example$loc, scale = example$scale,
            shape1 = example$shape1, shape2 = example$shape2, log = TRUE)
```

```{r amoroso-mix-p}
pAmorosoMix(2, w = example$w, loc = example$loc, scale = example$scale,
            shape1 = example$shape1, shape2 = example$shape2)
pAmorosoMix(2, w = example$w, loc = example$loc, scale = example$scale,
            shape1 = example$shape1, shape2 = example$shape2, lower.tail = FALSE)
pAmorosoMix(2, w = example$w, loc = example$loc, scale = example$scale,
            shape1 = example$shape1, shape2 = example$shape2, log.p = TRUE)
```

```{r amoroso-mix-q}
q_vec(qAmorosoMix, c(0.25, 0.5, 0.75), w = example$w, loc = example$loc,
      scale = example$scale, shape1 = example$shape1, shape2 = example$shape2)
q_vec(qAmorosoMix, c(0.25, 0.5, 0.75), w = example$w, loc = example$loc,
      scale = example$scale, shape1 = example$shape1, shape2 = example$shape2,
      lower.tail = FALSE)
q_vec(qAmorosoMix, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      loc = example$loc, scale = example$scale, shape1 = example$shape1,
      shape2 = example$shape2, log.p = TRUE)
```

```{r amoroso-mix-r}
draw_many(rAmorosoMix, list(w = example$w, loc = example$loc,
                            scale = example$scale, shape1 = example$shape1,
                            shape2 = example$shape2))
```

```{r amoroso-mix-plot}
df_amor <- do.call(rbind, lapply(amor_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dAmorosoMix, list(w = ps$w, loc = ps$loc, scale = ps$scale, shape1 = ps$shape1, shape2 = ps$shape2)), label = ps$label)
}))

ggplot(df_amor, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Amoroso mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

```{r amoroso-gpd}
amor_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), loc = c(0.5, 0.5, 0.5), scale = c(1.0, 1.3, 1.6), shape1 = c(2.5, 3.0, 4.0), shape2 = c(1.2, 1.2, 1.2), threshold = 2.8, tail_scale = 0.4, tail_shape = 0.2),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), loc = c(0.4, 0.6, 0.6), scale = c(1.1, 1.2, 1.5), shape1 = c(2.2, 2.8, 3.8), shape2 = c(1.1, 1.2, 1.3), threshold = 3.0, tail_scale = 0.35, tail_shape = 0.18),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), loc = c(0.6, 0.6, 0.6), scale = c(0.9, 1.1, 1.4), shape1 = c(2.0, 2.6, 3.5), shape2 = c(1.0, 1.1, 1.2), threshold = 2.5, tail_scale = 0.45, tail_shape = 0.22)
)
example <- amor_gpd_sets[[1]]
```

```{r amoroso-gpd-d}
dAmorosoMixGpd(3, w = example$w, loc = example$loc, scale = example$scale,
               shape1 = example$shape1, shape2 = example$shape2,
               threshold = example$threshold, tail_scale = example$tail_scale,
               tail_shape = example$tail_shape)
dAmorosoMixGpd(3, w = example$w, loc = example$loc, scale = example$scale,
               shape1 = example$shape1, shape2 = example$shape2,
               threshold = example$threshold, tail_scale = example$tail_scale,
               tail_shape = example$tail_shape, log = TRUE)
```

```{r amoroso-gpd-p}
pAmorosoMixGpd(3, w = example$w, loc = example$loc, scale = example$scale,
               shape1 = example$shape1, shape2 = example$shape2,
               threshold = example$threshold, tail_scale = example$tail_scale,
               tail_shape = example$tail_shape)
pAmorosoMixGpd(3, w = example$w, loc = example$loc, scale = example$scale,
               shape1 = example$shape1, shape2 = example$shape2,
               threshold = example$threshold, tail_scale = example$tail_scale,
               tail_shape = example$tail_shape, lower.tail = FALSE)
pAmorosoMixGpd(3, w = example$w, loc = example$loc, scale = example$scale,
               shape1 = example$shape1, shape2 = example$shape2,
               threshold = example$threshold, tail_scale = example$tail_scale,
               tail_shape = example$tail_shape, log.p = TRUE)
```

```{r amoroso-gpd-q}
q_vec(qAmorosoMixGpd, c(0.25, 0.5, 0.75), w = example$w, loc = example$loc,
      scale = example$scale, shape1 = example$shape1, shape2 = example$shape2,
      threshold = example$threshold, tail_scale = example$tail_scale,
      tail_shape = example$tail_shape)
q_vec(qAmorosoMixGpd, c(0.25, 0.5, 0.75), w = example$w, loc = example$loc,
      scale = example$scale, shape1 = example$shape1, shape2 = example$shape2,
      threshold = example$threshold, tail_scale = example$tail_scale,
      tail_shape = example$tail_shape, lower.tail = FALSE)
q_vec(qAmorosoMixGpd, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      loc = example$loc, scale = example$scale, shape1 = example$shape1,
      shape2 = example$shape2, threshold = example$threshold,
      tail_scale = example$tail_scale, tail_shape = example$tail_shape,
      log.p = TRUE)
```

```{r amoroso-gpd-r}
draw_many(rAmorosoMixGpd, example)
```

```{r amoroso-gpd-plot}
df_amor_gpd <- do.call(rbind, lapply(amor_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dAmorosoMixGpd, list(w = ps$w, loc = ps$loc, scale = ps$scale, shape1 = ps$shape1, shape2 = ps$shape2, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_amor_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Amoroso mixtures with GPD tail", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```





