---
title: "Laplace"
pkgdown:
  slug: "kernels/kernel-laplace"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\\VignetteIndexEntry{Laplace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
source("_load_pkg.R")
```

## Laplace {#v02-laplace}


### Laplace mixture kernel

A Laplace (double-exponential) component with location \(\ell\in\mathbb{R}\) and scale \(b>0\) has density
\[
f(y\mid \ell,b)=\frac{1}{2b}\exp\!\left(-\frac{|y-\ell|}{b}\right),\quad y\in\mathbb{R}.
\]

A finite Laplace mixture is
\[
f(y)=\sum_{j=1}^J w_j\,\text{Laplace}(y\mid \ell_j,b_j).
\]

**Parameter mapping (math \(\rightarrow\) code):** \(\ell_j\to\) `location[j]`, \(b_j\to\) `scale[j]`, \(w_j\to\) `w[j]`.

### Laplace mixture with GPD tail

The spliced kernel uses the Laplace mixture below \(u\) and a GPD tail above \(u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD (mixture kernel)

Laplace mixture density $f(x)=\sum_j w_j\,\text{Laplace}(x\mid \ell_j, b_j)$.

```{r laplace-mix}
grid <- seq(-4, 4, length.out = 400)
lap_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), location = c(0.0, 1.0, -2), scale = c(1.0, 0.9, 1.1)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), location = c(0.2, 1.2, -0.5), scale = c(0.9, 2, 1.0)),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), location = c(-0.2, 2, -1.0), scale = c(1.1, 0.95, 1.05))
)

example <- lap_sets[[1]]
```

```{r laplace-mix-d}
dLaplaceMix(0, w = example$w, location = example$location, scale = example$scale)
dLaplaceMix(0, w = example$w, location = example$location, scale = example$scale, log = TRUE)
```

```{r laplace-mix-p}
pLaplaceMix(0, w = example$w, location = example$location, scale = example$scale)
pLaplaceMix(0, w = example$w, location = example$location, scale = example$scale, lower.tail = FALSE)
pLaplaceMix(0, w = example$w, location = example$location, scale = example$scale, log.p = TRUE)
```

```{r laplace-mix-q}
q_vec(qLaplaceMix, c(0.25, 0.5, 0.75), w = example$w,
      location = example$location, scale = example$scale)
q_vec(qLaplaceMix, c(0.25, 0.5, 0.75), w = example$w,
      location = example$location, scale = example$scale, lower.tail = FALSE)
q_vec(qLaplaceMix, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      location = example$location, scale = example$scale, log.p = TRUE)
```

```{r laplace-mix-r}
draw_many(rLaplaceMix, list(w = example$w, location = example$location, scale = example$scale))
```

```{r laplace-mix-plot}
df_lap <- do.call(rbind, lapply(lap_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dLaplaceMix, list(w = ps$w, location = ps$location, scale = ps$scale)), label = ps$label)
}))

ggplot(df_lap, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Laplace mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

```{r laplace-gpd}
lap_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.4), location = c(-0.5, 1), scale = c(0.4, 0.6), threshold = 1.5, tail_scale = 0.35, tail_shape = 0.3),
  list(label = "Mix B", w = c(0.5, 0.5), location = c(0, 0.8), scale = c(0.5, 0.5), threshold = 1.2, tail_scale = 0.3, tail_shape = 0.25),
  list(label = "Mix C", w = c(0.7, 0.3), location = c(0.2, 1.5), scale = c(0.35, 0.7), threshold = 1.8, tail_scale = 0.4, tail_shape = 0.22)
)

example <- lap_gpd_sets[[1]]
```

```{r laplace-gpd-d}
dLaplaceMixGpd(1, w = example$w, location = example$location, scale = example$scale, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r laplace-gpd-p}
pLaplaceMixGpd(1, w = example$w, location = example$location, scale = example$scale, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r laplace-gpd-q}
q_vec(qLaplaceMixGpd, c(0.5, 0.9), w = example$w, location = example$location, scale = example$scale, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r laplace-gpd-r}
draw_many(rLaplaceMixGpd, example)
```

```{r laplace-gpd-plot}
df_lap_gpd <- do.call(rbind, lapply(lap_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dLaplaceMixGpd, list(w = ps$w, location = ps$location, scale = ps$scale, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_lap_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Laplace mixtures with GPD tail (different thresholds)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```





