---
title: "Inverse Gaussian (mixture kernel)"
pkgdown:
  slug: "kernels/kernel-inverse-gaussian-mixture"
---

```{r setup, include=FALSE, cache=FALSE}
source("_load_pkg.R")
```

## Inverse Gaussian (mixture kernel) {#v02-inverse-gaussian-mixture-kernel}


### Inverse Gaussian mixture kernel (finite components) {#inverse-gaussian-mixture-kernel-finite}

An inverse Gaussian component with mean \(\mu>0\) and shape \(\lambda>0\) has density
\[
f(y\mid \mu,\lambda)
=
\left(\frac{\lambda}{2\pi y^3}\right)^{1/2}
\exp\!\left(-\frac{\lambda (y-\mu)^2}{2\mu^2 y}\right),
\quad y>0.
\]

A finite inverse Gaussian mixture is
\[
f(y)=\sum_{j=1}^J w_j\,\text{IG}(y\mid \mu_j,\lambda_j),\quad y>0.
\]

**Parameter mapping (math \(\rightarrow\) code):** \(\mu_j\to\) `mean[j]`, \(\lambda_j\to\) `shape[j]`, \(w_j\to\) `w[j]`.

### Inverse Gaussian mixture with GPD tail

The spliced variant attaches a GPD tail above \(u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD

```{r invgauss-mix}
grid <- seq(0.1, 6, length.out = 400)
ig_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), mean = c(1.0, 1.5, 2.0), shape = c(2.0, 3.0, 4.0)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), mean = c(1.1, 1.6, 2.2), shape = c(2.2, 3.2, 4.2)),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), mean = c(0.9, 1.4, 2.1), shape = c(1.8, 2.8, 3.8))
)

example <- ig_sets[[1]]
```

```{r invgauss-mix-d}
dInvGaussMix(1, w = example$w, mean = example$mean, shape = example$shape)
dInvGaussMix(1.5, w = example$w, mean = example$mean, shape = example$shape, log = TRUE)
```

```{r invgauss-mix-p}
pInvGaussMix(1.5, w = example$w, mean = example$mean, shape = example$shape)
pInvGaussMix(1.5, w = example$w, mean = example$mean, shape = example$shape, lower.tail = FALSE)
pInvGaussMix(1.5, w = example$w, mean = example$mean, shape = example$shape, log.p = TRUE)
```

```{r invgauss-mix-q}
q_vec(qInvGaussMix, c(0.25, 0.5, 0.75), w = example$w, mean = example$mean,
      shape = example$shape)
q_vec(qInvGaussMix, c(0.25, 0.5, 0.75), w = example$w, mean = example$mean,
      shape = example$shape, lower.tail = FALSE)
q_vec(qInvGaussMix, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      mean = example$mean, shape = example$shape, log.p = TRUE)
```

```{r invgauss-mix-r}
draw_many(rInvGaussMix, list(w = example$w, mean = example$mean, shape = example$shape))
```

```{r invgauss-mix-plot}
df_ig <- do.call(rbind, lapply(ig_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dInvGaussMix, list(w = ps$w, mean = ps$mean, shape = ps$shape)), label = ps$label)
}))

ggplot(df_ig, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Inverse Gaussian mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

```{r invgauss-gpd}
ig_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.4), mean = c(1, 2.5), shape = c(2, 3), threshold = 2.5, tail_scale = 0.5, tail_shape = 0.25),
  list(label = "Mix B", w = c(0.5, 0.5), mean = c(1.5, 2), shape = c(2.5, 2.5), threshold = 2.0, tail_scale = 0.4, tail_shape = 0.2),
  list(label = "Mix C", w = c(0.7, 0.3), mean = c(1.2, 3), shape = c(3, 2), threshold = 3.0, tail_scale = 0.6, tail_shape = 0.18)
)

example <- ig_gpd_sets[[1]]
```

```{r invgauss-gpd-d}
dInvGaussMixGpd(2.5, w = example$w, mean = example$mean, shape = example$shape, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r invgauss-gpd-p}
pInvGaussMixGpd(2.5, w = example$w, mean = example$mean, shape = example$shape, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r invgauss-gpd-q}
q_vec(qInvGaussMixGpd, c(0.5, 0.9), w = example$w, mean = example$mean, shape = example$shape, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r invgauss-gpd-r}
draw_many(rInvGaussMixGpd, example)
```

```{r invgauss-gpd-plot}
df_ig_gpd <- do.call(rbind, lapply(ig_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dInvGaussMixGpd, list(w = ps$w, mean = ps$mean, shape = ps$shape, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_ig_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Inverse Gaussian mixtures with GPD tail", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [Introduction With Gpd Kernel](introduction-with-gpd-kernel.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->





