---
title: "Mathematical definitions and conventions"
pkgdown:
  slug: "kernels/mathematical-definitions-and-conventions"
---

```{r setup, include=FALSE}
source("_load_pkg.R")
```

## Mathematical definitions and conventions {#v02-mathematical-definitions-and-conventions}

This vignette documents the probability distributions exported by **DPmixGPD** and the corresponding `d/p/q/r` functions used throughout the package. For a continuous random variable \(Y\) with parameter scalars \(\boldsymbol{\theta}\), we use the standard quartet:

- **Density (PDF)**: \(f(y\mid\boldsymbol{\theta})\), implemented by `d*()`.
- **Distribution (CDF)**: \(F(y\mid\boldsymbol{\theta})=\Pr(Y\le y)\), implemented by `p*()`.
- **Quantile**: \(Q(p\mid\boldsymbol{\theta}) = \inf\{y: F(y\mid\boldsymbol{\theta})\ge p\}\), implemented by `q*()`.
- **Random generation**: \(Y = Q(U\mid\boldsymbol{\theta})\) with \(U\sim \text{Unif}(0,1)\), implemented by `r*()`.

All `d*()` functions optionally return log-densities via `log = TRUE`, i.e., \(\log f(y\mid\boldsymbol{\theta})\). This is numerically convenient because likelihoods multiply densities but add log-densities:
\[
\log\Big\{\prod_{i=1}^n f(y_i\mid\boldsymbol{\theta})\Big\}=\sum_{i=1}^n \log f(y_i\mid\boldsymbol{\theta}).
\]

All `p*()` functions support `lower.tail` and `log.p`:
\[
\Pr(Y>y)=1-F(y),\qquad \log F(y)\ \text{or}\ \log\{1-F(y)\}.
\]

The exported uppercase NIMBLE kernels are scalar in their first argument, and `r*()` kernels are implemented for `n = 1`. For vectorized R usage (including `n > 1`), use the lowercase wrapper families documented in the `*_lowercase` help topics.

```{r vectorized-dqrp}
w <- c(0.60, 0.40)
shape <- c(2.0, 5.0)
scale <- c(1.0, 2.0)

dGammaMix(seq(0.5, 2.0, length.out = 5), w = w, shape = shape, scale = scale)
qGammaMix(c(0.1, 0.5, 0.9), w = w, shape = shape, scale = scale)
rGammaMix(5, w = w, shape = shape, scale = scale)
```

### Finite mixtures

Several kernels are available as **finite mixtures**. Let \(f_j(\cdot\mid\boldsymbol{\theta}_j)\) denote the \(j\)th component density and let \(\boldsymbol{w}=(w_1,\ldots,w_J)\) be mixture weights with \(w_j\ge 0\) and \(\sum_{j=1}^J w_j=1\). The mixture density and CDF are

\[
f(y)=\sum_{j=1}^J w_j f_j(y\mid\boldsymbol{\theta}_j),
\qquad
F(y)=\sum_{j=1}^J w_j F_j(y\mid\boldsymbol{\theta}_j).
\]

Random generation proceeds by sampling a latent component index \(Z\sim\text{Categorical}(\boldsymbol{w})\), then sampling \(Y\mid Z=j\sim f_j(\cdot\mid\boldsymbol{\theta}_j)\). Mixture quantiles typically require numerical inversion of the mixture CDF.

### GPD tails and splicing

The **Generalized Pareto Distribution (GPD)** is used for tail modeling relative to a threshold \(u\). In the spliced kernels (the `*Gpd` functions), a bulk kernel is used below \(u\) and a GPD tail is attached above \(u\), with parameters \((u,\sigma,\xi)\). In code, these are represented by `threshold = u`, `tail_scale = \sigma`, and `tail_shape = \xi` (the standalone GPD functions use `scale` and `shape` for \(\sigma\) and \(\xi\)).










