---
title: "Summary: Distribution & Backend Reference"
pkgdown:
  slug: "kernels/summary-distribution-backend-reference"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\\VignetteIndexEntry{Summary: Distribution & Backend Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
source("_load_pkg.R")
```

## Summary: Distribution & Backend Reference {#v02-summary-distribution-backend-reference}

This section provides a compact reference for the distributions supported by **DPmixGPD** and the corresponding user-facing function families.  
The **CRP backend** uses base (single-kernel) functions, while the **SB backend** uses mixture (multi-component) functions. The “Type” column indicates whether each backend expects **scalar** parameters (single value per parameter) or **vectors** indexed by mixture component.


```{r dist-backend-summary, echo=FALSE, message=FALSE, warning=FALSE}
df <- data.frame(
  Distribution = c(
    "Normal","Gamma","Lognormal","Laplace","Inverse Gaussian","Amoroso","Cauchy",
    "GPD",
    "Normal + GPD","Gamma + GPD","Lognormal + GPD","Laplace + GPD","InvGauss + GPD","Amoroso + GPD"
  ),
  Parameters = c(
    "$\\mu,\\,\\sigma$",
    "$\\alpha,\\,\\beta$",
    "$\\mu,\\,\\sigma$",
    "$\\ell,\\,b$",
    "$\\mu,\\,\\lambda$",
    "$a,\\,\\theta,\\,\\alpha,\\,\\beta$",
    "$x_0,\\,\\gamma$",
    "$u,\\,\\sigma,\\,\\xi$",
    "$\\mu,\\,\\sigma,\\,u,\\,\\sigma,\\,\\xi$",
    "$\\alpha,\\,\\beta,\\,u,\\,\\sigma,\\,\\xi$",
    "$\\mu,\\,\\sigma,\\,u,\\,\\sigma,\\,\\xi$",
    "$\\ell,\\,b,\\,u,\\,\\sigma,\\,\\xi$",
    "$\\mu,\\,\\lambda,\\,u,\\,\\sigma,\\,\\xi$",
    "$a,\\,\\theta,\\,\\alpha,\\,\\beta,\\,u,\\,\\sigma,\\,\\xi$"
  ),
  `Arguments` = c(
    "`mean`, `sd`",
    "`shape`, `scale`",
    "`meanlog`, `sdlog`",
    "`location`, `scale`",
    "`mean`, `shape`",
    "`loc`, `scale`, `shape1`, `shape2`",
    "`location`, `scale`",
    "`threshold`, `scale`, `shape`",
    "`mean`, `sd`, `threshold`, `tail_scale`, `tail_shape`",
    "`shape`, `scale`, `threshold`, `tail_scale`, `tail_shape`",
    "`meanlog`, `sdlog`, `threshold`, `tail_scale`, `tail_shape`",
    "`location`, `scale`, `threshold`, `tail_scale`, `tail_shape`",
    "`mean`, `shape`, `threshold`, `tail_scale`, `tail_shape`",
    "`loc`, `scale`, `shape1`, `shape2`, `threshold`, `tail_scale`, `tail_shape`"
  ),
  `Type` = c(
    "scalars","scalars","scalars","scalars","scalars","scalars","scalars",
    "scalars",
    "scalars","scalars","scalars","scalars","scalars","scalars"
  ),
  `Function` = c(
    "`dnorm()`, `pnorm()`, `qnorm()`, `rnorm()`",
    "`dgamma()`, `pgamma()`, `qgamma()`, `rgamma()`",
    "`dlnorm()`, `plnorm()`, `qlnorm()`, `rlnorm()`",
    "`ddexp()`, `pdexp()`, `qdexp()`, `rdexp()`",
    "`dinvgauss()`, `pinvgauss()`, `qinvgauss()`, `rinvgauss()`",
    "`damoroso()`, `pamoroso()`, `qamoroso()`, `ramoroso()`",
    "`dcauchy()`, `pcauchy()`, `qcauchy()`, `rcauchy()`",
    "`dGpd()`, `pGpd()`, `qGpd()`, `rGpd()`",
    "`dNormGpd()`, `pNormGpd()`, `qNormGpd()`, `rNormGpd()`",
    "`dGammaGpd()`, `pGammaGpd()`, `qGammaGpd()`, `rGammaGpd()`",
    "`dLognormalGpd()`, `pLognormalGpd()`, `qLognormalGpd()`, `rLognormalGpd()`",
    "`dLaplaceGpd()`, `pLaplaceGpd()`, `qLaplaceGpd()`, `rLaplaceGpd()`",
    "`dInvGaussGpd()`, `pInvGaussGpd()`, `qInvGaussGpd()`, `rInvGaussGpd()`",
    "`dAmorosoGpd()`, `pAmorosoGpd()`, `qAmorosoGpd()`, `rAmorosoGpd()`"
  ),
  `Type.` = c(
    "vectors","vectors","vectors","vectors","vectors","vectors","vectors",
    "NA",
    "vectors","vectors + scalars","vectors + scalars","vectors + scalars","vectors + scalars","vectors + scalars"
  ),
  `Function.` = c(
    "`dNormMix()`, `pNormMix()`, `qNormMix()`, `rNormMix()`",
    "`dGammaMix()`, `pGammaMix()`, `qGammaMix()`, `rGammaMix()`",
    "`dLognormalMix()`, `pLognormalMix()`, `qLognormalMix()`, `rLognormalMix()`",
    "`dLaplaceMix()`, `pLaplaceMix()`, `qLaplaceMix()`, `rLaplaceMix()`",
    "`dInvGaussMix()`, `pInvGaussMix()`, `qInvGaussMix()`, `rInvGaussMix()`",
    "`dAmorosoMix()`, `pAmorosoMix()`, `qAmorosoMix()`, `rAmorosoMix()`",
    "`dCauchyMix()`, `pCauchyMix()`, `qCauchyMix()`, `rCauchyMix()`",
    "NA",
    "`dNormMixGpd()`, `pNormMixGpd()`, `qNormMixGpd()`, `rNormMixGpd()`",
    "`dGammaMixGpd()`, `pGammaMixGpd()`, `qGammaMixGpd()`, `rGammaMixGpd()`",
    "`dLognormalMixGpd()`, `pLognormalMixGpd()`, `qLognormalMixGpd()`, `rLognormalMixGpd()`",
    "`dLaplaceMixGpd()`, `pLaplaceMixGpd()`, `qLaplaceMixGpd()`, `rLaplaceMixGpd()`",
    "`dInvGaussMixGpd()`, `pInvGaussMixGpd()`, `qInvGaussMixGpd()`, `rInvGaussMixGpd()`",
    "`dAmorosoMixGpd()`, `pAmorosoMixGpd()`, `qAmorosoMixGpd()`, `rAmorosoMixGpd()`"
  ),
  check.names = FALSE
)

# Rename SB columns to match header labels cleanly
names(df)[4:7] <- c("Type", "Function", "Type", "Function")

knitr::kable(
  df,
  format = "html",
  escape = FALSE,
  align = rep("c", 7)
) |>
  kableExtra::kable_styling(full_width = FALSE, position = "center") |>
  kableExtra::add_header_above(c(" " = 3, "CRP" = 2, "SB" = 2)) |>
  kableExtra::row_spec(0, bold = TRUE)
```

### Notes 

- **CRP Backend** uses base (single-kernel) functions; parameters are scalar.

- **SB Backend** uses mixture/spliced mixture functions; component-specific parameters are typically vectors indexed by component \(j\); mixture weights `w` are required as vector.

- For **GPD-only**, SB entries are **NA** because there is no mixture-only GPD wrapper in the SB family.







