---
title: "DPmixGPD: Model specification and posterior computation"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{DPmixGPD: Model specification and posterior computation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: dpmixgpd.bib
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)
```

DPmixGPD is built around a practical modeling principle: the distributional structure that fits the *bulk* (where most observations live) is often a poor description of the *tail* (where rare extremes live). The package therefore models the bulk using a Dirichlet process mixture (DPM) and, optionally, models the upper tail using a generalized Pareto distribution (GPD), spliced in a way that yields a single proper density on the full support.

This vignette defines the full statistical model and the posterior computation used throughout the package. The workflow vignettes (unconditional, conditional, causal) use this specification but avoid restating it.

# Notation

Let $y_1,\dots,y_N$ be observed outcomes. When covariates are present, let $\boldsymbol{x}_i \in \mathbb{R}^p$ denote the $i$-th row of a design matrix $X$. Mixture components are indexed by $j$ (not $h$). The Dirichlet process concentration parameter is $\kappa>0$.

Write the bulk density and CDF as $f_{\mathrm{DPM}}(\cdot\mid\boldsymbol{x})$ and $F_{\mathrm{DPM}}(\cdot\mid\boldsymbol{x})$. Write the GPD density and CDF as $f_{\mathrm{GPD}}(\cdot\mid u(\boldsymbol{x}),\sigma(\boldsymbol{x}),\xi)$ and $F_{\mathrm{GPD}}(\cdot\mid u(\boldsymbol{x}),\sigma(\boldsymbol{x}),\xi)$.

# Bulk model: Dirichlet process mixture

## Hierarchical specification

The bulk model is a kernel mixture with a Dirichlet process prior on the mixing distribution [@ferguson1973]. In allocation form,
\[
  y_i \mid z_i,\{\Theta_j\}\;\sim\;k\!\left(y_i\mid\Theta_{z_i},\boldsymbol{x}_i\right),
  \qquad
  z_i\mid\boldsymbol{w}\;\sim\;\mathrm{Categorical}(w_1,w_2,\dots),
\]
\[
  \Theta_j\mid H\;\overset{\mathrm{iid}}{\sim}\;H,\qquad
  H\;\sim\;\mathrm{DP}(\kappa,H_0),
\]
where $k(\cdot\mid\Theta_j,\boldsymbol{x})$ is the chosen kernel family (Normal, Gamma, Lognormal, Laplace, Inverse Gaussian, Amoroso, etc.), $H_0$ is a base measure for the kernel parameters, and $\kappa$ controls the expected number of occupied clusters.

Marginally this implies an (infinite) mixture representation
\[
  f_{\mathrm{DPM}}(y\mid\boldsymbol{x})
  
  =
  \sum_{j=1}^{\infty} w_j\,k\!\left(y\mid\Theta_j,\boldsymbol{x}\right).
\]

## Stick-breaking representation and truncation

DPmixGPD provides a stick-breaking backend (`backend="sb"`) based on Sethuraman's construction [@sethuraman1994]. The weights are
\[
  w_j = V_j\prod_{\ell<j}(1-V_\ell),
  \qquad
  V_j\overset{\mathrm{iid}}{\sim}\mathrm{Beta}(1,\kappa),
  \qquad
  \Theta_j\overset{\mathrm{iid}}{\sim}H_0.
\]

In computation, the package uses a finite truncation level $J$ (`components`) and works with
\[
  f_{\mathrm{DPM}}(y\mid\boldsymbol{x})\;\approx\;\sum_{j=1}^{J} w_j\,k\!\left(y\mid\Theta_j,\boldsymbol{x}\right),
\]
where the final weight absorbs leftover mass. This is the standard blocked-Gibbs approximation for stick-breaking priors [@ishwaran2001].

Define counts
\[
  n_j = \sum_{i=1}^N\mathbb{I}(z_i=j),\qquad
  m_j = \sum_{i=1}^N\mathbb{I}(z_i>j).
\]
Then the stick variables have the conjugate full conditional
\[
  V_j\mid\boldsymbol{z},\kappa\;\sim\;\mathrm{Beta}(1+n_j,\;\kappa+m_j),
\]
and weights are reconstructed as above.

## CRP backend (partition-based view)

DPmixGPD also offers a Chinese restaurant process backend (`backend="crp"`), which integrates out $H$ and updates allocations under the induced random partition distribution [@blackwell1973; @neal2000]. In this view, observation $i$ joins an existing cluster with probability proportional to its current size, and starts a new cluster with probability proportional to $\kappa$, coupled with the base measure $H_0$.

# Tail model: generalized Pareto distribution

Let $u(\boldsymbol{x})$ be a (possibly covariate-dependent) threshold. For $y>u(\boldsymbol{x})$, define exceedance $e=y-u(\boldsymbol{x})$. The GPD model is
\[
  Y\mid Y>u(\boldsymbol{x}),\boldsymbol{x}\;\sim\;\mathrm{GPD}\big(u(\boldsymbol{x}),\sigma(\boldsymbol{x}),\xi\big),
\]
with scale $\sigma(\boldsymbol{x})>0$ and shape $\xi\in\mathbb{R}$. The density is
\[
f_{\mathrm{GPD}}\big(y\mid u(\boldsymbol{x}),\sigma(\boldsymbol{x}),\xi\big)
=
\begin{cases}
\dfrac{1}{\sigma(\boldsymbol{x})}\left(1+\xi\dfrac{y-u(\boldsymbol{x})}{\sigma(\boldsymbol{x})}\right)^{-1/\xi-1}, & \xi\neq 0,\\[10pt]
\dfrac{1}{\sigma(\boldsymbol{x})}\exp\!\left(-\dfrac{y-u(\boldsymbol{x})}{\sigma(\boldsymbol{x})}\right), & \xi=0,
\end{cases}
\]
with support constraint $1+\xi\{y-u(\boldsymbol{x})\}/\sigma(\boldsymbol{x})>0$. The CDF is
\[
F_{\mathrm{GPD}}\big(y\mid u(\boldsymbol{x}),\sigma(\boldsymbol{x}),\xi\big)
=
\begin{cases}
1-\left(1+\xi\dfrac{y-u(\boldsymbol{x})}{\sigma(\boldsymbol{x})}\right)^{-1/\xi}, & \xi\neq 0,\\[10pt]
1-\exp\!\left(-\dfrac{y-u(\boldsymbol{x})}{\sigma(\boldsymbol{x})}\right), & \xi=0.
\end{cases}
\]

The use of the GPD for exceedances is justified by classical peaks-over-threshold limit theory [@balkema1974; @pickands1975; @coles2001].

# The spliced bulk–tail model

DPmixGPD defines a single proper density by splicing the bulk DPM and the GPD tail at $u(\boldsymbol{x})$:
\[
f(y\mid\boldsymbol{x})
=
\begin{cases}
f_{\mathrm{DPM}}(y\mid\boldsymbol{x}), & y\le u(\boldsymbol{x}),\\[6pt]
\Big(1-F_{\mathrm{DPM}}\big(u(\boldsymbol{x})\mid\boldsymbol{x}\big)\Big)\,f_{\mathrm{GPD}}\big(y\mid u(\boldsymbol{x}),\sigma(\boldsymbol{x}),\xi\big), & y>u(\boldsymbol{x}).
\end{cases}
\]
The multiplicative factor $1-F_{\mathrm{DPM}}\{u(\boldsymbol{x})\mid\boldsymbol{x}\}$ ensures that the total tail probability mass matches the bulk survival beyond the threshold.

Equivalently, the spliced CDF is
\[
F(y\mid\boldsymbol{x})
=
\begin{cases}
F_{\mathrm{DPM}}(y\mid\boldsymbol{x}), & y\le u(\boldsymbol{x}),\\[6pt]
F_{\mathrm{DPM}}\big(u(\boldsymbol{x})\mid\boldsymbol{x}\big)
+\Big(1-F_{\mathrm{DPM}}\big(u(\boldsymbol{x})\mid\boldsymbol{x}\big)\Big)
\,F_{\mathrm{GPD}}\big(y\mid u(\boldsymbol{x}),\sigma(\boldsymbol{x}),\xi\big), & y>u(\boldsymbol{x}).
\end{cases}
\]

This section is the reference point for all later predictive quantities (density, survival, quantiles, exceedance probabilities).

# Covariate dependence: general link mode

In conditional models, DPmixGPD allows (selected) parameters to depend on $\boldsymbol{x}$ via a link representation. For a generic parameter $\theta_j(\boldsymbol{x})$ in link mode,
\[
  g\!\left(\theta_j(\boldsymbol{x})\right) = \eta_j(\boldsymbol{x}) = \boldsymbol{x}^\top\boldsymbol{\beta}_j,
\]
where $g(\cdot)$ is a link function and $\boldsymbol{\beta}_j$ is a component-specific regression coefficient. The package exposes this mechanism through `param_specs`, and it is not restricted to a single parameter: in principle, any parameter registered as linkable for a kernel can be placed in link mode.

The same idea applies to the tail module when a tail parameter is allowed to vary with $\boldsymbol{x}$ (for example, a covariate-dependent threshold or scale).

# Likelihood and posterior computation

## Likelihood

Under the spliced model, each observation contributes either a bulk likelihood term or a tail likelihood term, depending on whether it falls below or above the threshold. For a fixed threshold $u$ (no covariates), the full likelihood is
\[
  \mathcal{L}(\Theta,\boldsymbol{w},\sigma,\xi\mid\boldsymbol{y})
  =
  \prod_{i: y_i\le u} f_{\mathrm{DPM}}(y_i)
  \times
  \prod_{i: y_i>u} \Big(1-F_{\mathrm{DPM}}(u)\Big)\,f_{\mathrm{GPD}}(y_i\mid u,\sigma,\xi).
\]
With covariates, replace $u$ by $u(\boldsymbol{x}_i)$ and use $f_{\mathrm{DPM}}(\cdot\mid\boldsymbol{x}_i)$ and $F_{\mathrm{DPM}}(\cdot\mid\boldsymbol{x}_i)$.

## Posterior sampling (high-level algorithm)

DPmixGPD targets the joint posterior of mixture allocations, kernel parameters, weights, and (optionally) tail parameters.

For the stick-breaking backend, a typical blocked-Gibbs iteration updates:

1. Allocation indicators $z_i$ via categorical probabilities proportional to $w_j\,k(y_i\mid\Theta_j,\boldsymbol{x}_i)$, with splicing handled by the tail module when $y_i>u(\boldsymbol{x}_i)$.
2. Stick variables $V_j$ (and hence $w_j$) from their conjugate beta full conditionals.
3. Kernel parameters $\Theta_j$ from their full conditional distributions. When conjugacy is unavailable, the package uses Metropolis–Hastings steps inside the Gibbs loop.
4. Tail parameters $(u(\cdot),\sigma(\cdot),\xi)$ according to the user-specified `param_specs`. Fixed parameters are held constant; prior-based parameters are updated by conjugate or Metropolis steps depending on the chosen prior family.
5. The concentration parameter $\kappa$ if it is assigned a prior (typically a gamma prior), again via conjugate or Metropolis updates.

The CRP backend updates allocations under the partition distribution and then updates cluster parameters accordingly.

## Posterior predictive quantities

Posterior prediction is implemented through the S3 method `predict()` for `mixgpd_fit` objects. For any functional $g(\theta)$ of the model parameters, prediction follows the standard Bayesian recipe:
\[
  p(\tilde{y}\mid\boldsymbol{x},\mathcal{D}) = \int p(\tilde{y}\mid\boldsymbol{x},\theta)\,p(\theta\mid\mathcal{D})\,d\theta,
\]
approximated by Monte Carlo over posterior draws $\theta^{(s)}$.

In particular,

- density: $f(\tilde{y}\mid\boldsymbol{x},\mathcal{D}) \approx S^{-1}\sum_{s=1}^S f(\tilde{y}\mid\boldsymbol{x},\theta^{(s)})$,
- survival: $S(\tilde{y}\mid\boldsymbol{x},\mathcal{D}) = 1-F(\tilde{y}\mid\boldsymbol{x},\mathcal{D})$,
- quantiles: $Q(p\mid\boldsymbol{x},\mathcal{D})$ obtained by inverting the posterior predictive CDF draw-by-draw and summarizing across draws.

# Built-in example datasets

The source package ships several small datasets intended for examples and testing. Each dataset is an R object in `data/` and can be loaded with `data("<name>")`. All of them are lists with a standard structure:

- `y`: numeric outcome,
- `X`: covariates (either `NULL` for unconditional or a `data.frame` for conditional),
- `T`: treatment indicator (only for causal datasets),
- `meta`: metadata (sample size, support, number of true components, whether a tail was injected),
- `truth`: simulation truth (kernel family and parameters used to generate the dataset).

The workflow vignettes use these datasets to avoid generating new data.

```{r, echo=FALSE}
library(DPmixGPD)
data("nc_pos_tail200_k4")
str(nc_pos_tail200_k4$meta)
str(nc_pos_tail200_k4$truth)
```
