---
title: "DPmixGPD: Model fundamentals (bulk DPM + GPD tail)"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{DPmixGPD: Model fundamentals (bulk DPM + GPD tail)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: dpmixgpd.bib
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE
)
```

```{r}
library(DPmixGPD)
```

This vignette defines the statistical model underlying DPmixGPD. The three ingredients are:

1. a Dirichlet process mixture (DPM) for the *bulk* (central) part of the distribution,
2. a generalized Pareto distribution (GPD) model for the *tail* (extremes), and
3. a *proper* splicing construction that yields a valid density/CDF on the full support.

Other vignettes (unconditional, conditional, causal) refer back here rather than repeating these definitions.

# Bulk model: Dirichlet process mixture

Let $y_1,\dots,y_n$ denote outcomes. DPmixGPD uses a mixture representation
\[
  y_i \mid z_i=j \sim k(y_i \mid \Theta_j), \qquad
  z_i \sim \mathrm{Categorical}(w_1,w_2,\dots),
\]
so the bulk density is
\[
  f_{\mathrm{DPM}}(y)
  =
  \sum_{j=1}^\infty w_j \, k(y \mid \Theta_j).
\]

A Dirichlet process prior is imposed on the random mixing measure $G$ underlying the parameters $\Theta_j$:
\[
  G \sim \mathrm{DP}(\alpha, G_0), \qquad \Theta_j \overset{iid}{\sim} G.
\]
Here $\alpha>0$ is the concentration parameter controlling expected clustering, and $G_0$ is the base measure encoding prior beliefs about kernel parameters [@ferguson1973].

DPmixGPD supports two equivalent DP representations:

- stick-breaking ("SB") [@sethuraman1994], implemented with a finite truncation at $J$ components [@ishwaran2001];
- Chinese restaurant process ("CRP"), a partition-based formulation [@blackwell1973; @neal2000].

In SB form (truncated at $J$),
\[
  v_j \sim \mathrm{Beta}(1,\alpha),\quad
  w_1=v_1,\quad
  w_j=v_j\prod_{\ell<j}(1-v_\ell)\ (j\ge 2),\quad
  w_J = 1-\sum_{j=1}^{J-1} w_j.
\]

# Tail model: generalized Pareto distribution

For upper-tail modeling, fix a threshold $u$ and define exceedances $x=y-u>0$ for $y>u$. The GPD density for exceedances is
\[
  g(x\mid \sigma,\xi)
  =
  \frac{1}{\sigma}
  \left(1+\xi\frac{x}{\sigma}\right)^{-1/\xi - 1},
  \qquad x>0,\ \sigma>0,
\]
with support constraint $1+\xi x/\sigma>0$. The corresponding CDF is
\[
  G(x\mid\sigma,\xi)
  =
  1-\left(1+\xi\frac{x}{\sigma}\right)^{-1/\xi}
  \quad (\xi\ne 0),
\]
and in the limit $\xi\to 0$,
\[
  G(x\mid\sigma,0)=1-\exp(-x/\sigma).
\]
This GPD limit result for threshold exceedances is central in extreme value theory [@balkema1974; @pickands1975; @coles2001].

# Splicing: a proper bulk--tail distribution

DPmixGPD combines a bulk model $F_{\mathrm{DPM}}$ with a GPD exceedance model while ensuring the resulting full distribution integrates to 1.

Let $u$ denote the tail threshold. Define:
\[
  p_u = F_{\mathrm{DPM}}(u), \qquad \bar p_u = 1 - p_u.
\]
The spliced CDF is defined by:
\[
F(y) =
\begin{cases}
F_{\mathrm{DPM}}(y), & y \le u,\\[6pt]
p_u + \bar p_u \, G(y-u\mid\sigma,\xi), & y>u.
\end{cases}
\]
By construction, $F$ is nondecreasing, right-continuous, and satisfies $\lim_{y\to-\infty}F(y)=0$, $\lim_{y\to\infty}F(y)=1$. The corresponding density is:
\[
f(y) =
\begin{cases}
f_{\mathrm{DPM}}(y), & y \le u,\\[6pt]
\bar p_u \, g(y-u\mid\sigma,\xi), & y>u.
\end{cases}
\]
This "bulk + tail using all observations" Bayesian philosophy matches earlier semiparametric bulk--tail constructions [@frigessi2002; @behrens2004; @doNascimento2012].

# Posterior computation in DPmixGPD

DPmixGPD implements these models through NIMBLE-backed MCMC. At a high level, posterior inference alternates between:

1. allocation updates $z_i$ (component membership),
2. component parameter updates $\Theta_j$,
3. mixture weight updates (SB/CRP),
4. tail updates $(u,\sigma,\xi)$ if `GPD = TRUE`,
5. hyperparameter updates (e.g., $\alpha$) where stochastic.

All user-facing workflows produce a `mixgpd_fit` object (or a causal fit object wrapping two `mixgpd_fit`s), which then supports `predict()`, `fitted()`, and related S3 methods.

# References
