---
title: "Backends"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Backends}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# Two backends, same model

DPmixGPD offers two equivalent representations of the same mixture model:

- **SB (stick‑breaking)**
- **CRP (Chinese restaurant process)**

# SB vs CRP: conceptual comparison

| Aspect | SB | CRP |
|---|---|---|
| Components | fixed truncation $K$ | capped but adaptive occupancy |
| Computation | predictable, vectorized | flexible clustering |
| Interpretation | weights from stick‑breaking | cluster allocation process |

# When to use each

- Use **SB** when runtime and memory predictability matter.
- Use **CRP** when you want adaptive cluster occupancy within the same $K$.

# Computational tradeoffs

- SB often runs faster per iteration.
- CRP can mix better for clustering structure.

# Minimal example (non‑executed)

```{r backend-example}
bundle_sb <- build_nimble_bundle(y = y, kernel = "gamma", backend = "sb", J = 5)
bundle_crp <- build_nimble_bundle(y = y, kernel = "gamma", backend = "crp", J = 5)
```

This vignette is intentionally short. For model building, see [model-spec](model-spec.Rmd).
