---
title: "Backends: SB vs CRP"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Backends: SB vs CRP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 9000, nburnin = 2500, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Goal

Compare the two mixture backends:

- **SB**: stick-breaking (finite truncation), controlled by `components`.
- **CRP**: Chinese Restaurant Process (cluster allocations).

We use the same data and kernel, then compare fitted summaries.

## Data

```{r}
library(DPmixGPD)

n <- 90
y <- abs(rnorm(n)) + 0.2
```

## Fit SB

```{r}
bundle_sb <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit_sb <- run_mcmc_bundle_manual(bundle_sb, show_progress = FALSE)
```

## Fit CRP

```{r}
bundle_crp <- build_nimble_bundle(
  y = y,
  backend = "crp",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit_crp <- run_mcmc_bundle_manual(bundle_crp, show_progress = FALSE)
```

## Compare fitted summaries

```{r}
mean_sb <- predict(fit_sb, type = "mean", cred.level = 0.90, interval = "credible")$fit
mean_crp <- predict(fit_crp, type = "mean", cred.level = 0.90, interval = "credible")$fit

rbind(
  SB  = unlist(mean_sb[1, c("estimate","lower","upper")]),
  CRP = unlist(mean_crp[1, c("estimate","lower","upper")])
)
```

## Practical guidance

- Choose **SB** when you want explicit control over truncation (`components`) and stable computation.
- Choose **CRP** when you want the number of clusters to adapt more directly, at the cost of potentially more label/cluster dynamics.

## Troubleshooting (Backends)

- Compilation errors often point to unsupported kernel/backend combinations or reserved names.
- Start with small `niter` and one chain, then scale up.

```{r sessioninfo, include=FALSE}
sessionInfo()
```
