---
title: "Causal Modeling (Scoped)"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Causal Modeling (Scoped)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# Scope and estimands

DPmixGPD supports **distributional causal estimands** computed from **separately modeled outcome distributions** for treated and control groups. Examples include:

- **CQTE** (conditional quantile treatment effects)
- **Tail risk effects** (differences in extreme quantiles or tail probabilities)

# What is (and is not) identified

- **Identification** is **external** to this package.
- DPmixGPD **does not** guarantee causal validity.
- You are responsible for the design/assumptions that justify causal interpretation.

# Workflow (conceptual)

1. Split by treatment group.
2. Fit an outcome model to each group.
3. Compare the fitted distributions to compute contrasts.

```{r causal-outline}
# Pseudocode only (not executed)
fit0 <- build_nimble_bundle(y = y0, X = X0, kernel = "gamma", GPD = TRUE)
fit1 <- build_nimble_bundle(y = y1, X = X1, kernel = "gamma", GPD = TRUE)
# run MCMC for each, then compare predictive distributions
```

# What the contrasts mean

- **CQTE** compares $Q_1(p) - Q_0(p)$ at quantile level $p$.
- **Tail effects** compare exceedance probabilities or extreme quantiles.

# Responsible framing

This package **models outcomes** and **computes contrasts**. It does not perform identification, design, or sensitivity analysis.

# What this vignette does NOT cover

- Propensity score modeling
- Identification theory
- Heavy simulations

See legacy material in the pkgdown **Legacy** section for extended studies.
