---
title: "Causal workflow (two-arm outcome modeling)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal workflow (two-arm outcome modeling)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
library(nimble)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 700, nburnin = 200, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 12000, nburnin = 3000, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Goal

This vignette shows the causal "two-arm" workflow: fit an outcome model separately (or jointly, if your API supports it) for treated and control arms, then compute a treatment effect estimand.

We focus on *distributional* effects, such as the conditional quantile treatment effect (CQTE):

\[
\text{CQTE}(\tau \mid x) = Q_{Y(1)\mid X}(\tau \mid x) - Q_{Y(0)\mid X}(\tau \mid x).
\]

**Important:** the *difference of two densities is not a density* for the treatment effect. If you want the distribution of \(Y(1)-Y(0)\), simulate posterior predictive draws of \(Y(1)\) and \(Y(0)\) and subtract.

## Simulated data

```{r}
library(DPmixGPD)

n <- 160
x <- rnorm(n)
X <- data.frame(x = x)

# treatment assignment (no propensity score used here)
a <- rbinom(n, 1, 0.5)

# outcome with heterogeneous effect + heavy-ish noise
te <- 0.4 + 0.6 * (x > 0)
y0 <- 0.5 + 0.7 * x + abs(rnorm(n)) + 0.1
y1 <- y0 + te

y <- ifelse(a == 1, y1, y0)
```

## Fit arm-specific models

The simplest approach is to fit two separate models.

```{r}
X0 <- X[a == 0, , drop = FALSE]
X1 <- X[a == 1, , drop = FALSE]
y0_obs <- y[a == 0]
y1_obs <- y[a == 1]

bundle0 <- build_nimble_bundle(
  y = y0_obs,
  X = X0,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

bundle1 <- build_nimble_bundle(
  y = y1_obs,
  X = X1,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit0 <- run_mcmc_bundle_manual(bundle0, show_progress = FALSE)
fit1 <- run_mcmc_bundle_manual(bundle1, show_progress = FALSE)
```

## CQTE on a covariate grid

```{r}
new_X <- data.frame(x = seq(min(x), max(x), length.out = 25))

taus <- c(0.1, 0.5, 0.9)

cqte_list <- lapply(taus, function(tau) {
  q0 <- predict(fit0, x = new_X, type = "quantile", index = tau, cred.level = 0.90, interval = "credible")$fit
  q1 <- predict(fit1, x = new_X, type = "quantile", index = tau, cred.level = 0.90, interval = "credible")$fit

  data.frame(
    x = new_X$x,
    tau = tau,
    estimate = q1$estimate - q0$estimate,
    lower = q1$lower - q0$upper,
    upper = q1$upper - q0$lower
  )
})

cqte <- do.call(rbind, cqte_list)
head(cqte)
```

```{r}
plot(cqte$x[cqte$tau == 0.5], cqte$estimate[cqte$tau == 0.5], type = "l",
     xlab = "x", ylab = "CQTE", main = "Conditional quantile treatment effect")
lines(cqte$x[cqte$tau == 0.1], cqte$estimate[cqte$tau == 0.1])
lines(cqte$x[cqte$tau == 0.9], cqte$estimate[cqte$tau == 0.9])
legend("topleft", legend = paste0("tau=", taus), lty = 1, bty = "n")
```

## Notes and troubleshooting

- If you later add propensity score adjustment, keep it explicit and documented (it changes assumptions and estimands).
- Arm-specific fits are simple and transparent; a joint causal model can share priors/structure for efficiency.

```{r sessioninfo, include=FALSE}
sessionInfo()
```
