---
title: "CausalMixGPD: Bayesian Nonparametric Models with Optional Heavy-Tail Augmentation and Causal Extensions"
output: rmarkdown::html_vignette
bibliography: CausalMixGPD_JSS_refs.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{CausalMixGPD: Bayesian Nonparametric Models with Optional Heavy-Tail Augmentation and Causal Extensions}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE, eval=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4,
  out.width = "80%",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

```


```{r, include=FALSE, eval=TRUE}
library(CausalMixGPD)
library(ggplot2)

mcmc_fixed <- list(
  niter = 2000,
  nburnin = 500,
  thin = 3,
  nchains = 3,
  seed = 1
)
```
# Introduction

Causal inference is often summarized through average effects, but many
applications require distributional comparisons between treatment
regimes. In clinical trials, investigators may want to understand how a
treatment changes outcomes across the outcome scale, including effects
among individuals in the upper tail where events are rare but
consequences can be substantial. This shift from mean-based to
distribution-based causal questions makes causal analysis tightly linked
to how well we can estimate (and predict) outcome distributions under
competing treatment assignments.

A growing R ecosystem supports Bayesian causal inference
through flexible outcome modeling and posterior uncertainty
quantification. For example,
[bcf](https://cran.r-project.org/package=bcf) implements
Bayesian Causal Forests for binary treatments and continuous outcomes,
providing posterior inference for both average and heterogeneous
treatment effects [@bcf_pkg]. The package
[bartCause](https://cran.r-project.org/package=bartCause)
builds causal estimators around Bayesian additive regression trees,
using BART as a regression engine for treatment-response modeling and
causal effect estimation [@bartCause_pkg; @hill2011bart].
Complementing tree-based approaches,
[BCEE](https://cran.r-project.org/package=BCEE) implements
a Bayesian model-averaging strategy for causal effect estimation with
support for binary or continuous exposures and outcomes
[@BCEE_pkg; @talbot2015bcee]. A recurring practical difficulty in
quantile-based causal reporting is the quantile crossing issue, where
quantile curves at different probability levels can cross when they are
estimated separately. To avoid this, `CausalMixGPD` directly models
outcome densities and derives quantiles and other functionals from the
posterior predictive distributions.

In the context of density estimation and conditional density estimation, Dirichlet Process Mixtures (DPM) provide a
versatile nonparametric solution because they can capture multimodality and latent heterogeneity
without fixing the number of mixing components in advance. In the experimental R
ecosystem, Bayesian nonparametric mixture modeling is
well supported by software based on Dirichlet process (DP) ideas. For example,
[dirichletprocess](https://cran.r-project.org/package=dirichletprocess)
provides a flexible interface for specifying DPMs and
fitting them by posterior simulation, yielding posterior inference for
mixture weights, component labels, and predictive densities under
user-chosen kernels [@dirichletprocess_pkg]. The package
[DPpackage](https://cran.r-project.org/package=DPpackage)
offers a broad suite of Bayesian semi- and nonparametric models in
R, including DP-based formulations for marginal and conditional
densities, regression functions, and model-based predictive
distributions [@jara2011DPpackage]. More recently,
[BNPmix](https://cran.r-project.org/package=BNPmix)
emphasizes practical Bayesian nonparametric mixture approaches,
providing posterior inference for clustering structure and
density/regression estimation under DP and Pitman--Yor-type mixture
specifications [@canale2021BNPmix].

That clustering perspective is also important for the present package.
Because DP mixtures induce random partitions through latent allocation
variables, they naturally support model-based subgroup discovery in
addition to density estimation. Bayesian cluster analysis is typically
reported through label-invariant summaries such as pairwise
co-clustering probabilities [@Binder1978] and representative
partitions selected by Dahl's least-squares criterion
[@Dahl2006]. In regression settings, clustering itself can depend
on predictors: dependent Dirichlet process constructions allow
covariates to affect mixture weights, component-specific parameters, or
both [@MacEachern1999,MacEachern2000; @DeIorio2004; @Quintana2022],
and predictor-dependent stick-breaking schemes provide a practical route
for covariate-modulated cluster prevalence [@Ren2011]. Existing
software supports parts of this workflow, but it typically focuses on
bulk density estimation and clustering under a single kernel family
without an explicit adjustment for heavy tailed data. As a result,
applications that require both latent subgroup discovery and
tail extrapolation still need additional modeling outside the standard
Bayesian nonparametric mixture workflow.

However, it is important to note that estimating values far in the tail requires special care when data becomes scarce.
Additionally, using a single model in the presence of heavy tail data can distort the fit in the central region.
In such cases extreme value theory (EVT) provides a principled framework for
modeling and extrapolation of tail behavior~[@de2007extreme]. EVT characterizes the
asymptotic distribution of threshold exceedances, which can be used to
estimate tail probabilities and quantiles beyond the range of observed data~[@balkema1974residual].
Bayesian implementations of EVT are available in
R when posterior uncertainty quantification for tail
quantities are required. For example,
[evdbayes](https://cran.r-project.org/package=evdbayes)
provides Bayesian inference for standard extreme-value models via
posterior simulation, enabling posterior and posterior-predictive
summaries for extreme-event quantities [@evdbayes_pkg]. For
peaks-over-threshold (POT) approaches, the package
[POT](https://cran.r-project.org/package=POT) implements a
focused collection of tools for GPD modeling and threshold-based tail
analysis [@POT_pkg]. Finally, for extreme quantile estimation and
rare-event tail summaries,
[extremefit](https://cran.r-project.org/package=extremefit)
provides a published software implementation targeted specifically at
extreme-quantile estimation [@durrieu2018extremefit].

This article introduces an R package (`CausalMixGPD`) that
connects Bayesian nonparametric modeling with EVT, clustering,
and robust causal analyses in the presence of covariates (confounders).
The package provides three simultaneous capabilities:

- **Bayesian nonparametric (BNP) density modeling and prediction with optional GPD tail:**
   `CausalMixGPD` fits DPMs using a user-chosen kernel below a certain threshold (bulk part),
   and can optionally splice a generalized Pareto tail above the threshold (tail part) to enable the estimation of tail quantities
   and posterior predictive inference for rare-event summaries.
- **Clustering under predictor dependent mixtures:**
   A dedicated clustering interface for both bulk-only and spliced models,
   supporting weight dependence, parameter dependence, or both, and returning
   label-invariant clustering summaries through posterior similarity matrices
   and Dahl representative partitions [@Binder1978; @Dahl2006].
- **Causal estimation and prediction for randomized controlled trials (RCTs) and observational studies:**
   The final goal of this package is to provide causal estimands by modeling treatment-specific potential outcome
   densities and reporting contrasts as functionals of the posterior over densities. Framework supports both randomized and observational settings, with the latter using a flexible Bayesian propensity score model to adjust for confounding.

Overall, `CausalMixGPD` offers a unified workflow for
density estimation, prediction, and causal reporting when
tail behavior is of significance. The package is implemented in R with computationally intensive components
coded in C++ and integrated via the `nimble` framework for model specification and posterior simulation [@nimble_pkg].
The remainder of this article is organized as follows. The Background and
Preliminaries section introduces the key modeling components, including DPMs,
GPDs, and the spliced bulk-tail construction. The Package Overview section
describes the package's main functions and user interface. The two Data
analysis sections illustrate clustering and causal inference workflows on R
datasets. Finally, the Discussion section concludes with limitations and future
directions.

# Package Overview: Workflow, Model, and Inference

This section describes the main workflow provided by `CausalMixGPD`
for one-arm conditional modeling and prediction when a generalized
Pareto tail is spliced above a covariate-dependent threshold. The causal
extension reuses the same modeling, estimation, and prediction machinery
after introducing treatment-specific outcome models and a propensity
score component; we defer those additions to the causal section.

## Augmented DPM--GPD model, likelihood, and priors

Let the observed data be $$
  \mathcal{D}=\{(y_i,x_i)\}_{i=1}^n,
$$

```{r overview-data-reload, echo=TRUE, eval=FALSE}
data("nc_posX100_p3_k2", package = "CausalMixGPD")
dat <- data.frame(y = nc_posX100_p3_k2$y, nc_posX100_p3_k2$X)
```
with $x_i\in\mathbb{R}^p$ denoting the covariate vector (including an
intercept if desired). The target conditional distribution
$F(\cdot\mid x)$ and its spliced DPM--GPD construction are defined in
the Background and Preliminaries section, with the spliced CDF and density given in
the Background and Preliminaries section. Posterior computation in
`CausalMixGPD` approximates the bulk DPM by a finite truncation with
$J$ mixture components, $$
  f_{\mathrm{DP}}(y\mid x;\Theta)\;\approx\;\sum_{j=1}^{J} w_j\,k(y\mid x;\theta_j),
  \qquad j=1,\ldots,J,
$$ where $j$ indexes mixture components, $w_j\ge 0$, and
$\sum_{j=1}^J w_j=1$. Kernel parameters can be constant or
covariate-dependent through link specifications of the form

$$
  g\!\left(\psi_j(x)\right)=x^\top \beta_{j,\psi},
$$

where $g(\cdot)$ enforces parameter support (e.g.,
positive scale parameters).

```{r overview-mcmc-settings, echo=TRUE, eval=FALSE, results='hide'}
mcmc_fixed
```
```{r overview-fit-model, echo=TRUE, eval=FALSE}
fit <- dpmgpd(
  formula = y ~ x1 + x2 + x3,
  data = dat,
  backend = "sb",
  kernel = "lognormal",
  components = 5,
  mcmc = mcmc_fixed
)

```
```{r overview-fit-model-cache, echo=FALSE, eval=FALSE, include=FALSE}
# no-op: cached artifact loading removed
```
```{r overview-fit-model-results, echo=FALSE, eval=FALSE, include=FALSE}
if (!exists("fit", inherits = FALSE)) {
  fit <- dpmgpd(
    formula = y ~ x1 + x2 + x3,
    data = dat,
    backend = "sb",
    kernel = "lognormal",
    components = 5,
    mcmc = mcmc_fixed
  )
}
print(fit)
```
For the spliced model, let $u(x)$ be the threshold, $\sigma(x)>0$ the
tail scale, and $\xi\in\mathbb{R}$ the tail shape (as in
the spliced-model subsection). Define
$\delta_i=\mathbf{1}\{y_i>u(x_i)\}$. Using the spliced density
with the truncated bulk mixture, the likelihood is
$$
  L(\Theta,\Phi;\mathcal{D})
  =\prod_{i=1}^n f(y_i\mid x_i;\Theta,\Phi),
$$
where $\Theta$ collects the truncation-level bulk
parameters $\{w_j,\theta_j\}_{j=1}^J$, and $\Phi$ collects the tail
parameters governing $u(x)$, $\sigma(x)$, and $\xi$. Equivalently, the
log-likelihood can be written as
$$
  \log L(\Theta,\Phi;\mathcal{D})
  \;=&\;\sum_{i=1}^n (1-\delta_i)\,\log f_{\mathrm{DP}}(y_i\mid x_i;\Theta)
  \nonumber\\
  &\;+\;\sum_{i=1}^n \delta_i
  \Bigl[
    \log\!\bigl\{1-F_{\mathrm{DP}}(u(x_i)\mid x_i;\Theta)\bigr\}
    +\log f_{\mathrm{GPD}}\!\bigl(y_i\mid x_i;\Phi\bigr)
  \Bigr],
$$
where $f_{\mathrm{GPD}}$ is the GPD density and
$F_{\mathrm{DP}}(\cdot\mid x)$ is the bulk mixture CDF induced by the
truncated mixture.

Priors follow the decomposition implied by the bulk-tail factorization.
Under the SB backend, the
truncation-level weights are generated through Beta stick breaks
from the stick-breaking construction, which are represented in
`nimble` using the Beta distribution in model code (see
`dbeta`). Under the CRP backend, the
latent-cluster-label representation is implemented using the CRP
distribution in `nimble` (see `dCRP`), matching the predictive
representation of the CRP. Kernel parameters and tail parameters
are assigned weakly informative priors, with regression coefficients
$\beta_{j,\psi}$ arising through the link representation
described above. The same link construction is used for
covariate-dependent $u(x)$ and $\sigma(x)$ when those are modeled as
functions of $x$.

## MCMC algorithm, convergence diagnostics, and parameter extraction

Posterior simulation targets
$p(\Theta,\Phi\mid \mathcal{D}) \propto L(\Theta,\Phi;\mathcal{D})\,p(\Theta,\Phi)$
using the likelihood above and the prior structure
implied by the SB or CRP representation. Given a model specification,
`CausalMixGPD` runs MCMC through `nimble`
[@devalpine2017nimble; @nimble_pkg] under the chosen backend; SB uses
Beta stick breaks in model code (see `dbeta`) and CRP uses the CRP
distribution for latent cluster labels (see `dCRP`). Users control the number
of iterations, burn-in, thinning, and number of chains through the
`mcmc` argument in the wrappers.

Convergence checking is supported through standard multi-chain
diagnostics and trace-based visualizations. The model object supports
`summary()` for posterior summaries, `plot()` for diagnostic
plots (trace, running mean, autocorrelation, and multi-chain diagnostics
such as $\widehat{R}$ when multiple chains are used), and
`params()` as a lightweight extractor returning posterior mean
parameters reshaped to natural dimensions. Interval summaries used by
`predict()` can be chosen as equal-tailed credible intervals or HPD
intervals as described in the Posterior functionals subsection
[@Gelman2013BDA; @coda_pkg]. Diagnostic plotting relies on
established MCMC diagnostic tooling [@ggmcmc_pkg].

```{r overview-mcmc-diagnostics, echo=TRUE, eval=FALSE}
print(fit)
p_hat <- params(fit)
p_hat
```
```{r overview-mcmc-diagnostics-cache, echo=FALSE, eval=FALSE, include=FALSE}
# no-op: cached artifact loading removed
```
```{r overview-mcmc-diagnostics-results, echo=FALSE, eval=FALSE, include=FALSE}
if (!exists("p_hat", inherits = FALSE)) {
  p_hat <- params(fit)
}
print(fit)
p_hat
```
### Bulk-only special case (no GPD tail)

Setting `GPD = FALSE` removes the splicing step so that
$f(y\mid x;\Theta)\equiv f_{\mathrm{DP}}(y\mid x;\Theta)$, and
$L(\Theta,\Phi;\mathcal{D})$ reduces to the bulk-mixture likelihood in
$\Theta$. All estimation and prediction targets described below remain
available, with tail-specific terms omitted. The wrapper `dpmix()`
fits this bulk-only model, while `dpmgpd()` fits the augmented
(spliced) model.

## Posterior functionals, interval summaries and prediction

The estimands reported by `CausalMixGPD` are functionals of the
estimated conditional distribution $F(\cdot\mid x)$ from
the Background and Preliminaries section. In this article we focus on four conditional
functionals: density $f(y\mid x)$, survival $S(y\mid x)=1-F(y\mid x)$,
quantiles $Q(\tau\mid x)$, and mean-type summaries.

Let $(\Theta^{(s)},\Phi^{(s)})_{s=1}^S$ denote post-burn-in draws from
the posterior. For density, survival, and quantiles, `CausalMixGPD`
evaluates the corresponding draw-wise functional using the spliced
formulas for the spliced CDF, density, and quantile mapping (with
the truncated bulk mixture), and then summarizes across draws.
Specifically, for a generic scalar functional $T(\Theta,\Phi;x)$ (e.g.,
$f(y_0\mid x)$ or $Q(\tau\mid x)$), $$
  \widehat{T}(x)=\frac{1}{S}\sum_{s=1}^S T(\Theta^{(s)},\Phi^{(s)};x),
$$ and uncertainty is summarized either by equal-tailed credible
intervals (posterior quantiles) or by highest posterior density (HPD)
intervals (shortest intervals covering the target posterior mass).
Equal-tailed intervals and general Bayesian posterior summaries follow
standard practice [@Gelman2013BDA]; HPD intervals are computed from
the MCMC draws using established implementations [@coda_pkg].

Mean-type summaries are handled differently in the package. For
`type = "mean"` and `type = "rmean"`, `CausalMixGPD` uses
posterior predictive Monte Carlo within each posterior draw rather than
direct plug-in evaluation. For a fixed covariate value $x$, within draw
$s$ the package generates $$
  y_{1}^{(s)},\ldots,y_{M}^{(s)} \;\sim\; f(\cdot\mid x;\Theta^{(s)},\Phi^{(s)}),
$$ and computes $$
  \mu^{(s)}(x)=\frac{1}{M}\sum_{m=1}^M y_{m}^{(s)},
  \qquad
  \mu_{c}^{(s)}(x)=\frac{1}{M}\sum_{m=1}^M \min\{y_{m}^{(s)},c\},
$$ corresponding to the conditional mean and the restricted mean at
cutoff $c$, respectively. Posterior summaries and intervals are then
formed from $\{\mu^{(s)}(x)\}_{s=1}^S$ or $\{\mu_c^{(s)}(x)\}_{s=1}^S$.
Under the spliced model, the conditional mean is infinite when
$\xi\ge 1$; accordingly, `CausalMixGPD` flags `type = "mean"`
as infinite when there is posterior mass with $\xi\ge 1$, while
`type = "rmean"` remains finite by construction.

Using the same fitted model `fit` from
the Augmented DPM--GPD model subsection, we now compute posterior summaries
on the original training covariates (default behavior of
`predict()` when `newdata` is not supplied).

```{r overview-estimation-same-fit, echo=TRUE, eval=FALSE}
q_grid <- c(0.50, 0.90, 0.95, 0.99)
x_eval <- dat[1:20, c("x1", "x2", "x3")]

est_quant <- predict(fit, newdata = x_eval, type = "quantile", index = q_grid,
                     interval = "hpd", level = 0.95)
```
```{r overview-estimation-same-fit-cache, echo=FALSE, eval=FALSE, include=FALSE}
# no-op: cached artifact loading removed
```
```{r overview-estimation-same-fit-results, echo=FALSE, eval=FALSE, include=FALSE}
head(as.data.frame(est_quant$fit), 4)
```
For a new covariate value $x^\star$, prediction is based on the
posterior predictive distribution $$
  p(y^\star\mid x^\star,\mathcal{D})
  =\int f(y^\star\mid x^\star;\Theta,\Phi)\,p(\Theta,\Phi\mid \mathcal{D})\,d\Theta\,d\Phi.
$$ The `predict()` method returns posterior summaries for the same
functional targets as in the current posterior-functionals subsection. For
`type = "density"` and `type = "survival"`, the user supplies
a grid of $y$ values and the method returns pointwise posterior means
with interval bounds computed draw-by-draw. For
`type = "quantile"`, the user supplies quantile indices $\tau$ and
the method returns posterior summaries of $Q(\tau\mid x^\star)$ computed
using the spliced quantile representation in
. For `type = "mean"` and
`type = "rmean"`, prediction uses posterior predictive Monte Carlo
as described above, with $x$ replaced by $x^\star$ and with
user-controlled $M$ through `nsim_mean`.

```{r overview-predict-grid, echo=TRUE, eval=FALSE}
q_levels <- c(0.25, 0.50, 0.75)
x_new <- as.data.frame(lapply(
  dat[, c("x1", "x2", "x3")],
  quantile,
  probs = q_levels,
  na.rm = TRUE
))
rownames(x_new) <- c("q25", "q50", "q75")
y_grid <- seq(0, 10, length.out = 200)
p_grid <- c(0.50, 0.90, 0.95, 0.99)
```
```{r overview-predict-density, echo=TRUE, eval=FALSE}
pdens <- predict(fit, newdata = x_new, y = y_grid, type = "density",
                 interval = "credible", level = 0.95)
```
```{r overview-predict-survival, echo=TRUE, eval=FALSE}
psurv <- predict(fit, newdata = x_new, y = y_grid, type = "survival",
                 interval = "credible", level = 0.95)
```
```{r overview-predict-quantile, echo=TRUE, eval=FALSE}
pquant <- predict(fit, newdata = x_new, type = "quantile", index = p_grid,
                  interval = "hpd", level = 0.95)
```
```{r overview-predict-cache, echo=FALSE, eval=FALSE, include=FALSE}
# no-op: cached artifact loading removed
```
```{r overview-predict-grid-results, echo=FALSE, eval=FALSE, include=FALSE}
x_new
head(y_grid, 4)
p_grid
```
```{r overview-predict-density-results, echo=FALSE, eval=FALSE, include=FALSE}
head(as.data.frame(pdens$fit), 4)
```
```{r overview-predict-survival-results, echo=FALSE, eval=FALSE, include=FALSE}
head(as.data.frame(psurv$fit), 4)
```
```{r overview-predict-quantile-results, echo=FALSE, eval=FALSE, include=FALSE}
head(as.data.frame(pquant$fit), 4)
```
## Clustering extension

The mixture modeling framework introduced earlier induces a latent partition
of the observations through the latent cluster labels $z_i$. While the primary
modeling workflow of the package focuses on estimating conditional densities
and causal functionals, the clustering extension instead treats this latent
partition as the main object of interest. The package therefore provides a
dedicated clustering interface that fits the same mixture models but returns
clustering summaries derived from the posterior distribution of the latent
cluster labels.

The clustering workflow mirrors the design used throughout the package.
Users specify the response variable, optional covariates, the mixture kernel,
and whether the model should use the bulk-only or the spliced likelihood.
The resulting model is then fitted using the same MCMC framework used for the
conditional and causal models. Consequently, clustering models share the
same prior specification, kernel choices, and computational infrastructure as
the rest of the package.

Two clustering wrappers are provided. The first constructs mixture models
using only the bulk kernel distribution. The second uses the spliced
bulk--tail likelihood introduced earlier, allowing the clustering structure
to account for extreme observations through the generalized Pareto tail.
Both interfaces accept the same model specification arguments as the main
modeling functions, ensuring that clustering models can be constructed using
the same syntax and options.

Three forms of covariate dependence are supported in clustering models.
First, covariates may influence the mixture weights while the component
distributions remain invariant across observations. In this configuration,
clusters represent subpopulations whose prevalence varies with predictors.
Second, the mixture weights may remain global while the component parameters
depend on covariates through the same link structures used in the conditional
density models. In this case clusters represent groups with distinct
covariate-response relationships. Third, both the mixture weights and the
component parameters may depend on covariates simultaneously, allowing
clusters to capture both changing prevalence and covariate-dependent
distributional behavior.

Posterior inference for clustering models uses the same sampling procedure as
the main mixture models, jointly updating the component parameters and the
latent cluster labels. Let
$z^{(s)}=\bigl(z_1^{(s)},\ldots,z_n^{(s)}\bigr)$, $s=1,\ldots,S$, denote the
retained MCMC draws of the latent cluster-label vector after burn-in and thinning.
Because mixture components are only identified up to label permutation,
`CausalMixGPD` reports label-invariant clustering summaries based on
pairwise co-clustering probabilities and a representative partition
[@Binder1978; @Dahl2006].

The primary uncertainty summary is the posterior similarity matrix (PSM),
computed entrywise as
$$
  \widehat S_{i\ell}
  =
  \frac{1}{S}\sum_{s=1}^S \mathbf{1}\!\left\{z_i^{(s)} = z_\ell^{(s)}\right\},
  \qquad i,\ell=1,\ldots,n.
$$
Thus, $\widehat S_{i\ell}$ is the posterior probability that observations $i$
and $\ell$ belong to the same cluster.

A single representative partition is then extracted by selecting the sampled
partition whose adjacency matrix is closest to the PSM under squared loss
[@Dahl2006]:
$$
  s^\star
  =
  \arg\min_{1 \le s \le S}
  \sum_{i=1}^n \sum_{\ell=1}^n
  \left(
    \mathbf{1}\!\left\{z_i^{(s)} = z_\ell^{(s)}\right\}
    - \widehat S_{i\ell}
  \right)^2.
$$
The representative labels are $\hat z_i = z_i^{(s^\star)}$, relabeled to
consecutive integers $1,\ldots,K$, and the reported cluster sizes are
$n_k = \sum_{i=1}^n \mathbf{1}\!\left\{\hat z_i = k\right\}$.

When `return_scores = TRUE`, the package also returns a training
membership score matrix. If $C_k=\{\ell:\hat z_\ell = k\}$ denotes cluster $k$
in the representative partition, the unnormalized score for observation $i$ and
cluster $k$ is
$$
  q_{ik}
  =
  \frac{1}{|C_k|}
  \sum_{\ell \in C_k} \widehat S_{i\ell},
$$
and the normalized training score is
$$
  p_{ik}
  =
  \frac{q_{ik}}{\sum_{h=1}^K q_{ih}}.
$$
The hard representative label is $\arg\max_{1 \le k \le K} p_{ik}$, while a
simple cluster-certainty summary is $\max_{1 \le k \le K} p_{ik}$.

For a new observation $(y_i^{\mathrm{new}}, x_i^{\mathrm{new}})$, the package
first forms within-draw component scores
$$
  r_{ij}^{(s)}
  \propto
  \omega_j^{(s)}\!\left(x_i^{\mathrm{new}}\right)
  f_j^{(s)}\!\left(y_i^{\mathrm{new}} \mid x_i^{\mathrm{new}}\right),
  \qquad j=1,\ldots,J,
$$
where $\omega_j^{(s)}(x)$ is the draw-specific gating weight when that quantity
is available; otherwise the implementation falls back to sampled global weights,
when present, or to the empirical component frequencies in draw $s$. For
occupied components, these scores are mapped to representative clusters through
$$
  m_{jc}^{(s)}
  =
  \frac{1}{n_j^{(s)}}
  \sum_{\ell:\,z_\ell^{(s)}=j}
  \mathbf{1}\!\left\{\hat z_\ell = c\right\},
  \qquad c=1,\ldots,K,
$$
and then averaged over posterior draws:
$$
  p_{ic}^{\mathrm{new}}
  =
  \frac{1}{S}
  \sum_{s=1}^S \sum_{j=1}^J r_{ij}^{(s)} m_{jc}^{(s)}.
$$
Empty sampled components are assigned a uniform map over the $K$ representative
clusters. These are the quantities returned internally by
`predict.dpmixgpd_cluster_fit()`.
After row normalization, the predicted cluster label is
$\arg\max_{1 \le c \le K} p_{ic}^{\mathrm{new}}$.

The clustering workflow is illustrated below using one of the example data
sets included with the package.

```{r cluster_example_fit, eval=FALSE}
data("nc_realX100_p3_k2", package = "CausalMixGPD")

dat_cl <- data.frame(y = nc_realX100_p3_k2$y, nc_realX100_p3_k2$X)

fit_cluster <- dpmix.cluster(
  y ~ x1 + x2 + x3,
  data       = dat_cl,
  kernel     = "normal",
  type       = "param",
  components = 8,
  mcmc = mcmc_fixed
)
```
Posterior clustering summaries can be obtained directly from the fitted
object. The following code returns the PSM, the representative training
partition, the training cluster sizes, and the associated certainty scores.

```{r cluster_psm_example, eval=FALSE}
cluster_psm   <- predict(fit_cluster, type = "psm")
cluster_train <- predict(fit_cluster, type = "label", return_scores = TRUE)

cluster_sizes <- data.frame(
  cluster = names(summary(cluster_train)$cluster_sizes),
  n = as.integer(summary(cluster_train)$cluster_sizes)
)
cluster_certainty <- apply(cluster_train$scores, 1, max)
```
Cluster assignments for new observations are obtained from the posterior
predictive score matrix in the same interface.

```{r cluster_prediction_example, eval=FALSE}
cluster_new <- predict(
  fit_cluster,
  newdata = dat_cl,
  type    = "label",
  return_scores = TRUE
)

cluster_new_certainty <- apply(cluster_new$scores, 1, max)
```
The same workflow applies to the spliced mixture model, which incorporates
the generalized Pareto tail component when clustering data with heavy-tailed
behavior.

```{r cluster_spliced_example, eval=FALSE}
fit_cluster_spliced <- dpmgpd.cluster(
  y ~ x1 + x2 + x3,
  data       = dat_cl,
  kernel     = "normal",
  type       = "both",
  components = 8,
  mcmc = mcmc_fixed
)
```
## Causal extension and propensity score augmentation

This package extends the one-arm conditional mixture regression
framework to two-arm causal studies by fitting an outcome model under
treatment and under control, optionally augmenting the outcome
covariates with a PS summary. Throughout this section
we write the model in its default *augmented spliced* form (bulk
mixture + GPD tail), using the same conditional distributional building
blocks introduced earlier (so we do not restate the spliced density; see
 and the corresponding conditional functionals
defined in the previous sections).

### Data, model components, and likelihood factorization

Let $$
  \mathcal{D}=\{(y_i,a_i,\boldsymbol{x}_i)\}_{i=1}^n,
  \qquad a_i\in\{0,1\},
$$ where $a_i=1$ denotes treatment and $a_i=0$ denotes control. The
causal interface consists of two pieces.

First, when PS augmentation is enabled, a binary regression model is
used for the treatment mechanism $
  a_i \mid \boldsymbol{x}_i, \boldsymbol{\gamma}
  &\sim \mathrm{Bernoulli}\big(e_{\boldsymbol{\gamma}}(\boldsymbol{x}_i)\big),
  h\!\left(e_{\boldsymbol{\gamma}}(\boldsymbol{x}_i)\right)
  &= \gamma_0 + \boldsymbol{x}_i^\top\boldsymbol{\gamma},
  \qquad h\in\{\text{logit},\text{probit}\},
$ which is consistent with the definition and balancing
property in --. The package also
supports a `"naive"` option, in which $e(\boldsymbol{x})$ is
taken to be constant (no covariate dependence).

Second, the outcome model is fit separately within each arm. Denote by
$F_a(\,\cdot\mid\boldsymbol{r})$ the fitted conditional outcome
distribution under arm $a\in\{0,1\}$, where $\boldsymbol{r}$ is the
covariate vector used by the outcome model. With PS augmentation, the
package constructs an *augmented* covariate vector $
  \boldsymbol{r}(\boldsymbol{x})
  =
  \Big(1,\ \boldsymbol{x}^\top,\ \psi\big(\widehat e(\boldsymbol{x})\big)\Big)^\top,
$ where $\widehat e(\boldsymbol{x})$ is a posterior summary
of the PS (posterior mean or median, controlled by
`ps_summary`), and $\psi(\cdot)$ is either
$\psi(p)=\text{logit}(p)$ or $\psi(p)=p$ (controlled by
`ps_scale`). When PS augmentation is disabled,
$\boldsymbol{r}(\boldsymbol{x})=(1,\boldsymbol{x}^\top)^\top$.

Within each arm, the conditional density $f_a(y\mid \boldsymbol{r})$
follows the same mixture regression + optional GPD tail specification
described in the one-arm section (kernel choice, link structure
$g(\cdot)$ for covariate-dependent parameters, and either `"sb"`
or `"crp"` backend for the DPM). Let
$\boldsymbol{\Theta}_a$ collect the arm-specific mixture and tail
parameters. The arm-wise likelihood contributions are $$
  L_a(\boldsymbol{\Theta}_a;\mathcal{D})
  =
  \prod_{i:\,a_i=a} f_a\!\left(y_i \mid \boldsymbol{r}(\boldsymbol{x}_i);
  \boldsymbol{\Theta}_a\right),
  \qquad a\in\{0,1\},
$$ with $f_a(\cdot\mid\cdot)$ given by the spliced construction already
defined (bulk mixture below threshold and GPD exceedances above
threshold). Practically, the implementation uses a two-stage workflow:
(i) fit the PS model using
--, then (ii)
compute $\widehat e(\boldsymbol{x}_i)$ and build
$\boldsymbol{r}(\boldsymbol{x}_i)$ in
 before fitting the two
arm-specific outcome models.

### Prior specification (PS then outcome)

For the PS regression coefficients, the default prior is independent
Normal, $$
  \gamma_0 \sim \mathcal{N}(m_0,s_0^2),
  \qquad
  \gamma_k \sim \mathcal{N}(m_0,s_0^2),
$$ with defaults controlled by `ps_prior`.

For each arm's outcome model, priors follow the same structure as the
one-arm case: a DPM prior (with either truncated stick-breaking
or a CRP latent-cluster-label representation, implemented in `nimble`
[@devalpine2017nimble; @nimble_pkg]) and kernel-/tail-specific priors
on the bulk and GPD parameters. Since the causal fit is simply a pair of
one-arm fits (treated and control), all customization options for
kernels, links, and tail inclusion apply arm-wise.
For quick lookup of these controls and their code-level defaults, see
the supplementary workflow maps (Figures~fig:supp-workflow-map
and~fig:supp-advanced-map) and option matrices in
Tables~tab:supp-core-controls--tab:supp-advanced-controls.

### MCMC algorithm, diagnostics, and parameter extraction

The causal fit runs MCMC in two stages, each with its own iteration
controls.

(i) PS stage: posterior simulation for $\boldsymbol{\gamma}$ in
  the propensity-score model using
    `mcmc_ps` settings. The PS posterior draws are summarized to
    obtain $\widehat e(\boldsymbol{x}_i)$, which is optionally
    transformed via `ps_scale` and appended to the outcome
  covariates.

(ii) Outcome stage: arm-specific posterior simulation for
     $\boldsymbol{\Theta}_0$ and $\boldsymbol{\Theta}_1$ using
     `mcmc_outcome` settings, with either `"sb"` or
     `"crp"` backend.

For routine convergence checks, users can inspect trace plots and
autocorrelation for monitored parameters and (when running multiple
chains) compare between-chain and within-chain mixing. The package
stores MCMC samples in the fitted object, and standard diagnostics can
be applied after converting samples to `coda` objects (e.g.,
effective sample size, Geweke diagnostics, and HPD intervals via
`coda::HPDinterval`; see the `coda` reference at the end
of this message). Posterior parameter summaries can be extracted via the
exported `params()` method, which returns arm-specific parameter
tables for causal fits.

### Estimation and interval summaries for causal contrasts

All reported causal quantities are computed as functionals of the fitted
arm-specific conditional distributions. Let $T_a(\boldsymbol{x})$ denote
any conditional functional under arm $a$ (density, survival, quantile,
or mean/restricted mean), computed from
$F_a(\cdot\mid\boldsymbol{r}(\boldsymbol{x}))$ using the same
definitions as in the one-arm case. The corresponding conditional causal
contrast is $$
  \Delta_T(\boldsymbol{x}) = T_1(\boldsymbol{x}) - T_0(\boldsymbol{x}).
$$ The package computes $\Delta_T(\boldsymbol{x})$ *draw-by-draw*
from posterior samples, then summarizes $\Delta_T(\boldsymbol{x})$ by a
posterior mean/median and uncertainty intervals. Equal-tailed credible
intervals are formed by posterior quantiles, while HPD intervals are
computed using `coda::HPDinterval`.

The main causal estimators provided by the package are:

(a) Conditional effects at covariate profiles. For a user-supplied set
    of covariate rows $\{\boldsymbol{x}^\star_m\}_{m=1}^M$, $$
      \mathrm{CATE}(\boldsymbol{x}^\star_m)=\mu_1(\boldsymbol{x}^\star_m)-\mu_0(\boldsymbol{x}^\star_m),
      \qquad
      \mathrm{CQTE}(\tau\mid \boldsymbol{x}^\star_m)
      =
      Q_1(\tau\mid \boldsymbol{x}^\star_m)-Q_0(\tau\mid \boldsymbol{x}^\star_m),
    $$ where
    $\mu_a(\boldsymbol{x})=\mathbb{E}(Y\mid A=a,\boldsymbol{X}=\boldsymbol{x})$
    and $Q_a(\tau\mid\boldsymbol{x})$ is the conditional quantile
    function. In heavy-tailed settings, $\mu_a(\boldsymbol{x})$ can be
    unstable when the implied tail shape yields infinite mean; the
    package therefore supports both `type="mean"` and
    `type="rmean"` for (restricted) mean-based effects, with
    restricted means computed as
    $\mathbb{E}[\min\{Y,c\}\mid A=a,\boldsymbol{X}=\boldsymbol{x}]$ at
    user-specified cutoff $c$.

(b) Marginal (standardized) summaries. The package reports marginal ATE
    and marginal QTE-style summaries by empirical standardization over
covariate profiles:
$$
      \mathrm{ATE}
      &= \frac{1}{n}\sum_{i=1}^n \Big\{\mu_1(\boldsymbol{x}_i)-\mu_0(\boldsymbol{x}_i)\Big\},
      \mathrm{QTE}(\tau)
      &= Q_1^{\mathrm{m}}(\tau)-Q_0^{\mathrm{m}}(\tau),\qquad
      Q_a^{\mathrm{m}}(\tau)=\inf\{y:\,F_a^{\mathrm{m}}(y)\ge \tau\},\quad
      F_a^{\mathrm{m}}(y)=\frac{1}{n}\sum_{i=1}^n F_a(y\mid \boldsymbol{x}_i).
$$
Treated-standardized analogues (ATT, QTT) are obtained
    by replacing the full-sample covariate empirical distribution with
    the treated-subsample empirical distribution $\{i:a_i=1\}$. In the
    current implementation, `ate()` and `att()` aggregate
    draw-wise conditional mean contrasts across rows, while
    `qte()` and `qtt()` compute draw-wise arm-specific
    marginal quantiles and then contrast them.

### Posterior prediction at new covariate profiles

Prediction in the causal setup proceeds by producing arm-specific
predictive summaries at new covariates and then differencing when
required. For new covariates $\boldsymbol{x}^\star$, the package first
predicts the PS (if enabled), forms
$\boldsymbol{r}(\boldsymbol{x}^\star)$ via
the chosen PS augmentation, and then evaluates the desired
distributional functional under both arms. For `type = "mean"`
and `type = "quantile"`, `predict()` returns
treated-minus-control summaries and stores arm-specific predictions as
attributes; for `type = "density"`, `"survival"`, and
`"prob"`, it returns arm-specific summaries directly.

A recommended way to build an interpretable prediction grid is to use
coordinate-wise empirical quartiles. Let $q_{k,p}$ be the $p$th sample
quantile of covariate $x_k$ for $p\in\{0.25,0.5,0.75\}$. The grid $$
  \mathcal{X}_\star = \prod_{k=1}^p \{q_{k,0.25},q_{k,0.5},q_{k,0.75}\}
$$ yields $3^p$ representative covariate profiles at which
$\mathrm{CATE}$, $\mathrm{CQTE}$, survival contrasts, and other
predictive contrasts can be reported.

### User-facing workflow (code)

The following example uses the formula interface and the high-level
wrapper, fitting the augmented spliced model by default.

```{r causal-data, echo=TRUE, eval=FALSE}
data("causal_pos500_p3_k2", package = "CausalMixGPD")
dat <- causal_pos500_p3_k2
df  <- data.frame(y = dat$y, A = dat$A, dat$X)
```
```{r causal-fit-model, echo=TRUE, eval=FALSE}
cfit <- dpmgpd.causal(
  formula = y ~ x1 + x2 + x3,
  data = df,
  treat = "A",
  backend = "sb",
  kernel = "lognormal",
  components = 5,
  PS = "logit",
  ps_scale = "logit",
  ps_summary = "mean",
  mcmc_outcome = mcmc_fixed,
  mcmc_ps = mcmc_fixed
)
```
The fitted model object `cfit` contains the MCMC samples for both the PS and
the arm-specific outcome models, and can be used to compute causal estimands and
predictive summaries.

To estimate the ATE and QTE at the empirical covariate distribution, use the
`ate()` and `qte()` functions, which compute the
standardized mean and quantile contrasts.

```{r causal-te-code, echo=TRUE, eval=FALSE}
ate_hat <- ate(cfit, level = 0.90, interval = "credible")
qte_hat <- qte(cfit, probs = c(0.50, 0.90, 0.95), level = 0.90, interval = "hpd")
```
For subpopulation effects, the `cate()` and `cqte()` functions compute conditional contrasts at user-specified covariate rows.

```{r causal-predict-grid, echo=TRUE, eval=FALSE}
qs    <- c(0.25, 0.50, 0.75)
Xgrid <- expand.grid(lapply(df[c("x1","x2","x3")], quantile, probs = qs))
```
The following code then produces conditional quantile and survival predictions at the grid points, returning treated-minus-control contrasts for quantiles and arm-specific summaries for survival.

```{r causal-predict-quantile-survival-code, echo=TRUE, eval=FALSE}
pred_q <- predict(cfit, newdata = Xgrid, type = "quantile", p = c(0.50, 0.90))
pred_s <- predict(cfit, newdata = Xgrid, type = "survival", y = rep(4, nrow(Xgrid)))
```
### Bulk-only causal mixtures (no GPD tail)

To run the same two-arm regression without a GPD tail, use the bulk-only
wrapper `dpmix.causal()` (or set `GPD=FALSE` at bundle
construction). All causal estimand and prediction functions
(`ate`, `qte`, `cate`, `cqte`, and
`predict`) operate the same way, now interpreting
$F_a(\cdot\mid\boldsymbol{x})$ as the bulk mixture distribution.

# Discussion

CausalMixGPD combines the flexibility of central distribution modeling, the rigor of upper-tail extrapolation, and the ability to compare the full outcome distributions for causal inference in a unified framework. It allows the full flexibility of the non-parametric bulk, the addition of GPD structure in the tails, and the ability to perform a unified inference for the posterior summaries.

The article focuses on the semiparametric aspect of the model, but the software has a wider scope that allows the user to perform a variety of tasks beyond the GPD splicing. It allows the user to perform stick-breaking and Chinese restaurant processes, different kernel families for real-line and positive outcomes, arm-specific causal inference, GPD splicing, and a clustering framework that uses the same machinery as the mixture fitting.

The clustering framework allows the dependence to be modeled through the weights, the atoms, or a combination of both, which is consistent with the dependent DP regression clustering framework. The posterior uncertainty in the clustering partitions is also performed by displaying the similarity matrices and the representative partitions, making the clustering task interpretable even in the presence of label switching in the MCMC samples.

The clustering, bulk tail, and posterior predictive analyses are all connected in a single workflow. This allows for a smooth transition from exploratory subgroup analyses to tail risk assessment and causal inference.

Some Limitations. One limitation of this method is that it is computationally expensive. MCMC for mixture models with covariate-dependent parameters and tail components is computationally expensive, especially when fitting separate chains for each arm, including separate models for each outcome, and including a series of PS estimation models. Larger sample sizes and more dimensions will be even more expensive, especially for sensitivity analyses and model comparison.

Another limitation of this method is that it is a spliced model, and this has all of the tuning parameters of a Dirichlet process mixture and a PoT model. There is a degree of uncertainty in the results, depending on the choice of kernel, truncation, and tail parameters. GPD is available for EVT for high thresholds, but for finite sample tail estimation, there must be sufficient exceedances. Similarly, clustering in a spliced model has similar behavior: strong overlap, weak predictor effects, and/or insufficient exceedances all contribute to a high degree of uncertainty in clustering, where the posterior similarity matrix is more informative than a clustering output.

The causal implementation is currently limited to binary treatments and continuous outcomes, assuming standard identification conditions of consistency, ignorability, and overlap. Propensity score augmentation is helpful for adjustments in observational studies, but it is no replacement for good study design, overlap, and sensitivity analyses for unmeasured confounding. It is not a framework for censored survival, multivariate outcomes, longitudinal treatments, or interference.



# References {#refs}




