---
title: "Causal Modeling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "cache/causal-",
  warning = FALSE,
  message = FALSE
)
```

# Overview

This vignette shows:

- arm-specific outcome fits
- how the propensity model enters
- QTE/ATE-ready prediction interfaces
- high-level assumptions

# Build a causal bundle

```{r}
library(DPmixGPD)
library(nimble)
use_cached_fit <- TRUE
.fit_path <- function(name) {
  path <- system.file("extdata", name, package = "DPmixGPD")
  if (path == "") path <- file.path("inst", "extdata", name)
  path
}
fit_causal_con <- readRDS(.fit_path("fit_causal_con.rds"))
fit_causal_trt <- readRDS(.fit_path("fit_causal_trt.rds"))
fit_causal_meta <- readRDS(.fit_path("fit_causal_meta.rds"))
fit_causal_small <- list(
  ps_fit = NULL,
  outcome_fit = list(con = fit_causal_con, trt = fit_causal_trt),
  bundle = fit_causal_meta,
  call = NULL
)
class(fit_causal_small) <- "dpmixgpd_causal_fit"
set.seed(1)
dat <- sim_causal_cqte(n = 120)
X <- dat$X
T <- dat$t
y <- dat$y

cb <- build_causal_bundle(
  y = y,
  X = X,
  T = T,
  backend = c("sb", "sb"),
  kernel = c("normal", "normal"),
  GPD = c(FALSE, FALSE),
  J = c(8, 8),
  mcmc_outcome = list(niter = 200, nburnin = 50, thin = 1, nchains = 1, seed = 1),
  mcmc_ps = list(niter = 200, nburnin = 50, thin = 1, nchains = 1, seed = 1)
)
```

# Run MCMC

```{r}
if (use_cached_fit) {
  cf <- fit_causal_small
} else {
  cf <- run_mcmc_causal(cb, show_progress = FALSE)
}
summary(cf$outcome_fit$trt)
```

# Arm-specific prediction

```{r}
pr_trt <- predict(cf$outcome_fit$trt, type = "quantile", p = c(0.5, 0.9), newdata = X)
pr_con <- predict(cf$outcome_fit$con, type = "quantile", p = c(0.5, 0.9), newdata = X)

pr_trt$fit
pr_con$fit
```

# QTE scaffolding

```{r}
qt <- cqte(cf, probs = c(0.5, 0.9), newdata = X)

qt$fit
```

Plot QTE across quantiles:

```{r}
qs <- c(0.1, 0.5, 0.9)
qt2 <- cqte(cf, probs = qs, newdata = X)
plot(qs, as.numeric(qt2$fit[1, ]), type = "b",
     xlab = "quantile", ylab = "QTE",
     main = "Quantile treatment effect")
```

# Assumptions (high level)

- Ignorability: treatment is independent of potential outcomes given covariates.
- Positivity: treatment assignment has support across covariate space.
- Correct specification of the propensity model form.

These assumptions are not verified automatically; the causal API provides
scaffolding that can be extended with diagnostics and sensitivity analysis.
