---
title: "Causal workflow (two-arm outcome modeling)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal workflow (two-arm outcome modeling)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 12000, nburnin = 3000, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Goal

This vignette demonstrates the causal inference workflow using `build_causal_bundle()` and `run_mcmc_causal()`. We compute distributional treatment effects, including:

- **Average Treatment Effect (ATE)**: \(E[Y(1) - Y(0) \mid X]\)
- **Quantile Treatment Effect (QTE)**: \(Q_{Y(1)}(\tau) - Q_{Y(0)}(\tau)\)

## Simulated data

```{r}
library(DPmixGPD)

n <- 80
X <- data.frame(x = rnorm(n))

# Treatment assignment (with propensity score)
T_ind <- rbinom(n, 1, plogis(0.2 + 0.5 * X$x))

# Outcome with heterogeneous treatment effect
y0 <- 0.5 + 0.7 * X$x + abs(rnorm(n)) + 0.1
te <- 0.4 + 0.6 * (X$x > 0)  # Treatment effect varies with x
y1 <- y0 + te
y <- ifelse(T_ind == 1, y1, y0)
```

## Build causal bundle

The `build_causal_bundle()` function creates a unified structure for:
- Propensity score (PS) model (optional)
- Control arm outcome model
- Treated arm outcome model

```{r}
bundle <- build_causal_bundle(
  y = y,
  X = X,
  T = T_ind,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  PS = "logit",  # Logistic regression for PS
  design = "observational",
  mcmc_outcome = mcmc,
  mcmc_ps = mcmc
)

bundle
```

## Run MCMC

```{r}
fit <- run_mcmc_causal(bundle, show_progress = FALSE)
fit
```

## Average Treatment Effect (ATE)

```{r}
# ATE at observed covariate values
ate_result <- ate(fit, interval = "credible", nsim_mean = 100)
head(ate_result$fit)

# ATE on a new covariate grid
X_new <- data.frame(x = seq(min(X$x), max(X$x), length.out = 20))
ate_grid <- ate(fit, newdata = X_new, interval = "credible", nsim_mean = 100)

# Plot ATE with credible intervals
plot(X_new$x, ate_grid$fit, type = "l", lwd = 2, col = "blue",
     xlab = "Covariate x", ylab = "ATE", 
     main = "Average Treatment Effect",
     ylim = range(c(ate_grid$fit, ate_grid$lower, ate_grid$upper), na.rm = TRUE))
lines(X_new$x, ate_grid$lower, lty = 2, col = "blue")
lines(X_new$x, ate_grid$upper, lty = 2, col = "blue")
abline(h = 0, lty = 3, col = "gray")
legend("topright", legend = c("ATE", "90% Credible Interval"), 
       lty = c(1, 2), col = c("blue", "blue"), bty = "n")
```

## Quantile Treatment Effect (QTE)

```{r}
# QTE at multiple quantiles
probs <- c(0.1, 0.5, 0.9)
qte_result <- qte(fit, probs = probs, interval = "credible")
# qte_result$fit is a matrix: rows = observations, cols = quantiles
head(qte_result$fit)

# QTE on a covariate grid
qte_grid <- qte(fit, probs = probs, newdata = X_new, interval = "credible")

# Plot QTE curves
# qte_grid$fit is a matrix with nrow(X_new) rows and length(probs) columns
tau_colors <- c("red", "blue", "green")
names(tau_colors) <- as.character(probs)

plot(X_new$x, qte_grid$fit[, 2],  # median (0.5) is column 2
     type = "l", lwd = 2, col = tau_colors["0.5"],
     xlab = "Covariate x", ylab = "QTE", 
     main = "Quantile Treatment Effects",
     ylim = range(qte_grid$fit, na.rm = TRUE))
lines(X_new$x, qte_grid$fit[, 1], lwd = 2, col = tau_colors["0.1"])  # 0.1 is column 1
lines(X_new$x, qte_grid$fit[, 3], lwd = 2, col = tau_colors["0.9"])  # 0.9 is column 3
abline(h = 0, lty = 3, col = "gray")
legend("topright", legend = paste0("τ = ", probs), 
       lty = 1, lwd = 2, col = tau_colors, bty = "n")
```

## Summary and diagnostics

```{r}
summary(fit)
```

## Notes and troubleshooting

- **Propensity score**: Set `PS = FALSE` for RCT designs, or use `PS = "logit"`/`"probit"` for observational studies.
- **Backend selection**: Use `backend = "sb"` for stable computation, `backend = "crp"` for adaptive clustering.
- **Treatment effect interpretation**: ATE and QTE are differences; ensure both arms have converged MCMC chains.

```{r sessioninfo, include=FALSE}
sessionInfo()
```
