---
title: "Predictor-Dependent Clustering with CausalMixGPD"
output: rmarkdown::html_vignette
bibliography: CausalMixGPD.bib
link-citations: true
vignette: >
  %\VignetteIndexEntry{Predictor-Dependent Clustering with CausalMixGPD}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.8,
  out.width = "85%",
  fig.align = "center",
  message = FALSE,
  warning = FALSE
)

.pkg_extdata <- function(name) {
  f <- system.file("extdata", name, package = "CausalMixGPD")
  if (nzchar(f)) {
    return(f)
  }
  vdir <- knitr::current_input(dir = TRUE)
  pkg_root <- normalizePath(file.path(vdir, ".."), winslash = "/", mustWork = TRUE)
  alt <- file.path(pkg_root, "inst", "extdata", name)
  if (file.exists(alt)) {
    return(normalizePath(alt, winslash = "/", mustWork = TRUE))
  }
  stop("Missing extdata: ", name, ". Run data-raw/build_vignette_fits.R", call. = FALSE)
}
```

# Introduction

The clustering interface in `CausalMixGPD` uses the same Bayesian nonparametric mixture machinery as the density and causal workflows, but the inferential focus changes. Instead of treating the fitted distribution only as a prediction engine, the clustering workflow treats the latent partition induced by the mixture model as a primary object of interest. The output is therefore organized around label-invariant partition summaries such as the posterior similarity matrix (PSM), representative cluster labels, cluster sizes, and membership scores [@Binder1978; @Dahl2006].

This clustering functionality is important because Dirichlet process mixtures do more than estimate smooth densities. Their latent allocation variables induce random partitions, so they can be used for subgroup discovery while still allowing flexible within-cluster outcome models. The software article presents clustering as one of the three main capabilities of the package, alongside one-arm modeling and causal inference.

This vignette gives a complete introduction to the clustering workflow. We first define the latent-partition notation and the three forms of predictor dependence supported by the package. We then fit a clustering model to a bundled dataset, extract label-invariant summaries, and predict cluster labels for new observations. We close with a reference section describing the main clustering-specific customization options.

# Notation and short theory background

## Mixture models and latent cluster labels

Let $Y$ denote the response and let $X$ denote the predictor vector. In a mixture model, the conditional density can be written abstractly as

$$
f(y \mid x) = \sum_{j=1}^{\infty} w_j(x) k(y \mid x, \theta_j),
$$

where the weights may or may not depend on predictors and the component-specific parameters may or may not depend on predictors. Introduce a latent allocation variable $z_i$ for observation $i$, with $z_i = j$ meaning that observation $i$ is assigned to component $j$. Then the conditional response model can be written as

$$
y_i \mid z_i, x_i, \{\theta_j\} \sim k(\cdot \mid x_i, \theta_{z_i}).
$$

The collection $\Pi = \{z_1, \ldots, z_n\}$ determines a random partition of the observations. In Bayesian nonparametric clustering, the posterior distribution of this partition is the central inferential object.

## Why labels are not directly interpretable

Mixture component labels are not identifiable across MCMC draws. One posterior draw may label a cluster as component 1 while another labels the same substantive group as component 4. Because of this label-switching phenomenon, cluster summaries should not be based on raw component labels alone.

Instead, the package reports label-invariant summaries. The most important one is the posterior similarity matrix, whose $(i,\ell)$ entry is

$$
S_{i\ell} = \Pr(z_i = z_\ell \mid \text{data}).
$$

This is estimated by averaging co-clustering indicators over retained posterior draws. A representative hard partition is then obtained using Dahl's least-squares criterion [@Dahl2006].

## Predictor dependence in clustering

The package supports three distinct clustering modes, each corresponding to a different way in which predictors can affect the mixture model.

### Weight dependence

In the weight-dependent mode,

$$
f(y \mid x) = \sum_{j=1}^{\infty} w_j(x) k(y \mid \theta_j).
$$

Here the cluster-specific kernel parameters are shared, but the prevalence of the clusters changes with predictors. This is useful when the goal is to let covariates modify which latent groups are more likely without changing the within-group model form.

### Parameter dependence

In the parameter-dependent mode,

$$
f(y \mid x) = \sum_{j=1}^{\infty} w_j k(y \mid x, \theta_j).
$$

The weights are global, but each cluster has its own predictor-response relationship. This is useful when clusters represent different regression regimes.

### Weight and parameter dependence

In the most flexible mode,

$$
f(y \mid x) = \sum_{j=1}^{\infty} w_j(x) k(y \mid x, \theta_j).
$$

Here both cluster prevalence and within-cluster behavior can vary with predictors.

## Main clustering summaries returned by the package

The clustering wrappers produce fitted objects whose predictions are cluster-focused rather than density-focused.

- `predict(type = "psm")` returns the posterior similarity matrix.
- `predict(type = "label")` returns representative labels and, when available, cluster membership scores.
- `summary()` reports cluster sizes and profile summaries.
- `plot()` can display the PSM, cluster sizes, cluster certainty, and representative-cluster summaries.

These summaries allow the user to distinguish between strong clustering structure and uncertain partitions. In many applications, the PSM is more informative than a single hard partition because it shows which observations cluster together consistently and which do not.

# Function map for the clustering workflow

The main exported functions used in this vignette are:

- `dpmix.cluster()` for bulk-only clustering.
- `dpmgpd.cluster()` for clustering with an active GPD tail.
- `predict()` with `type = "psm"` or `type = "label"`.
- `summary()` for cluster sizes and profiles.
- `plot()` for PSM heatmaps and cluster-summary graphics.

# Package setup

```{r libraries}
library(CausalMixGPD)
library(ggplot2)
```

```{r mcmc-settings}
mcmc_vig <- list(
  niter = 1200,
  nburnin = 300,
  thin = 2,
  nchains = 2,
  seed = 2026
)
```

# Data

We use the bundled dataset `nc_realX100_p3_k2`, which contains a real-valued response and three predictors.

```{r load-data}
data("nc_realX100_p3_k2", package = "CausalMixGPD")
dat_cl <- data.frame(
  y = nc_realX100_p3_k2$y,
  nc_realX100_p3_k2$X
)

str(dat_cl)
summary(dat_cl)
```

For illustration, we form a simple train/test split so that we can show both training-cluster summaries and out-of-sample label prediction.

```{r split-data}
clust_out <- readRDS(.pkg_extdata("clustering_outputs.rds"))
train_dat <- clust_out$train_dat
test_dat <- clust_out$test_dat

nrow(train_dat)
nrow(test_dat)
```

A quick visualization of the response helps motivate the use of a flexible mixture model.

```{r exploratory-plots, fig.height=4.8, fig.width=7}
ggplot(train_dat, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 25,
                 fill = "grey85", colour = "grey35") +
  geom_density(linewidth = 0.9) +
  labs(x = "y", y = "Density", title = "Training-sample response distribution") +
  theme_minimal()
```

# Fitting a clustering model

The bulk-only clustering wrapper is `dpmix.cluster()`. It uses the same general mixture machinery as the one-arm interface, but returns a clustering-oriented fitted object.

```{r fit-cluster}
fit_cluster <- clust_out$fit_cluster
```

The choice `type = "both"` requests dependence in both the mixing weights and the component-specific parameters. In practice, this is the most flexible clustering specification because it allows both cluster prevalence and cluster-specific regression behavior to vary with covariates.

# Posterior similarity matrix

The first label-invariant summary is the PSM.

```{r psm-prediction}
z_train_psm <- clust_out$psm_obj
z_train_psm
```

```{r psm-plot, fig.height=5.2, fig.width=7}
plot(z_train_psm, type = "summary")
```

The block structure in the PSM is the main diagnostic for cluster separation. Darker within-block regions indicate pairs of observations that co-cluster consistently across posterior draws, while diffuse or weak block boundaries indicate uncertainty about the partition.

# Representative training partition

The next step is to extract representative cluster labels for the training sample.

```{r label-prediction-train}
z_train_lab <- clust_out$train_lab
z_train_lab
```

```{r label-plot-train, fig.height=5, fig.width=7}
plot(z_train_lab, type = "summary")
```

The representative partition provides a single interpretable clustering summary on the response scale. It is convenient for reporting, but it should be read together with the PSM because the latter conveys the posterior uncertainty in the clustering structure.

# Predicting labels for new observations

A distinctive feature of the package is that it also supports cluster prediction for new observations.

```{r label-prediction-test}
z_test <- clust_out$test_lab
z_test
```

```{r test-cluster-profiles}
clust_out$cluster_profiles
```

```{r sizes-plot-test, fig.height=4.8, fig.width=7}
plot(z_test, type = "sizes")
```

This prediction step maps new observations into the space of the representative training clusters. That is useful in applications where the discovered subgroup structure needs to be carried forward to new units without refitting the entire model.

# Analysis summary

The clustering workflow in `CausalMixGPD` is most informative when the user interprets it as a posterior partition analysis rather than as a hard unsupervised classification algorithm. The PSM shows which observations reliably belong together, the representative labels provide a convenient summary partition, and the test-set predictions extend that partition to new observations.

The three predictor-dependence modes are scientifically meaningful rather than cosmetic. Weight dependence is useful when covariates mainly affect subgroup prevalence. Parameter dependence is useful when clusters correspond to different regression surfaces. The combined mode is appropriate when both mechanisms are plausible. In all three cases, the package reports label-invariant summaries so that posterior uncertainty about the partition is not hidden behind arbitrary component labels.

# Customization options for clustering models

The software appendix gives a compact map of the clustering-specific controls; the most important ones are summarized here.

## Clustering mode

`type = "weights"` places predictor dependence in the mixing weights.

`type = "param"` places predictor dependence in the component-specific kernel parameters.

`type = "both"` allows both forms of dependence simultaneously.

## Wrapper choice

`dpmix.cluster()` fits the bulk-only clustering model.

`dpmgpd.cluster()` adds an active GPD tail and should be considered when tail behavior is part of the clustering problem.

## Kernel, backend, and truncation

`kernel` selects the bulk family, subject to the support of the response.

`components` controls the truncation size. In clustering problems, under-truncation can hide meaningful subgroup structure, so this argument deserves explicit attention.

The package handles some backend details internally based on the clustering mode, but users still choose the scientific dependence structure through `type`.

## Monitoring and output controls

`monitor`, `monitor_latent`, and `monitor_v` control how much posterior state is retained.

`predict(type = "psm")` and `predict(type = "label")` determine whether the user wants a matrix-valued co-clustering summary or representative labels.

`return_scores = TRUE` can be used when membership scores or certainty summaries are needed.

`summary()` reports cluster sizes and within-cluster profiles, while `plot()` supports PSM heatmaps and label-summary graphics such as cluster sizes and certainty.

# Session information

```{r session-info}
sessionInfo()
```
