---
title: "Conditional Modeling"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Conditional Modeling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# Goal

Conditional modeling lets **distributional features** (not just means) vary with covariates. This is **not** mean regression; it is full distributional regression.

# Example data

```{r data}
data("nc_posX100_p3_k2")
X <- as.matrix(nc_posX100_p3_k2$X)
y <- nc_posX100_p3_k2$y
```

# Conditional mixture

```{r cond}
bundle_cond <- build_nimble_bundle(
  y = y,
  X = X,
  kernel = "lognormal",
  backend = "sb",
  GPD = TRUE,
  J = 4,
  mcmc = list(niter = 250, nburnin = 50, nchains = 1)
)
fit_cond <- run_mcmc_bundle_manual(bundle_cond, show_progress = FALSE)
```

# Interpretation across quantiles

```{r cond-pred}
x_new <- X[1:10, , drop = FALSE]
q_pred <- predict(fit_cond, x = x_new, type = "quantile", index = c(0.1, 0.5, 0.9))
plot(q_pred)
```

# Covariates and tails

By allowing tail parameters to depend on $X$, DPmixGPD can capture **heterogeneous tail risk** (e.g., extreme outcomes becoming more likely in certain covariate regimes).

# Connection to unconditional models

The unconditional case is the special case with **no covariates**. See [unconditional](unconditional.Rmd).

# What this is NOT

- It is **not** treatment effect modeling.
- It is **not** identification or causal inference.

For causal contrasts, see [causal](causal.Rmd).
