---
<<<<<<< HEAD
<<<<<<< HEAD
title: "Conditional models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Conditional models}
=======
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
title: "Conditional Modeling"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Conditional Modeling}
<<<<<<< HEAD
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
<<<<<<< HEAD
<<<<<<< HEAD
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
library(nimble)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 600, nburnin = 150, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 9000, nburnin = 2500, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Goal

Fit a conditional model (with covariates) and generate predictions on a grid of new covariate values.

## Simulated data

```{r}
library(DPmixGPD)

n <- 120
x <- rnorm(n)
X <- data.frame(x = x)

# simple signal + heavy-ish noise
y <- 0.5 + 0.8 * x + abs(rnorm(n)) + 0.1
```

## Fit

```{r}
bundle <- build_nimble_bundle(
  y = y,
  X = X,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE)
fit
```

## Fitted values (training data)

```{r}
f <- fitted(fit, type = "mean", level = 0.90)
head(f)

# residual summary
summary(f$residuals)
```

## Predictions on new data

```{r}
new_X <- data.frame(x = seq(min(x), max(x), length.out = 25))

pred_mean <- predict(fit, x = new_X, type = "mean", cred.level = 0.90, interval = "credible")
pred_med  <- predict(fit, x = new_X, type = "median", cred.level = 0.90, interval = "credible")

head(pred_mean$fit)
head(pred_med$fit)
```

## Quantile curves

```{r}
q_levels <- c(0.1, 0.5, 0.9)
q_fits <- lapply(q_levels, function(tau) {
  predict(fit, x = new_X, type = "quantile", index = tau, cred.level = 0.90, interval = "credible")$fit
})

# Bind into a simple long data.frame for plotting
q_df <- do.call(rbind, Map(function(tau, df) {
  data.frame(x = new_X$x, tau = tau, estimate = df$estimate, lower = df$lower, upper = df$upper)
}, q_levels, q_fits))

head(q_df)
```

```{r}
# Minimal base plotting to avoid extra dependencies
plot(q_df$x[q_df$tau == 0.5], q_df$estimate[q_df$tau == 0.5], type = "l",
     xlab = "x", ylab = "Predicted quantile", main = "Conditional quantile curves")
lines(q_df$x[q_df$tau == 0.1], q_df$estimate[q_df$tau == 0.1])
lines(q_df$x[q_df$tau == 0.9], q_df$estimate[q_df$tau == 0.9])
legend("topleft", legend = paste0("tau=", q_levels), lty = 1, bty = "n")
```

## Notes

- If you see a NIMBLE error mentioning keywords (e.g., `if`), rename covariate columns.
- Increase `niter` and run multiple chains for serious inference.

```{r sessioninfo, include=FALSE}
sessionInfo()
```
=======
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  comment = NA,
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# Goal

Conditional modeling lets **distributional features** (not just means) vary with covariates. This is **not** mean regression; it is full distributional regression.

# Example data

```{r data}
data("nc_posX100_p3_k2")
X <- as.matrix(nc_posX100_p3_k2$X)
y <- nc_posX100_p3_k2$y
```

# Conditional mixture

```{r cond}
bundle_cond <- build_nimble_bundle(
  y = y,
  X = X,
  kernel = "lognormal",
  backend = "sb",
  GPD = TRUE,
  J = 4,
  mcmc = list(niter = 250, nburnin = 50, nchains = 1)
)
fit_cond <- run_mcmc_bundle_manual(bundle_cond, show_progress = FALSE)
```

# Interpretation across quantiles

```{r cond-pred}
x_new <- X[1:10, , drop = FALSE]
q_pred <- predict(fit_cond, x = x_new, type = "quantile", index = c(0.1, 0.5, 0.9))
plot(q_pred)
```

# Covariates and tails

By allowing tail parameters to depend on $X$, DPmixGPD can capture **heterogeneous tail risk** (e.g., extreme outcomes becoming more likely in certain covariate regimes).

# Connection to unconditional models

The unconditional case is the special case with **no covariates**. See [unconditional](unconditional.Rmd).

# What this is NOT

- It is **not** treatment effect modeling.
- It is **not** identification or causal inference.

For causal contrasts, see [causal](causal.Rmd).
<<<<<<< HEAD
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
