---
title: "Available Distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Available Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)

library(ggplot2)
library(kableExtra)
library(DPmixGPD)
set.seed(1)

options(digits = 3)
knitr::opts_chunk$set(digits = 3)
```

## Overview

DPmixGPD supports multiple bulk kernels for mixture components and can optionally
splice a Generalized Pareto Distribution (GPD) tail beyond a threshold.

All examples below use **vectorized R wrappers** (lowercase) only.

The `d*`, `p*`, `q*`, `r*` prefixes mean:

- `d*`: density (PDF); supports `log` where available.
- `p*`: distribution (CDF); supports `lower.tail` and `log.p`.
- `q*`: quantile; supports `lower.tail` and `log.p`.
- `r*`: random generation; supports `n`.

GPD variants add `threshold`, `tail_scale`, and `tail_shape`.

See **Complete Function Reference** at the end for the full `d/p/q/r` sets (R + NIMBLE) and Rd links.

```{r helpers, include=FALSE}
format_value <- function(v, digits = 3) {
  if (length(v) > 1) {
    paste0("(", paste(formatC(v, digits = digits, format = "f", drop0trailing = TRUE), collapse = ", "), ")")
  } else {
    formatC(v, digits = digits, format = "f", drop0trailing = TRUE)
  }
}

param_label <- function(params, digits = 3) {
  paste(names(params), "=", vapply(params, format_value, character(1), digits = digits),
        collapse = ", ")
}

plot_density_overlay <- function(grid, sets, title) {
  df <- do.call(rbind, lapply(sets, function(s) {
    y <- do.call(s$fn, c(list(x = grid), s$params))
    data.frame(x = grid, density = y, label = s$label)
  }))

  ggplot(df, aes(x = x, y = density, color = label)) +
    geom_line(linewidth = 1) +
    labs(x = "x", y = "Density", title = title) +
    theme_minimal() +
    theme(legend.position = "top")
}

param_table <- function(title, sets) {
  df <- data.frame(
    Label = vapply(sets, function(s) s$label, character(1)),
    Parameters = vapply(sets, function(s) param_label(s$params), character(1)),
    check.names = FALSE
  )

  knitr::kable(df, align = "c", caption = title, format = "html") %>%
    kable_styling(bootstrap_options = c("striped", "hover"),
                  full_width = FALSE, position = "center")
}

demo_table <- function(title, variants, x, prob = 0.8, n = 3) {
  df <- data.frame(
    Variant = vapply(variants, function(v) v$label, character(1)),
    `d(x)` = vapply(variants, function(v) {
      do.call(v$d, c(list(x = x), v$params))
    }, numeric(1)),
    `p(x)` = vapply(variants, function(v) {
      do.call(v$p, c(list(q = x), v$params))
    }, numeric(1)),
    `q(p)` = vapply(variants, function(v) {
      do.call(v$q, c(list(p = prob), v$params))
    }, numeric(1)),
    `r(n)` = vapply(variants, function(v) {
      vals <- do.call(v$r, c(list(n = n), v$params))
      paste(formatC(vals, digits = 3, format = "f", drop0trailing = TRUE), collapse = ", ")
    }, character(1)),
    check.names = FALSE
  )

  knitr::kable(df, align = "c", caption = title, format = "html") %>%
    kable_styling(bootstrap_options = c("striped", "hover"),
                  full_width = FALSE, position = "center")
}

link_base <- function() {
  out_dir <- knitr::opts_knit$get("output.dir")
  if (!is.null(out_dir)) {
    out_dir_norm <- gsub("\\\\", "/", out_dir)
    if (grepl("docs/articles", out_dir_norm)) {
      return("../reference/")
    }
  }
  if (dir.exists("docs/reference")) {
    return("../docs/reference/")
  }
  "../reference/"
}

rd_link <- function(name, label = NULL) {
  if (is.null(label)) {
    label <- name
  }
  sprintf("[%s](%s%s.html)", label, link_base(), name)
}
```

## Generalized Pareto (GPD)

**Functions**

- Base: `dgpd`, `pgpd`, `qgpd`, `rgpd`

```{r gpd-params, results="asis"}
sets <- list(
  list(label = "GPD A", fn = dgpd, params = list(threshold = 1.0, scale = 0.6, shape = 0.15)),
  list(label = "GPD B", fn = dgpd, params = list(threshold = 1.5, scale = 0.5, shape = 0.25)),
  list(label = "GPD C", fn = dgpd, params = list(threshold = 2.0, scale = 0.7, shape = 0.10))
)

param_table("GPD parameter sets", sets)
```

```{r gpd-demo, results="asis"}
variants <- list(
  list(label = "GPD A", d = dgpd, p = pgpd, q = qgpd, r = rgpd,
       params = list(threshold = 1.0, scale = 0.6, shape = 0.15)),
  list(label = "GPD B", d = dgpd, p = pgpd, q = qgpd, r = rgpd,
       params = list(threshold = 1.5, scale = 0.5, shape = 0.25)),
  list(label = "GPD C", d = dgpd, p = pgpd, q = qgpd, r = rgpd,
       params = list(threshold = 2.0, scale = 0.7, shape = 0.10))
)

demo_table("GPD function demo (x = 2.0, p = 0.8, n = 3)", variants, x = 2.0)
```

```{r gpd-plot, echo=FALSE}
grid <- seq(0.0, 12.0, length.out = 400)
print(plot_density_overlay(grid, sets, "GPD densities "))
```

## Normal

**Functions**

- Base: `dnorm`, `pnorm`, `qnorm`, `rnorm`
- Base + GPD: `dnormgpd`, `pnormgpd`, `qnormgpd`, `rnormgpd`
- Mixture: `dnormmix`, `pnormmix`, `qnormmix`, `rnormmix`
- Mixture + GPD: `dnormmixgpd`, `pnormmixgpd`, `qnormmixgpd`, `rnormmixgpd`

```{r norm-params, results="asis"}
sets <- list(
  list(label = "Base", fn = dnorm, params = list(mean = 0, sd = 1)),
  list(label = "Base + GPD", fn = dnormgpd,
       params = list(mean = 0, sd = 1, threshold = 1.5, tail_scale = 0.5, tail_shape = 0.2)),
  list(label = "Mixture", fn = dnormmix,
       params = list(w = c(0.7, 0.3), mean = c(-1, 1.5), sd = c(1.0, 0.7))),
  list(label = "Mixture + GPD", fn = dnormmixgpd,
       params = list(w = c(0.7, 0.3), mean = c(-1, 1), sd = c(1.0, 0.7),
                     threshold = 1.5, tail_scale = 0.5, tail_shape = 0.2))
)

param_table("Normal parameter sets", sets)
```

```{r norm-demo, results="asis"}
variants <- list(
  list(label = "Base", d = dnorm, p = pnorm, q = qnorm, r = rnorm,
       params = list(mean = 0, sd = 1)),
  list(label = "Base + GPD", d = dnormgpd, p = pnormgpd, q = qnormgpd, r = rnormgpd,
       params = list(mean = 0, sd = 1, threshold = 1.5, tail_scale = 0.5, tail_shape = 0.2)),
  list(label = "Mixture", d = dnormmix, p = pnormmix, q = qnormmix, r = rnormmix,
       params = list(w = c(0.7, 0.3), mean = c(-1, 1.5), sd = c(1.0, 0.7))),
  list(label = "Mixture + GPD", d = dnormmixgpd, p = pnormmixgpd, q = qnormmixgpd, r = rnormmixgpd,
       params = list(w = c(0.7, 0.3), mean = c(-1, 1), sd = c(1.0, 0.7),
                     threshold = 1.5, tail_scale = 0.5, tail_shape = 0.2))
)

demo_table("Normal function demo (x = 0.5, p = 0.8, n = 3)", variants, x = 0.5)
```

```{r norm-plot, echo=FALSE}
grid <- seq(-6, 6, length.out = 400)
print(plot_density_overlay(grid, sets, "Normal densities "))
```

## Gamma

**Functions**

- Base: `dgamma`, `pgamma`, `qgamma`, `rgamma`
- Base + GPD: `dgammagpd`, `pgammagpd`, `qgammagpd`, `rgammagpd`
- Mixture: `dgammamix`, `pgammamix`, `qgammamix`, `rgammamix`
- Mixture + GPD: `dgammamixgpd`, `pgammamixgpd`, `qgammamixgpd`, `rgammamixgpd`

```{r gamma-params, results="asis"}
sets <- list(
  list(label = "Base", fn = dgamma, params = list(shape = 2.5, scale = 1.0)),
  list(label = "Base + GPD", fn = dgammagpd,
       params = list(shape = 3, scale = 1.0, threshold = 5.0, tail_scale = 1.0, tail_shape = 0.2)),
  list(label = "Mixture", fn = dgammamix,
       params = list(w = c(0.6, 0.4), shape = c(2, 5), scale = c(1.0, 0.7))),
  list(label = "Mixture + GPD", fn = dgammamixgpd,
       params = list(w = c(0.6, 0.4), shape = c(2, 5), scale = c(1.0, 0.7),
                     threshold = 5.0, tail_scale = 1.0, tail_shape = 0.2))
)

param_table("Gamma parameter sets", sets)
```

```{r gamma-demo, results="asis"}
variants <- list(
  list(label = "Base", d = dgamma, p = pgamma, q = qgamma, r = rgamma,
       params = list(shape = 2.5, scale = 1.0)),
  list(label = "Base + GPD", d = dgammagpd, p = pgammagpd, q = qgammagpd, r = rgammagpd,
       params = list(shape = 3, scale = 1.0, threshold = 5.0, tail_scale = 1.0, tail_shape = 0.2)),
  list(label = "Mixture", d = dgammamix, p = pgammamix, q = qgammamix, r = rgammamix,
       params = list(w = c(0.6, 0.4), shape = c(2, 5), scale = c(1.0, 0.7))),
  list(label = "Mixture + GPD", d = dgammamixgpd, p = pgammamixgpd, q = qgammamixgpd, r = rgammamixgpd,
       params = list(w = c(0.6, 0.4), shape = c(2, 5), scale = c(1.0, 0.7),
                     threshold = 5.0, tail_scale = 1.0, tail_shape = 0.2))
)

demo_table("Gamma function demo (x = 3.0, p = 0.8, n = 3)", variants, x = 3.0)
```

```{r gamma-plot, echo=FALSE}
grid <- seq(0.01, 12, length.out = 400)
print(plot_density_overlay(grid, sets, "Gamma densities "))
```

## Lognormal

**Functions**

- Base: `dlnorm`, `plnorm`, `qlnorm`, `rlnorm`
- Base + GPD: `dlognormalgpd`, `plognormalgpd`, `qlognormalgpd`, `rlognormalgpd`
- Mixture: `dlognormalmix`, `plognormalmix`, `qlognormalmix`, `rlognormalmix`
- Mixture + GPD: `dlognormalmixgpd`, `plognormalmixgpd`, `qlognormalmixgpd`, `rlognormalmixgpd`

```{r logn-params, results="asis"}
sets <- list(
  list(label = "Base", fn = dlnorm, params = list(meanlog = 0.2, sdlog = 0.5)),
  list(label = "Base + GPD", fn = dlognormalgpd,
       params = list(meanlog = 0.2, sdlog = 0.5, threshold = 3.0, tail_scale = 0.8, tail_shape = 0.2)),
  list(label = "Mixture", fn = dlognormalmix,
       params = list(w = c(0.6, 0.4), meanlog = c(0, 0.6), sdlog = c(0.4, 0.5))),
  list(label = "Mixture + GPD", fn = dlognormalmixgpd,
       params = list(w = c(0.6, 0.4), meanlog = c(0, 0.6), sdlog = c(0.4, 0.5),
                     threshold = 3.0, tail_scale = 0.8, tail_shape = 0.2))
)

param_table("Lognormal parameter sets", sets)
```

```{r logn-demo, results="asis"}
variants <- list(
  list(label = "Base", d = dlnorm, p = plnorm, q = qlnorm, r = rlnorm,
       params = list(meanlog = 0.2, sdlog = 0.5)),
  list(label = "Base + GPD", d = dlognormalgpd, p = plognormalgpd, q = qlognormalgpd, r = rlognormalgpd,
       params = list(meanlog = 0.2, sdlog = 0.5, threshold = 3.0, tail_scale = 0.8, tail_shape = 0.2)),
  list(label = "Mixture", d = dlognormalmix, p = plognormalmix, q = qlognormalmix, r = rlognormalmix,
       params = list(w = c(0.6, 0.4), meanlog = c(0, 0.6), sdlog = c(0.4, 0.5))),
  list(label = "Mixture + GPD", d = dlognormalmixgpd, p = plognormalmixgpd,
       q = qlognormalmixgpd, r = rlognormalmixgpd,
       params = list(w = c(0.6, 0.4), meanlog = c(0, 0.6), sdlog = c(0.4, 0.5),
                     threshold = 3.0, tail_scale = 0.8, tail_shape = 0.2))
)

demo_table("Lognormal function demo (x = 1.2, p = 0.8, n = 3)", variants, x = 1.2)
```

```{r logn-plot, echo=FALSE}
grid <- seq(0.01, 10, length.out = 400)
print(plot_density_overlay(grid, sets, "Lognormal densities "))
```

## Laplace

**Functions**

- Base: `ddexp`, `pdexp`, `qdexp`, `rdexp`
- Base + GPD: `dlaplacegpd`, `plaplacegpd`, `qlaplacegpd`, `rlaplacegpd`
- Mixture: `dlaplacemix`, `plaplacemix`, `qlaplacemix`, `rlaplacemix`
- Mixture + GPD: `dlaplacemixgpd`, `plaplacemixgpd`, `qlaplacemixgpd`, `rlaplacemixgpd`

```{r laplace-params, results="asis"}
sets <- list(
  list(label = "Base", fn = nimble::ddexp, params = list(location = 0, scale = 0.8)),
  list(label = "Base + GPD", fn = dlaplacegpd,
       params = list(location = 0, scale = 0.8, threshold = 2.0, tail_scale = 0.5, tail_shape = 0.2)),
  list(label = "Mixture", fn = dlaplacemix,
       params = list(w = c(0.7, 0.3), location = c(0, 0.8), scale = c(0.5, 0.8))),
  list(label = "Mixture + GPD", fn = dlaplacemixgpd,
       params = list(w = c(0.7, 0.3), location = c(0, 0.8), scale = c(0.5, 0.8),
                     threshold = 2.0, tail_scale = 0.5, tail_shape = 0.2))
)

param_table("Laplace parameter sets", sets)
```

```{r laplace-demo, results="asis"}
variants <- list(
  list(label = "Base", d = nimble::ddexp, p = nimble::pdexp, q = nimble::qdexp, r = nimble::rdexp,
       params = list(location = 0, scale = 0.8)),
  list(label = "Base + GPD", d = dlaplacegpd, p = plaplacegpd, q = qlaplacegpd, r = rlaplacegpd,
       params = list(location = 0, scale = 0.8, threshold = 2.0, tail_scale = 0.5, tail_shape = 0.2)),
  list(label = "Mixture", d = dlaplacemix, p = plaplacemix, q = qlaplacemix, r = rlaplacemix,
       params = list(w = c(0.7, 0.3), location = c(0, 0.8), scale = c(0.5, 0.8))),
  list(label = "Mixture + GPD", d = dlaplacemixgpd, p = plaplacemixgpd,
       q = qlaplacemixgpd, r = rlaplacemixgpd,
       params = list(w = c(0.7, 0.3), location = c(0, 0.8), scale = c(0.5, 0.8),
                     threshold = 2.0, tail_scale = 0.5, tail_shape = 0.2))
)

demo_table("Laplace function demo (x = 0.5, p = 0.8, n = 3)", variants, x = 0.5)
```

```{r laplace-plot, echo=FALSE}
grid <- seq(-6, 6, length.out = 400)
print(plot_density_overlay(grid, sets, "Laplace densities "))
```

## Inverse Gaussian

**Functions**

- Base: `dinvgauss`, `pinvgauss`, `qinvgauss`, `rinvgauss`
- Base + GPD: `dinvgaussgpd`, `pinvgaussgpd`, `qinvgaussgpd`, `rinvgaussgpd`
- Mixture: `dinvgaussmix`, `pinvgaussmix`, `qinvgaussmix`, `rinvgaussmix`
- Mixture + GPD: `dinvgaussmixgpd`, `pinvgaussmixgpd`, `qinvgaussmixgpd`, `rinvgaussmixgpd`

```{r ig-params, results="asis"}
sets <- list(
  list(label = "Base", fn = dinvgauss, params = list(mean = 1.2, shape = 3.0)),
  list(label = "Base + GPD", fn = dinvgaussgpd,
       params = list(mean = 1.2, shape = 3.0, threshold = 2.0, tail_scale = 0.5, tail_shape = 0.2)),
  list(label = "Mixture", fn = dinvgaussmix,
       params = list(w = c(0.6, 0.4), mean = c(1.0, 2.0), shape = c(2.0, 4.0))),
  list(label = "Mixture + GPD", fn = dinvgaussmixgpd,
       params = list(w = c(0.6, 0.4), mean = c(1.0, 2.0), shape = c(2.0, 4.0),
                     threshold = 2.0, tail_scale = 0.5, tail_shape = 0.2))
)

param_table("Inverse Gaussian parameter sets", sets)
```

```{r ig-demo, results="asis"}
variants <- list(
  list(label = "Base", d = dinvgauss, p = pinvgauss, q = qinvgauss, r = rinvgauss,
       params = list(mean = 1.2, shape = 3.0)),
  list(label = "Base + GPD", d = dinvgaussgpd, p = pinvgaussgpd, q = qinvgaussgpd, r = rinvgaussgpd,
       params = list(mean = 1.2, shape = 3.0, threshold = 2.0, tail_scale = 0.5, tail_shape = 0.2)),
  list(label = "Mixture", d = dinvgaussmix, p = pinvgaussmix, q = qinvgaussmix, r = rinvgaussmix,
       params = list(w = c(0.6, 0.4), mean = c(1.0, 2.0), shape = c(2.0, 4.0))),
  list(label = "Mixture + GPD", d = dinvgaussmixgpd, p = pinvgaussmixgpd,
       q = qinvgaussmixgpd, r = rinvgaussmixgpd,
       params = list(w = c(0.6, 0.4), mean = c(1.0, 2.0), shape = c(2.0, 4.0),
                     threshold = 2.0, tail_scale = 0.5, tail_shape = 0.2))
)

demo_table("Inverse Gaussian function demo (x = 1.5, p = 0.8, n = 3)", variants, x = 1.5)
```

```{r ig-plot, echo=FALSE}
grid <- seq(0.01, 10, length.out = 400)
print(plot_density_overlay(grid, sets, "Inverse Gaussian densities "))
```

## Amoroso

**Functions**

- Base: `damoroso`, `pamoroso`, `qamoroso`, `ramoroso`
- Base + GPD: `damorosogpd`, `pamorosogpd`, `qamorosogpd`, `ramorosogpd`
- Mixture: `damorosomix`, `pamorosomix`, `qamorosomix`, `ramorosomix`
- Mixture + GPD: `damorosomixgpd`, `pamorosomixgpd`, `qamorosomixgpd`, `ramorosomixgpd`

```{r amor-params, results="asis"}
sets <- list(
  list(label = "Base", fn = damoroso,
       params = list(loc = 0, scale = 1.2, shape1 = 2.5, shape2 = 1.2)),
  list(label = "Base + GPD", fn = damorosogpd,
       params = list(loc = 0, scale = 1.2, shape1 = 2.5, shape2 = 1.2,
                     threshold = 2.5, tail_scale = 0.4, tail_shape = 0.2)),
  list(label = "Mixture", fn = damorosomix,
       params = list(w = c(0.6, 0.4), loc = c(0, 0), scale = c(1.0, 1.5),
                     shape1 = c(2, 3), shape2 = c(1.2, 1.2))),
  list(label = "Mixture + GPD", fn = damorosomixgpd,
       params = list(w = c(0.6, 0.4), loc = c(0, 0), scale = c(1.0, 1.5),
                     shape1 = c(2, 3), shape2 = c(1.2, 1.2),
                     threshold = 2.5, tail_scale = 0.4, tail_shape = 0.2))
)

param_table("Amoroso parameter sets", sets)
```

```{r amor-demo, results="asis"}
variants <- list(
  list(label = "Base", d = damoroso, p = pamoroso, q = qamoroso, r = ramoroso,
       params = list(loc = 0, scale = 1.2, shape1 = 2.5, shape2 = 1.2)),
  list(label = "Base + GPD", d = damorosogpd, p = pamorosogpd, q = qamorosogpd, r = ramorosogpd,
       params = list(loc = 0, scale = 1.2, shape1 = 2.5, shape2 = 1.2,
                     threshold = 2.5, tail_scale = 0.4, tail_shape = 0.2)),
  list(label = "Mixture", d = damorosomix, p = pamorosomix, q = qamorosomix, r = ramorosomix,
       params = list(w = c(0.6, 0.4), loc = c(0, 0), scale = c(1.0, 1.5),
                     shape1 = c(2, 3), shape2 = c(1.2, 1.2))),
  list(label = "Mixture + GPD", d = damorosomixgpd, p = pamorosomixgpd,
       q = qamorosomixgpd, r = ramorosomixgpd,
       params = list(w = c(0.6, 0.4), loc = c(0, 0), scale = c(1.0, 1.5),
                     shape1 = c(2, 3), shape2 = c(1.2, 1.2),
                     threshold = 2.5, tail_scale = 0.4, tail_shape = 0.2))
)

demo_table("Amoroso function demo (x = 1.8, p = 0.8, n = 3)", variants, x = 1.8)
```

```{r amor-plot, echo=FALSE}
grid <- seq(0.01, 10, length.out = 400)
print(plot_density_overlay(grid, sets, "Amoroso densities "))
```

## Cauchy

**Functions**

- Base: `dcauchy_vec`, `pcauchy_vec`, `qcauchy_vec`, `rcauchy_vec`
- Mixture: `dcauchymix`, `pcauchymix`, `qcauchymix`, `rcauchymix`

```{r cauchy-params, results="asis"}
sets <- list(
  list(label = "Base", fn = dcauchy_vec, params = list(location = 0, scale = 1.0)),
  list(label = "Mixture", fn = dcauchymix,
       params = list(w = c(0.6, 0.4), location = c(-1, 1), scale = c(1.0, 1.5)))
)

param_table("Cauchy parameter sets", sets)
```

```{r cauchy-demo, results="asis"}
variants <- list(
  list(label = "Base", d = dcauchy_vec, p = pcauchy_vec, q = qcauchy_vec, r = rcauchy_vec,
       params = list(location = 0, scale = 1.0)),
  list(label = "Mixture", d = dcauchymix, p = pcauchymix, q = qcauchymix, r = rcauchymix,
       params = list(w = c(0.6, 0.4), location = c(-1, 1), scale = c(1.0, 1.5)))
)

demo_table("Cauchy function demo (x = 0.5, p = 0.8, n = 3)", variants, x = 0.5)
```

```{r cauchy-plot, echo=FALSE}
grid <- seq(-8, 8, length.out = 400)
print(plot_density_overlay(grid, sets, "Cauchy densities "))
```

## Complete Function Reference

```{r full-ref, echo=FALSE, results="asis"}
ref_df <- data.frame(
  Distribution = c(
    "GPD",
    rep("Normal", 4),
    rep("Gamma", 4),
    rep("Lognormal", 4),
    rep("Laplace", 4),
    rep("Inverse Gaussian", 4),
    rep("Amoroso", 4),
    rep("Cauchy", 2)
  ),
  Variant = c(
    "Standalone",
    "Base", "Base + GPD", "Mixture", "Mixture + GPD",
    "Base", "Base + GPD", "Mixture", "Mixture + GPD",
    "Base", "Base + GPD", "Mixture", "Mixture + GPD",
    "Base", "Base + GPD", "Mixture", "Mixture + GPD",
    "Base", "Base + GPD", "Mixture", "Mixture + GPD",
    "Base", "Base + GPD", "Mixture", "Mixture + GPD",
    "Base", "Mixture"
  ),
  `Rd page` = c(
    rd_link("gpd"),
    "stats::Normal",
    rd_link("normal_gpd"),
    rd_link("normal_mix"),
    rd_link("normal_mixgpd"),
    "stats::GammaDist",
    rd_link("gamma_gpd"),
    rd_link("gamma_mix"),
    rd_link("gamma_mixgpd"),
    "stats::Lognormal",
    rd_link("lognormal_gpd"),
    rd_link("lognormal_mix"),
    rd_link("lognormal_mixgpd"),
    "nimble::ddexp",
    rd_link("laplace_gpd"),
    rd_link("laplace_mix"),
    rd_link("laplace_MixGpd"),
    rd_link("InvGauss"),
    rd_link("InvGauss_gpd"),
    rd_link("InvGauss_mix"),
    rd_link("InvGauss_mixgpd"),
    rd_link("amoroso"),
    rd_link("amoroso_gpd"),
    rd_link("amoroso_mix"),
    rd_link("amoroso_mixgpd"),
    rd_link("cauchy"),
    rd_link("cauchy_mix")
  ),
  `NIMBLE call` = c(
    "dGpd / pGpd / qGpd / rGpd",
    "dnorm / pnorm / qnorm / rnorm",
    "dNormGpd / pNormGpd / qNormGpd / rNormGpd",
    "dNormMix / pNormMix / qNormMix / rNormMix",
    "dNormMixGpd / pNormMixGpd / qNormMixGpd / rNormMixGpd",
    "dgamma / pgamma / qgamma / rgamma",
    "dGammaGpd / pGammaGpd / qGammaGpd / rGammaGpd",
    "dGammaMix / pGammaMix / qGammaMix / rGammaMix",
    "dGammaMixGpd / pGammaMixGpd / qGammaMixGpd / rGammaMixGpd",
    "dlnorm / plnorm / qlnorm / rlnorm",
    "dLognormalGpd / pLognormalGpd / qLognormalGpd / rLognormalGpd",
    "dLognormalMix / pLognormalMix / qLognormalMix / rLognormalMix",
    "dLognormalMixGpd / pLognormalMixGpd / qLognormalMixGpd / rLognormalMixGpd",
    "ddexp / pdexp / qdexp / rdexp",
    "dLaplaceGpd / pLaplaceGpd / qLaplaceGpd / rLaplaceGpd",
    "dLaplaceMix / pLaplaceMix / qLaplaceMix / rLaplaceMix",
    "dLaplaceMixGpd / pLaplaceMixGpd / qLaplaceMixGpd / rLaplaceMixGpd",
    "dInvGauss / pInvGauss / qInvGauss / rInvGauss",
    "dInvGaussGpd / pInvGaussGpd / qInvGaussGpd / rInvGaussGpd",
    "dInvGaussMix / pInvGaussMix / qInvGaussMix / rInvGaussMix",
    "dInvGaussMixGpd / pInvGaussMixGpd / qInvGaussMixGpd / rInvGaussMixGpd",
    "dAmoroso / pAmoroso / qAmoroso / rAmoroso",
    "dAmorosoGpd / pAmorosoGpd / qAmorosoGpd / rAmorosoGpd",
    "dAmorosoMix / pAmorosoMix / qAmorosoMix / rAmorosoMix",
    "dAmorosoMixGpd / pAmorosoMixGpd / qAmorosoMixGpd / rAmorosoMixGpd",
    "dCauchy / pCauchy / qCauchy / rCauchy",
    "dCauchyMix / pCauchyMix / qCauchyMix / rCauchyMix"
  ),
  `R call` = c(
    "dgpd / pgpd / qgpd / rgpd",
    "dnorm / pnorm / qnorm / rnorm",
    "dnormgpd / pnormgpd / qnormgpd / rnormgpd",
    "dnormmix / pnormmix / qnormmix / rnormmix",
    "dnormmixgpd / pnormmixgpd / qnormmixgpd / rnormmixgpd",
    "dgamma / pgamma / qgamma / rgamma",
    "dgammagpd / pgammagpd / qgammagpd / rgammagpd",
    "dgammamix / pgammamix / qgammamix / rgammamix",
    "dgammamixgpd / pgammamixgpd / qgammamixgpd / rgammamixgpd",
    "dlnorm / plnorm / qlnorm / rlnorm",
    "dlognormalgpd / plognormalgpd / qlognormalgpd / rlognormalgpd",
    "dlognormalmix / plognormalmix / qlognormalmix / rlognormalmix",
    "dlognormalmixgpd / plognormalmixgpd / qlognormalmixgpd / rlognormalmixgpd",
    "ddexp / pdexp / qdexp / rdexp",
    "dlaplacegpd / plaplacegpd / qlaplacegpd / rlaplacegpd",
    "dlaplacemix / plaplacemix / qlaplacemix / rlaplacemix",
    "dlaplacemixgpd / plaplacemixgpd / qlaplacemixgpd / rlaplacemixgpd",
    "dinvgauss / pinvgauss / qinvgauss / rinvgauss",
    "dinvgaussgpd / pinvgaussgpd / qinvgaussgpd / rinvgaussgpd",
    "dinvgaussmix / pinvgaussmix / qinvgaussmix / rinvgaussmix",
    "dinvgaussmixgpd / pinvgaussmixgpd / qinvgaussmixgpd / rinvgaussmixgpd",
    "damoroso / pamoroso / qamoroso / ramoroso",
    "damorosogpd / pamorosogpd / qamorosogpd / ramorosogpd",
    "damorosomix / pamorosomix / qamorosomix / ramorosomix",
    "damorosomixgpd / pamorosomixgpd / qamorosomixgpd / ramorosomixgpd",
    "dcauchy_vec / pcauchy_vec / qcauchy_vec / rcauchy_vec",
    "dcauchymix / pcauchymix / qcauchymix / rcauchymix"
  ),
  `Vector params` = c(
    "",
    "", "", "w, mean, sd", "w, mean, sd",
    "", "", "w, shape, scale", "w, shape, scale",
    "", "", "w, meanlog, sdlog", "w, meanlog, sdlog",
    "", "", "w, location, scale", "w, location, scale",
    "", "", "w, mean, shape", "w, mean, shape",
    "", "", "w, loc, scale, shape1, shape2", "w, loc, scale, shape1, shape2",
    "", "w, location, scale"
  ),
  `Scalar params` = c(
    "threshold, scale, shape",
    "mean, sd",
    "mean, sd, threshold, tail_scale, tail_shape",
    "",
    "threshold, tail_scale, tail_shape",
    "shape, scale",
    "shape, scale, threshold, tail_scale, tail_shape",
    "",
    "threshold, tail_scale, tail_shape",
    "meanlog, sdlog",
    "meanlog, sdlog, threshold, tail_scale, tail_shape",
    "",
    "threshold, tail_scale, tail_shape",
    "location, scale",
    "location, scale, threshold, tail_scale, tail_shape",
    "",
    "threshold, tail_scale, tail_shape",
    "mean, shape",
    "mean, shape, threshold, tail_scale, tail_shape",
    "",
    "threshold, tail_scale, tail_shape",
    "loc, scale, shape1, shape2",
    "loc, scale, shape1, shape2, threshold, tail_scale, tail_shape",
    "",
    "threshold, tail_scale, tail_shape",
    "location, scale",
    ""
  ),
  check.names = FALSE
)

print(knitr::kable(ref_df, align = "c", caption = "Complete distribution reference",
                   escape = FALSE, format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE, position = "center") %>%
  kableExtra::collapse_rows(columns = 1, valign = "top"))
```

## NIMBLE vs R functions

- **NIMBLE (CamelCase)**: scalar functions intended for model code.
- **R (lowercase)**: vectorized wrappers intended for analysis and plotting.
- The same parameter names are used; R versions vectorize over the first argument.

```{r sessioninfo, include=FALSE}
sessionInfo()
```


