---
<<<<<<< HEAD
title: "Available distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Available distributions}
=======
title: "Distributions"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Distributions}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
<<<<<<< HEAD
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
set.seed(1)
```

## Goal

DPmixGPD supports multiple *bulk kernels* for the mixture components, and can optionally splice a Generalized Pareto tail (GPD) beyond a threshold.

This vignette:

- summarizes the kernel choices,
- shows how the kernel choice appears in the API,
- gives quick distribution sanity checks using base R `d/p/q/r` functions.

## Kernel summary

Below is a compact summary of common bulk kernels and their usual parameters.

```{r}
knitr::kable(
  data.frame(
    kernel = c("normal", "gamma", "lognormal", "cauchy", "laplace", "invgauss", "amoroso"),
    typical_parameters = c(
      "mean, sd",
      "shape, rate",
      "meanlog, sdlog",
      "location, scale",
      "location, scale",
      "mean, shape",
      "location/scale/shape (family-specific)"
    )
  ),
  caption = "Common bulk kernels (names may vary by version)"
)
```

## Where you choose the kernel

The kernel is selected in `build_nimble_bundle()` via the `kernel` argument.

```{r, eval=FALSE}
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",      # or "crp"
  kernel  = "lognormal",
  GPD     = TRUE,
  components = 6,
  mcmc = list(niter = 500, nburnin = 100, thin = 2, nchains = 1, seed = 1)
)
```

## Quick distribution checks

These plots are *not* from the DP mixture model; they are just sanity checks for the underlying parametric families.

### Normal

```{r}
x <- seq(-4, 4, length.out = 400)
plot(x, dnorm(x, mean = 0, sd = 1), type = "l", xlab = "x", ylab = "density", main = "Normal(0,1) density")
```

### Gamma

```{r}
x <- seq(0, 12, length.out = 400)
plot(x, dgamma(x, shape = 2, rate = 1), type = "l", xlab = "x", ylab = "density", main = "Gamma(shape=2, rate=1) density")
```

### Lognormal

```{r}
x <- seq(0, 12, length.out = 400)
plot(x, dlnorm(x, meanlog = 0, sdlog = 0.6), type = "l", xlab = "x", ylab = "density", main = "Lognormal(meanlog=0, sdlog=0.6) density")
```

## Notes

- Mixture models use these families as *components*; the overall fitted distribution can be much more flexible.
- Turning `GPD = TRUE` adds EVT-motivated tail behavior beyond a threshold.
- When in doubt, start with `kernel = "normal"` for debugging, then switch to heavier-tailed kernels.

```{r sessioninfo, include=FALSE}
sessionInfo()
```
=======
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# Kernel families supported

DPmixGPD provides mixture kernels with optional GPD tails. Each kernel has a **bulk** version (e.g., `dGammaMix`) and a **bulk+GPD** version (e.g., `dGammaMixGpd`).

## Full list of kernels and parameters

| Kernel | Support | Parameters | Notes |
|---|---|---|---|
| Normal | $\mathbb{R}$ | `mean`, `sd` | Symmetric, light tails |
| Gamma | $(0,\infty)$ | `shape`, `scale` | Right‑skewed, positive support |
| Lognormal | $(0,\infty)$ | `meanlog`, `sdlog` | Log‑scale parameters |
| Laplace | $\mathbb{R}$ | `location`, `scale` | Sharper peak, exponential tails |
| Inverse Gaussian | $(0,\infty)$ | `mean`, `shape` | Positive, heavy right tail |
| Amoroso | $(0,\infty)$ (by default) | `loc`, `scale`, `shape1`, `shape2` | Very flexible; `scale` can flip support |
| Cauchy | $\mathbb{R}$ | `location`, `scale` | Heavy tails, weak moments |

**GPD tail parameters** (for `*MixGpd` functions):

- `threshold` (splicing point $u$)
- `tail_scale` ($\sigma$)
- `tail_shape` ($\xi$)

# Strengths and weaknesses (center vs tail)

- **Normal**: clean center, light tails; not ideal for heavy tails.
- **Gamma / Lognormal / InvGauss**: good for positive, right‑skewed outcomes.
- **Laplace**: sharp center, exponential tails; robust to mild outliers.
- **Cauchy**: extreme tails; use only when outliers dominate.
- **Amoroso**: most flexible; can mimic several positive‑support families.

# How kernels interact with GPD tails

The GPD tail is **spliced** above `threshold` so the bulk kernel governs central behavior while the GPD handles extreme exceedances. This isolates tail modeling from the bulk and avoids overfitting the center.

# Parameterization warnings

- **Gamma** uses `shape` and `scale` (not rate).
- **Lognormal** uses `meanlog` and `sdlog` (log‑scale parameters).
- **Laplace** uses `location` and `scale`.
- **Inverse Gaussian** uses `mean` and `shape`.
- **Amoroso** uses `loc`, `scale`, `shape1`, `shape2` and can flip support if `scale < 0`.

Always verify parameter meanings in the kernel help pages:

```{r kernel-help}
?dGammaMix
?dLognormalMix
?dLaplaceMix
?dInvGaussMix
?dAmorosoMix
?dCauchyMix
```

# What this vignette does NOT cover

- DP mixture modeling
- MCMC
- Backends
- Causal interpretation

Those are covered in the dedicated vignettes listed in [introduction](introduction.Rmd).
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
