---
title: "Available Distributions"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Available Distributions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(ggplot2)
library(kableExtra)
library(DPmixGPD)
set.seed(1)
```

## Overview

DPmixGPD supports multiple bulk kernels for the mixture components and can optionally splice a Generalized Pareto Distribution (GPD) tail beyond a threshold.

## Theory (brief)

Each kernel defines the component density $K(y; \\theta)$. The mixture density
is a convex combination of kernel components, and the GPD tail replaces the
bulk kernel for $y > u$ when tail augmentation is enabled. This vignette focuses
on the bulk kernel shapes.

## Kernel Summary

```{r}
kernel_df <- data.frame(
  Kernel = c("normal", "gamma", "lognormal", "cauchy", "laplace", "invgauss", "amoroso"),
  Parameters = c(
    "mean, sd",
    "shape, rate",
    "meanlog, sdlog",
    "location, scale",
    "location, scale",
    "mean, shape",
    "location, scale, shape"
  )
)

kable(kernel_df, align = "c", caption = "Available Bulk Kernels") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE, position = "center")
```

## Kernel Selection

The kernel is specified in `build_nimble_bundle()` via the `kernel` argument.

```{r, eval=FALSE}
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel  = "lognormal",
  GPD     = TRUE,
  components = 6,
  mcmc = list(niter = 500, nburnin = 100, thin = 2, nchains = 1, seed = 1)
)
```

## Distribution Visualizations

```{r}
x_all <- seq(-4, 4, length.out = 300)
x_pos <- seq(0.01, 8, length.out = 300)

dens_list <- list(
  Normal = data.frame(x = x_all, density = dNormMix(x_all, w = 1, mean = 0, sd = 1)),
  Cauchy = data.frame(x = x_all, density = dCauchyMix(x_all, w = 1, location = 0, scale = 1)),
  Laplace = data.frame(x = x_all, density = dLaplaceMix(x_all, w = 1, location = 0, scale = 1)),
  Gamma = data.frame(x = x_pos, density = dGammaMix(x_pos, w = 1, shape = 2, scale = 1)),
  Lognormal = data.frame(x = x_pos, density = dLognormalMix(x_pos, w = 1, meanlog = 0, sdlog = 0.6)),
  InvGauss = data.frame(x = x_pos, density = dInvGaussMix(x_pos, w = 1, mean = 1, shape = 2)),
  Amoroso = data.frame(x = x_pos, density = dAmorosoMix(x_pos, w = 1, loc = 0, scale = 1, shape1 = 1.5, shape2 = 2))
)

plot_df <- do.call(rbind, lapply(names(dens_list), function(name) {
  df <- dens_list[[name]]
  df$Kernel <- name
  df
}))

ggplot(plot_df, aes(x = x, y = density, color = Kernel)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ Kernel, scales = "free") +
  labs(x = "x", y = "Density", title = "Bulk Kernel Shapes (Vectorized DPmixGPD Functions)") +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r sessioninfo, include=FALSE}
sessionInfo()
```
