---
<<<<<<< HEAD
<<<<<<< HEAD
title: "DPmixGPD: quick start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DPmixGPD: quick start}
=======
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
title: "Introduction"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Introduction}
<<<<<<< HEAD
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
<<<<<<< HEAD
<<<<<<< HEAD
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
library(nimble)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Goal

This vignette is a fast, end-to-end tour:

- build a model specification,
- run MCMC,
- extract fitted values and predictions.

**Runtime:** designed to run quickly in "FAST" mode.

## What DPmixGPD models

DPmixGPD fits flexible mixture models for the *bulk* of the distribution and can splice a Generalized Pareto tail beyond a threshold. In practice, you use it when:

- the center of the data is not well described by a single parametric family, and
- the extreme right tail needs principled extrapolation.

## A minimal run

```{r minimal-run}
library(DPmixGPD)

# A toy heavy-ish tail sample
n <- 80
y <- abs(rnorm(n)) + 0.15

bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE)
fit
```

## Fitted values and residuals

```{r fitted}
f <- fitted(fit, type = "mean", level = 0.90)
head(f)

# quick residual sanity check
summary(f$residuals)
```

## Predictions

```{r predict}
# For unconditional models, predict() returns population-level summaries
pred_mean <- predict(fit, type = "mean", cred.level = 0.90, interval = "credible")
pred_q90  <- predict(fit, type = "quantile", index = 0.90, cred.level = 0.90, interval = "credible")

pred_mean$fit
pred_q90$fit
```

## A quick diagnostic plot

```{r plot-fit}
# Plot methods may vary by version; keep this simple.
# If your plot() method supports a family argument, trace plots are the safest.
try(plot(fit, family = "trace"), silent = TRUE)
```

## Troubleshooting

- **"keywords: if" from NIMBLE**: a covariate column is named `if` (or another reserved keyword). Rename columns (e.g., `if` -> `x_if`).
- **"No space left on device" during build/check**: set `TMPDIR`/`TEMP`/`TMP` to a drive with free space (e.g., `D:/Rtmp`).
- **Coverage looks low**: code generation (e.g., `eval(parse())`) is hard to trace. Focus tests on user-facing functions.

## Where to go next

- See **Model specification** for all options.
- See **Unconditional** for density and tail diagnostics.
- See **Backends** for SB vs CRP comparisons.

```{r sessioninfo, include=FALSE}
sessionInfo()
```
=======
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  comment = NA,
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# What problem does DPmixGPD solve?

DPmixGPD models **entire outcome distributions**, not just means. It is designed for:

- **Distributional modeling** (shape, skew, multi-modality)
- **Extremes and tail risk** (via optional GPD tails)
- **Causal contrast of distributions** when identification is provided externally

If your outcome is heavy‑tailed, skewed, or multi‑modal, DPmixGPD provides a flexible mixture‑plus‑tail approach that can be used in unconditional, conditional, or group‑comparison settings.

# One unified framework

DPmixGPD is one engine with three common use cases:

1. **Unconditional**: fit a flexible mixture to the bulk of the data.
2. **Conditional**: let mixture parameters depend on covariates via links.
3. **Causal**: fit **separate outcome models by treatment group** and compare distributions.

The model structure stays the same:

- **Bulk**: Dirichlet process mixture over a kernel family.
- **Tail (optional)**: splice a Generalized Pareto Distribution (GPD) above a threshold.

# What “causal” means in this package

DPmixGPD **does not identify causal effects on its own**. It provides tools to **model outcome distributions** for treated and control groups and then compute **distributional contrasts** (e.g., CQTE) **if** your identification assumptions are satisfied outside the package.

This package:

- **does** fit outcome models and compute contrasts
- **does not** supply identification, randomization, or validity guarantees

# Motivating example (conceptual)

Imagine a policy that changes not only the **mean** of earnings but also the **tail risk** (very high or very low outcomes). A mean‑only analysis would miss distributional shifts. DPmixGPD lets you fit flexible outcome distributions for each group and compare effects at specific quantiles or in the tail.

# Where to go next

- **Distributions**: [distributions](distributions.Rmd) for kernel choices and parameter meanings.
- **Model specification**: [model-spec](model-spec.Rmd) for what you build vs what is generated.
- **MCMC workflow**: [mcmc-workflow](mcmc-workflow.Rmd) for responsible inference.
- **Worked examples**:
  - [unconditional](unconditional.Rmd)
  - [conditional](conditional.Rmd)
  - [causal](causal.Rmd)
- **Backends**: [backends](backends.Rmd)
- **S3 reference**: [reference-s3](reference-s3.Rmd)
<<<<<<< HEAD
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
