---
title: "DPmixGPD: Quick Start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DPmixGPD: Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(DPmixGPD)
library(ggplot2)
library(kableExtra)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Overview

This vignette provides an end-to-end introduction to the DPmixGPD workflow:

- Build a model specification
- Run MCMC sampling
- Extract fitted values and predictions

## Theory (brief)

DPmixGPD models the bulk of a distribution with a Dirichlet process (DP) mixture,
then optionally splices a Generalized Pareto Distribution (GPD) tail beyond a
threshold $u$. For outcomes $y_i$ and kernel $K$, the bulk model is
$$
f(y_i) = \\int K(y_i; \\theta)\\, dG(\\theta), \\quad G \\sim \\mathrm{DP}(\\alpha, G_0).
$$
When a tail is included, the density is replaced for $y > u$ by a GPD tail with
scale and shape parameters, preserving continuity at $u$.

## Model Description

DPmixGPD fits flexible mixture models for the bulk of the distribution and can splice a Generalized Pareto Distribution (GPD) tail beyond a threshold. This approach is appropriate when:
 
- The center of the data is not well described by a single parametric family
- The extreme right tail requires principled extrapolation

## Minimal Example

```{r minimal-run, results='hide'}
library(DPmixGPD)

data("faithful", package = "datasets")
y <- faithful$eruptions

bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE)
```

## Fitted Values and Residuals

```{r fitted}
f <- fitted(fit, type = "mean", level = 0.90)
head(f)
summary(f$residuals)
```

## Predictions

```{r predict}
pred_mean <- predict(fit, type = "mean", cred.level = 0.90, interval = "credible")
pred_q90  <- predict(fit, type = "quantile", index = 0.90, cred.level = 0.90, interval = "credible")

pred_mean$fit
pred_q90$fit
```

## Diagnostic Plots

```{r plot-fit}
if (requireNamespace("ggmcmc", quietly = TRUE) && requireNamespace("coda", quietly = TRUE)) {
  plot(fit)
} else {
  message("Plotting requires 'ggmcmc' and 'coda' packages.")
}
```

## Troubleshooting

- **NIMBLE keyword error**: Rename covariate columns that use reserved keywords (e.g., `if` to `x_if`).
- **Disk space error**: Set `TMPDIR`/`TEMP`/`TMP` to a drive with sufficient free space.

## Next Steps

- **Model Specification**: Complete documentation of all options
- **Unconditional Models**: Density estimation and tail diagnostics
- **Backends**: Comparison of SB and CRP backends

```{r sessioninfo, include=FALSE}
sessionInfo()
```
