---
title: "DPmixGPD: quick start"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{DPmixGPD: quick start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Goal

This vignette is a fast, end-to-end tour:

- build a model specification,
- run MCMC,
- extract fitted values and predictions.

**Runtime:** designed to run quickly in "FAST" mode.

## What DPmixGPD models

DPmixGPD fits flexible mixture models for the *bulk* of the distribution and can splice a Generalized Pareto tail beyond a threshold. In practice, you use it when:

- the center of the data is not well described by a single parametric family, and
- the extreme right tail needs principled extrapolation.

## A minimal run

```{r minimal-run}
library(DPmixGPD)

# A toy heavy-ish tail sample
n <- 80
y <- abs(rnorm(n)) + 0.15

bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE)
fit
```

## Fitted values and residuals

```{r fitted}
f <- fitted(fit, type = "mean", level = 0.90)
head(f)

# quick residual sanity check
summary(f$residuals)
```

## Predictions

```{r predict}
# For unconditional models, predict() returns population-level summaries
pred_mean <- predict(fit, type = "mean", cred.level = 0.90, interval = "credible")
pred_q90  <- predict(fit, type = "quantile", index = 0.90, cred.level = 0.90, interval = "credible")

pred_mean$fit
pred_q90$fit
```

## A quick diagnostic plot

```{r plot-fit}
# Plot methods may vary by version; keep this simple.
# If your plot() method supports a family argument, trace plots are the safest.
try(plot(fit, family = "trace"), silent = TRUE)
```

## Troubleshooting

- **"keywords: if" from NIMBLE**: a covariate column is named `if` (or another reserved keyword). Rename columns (e.g., `if` -> `x_if`).
- **"No space left on device" during build/check**: set `TMPDIR`/`TEMP`/`TMP` to a drive with free space (e.g., `D:/Rtmp`).
- **Coverage looks low**: code generation (e.g., `eval(parse())`) is hard to trace. Focus tests on user-facing functions.

## Where to go next

- See **Model specification** for all options.
- See **Unconditional** for density and tail diagnostics.
- See **Backends** for SB vs CRP comparisons.

```{r sessioninfo, include=FALSE}
sessionInfo()
```
