---
title: "Kernel reference: Lognormal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kernel reference: Lognormal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  cache = TRUE,
  cache.path = "../.cache/vignettes/"
)
library(DPmixGPD)
```

The **Lognormal** kernel is parameterized by `meanlog` and `sdlog` (i.e., `log(y)` is Normal).

## Exported mixture helpers

- `dLognormalMix(x, meanlog, sdlog, w, log = FALSE)`
- `pLognormalMix(q, meanlog, sdlog, w, lower.tail = TRUE, log.p = FALSE)`
- `qLognormalMix(p, meanlog, sdlog, w, lower.tail = TRUE, log.p = FALSE)`
- `rLognormalMix(n, meanlog, sdlog, w)`

For the full catalog (including CRP utilities and GPD splicing), see:

- [Available Distributions](v02-available-distributions.html)

## Using Lognormal in a model

```{r}
y <- abs(stats::rnorm(50)) + 0.1
bundle <- build_nimble_bundle(
  y = y,
  backend = "crp",
  kernel  = "lognormal",
  GPD     = TRUE,
  components = 6
)
```



