---
title: "Kernel reference: Normal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kernel reference: Normal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
library(DPmixGPD)
```

The **Normal** kernel is parameterized by mean `mean` and standard deviation `sd`.

## Exported mixture helpers

For the stick-breaking (SB) backend, the package provides vectorized mixture helpers:

- `dNormMix(x, mean, sd, w, log = FALSE)`
- `pNormMix(q, mean, sd, w, lower.tail = TRUE, log.p = FALSE)`
- `qNormMix(p, mean, sd, w, lower.tail = TRUE, log.p = FALSE)`
- `rNormMix(n, mean, sd, w)`

For a full catalog (including CRP utilities and GPD splicing), see:

- [Available Distributions](v02-available-distributions.html)

## Using Normal in a model

```{r}
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel  = "normal",
  GPD     = TRUE,
  Kmax    = 6
)
```
