---
title: "Kernel reference: Laplace"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kernel reference: Laplace}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE,
  cache = TRUE,
  cache.path = "../.cache/vignettes/"
)
library(DPmixGPD)
```

The **Laplace** kernel is parameterized by `location` and `scale`.

## Exported mixture helpers

- `dLaplaceMix(x, location, scale, w, log = FALSE)`
- `pLaplaceMix(q, location, scale, w, lower.tail = TRUE, log.p = FALSE)`
- `qLaplaceMix(p, location, scale, w, lower.tail = TRUE, log.p = FALSE)`
- `rLaplaceMix(n, location, scale, w)`

For the full catalog (including CRP utilities and GPD splicing), see:

- [Available Distributions](v02-available-distributions.html)

## Using Laplace in a model

```{r}
y <- abs(stats::rnorm(50)) + 0.1
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel  = "laplace",
  GPD     = TRUE,
  components = 6
)
```



