---
title: "MCMC workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCMC workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
library(nimble)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 6000, nburnin = 2000, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Goal

This vignette shows a practical MCMC loop: build -> run -> inspect -> extract.

**Runtime:** fast in `FAST` mode.

## Build and run

```{r}
library(DPmixGPD)

y <- abs(rnorm(60)) + 0.15
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE)
fit
```

## Print and summary

```{r}
print(fit)
summary(fit)
```

## Plots

Different versions of the package may support different plot families; `trace` is usually the safest.

```{r}
try(plot(fit, family = "trace"), silent = TRUE)
```

## Extract posterior draws

The exact storage depends on how your fit object is organized, but most workflows expose a draws object.

```{r}
# Common patterns
if (!is.null(fit$mcmc$samples)) {
  s <- fit$mcmc$samples
  mat <- try(as.matrix(s), silent = TRUE)
  if (!inherits(mat, "try-error")) {
    dim(mat)
    colnames(mat)[1:min(20, ncol(mat))]
  }
}
```

## Re-running with different MCMC settings

A helpful pattern is: build once, rerun MCMC with different settings. If your API supports it, use it.
Otherwise rebuild with a different `mcmc` list.

```{r, eval=FALSE}
# Example (pseudo):
# bundle2 <- update_mcmc(bundle, niter = 8000, nburnin = 2000)
# fit2 <- run_mcmc_bundle_manual(bundle2)
```

## Troubleshooting

- **Compilation errors**: call `printErrors()` if NIMBLE suggests it.
- **Slow runs**: start with `FAST <- TRUE`, then scale up.
- **Multi-chain seeds**: prefer a vector of seeds (one per chain).

## Next

- See **Unconditional** and **Conditional** for prediction and diagnostics.

```{r sessioninfo, include=FALSE}
sessionInfo()
```
