---
<<<<<<< HEAD
title: "MCMC workflow"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{MCMC workflow}
=======
title: "MCMC Workflow"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{MCMC Workflow}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
<<<<<<< HEAD
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
library(nimble)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 6000, nburnin = 2000, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Goal

This vignette shows a practical MCMC loop: build -> run -> inspect -> extract.

**Runtime:** fast in `FAST` mode.

## Build and run

```{r}
library(DPmixGPD)

y <- abs(rnorm(60)) + 0.15
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE)
fit
```

## Print and summary

```{r}
print(fit)
summary(fit)
```

## Plots

Different versions of the package may support different plot families; `trace` is usually the safest.

```{r}
try(plot(fit, family = "trace"), silent = TRUE)
```

## Extract posterior draws

The exact storage depends on how your fit object is organized, but most workflows expose a draws object.

```{r}
# Common patterns
if (!is.null(fit$mcmc$samples)) {
  s <- fit$mcmc$samples
  mat <- try(as.matrix(s), silent = TRUE)
  if (!inherits(mat, "try-error")) {
    dim(mat)
    colnames(mat)[1:min(20, ncol(mat))]
  }
}
```

## Re-running with different MCMC settings

A helpful pattern is: build once, rerun MCMC with different settings. If your API supports it, use it.
Otherwise rebuild with a different `mcmc` list.

```{r, eval=FALSE}
# Example (pseudo):
# bundle2 <- update_mcmc(bundle, niter = 8000, nburnin = 2000)
# fit2 <- run_mcmc_bundle_manual(bundle2)
```

## Troubleshooting

- **Compilation errors**: call `printErrors()` if NIMBLE suggests it.
- **Slow runs**: start with `FAST <- TRUE`, then scale up.
- **Multi-chain seeds**: prefer a vector of seeds (one per chain).

## Next

- See **Unconditional** and **Conditional** for prediction and diagnostics.

```{r sessioninfo, include=FALSE}
sessionInfo()
```
=======
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# What MCMC settings mean

Key knobs in `mcmc = list(...)`:

- `niter`: total iterations
- `nburnin`: burn‑in discarded
- `thin`: thinning interval
- `nchains`: number of chains
- `seed`: reproducibility

Short runs are useful for **debugging**, not for inference.

# What short runs can and cannot tell you

**Short runs can**:

- verify model compiles
- check for coding mistakes
- produce quick sanity plots

**Short runs cannot**:

- guarantee convergence
- stabilize tail estimates
- justify causal or distributional claims

# Basic diagnostics philosophy

You should check at least:

- trace plots for major parameters
- mixing of mixture weights and kernel parameters
- stability of posterior predictive summaries

# Common failure modes

- **Compilation errors**: often missing data dimensions or unsupported kernels.
- **Poor mixing**: can be due to weak identifiability or too few components.
- **Identifiability issues**: especially with flexible kernels or tail parameters.
- **Over‑complex models**: too many components for small $n$.

# Practical guidance

- Start with **small** `J` (e.g., 3–5) and increase only if needed.
- Use **simple kernels** first (e.g., gamma/normal) before amoroso or cauchy.
- Treat tail parameters as high‑variance; stabilize with enough data above the threshold.

Example workflow (non‑executed):

```{r mcmc-example}
bundle <- build_nimble_bundle(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = TRUE,
  J = 5,
  mcmc = list(niter = 1000, nburnin = 250, nchains = 2)
)
fit <- run_mcmc_bundle_manual(bundle)
plot(fit, family = c("traceplot", "running"))
```

# When *not* to trust results

- single‑chain, very short runs
- widely divergent traces
- effective sample sizes near zero
- tail estimates based on very few exceedances

This package gives you flexible models; it does not replace careful MCMC practice.
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
