---
title: "MCMC Workflow"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{MCMC Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# What MCMC settings mean

Key knobs in `mcmc = list(...)`:

- `niter`: total iterations
- `nburnin`: burn‑in discarded
- `thin`: thinning interval
- `nchains`: number of chains
- `seed`: reproducibility

Short runs are useful for **debugging**, not for inference.

# What short runs can and cannot tell you

**Short runs can**:

- verify model compiles
- check for coding mistakes
- produce quick sanity plots

**Short runs cannot**:

- guarantee convergence
- stabilize tail estimates
- justify causal or distributional claims

# Basic diagnostics philosophy

You should check at least:

- trace plots for major parameters
- mixing of mixture weights and kernel parameters
- stability of posterior predictive summaries

# Common failure modes

- **Compilation errors**: often missing data dimensions or unsupported kernels.
- **Poor mixing**: can be due to weak identifiability or too few components.
- **Identifiability issues**: especially with flexible kernels or tail parameters.
- **Over‑complex models**: too many components for small $n$.

# Practical guidance

- Start with **small** `J` (e.g., 3–5) and increase only if needed.
- Use **simple kernels** first (e.g., gamma/normal) before amoroso or cauchy.
- Treat tail parameters as high‑variance; stabilize with enough data above the threshold.

Example workflow (non‑executed):

```{r mcmc-example}
bundle <- build_nimble_bundle(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = TRUE,
  J = 5,
  mcmc = list(niter = 1000, nburnin = 250, nchains = 2)
)
fit <- run_mcmc_bundle_manual(bundle)
plot(fit, family = c("traceplot", "running"))
```

# When *not* to trust results

- single‑chain, very short runs
- widely divergent traces
- effective sample sizes near zero
- tail estimates based on very few exceedances

This package gives you flexible models; it does not replace careful MCMC practice.
