---
title: "Model Specification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
library(ggplot2)
library(kableExtra)
set.seed(1)
```

## Overview

This vignette documents the inputs that define a DPmixGPD model and how they map to a compiled NIMBLE bundle.

## Core Constructor

Most workflows begin with `build_nimble_bundle()`.

```{r}
library(DPmixGPD)

y <- abs(rnorm(40)) + 0.2
X <- data.frame(x = rnorm(40))

bundle <- build_nimble_bundle(
  y = y,
  X = X,
  backend = "sb",
  kernel  = "normal",
  GPD     = TRUE,
  components = 6,
  mcmc = list(niter = 200, nburnin = 50, thin = 2, nchains = 1, seed = 1)
)

bundle
```

## Key Arguments

```{r, echo=FALSE}
args_df <- data.frame(
  Argument = c("backend", "backend", "kernel", "GPD", "GPD"),
  Value = c("sb", "crp", "(various)", "TRUE", "FALSE"),
  Description = c(
    "Stick-breaking mixture (finite truncation)",
    "Chinese Restaurant Process mixture",
    "Bulk component family (normal, gamma, lognormal, ...)",
    "Splice GPD tail beyond threshold",
    "Bulk-only model"
  )
)

kable(args_df, align = "c", caption = "Model Specification Arguments") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE, position = "center")
```

## Model Types

| Type | Specification | Prediction Behavior |
|------|---------------|---------------------|
| **Unconditional** | Omit `X` or pass `X = NULL` | Population-level, replicated across observations |
| **Conditional** | Provide `X` | Computed per-row of `X` |

## Input Requirements

- For conditional models, `length(y)` must equal `nrow(X)`
- Covariates must be numeric for regression-style links
- Covariate names must not use NIMBLE reserved keywords (`if`, `for`, `while`, etc.)

```{r, eval=FALSE}
# Rename reserved keywords
names(X)[names(X) == "if"] <- "x_if"
```

