---
<<<<<<< HEAD
title: "Model specification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model specification}
=======
title: "Model Specification"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Model Specification}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
<<<<<<< HEAD
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
set.seed(1)
```

## Goal

This vignette documents the inputs that define a DPmixGPD model and how they map to a compiled NIMBLE bundle.

**Runtime:** fast (no long MCMC).

## The core constructor

Most workflows start with `build_nimble_bundle()`.

```{r}
library(DPmixGPD)

y <- abs(rnorm(40)) + 0.2
X <- data.frame(x = rnorm(40))

bundle <- build_nimble_bundle(
  y = y,
  X = X,
  backend = "sb",
  kernel  = "normal",
  GPD     = TRUE,
  components = 6,
  mcmc = list(niter = 200, nburnin = 50, thin = 2, nchains = 1, seed = 1)
)

bundle
```

## Key arguments

### `backend`

- `"sb"`: stick-breaking mixture (finite truncation).
- `"crp"`: Chinese Restaurant Process mixture.

### `kernel`

Controls the bulk component family (e.g., `"normal"`, `"gamma"`, `"lognormal"`, ...).

### `GPD`

- `TRUE`: splice a GPD tail beyond a threshold.
- `FALSE`: bulk-only model.

## Conditional vs unconditional

- **Unconditional**: omit `X` (or pass `X = NULL`). Predictions are population-level and replicated across observations.
- **Conditional**: provide `X`. Predictions are computed per-row of `X`.

## Common input hazards

### Reserved variable names (NIMBLE keywords)

If a covariate column is named `if`, `for`, `while`, etc., NIMBLE compilation fails.

```{r, eval=FALSE}
X <- data.frame(if = rnorm(40))
# Fix:
names(X)[names(X) == "if"] <- "x_if"
```

### Shape checks

- `length(y)` must equal `nrow(X)` for conditional models.
- Covariates should be numeric for regression-style links.

## Validity and error messages

Good packages fail loudly and early. If your build step does not yet validate inputs, consider adding checks that:

- confirm supported `backend` and `kernel` values,
- prevent reserved names,
- validate the `mcmc` list.

## Next

- See **MCMC workflow** for running chains and extracting samples.
- See **Backends** for SB vs CRP differences.

=======
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# What is a “bundle”?

A **bundle** is the complete, pre‑MCMC object created by `build_nimble_bundle()` (or the causal builders). It includes:

- NIMBLE model code
- data/constants/dimensions
- initialization function
- monitor list
- stored MCMC settings

It is intentionally **pre‑run**. You execute MCMC later with `run_mcmc_bundle_manual()`.

# What users specify vs. what is generated

**You specify**:

- data (`y`, and optionally `X` or propensity scores)
- kernel family and GPD option
- backend (`"sb"` or `"crp"`)
- truncation (`J`/`components`)
- MCMC settings (stored in the bundle)
- optional `param_specs` overrides

**The package generates**:

- mixture code for the chosen kernel
- link‑mode covariate structure
- priors (default or user‑overridden)
- monitors and dimensions

# Role of priors and links

DPmixGPD supports **link‑mode** parameters. When a parameter is linked to covariates, the model introduces regression coefficients named `beta_<param>` (component‑specific) and optionally `beta_ps_<param>` (propensity score covariate). Priors for these coefficients are set via `param_specs`.

Example (non‑executed):

```{r bundle-example}
bundle <- build_nimble_bundle(
  y = y,
  X = X,
  kernel = "gamma",
  backend = "crp",
  GPD = TRUE,
  J = 5,
  mcmc = list(niter = 200, nburnin = 50, nchains = 1)
)
```

# Conditional structure

When covariates are present (`X`), selected parameters are encoded as

$$\theta_{ik} = g^{-1}(X_i^\top \beta_{k})$$

where $k$ indexes mixture components and $g$ is the link function (e.g., identity or log).

# How causal models reuse the same structure

Causal workflows **reuse the same mixture model**. The difference is that the outcome model is fit **separately by treatment group**, and contrasts are computed across the resulting distributions. The identification assumptions are external to the package.

# Internal naming conventions (high‑level)

- Component regression coefficients: `beta_<param>[1:K,1:P]`
- Propensity score coefficients (if used): `beta_ps_<param>[1:K]`
- GPD threshold link coefficients: `beta_threshold[1:P]`
- GPD tail scale link coefficients: `beta_tail_scale[1:P]`

These names appear in model code, monitors, and posterior summaries.

# What this vignette does NOT cover

- Diagnostics or convergence assessment
- Long examples
- Interpretation of effects

For MCMC practice, see [mcmc-workflow](mcmc-workflow.Rmd).
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
