---
title: "Model Specification"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model Specification}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
library(ggplot2)
library(kableExtra)
set.seed(1)
```

## Overview

This vignette describes the probabilistic model behind DPmixGPD and how user
inputs map to a compiled NIMBLE bundle.

## Theory (core model)

### Bulk mixture

For outcomes $y_i$ and kernel $K$, the bulk distribution is modeled as a DP
mixture:
$$
f(y_i) = \\int K(y_i; \\theta)\\, dG(\\theta), \\quad G \\sim \\mathrm{DP}(\\alpha, G_0).
$$
The DP prior on $G$ induces clustering of component parameters and allows the
mixture to adapt to unknown distributional shapes.

### Tail augmentation (optional)

When $\\mathrm{GPD} = \\mathrm{TRUE}$, a Generalized Pareto tail replaces the
bulk kernel beyond a threshold $u$:
$$
f(y) =
\\begin{cases}
f_{\\text{bulk}}(y), & y \\le u \\\\
f_{\\text{GPD}}(y; u, \\sigma, \\xi), & y > u
\\end{cases}
$$
with continuity enforced at the threshold.

### Conditional extension

Conditional models allow kernel parameters to depend on covariates $x_i$:
$$
f(y_i \\mid x_i) = \\int K(y_i; \\theta(x_i))\\, dG(\\theta).
$$
The linking structure is kernel-specific and implemented in the model builder.

## Core constructor

Most workflows begin with `build_nimble_bundle()`. The bundle includes compiled
NIMBLE code and MCMC settings.

```{r}
library(DPmixGPD)

data("faithful", package = "datasets")
y <- faithful$eruptions

bundle_uncond <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel  = "normal",
  GPD     = TRUE,
  components = 6,
  mcmc = list(niter = 200, nburnin = 50, thin = 2, nchains = 1, seed = 1)
)
```

## Conditional example

```{r}
data("mtcars", package = "datasets")
df <- mtcars
X <- df[, c("wt", "hp")]
X <- as.data.frame(X)
y <- df$mpg

bundle_cond <- build_nimble_bundle(
  y = y,
  X = X,
  backend = "sb",
  kernel  = "normal",
  GPD     = TRUE,
  components = 6,
  mcmc = list(niter = 200, nburnin = 50, thin = 2, nchains = 1, seed = 1)
)
```

## Key arguments

```{r, echo=FALSE}
args_df <- data.frame(
  Argument = c("backend", "backend", "kernel", "GPD", "GPD", "components", "alpha_random"),
  Value = c("sb", "crp", "(various)", "TRUE", "FALSE", "K", "TRUE/FALSE"),
  Description = c(
    "Stick-breaking mixture (finite truncation)",
    "Chinese Restaurant Process mixture",
    "Bulk component family (normal, gamma, lognormal, ...)",
    "Splice GPD tail beyond threshold",
    "Bulk-only model",
    "Truncation level for SB (maximum components)",
    "Learn concentration parameter (DP strength)"
  )
)

kable(args_df, align = "c", caption = "Model Specification Arguments") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE, position = "center")
```

## Model types

| Type | Specification | Prediction Behavior |
|------|---------------|---------------------|
| **Unconditional** | Omit `X` or pass `X = NULL` | Population-level, replicated across observations |
| **Conditional** | Provide `X` | Computed per-row of `X` |

## Input requirements

- For conditional models, `length(y)` must equal `nrow(X)`
- Covariates must be numeric for regression-style links
- Covariate names must not use NIMBLE reserved keywords (`if`, `for`, `while`, etc.)

```{r, eval=FALSE}
# Rename reserved keywords
names(X)[names(X) == "if"] <- "x_if"
```


