---
title: "S3 Methods Reference"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{S3 Methods Reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# Core S3 methods

| Method | Object | Purpose | Output |
|---|---|---|---|
| `print()` | `dpmixgpd_bundle` | quick bundle summary | concise configuration | 
| `summary()` | `dpmixgpd_bundle` | detailed bundle summary | lists code/data/monitors |
| `print()` | `mixgpd_fit` | fit summary | kernel/backends/mcmc info |
| `summary()` | `mixgpd_fit` | posterior summary | parameter tables |
| `plot()` | `mixgpd_fit` | diagnostics | trace, running mean, autocorrelation |
| `predict()` | `mixgpd_fit` | posterior predictive | density, quantile, mean, survival |

# Parameter naming conventions

Common parameter names in outputs:

- `w` (mixture weights)
- `alpha` (concentration)
- kernel parameters (e.g., `mean`, `sd`, `shape`, `scale`)
- link coefficients: `beta_<param>`
- GPD links: `beta_threshold`, `beta_tail_scale`

# Typical usage patterns

```{r s3-example}
fit <- run_mcmc_bundle_manual(bundle)
print(fit)
summary(fit)
plot(fit, family = c("traceplot", "running"))
q_pred <- predict(fit, type = "quantile", index = c(0.1, 0.5, 0.9))
plot(q_pred)
```

# Notes

- `predict()` returns a structured object with `fit` and interval summaries.
- `plot()` accepts families such as `traceplot`, `running`, and `autocorrelation`.
- Use `params()` for convenient access to posterior draws in original scales.
