---
title: "Unconditional Models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unconditional Models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(ggplot2)
library(kableExtra)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 8000, nburnin = 2000, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Overview

This vignette demonstrates fitting an unconditional DP mixture model with optional GPD tail augmentation.

## Theory (brief)

Unconditional models treat observations as exchangeable draws from a mixture
distribution. The DP prior on the mixing measure $G$ yields flexible density
estimation:
$$
y_i \\mid G \\sim \\int K(y_i; \\theta)\\, dG(\\theta), \\quad G \\sim \\mathrm{DP}(\\alpha, G_0).
$$
When $\\mathrm{GPD} = \\mathrm{TRUE}$, a GPD tail replaces the bulk kernel beyond a
threshold $u$ to stabilize tail behavior.

## Model Fitting

```{r, results='hide'}
library(DPmixGPD)

data("faithful", package = "datasets")
y <- faithful$eruptions

bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE)
```

## Fitted Summaries

```{r}
f_mean <- fitted(fit, type = "mean", level = 0.90)
head(f_mean)

f_med <- fitted(fit, type = "median", level = 0.90)
head(f_med)
```

## Posterior Predictive Summaries

```{r}
pred_mean <- predict(fit, type = "mean", cred.level = 0.90, interval = "credible")
pred_q95  <- predict(fit, type = "quantile", index = 0.95, cred.level = 0.90, interval = "credible")

pred_mean$fit
pred_q95$fit
```

## Residual Analysis

```{r}
res <- f_mean$residuals
summary(res)
```

## Diagnostic Plots

```{r}
if (requireNamespace("ggmcmc", quietly = TRUE) && requireNamespace("coda", quietly = TRUE)) {
  plot(fit)
} else {
  message("Plotting requires 'ggmcmc' and 'coda' packages.")
}
```

```{r sessioninfo, include=FALSE}
sessionInfo()
```

