---
title: "DPmixGPD: Unconditional workflow (bulk + optional GPD tail)"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{DPmixGPD: Unconditional workflow (bulk + optional GPD tail)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: dpmixgpd.bib
link-citations: true
---

```{r setup, include=FALSE}
options(dpmixgpd.knitr.kable = TRUE)
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)

.extdata <- function(fname) {
  # Approach 1: Prefer local artifacts in the source tree.
  local_paths <- c(
    file.path("inst", "extdata", fname),
    file.path("..", "inst", "extdata", fname),
    file.path(getwd(), "inst", "extdata", fname)
  )
  for (path in local_paths) {
    if (file.exists(path)) {
      return(normalizePath(path, winslash = "/", mustWork = TRUE))
    }
  }

  # Approach 2: Try system.file() for installed packages.
  path <- system.file("extdata", fname, package = "DPmixGPD")
  if (nzchar(path) && file.exists(path)) {
    return(path)
  }

  # Approach 3: Use find.package() to locate package root.
  tryCatch({
    pkg_path <- find.package("DPmixGPD")
    path <- file.path(pkg_path, "extdata", fname)
    if (file.exists(path)) {
      return(normalizePath(path, winslash = "/", mustWork = TRUE))
    }
  }, error = function(e) {})

  stop("Cannot find artifact '", fname, "' in any expected location.\n",
       "Please run tools/make_vignette_fits.R to generate precomputed artifacts.")
}
```

This vignette shows the end-to-end workflow for an unconditional model: build a bundle, run posterior sampling, and compute distributional predictions. The mathematical specification of the DPM bulk model and the spliced GPD tail is defined once in **“Model specification and posterior computation”** (the `basic` vignette). Here we focus on usage.

DPmixGPD ships small example datasets; we use `nc_pos_tail200_k4`, which is a positive-support dataset with an injected upper tail (so `GPD=TRUE` is meaningful).

```{r}
library(DPmixGPD)
data("nc_pos_tail200_k4")
y <- nc_pos_tail200_k4$y
```

Model fit (code shown for reproducibility, but not executed during CRAN checks)

```r
bundle <- build_nimble_bundle(
  y = y,
  X = NULL,
  backend = "sb",         # "sb" or "crp"
  kernel = "lognormal",  # choose a positive-support kernel
  GPD = TRUE,
  components = 4,
  mcmc = list(niter = 2000, nburnin = 500, thin = 5, nchains = 1, seed = 101)
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = TRUE)
```

Predictions

```r
grid <- seq(min(y), max(y) * 1.5, length.out = 200)

dhat <- predict(fit, y = grid, type = "density")
shat <- predict(fit, y = grid, type = "survival")
qs   <- predict(fit, type = "quantile", p = c(0.5, 0.9, 0.95, 0.99), cred.level = 0.90)
```

Results (precomputed)

The table and figure below are precomputed from the shipped dataset to keep CRAN checks fast and deterministic. To reproduce them locally, run `tools/make_vignette_artifacts.R` and rebuild the vignettes.

```{r}
tbl <- utils::read.csv(.extdata("unconditional_quantiles.csv"))
knitr::kable(tbl, digits = 3)
```

```{r, results='asis'}
cat(
  sprintf(
    "![Posterior predictive density for the unconditional fit (precomputed).](%s)\n",
    knitr::image_uri(.extdata("unconditional_density.png"))
  )
)
```

Notes on customization

1. Kernel choice: use `kernel_support_table()` to see which kernels match your support (real line vs positive support) and which parameters are available.

2. Tail control: setting `GPD=TRUE` activates the spliced tail module (see the `basic` vignette for the exact spliced density). Tail priors and threshold control are provided via `param_specs`.

3. Backend choice: stick-breaking (`backend="sb"`) is often the most stable default; the CRP backend is available when you prefer partition-based updates.
