---
<<<<<<< HEAD
title: "Unconditional models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unconditional models}
=======
title: "Unconditional Modeling"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Unconditional Modeling}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
<<<<<<< HEAD
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5
)
library(nimble)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 8000, nburnin = 2000, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Goal

Fit an unconditional DP mixture (optionally with a GPD tail) and examine fitted values and basic diagnostics.

## Fit

```{r}
library(DPmixGPD)

n <- 100
y <- abs(rnorm(n)) + 0.2

bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE)
fit
```

## Fitted summaries

```{r}
# For unconditional models, fitted values replicate a population-level location
f_mean <- fitted(fit, type = "mean", level = 0.90)
head(f_mean)

f_med <- fitted(fit, type = "median", level = 0.90)
head(f_med)
```

## Posterior predictive summaries

```{r}
pred_mean <- predict(fit, type = "mean", cred.level = 0.90, interval = "credible")
pred_q95  <- predict(fit, type = "quantile", index = 0.95, cred.level = 0.90, interval = "credible")

pred_mean$fit
pred_q95$fit
```

## Residual check

```{r}
res <- f_mean$residuals
summary(res)
```

## Plot

```{r}
try(plot(fit, family = "trace"), silent = TRUE)
```

## Notes

- For heavier tail behavior, use `GPD = TRUE` and increase MCMC iterations.
- For bulk-only comparisons, set `GPD = FALSE`.

```{r sessioninfo, include=FALSE}
sessionInfo()
```
=======
  comment = NA,
  fig.width = 7,
  fig.height = 4.5,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
```

# Goal

Unconditional modeling fits a flexible mixture to the **bulk** distribution, with an optional **GPD tail** for extremes. This is the base case for DPmixGPD.

# Example data

```{r data}
data("nc_pos200_k3")
y <- nc_pos200_k3$y
```

# Bulk‑only mixture

```{r bulk-only}
bundle_bulk <- build_nimble_bundle(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = FALSE,
  J = 4,
  mcmc = list(niter = 200, nburnin = 50, nchains = 1)
)
fit_bulk <- run_mcmc_bundle_manual(bundle_bulk, show_progress = FALSE)
```

# Bulk + GPD tail

```{r bulk-gpd}
bundle_tail <- build_nimble_bundle(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = TRUE,
  J = 4,
  mcmc = list(niter = 200, nburnin = 50, nchains = 1)
)
fit_tail <- run_mcmc_bundle_manual(bundle_tail, show_progress = FALSE)
```

# Distributional interpretation

- The **bulk** mixture captures multi‑modality or skewness.
- The **GPD tail** isolates extreme exceedances and stabilizes tail risk estimates.

# One “money” plot (bulk vs tail)

The plot below compares a **bulk‑only** kernel to a **bulk+GPD** splice using fixed parameters
(no MCMC required for the illustration).

```{r money-plot, eval=TRUE}
grid <- seq(0, 8, length.out = 200)
bulk <- dGammaMix(grid, w = c(0.7, 0.3), shape = c(2, 6), scale = c(1, 0.4))
tail <- vapply(
  grid,
  function(x) dGammaMixGpd(
    x,
    w = c(0.7, 0.3),
    shape = c(2, 6),
    scale = c(1, 0.4),
    threshold = 4,
    tail_scale = 1.0,
    tail_shape = 0.2
  ),
  numeric(1)
)
plot(grid, bulk, type = "l", lwd = 2, col = "steelblue", ylab = "density", xlab = "y")
lines(grid, tail, lwd = 2, col = "firebrick")
legend("topright", legend = c("Bulk only", "Bulk + GPD"), col = c("steelblue", "firebrick"), lwd = 2, bty = "n")
```

# Why the GPD tail exists

Mixture kernels are excellent for the **center** of the distribution but can be unstable in the far tail. The GPD tail gives a principled extreme‑value model while keeping the bulk flexible.

# Next steps

- For covariates, see [conditional](conditional.Rmd).
- For careful inference, see [mcmc-workflow](mcmc-workflow.Rmd).
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
