---
title: "0. Start Here"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{0. Start Here}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  eval = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
set.seed(1)

quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == "windows") "NUL" else "/dev/null"
  out <- NULL
  utils::capture.output(
    out <- eval(substitute(expr), envir = parent.frame()),
    file = nullfile
  )
  out
}
```

# Start Here

This vignette gives a minimal, fully working workflow for an **unconditional** model and a **conditional** model. Everything uses short MCMC runs so the vignette renders quickly.

---

## Unconditional Model (CRP, bulk-only)

```{r start-uncond-data}
data("nc_pos200_k3")
y <- nc_pos200_k3$y
```

```{r start-uncond-bundle}
bundle_uncond <- build_nimble_bundle(
  y = y,
  backend = "crp",
  kernel = "gamma",
  GPD = FALSE,
  components = 5,
  mcmc = list(niter = 200, nburnin = 50, thin = 1, nchains = 1, seed = 1)
)
```

```{r start-uncond-fit}
fit_uncond <- quiet_mcmc(run_mcmc_bundle_manual(bundle_uncond, show_progress = FALSE))
summary(fit_uncond)
```

```{r start-uncond-predict}
pred_q <- predict(fit_uncond, type = "quantile", index = c(0.5, 0.9), interval = "credible")
head(pred_q$fit)
plot(pred_q)
```

---

## Conditional Model (SB, bulk-only)

```{r start-cond-data}
data("nc_posX100_p3_k2")
yc <- nc_posX100_p3_k2$y
X <- as.matrix(nc_posX100_p3_k2$X)
```

```{r start-cond-bundle}
bundle_cond <- build_nimble_bundle(
  y = yc,
  X = X,
  backend = "sb",
  kernel = "lognormal",
  GPD = FALSE,
  components = 5,
  mcmc = list(niter = 250, nburnin = 50, thin = 1, nchains = 1, seed = 2)
)
```

```{r start-cond-fit}
fit_cond <- quiet_mcmc(run_mcmc_bundle_manual(bundle_cond, show_progress = FALSE))
summary(fit_cond)
```

```{r start-cond-predict}
x_new <- X[1:20, , drop = FALSE]
pred_mean <- predict(fit_cond, x = x_new, type = "mean", interval = "credible", nsim_mean = 200)
head(pred_mean$fit)
plot(pred_mean)
```

---

## Useful S3 Methods

```{r start-s3}
params(fit_uncond)
plot(fit_uncond, family = c("traceplot", "running"))
```

---

## Next Steps

- `vignette 1`: package overview and terminology
- `vignette 5`: full three-phase workflow (spec ? bundle ? MCMC)
- `vignette 6-13`: unconditional/conditional models with and without GPD
- `vignette 14-19`: causal workflows
