---
title: "Start Here: What DPmixGPD Does"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Start Here: What DPmixGPD Does}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "cache/v00-start-here-",
  fig.width = 6,
  warning = FALSE,
  message = FALSE
)
library(DPmixGPD)
library(nimble)
use_cached_fit <- TRUE
.fit_path <- function(name) {
  path <- system.file("extdata", name, package = "DPmixGPD")
  if (path == "") path <- file.path("inst", "extdata", name)
  path
}
fit_small <- readRDS(.fit_path("fit_small.rds"))
library(ggplot2)
```

What you'll learn: how DPmixGPD splits bulk and tail modelling, what outputs you get, and where to go next.

## What problem are we solving?

Standard single-distribution fits throw up their hands when data are sparse in the tail: the bulk dominates, heavy excursions are ignored, and probabilistic forecasts understate rare events. DPmixGPD targets precisely that regime by pairing a flexible Dirichlet process mixture in the bulk with a dedicated generalized Pareto tail beyond a threshold so you can keep inference honest on extremes.

## What DPmixGPD fits

- **Bulk model:** DP mixture over kernels (`normal`, `gamma`, `lognormal`, `Laplace`, `inverse Gaussian`, `Amoroso`) delivers adaptively shaped density/conditional models.
- **Tail model:** a Generalized Pareto Distribution is glued on at a learned threshold, with smooth transitions and optional covariate links.
- **Prediction:** the same fitted object can forecast densities, survival curves, quantiles, and posterior predictive draws.

## Two workflows

### Density / regression

You build a bundle, run MCMC once, and derive density or quantile estimates via `predict()` for new covariates.

### Causal

You fit the propensity score plus two outcome arms (treated `trt` and control `con`), then compare arm-specific predictions to produce CQTE curves.

## What you get out

- Posterior predictive distribution (density/survival).
- Quantiles at arbitrary probabilities.
- Tail index (shape/scale) and threshold diagnostics.
- Standard trace/ESS summaries, block-specific plots, and credible bands on predictions.

## What to read next

- [v02-single-outcome-modeling.Rmd](v02-single-outcome-modeling.html) for a canonical density recipe.
- [v05-causal-cqte.Rmd](v05-causal-cqte.html) for the CQTE workflow.
- [v08-prediction-and-exports.Rmd](v08-prediction-and-exports.html) for exporting predictions.

## Bulk + tail diagram

```{r diagram, fig.cap="DPmixGPD concept: flexible bulk + GPD tail stitched at a threshold."}
bulk <- data.frame(x = seq(0, 10, length.out = 200))
bulk$y <- dlnorm(bulk$x, meanlog = 0.5, sdlog = 0.4) * 0.8
tail <- data.frame(x = seq(8, 20, length.out = 3))
tail$y <- vapply(
  tail$x,
  DPmixGPD::dGpd,
  numeric(1),
  threshold = 10,
  scale = 2,
  shape = 0.3
)
ggplot() +
  geom_area(data = bulk, aes(x = x, y = y), fill = "#66c2a5", alpha = 0.7) +
  geom_line(data = tail, aes(x = x, y = y), color = "#fc8d62", size = 1.2) +
  geom_vline(xintercept = 10, linetype = "dashed") +
  annotate("text", x = 12, y = 0.15, label = "GPD tail", color = "#fc8d62") +
  annotate("text", x = 4, y = 0.25, label = "DP mixture bulk", color = "#66c2a5") +
  labs(x = "Outcome", y = "Density")
```

## 10-line example

```{r quick-demo}
set.seed(7)
y <- sim_bulk_tail(90, tail_prob = 0.2)
J <- 6
if (use_cached_fit) {
  fit <- fit_small
} else {
  bundle <- build_nimble_bundle(
    y = y,
    backend = "sb",
    kernel = "lognormal",
    GPD = TRUE,
    J = J,
    mcmc = list(niter = 200, nburnin = 50, thin = 2, nchains = 2, seed = c(1, 2))
  )
  fit <- run_mcmc_bundle_manual(bundle)
}
print(fit)
summary(fit, component = "tail")
```

## Prediction and diagnostics

```{r predictions, fig.cap="Density prediction on held-out grid (posterior mean)."}
pred <- predict(fit, type = "survival", y = seq(0, max(y) + 5, length.out = 100))
surv_df <- data.frame(y = pred$grid, survival = as.numeric(pred$fit[1, ]))
ggplot(surv_df, aes(x = y, y = survival)) +
  geom_line(color = "#7b3294") +
  labs(title = "Posterior survival curve", y = "Survival", x = "y")
```
```{r quantiles}
predict(fit, type = "quantile", p = c(.5, .8, .95))
```

## Session info

```{r session}
sessionInfo()
```
