---
title: "Installation, Reproducibility, and Package Structure"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Installation, Reproducibility, and Package Structure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "cache/01-installation-",
  fig.width = 6,
  warning = FALSE,
  message = FALSE
)
library(DPmixGPD)
library(nimble)
use_cached_fit <- TRUE
.fit_path <- function(name) {
  path <- system.file("extdata", name, package = "DPmixGPD")
  if (path == "") path <- file.path("inst", "extdata", name)
  path
}
fit_small <- readRDS(.fit_path("fit_small.rds"))
library(ggplot2)
```

What you'll learn: how to install DPmixGPD consistently, reproduce MCMC draws via seeds and thinning, and locate the bundle/prediction/engine pieces inside the package.

## Installation options

- **CRAN** (preferred for most users): run `install.packages("DPmixGPD")` after the package appears.
- **GitHub** (latest features):

```r
if (!requireNamespace("remotes", quietly = TRUE)) remotes::install.packages("remotes")
remotes::install_github("fsu-dpmixgpd/DPmixGPD")  # placeholder repo URL
```

- **Nimble note:** DPmixGPD relies on Nimble's code generation. `nimble::nimbleOptions` can control compilation flags, so set reproducible values before running `run_mcmc_bundle_manual()`.

## Reproducibility checklist

- Pass `seed` values for `build_nimble_bundle()` and `run_mcmc_bundle_manual()` chains (`mcmc$niter`, `nburnin`, `nchains`, `thin` as deterministic lists).
- Fix `plan()` for futures if using `future::plan()` in parallel chains; the package uses `future.apply` under the hood.
- Save the returned object (`mixgpd_fit`) and rerun summaries/predict using the same seed list to avoid jitter.

## Package architecture map

| Area | Key functions | Purpose |
| --- | --- | --- |
| Bundle builder | `build_nimble_bundle()`, `build_causal_bundle()` | Creates Nimble nodes and mixture layout (SB or CRP) with J components, a kernel, and optional GPD tail. |
| Prediction + diagnostics | `predict.mixgpd_fit()`, `cqte()`, `summary.mixgpd_fit()` | Compute quantiles, plotting data, and tail comparisons using the same mixture weights as estimation. |
| Kernel registry | `get_kernel_registry()`, `init_kernel_registry()` | Lists the supported kernels (e.g., `gamma`, `lognormal`, `normal`) and their parameter shapes. |
| Backend engine | `run_mcmc_bundle_manual()`, `run_mcmc_causal()` | Executes Nimble MCMC with SB or CRP latent structures plus any causal stacking. |
| Utilities | `sim_*()` helpers, `get_tail_registry()` | Simulated datasets for tutorials and tail-specific metadata. |

## Common pitfalls

| Symptom | Likely cause | Fix |
| --- | --- | --- |
| "Model not fully initialized" | Some latent nodes were left `NULL` | Supply explicit `components`/`J`, `GPD`, and kernel priors in the bundle call before running MCMC. |
| Dimension mismatch in newdata | Covariate names differ or missing | Always pass a `data.frame` whose column names match the ones used in `build_nimble_bundle()` (case sensitive). |
| Positivity warning in causal | Overlap poor for treated vs control | Check propensity scores and consider trimming rare regions before drawing CQTE curves. |

## Quick reproducible example

```{r reproducible-example}
y <- sim_bulk_tail(n = 120, seed = 42)
if (use_cached_fit) {
  fit <- fit_small
} else {
  bundle <- build_nimble_bundle(
    y = y,
    backend = "sb",
    kernel = "gamma",
    GPD = TRUE,
    J = 5,
    mcmc = list(niter = 200, nburnin = 50, thin = 2, nchains = 2, seed = c(1, 2))
  )
  fit <- run_mcmc_bundle_manual(bundle)
}
print(fit)
```

The `print()` output reiterates the backend, kernel, J components, and the `GPD` flag (TRUE/FALSE). The bundle bundler never leaves the latent parameter initialization to chance.

## Diagnostic plot

```{r diag-density}
data.frame(y = y) |>
  ggplot(aes(x = y)) +
  geom_density(fill = "steelblue", alpha = 0.4) +
  labs(title = "Input data density", x = "y", y = "Density")
```

## Prediction snippet

```{r prediction-example}
preds <- predict(fit, type = "quantile", p = c(0.5, 0.9, 0.95))
print(preds)
```

## Troubleshooting checklist

1. Confirm the target data fit the kernel domain (e.g., positive for `gamma`, `lognormal`).
2. Inspect `fit$threshold` and `fit$tail_shape` for GPD instabilities.
3. Re-run with more burn-in and thinning when chains mix poorly.

## Session info

```{r session-info}
sessionInfo()
```
