---
title: "1. Introduction to DPmixGPD"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{1. Introduction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 7,
  fig.height = 5,
  eval = FALSE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
```

# DPmixGPD: Bayesian Nonparametric Mixture Models with GPD Tails

## Overview

**DPmixGPD** is an R package for Bayesian nonparametric Dirichlet Process mixture models with flexible kernels and optional Generalized Pareto Distribution (GPD) tail augmentation.

### Key Features

- **Flexible mixture kernels**: normal, gamma, lognormal, laplace, invgauss, amoroso, cauchy
- **Two MCMC backends**: Stick-Breaking (SB) and Chinese Restaurant Process (CRP)
- **Optional GPD tail**: Enhance tail probability estimation
- **Conditional modeling**: Add covariates via parameter linking
- **Causal workflows**: Propensity score adjustment with multiple PS strategies
- **Comprehensive prediction**: density, survival, quantile, sample, mean

---

## Model Structure

### Single-Outcome Model

**Bulk**: Dirichlet Process mixture
$$Y_i \sim \sum_{k=1}^{\infty} w_k f_k(y_i; \theta_k)$$

**Tail (optional)**: GPD beyond threshold
$$P(Y > u | Y > u) \sim \text{GPD}(\text{scale}, \text{shape})$$

### Causal Model

Two outcome arms with optional propensity score (PS) adjustment:
- **Control** ($T=0$): $Y_i^{(0)} \sim \text{DPM}_{\text{con}}(X_i, PS_i)$
- **Treated** ($T=1$): $Y_i^{(1)} \sim \text{DPM}_{\text{trt}}(X_i, PS_i)$

---

## MCMC Backends

### Stick-Breaking (SB)
- Fixed number of mixture components $J$
- Exact truncation at $J$ components
- Faster per-iteration computation
- Use when computational cost is primary concern

### Chinese Restaurant Process (CRP)
- Truncation at $K_{\max}$ for sampling
- Adaptively discovers number of clusters
- More flexible cluster allocation
- Default for exploratory analysis

---

## Propensity Score (PS) Models

**Three PS strategies** (causal workflows only):

1. **`PS="logit"`** (default): Bayesian logistic regression
2. **`PS="probit"`**: Bayesian probit regression
3. **`PS="naive"`**: Gaussian naive Bayes classifier
4. **`PS=FALSE`**: No PS; outcomes depend on $X$ only

When PS is estimated, scores are injected into outcome model covariates for confounding adjustment.

---

## Workflow Overview

### Three-Phase Model Building

**Phase 1: Specification**
```r
spec <- compile_model_spec(y, kernel="gamma", backend="crp", GPD=TRUE)
```

**Phase 2: Bundle**
```r
bundle <- build_nimble_bundle(spec)
```

**Phase 3: MCMC**
```r
fit <- run_mcmc_bundle_manual(bundle, niter=1000, nburnin=100)
```

---

## Quick Start

```{r intro-quickstart}
# Load packaged dataset
data("nc_pos200_k3")
y <- nc_pos200_k3$y

# Build unconditional model: bulk-only, SB backend
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "gamma",
  GPD = FALSE,
  components = 6,
  mcmc = list(niter = 200, nburnin = 50, nchains = 1)
)

# Print bundle
print(bundle)
```

---

## Outline of This Vignette Series

1. **Introduction** (this vignette)
2. **Available Distributions**: dqrp functions, 10k samples, visualizations
3. **Basic Model Workflow**: compile → bundle → MCMC
4. **Unconditional DPmix (CRP)**: bulk-only, no covariates
5. **Unconditional DPmix (SB)**: bulk-only, no covariates
6. **Unconditional DPmixGPD (CRP)**: bulk+tail, no covariates
7. **Unconditional DPmixGPD (SB)**: bulk+tail, no covariates
8. **Conditional DPmix (CRP)**: bulk-only, with covariates
9. **Conditional DPmix (SB)**: bulk-only, with covariates
10. **Conditional DPmixGPD (CRP)**: bulk+tail, with covariates
11. **Conditional DPmixGPD (SB)**: bulk+tail, with covariates
12. **Causal (Same Backend CRP)**: PS + treatment effects, CRP
13. **Causal (Same Backend SB)**: PS + treatment effects, SB
14. **Causal (Different Backends CRP)**: Per-arm backends, CRP
15. **Causal (Different Backends SB)**: Per-arm backends, SB

---

## Dependencies

```{r intro-deps}
packageVersion("DPmixGPD")
packageVersion("nimble")
packageVersion("ggplot2")
```

Each vignette demonstrates complete workflows with reproducible examples.
