---
title: "3. Basic Workflow: Model Specification, Bundle, & MCMC"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{3. Basic Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
set.seed(42)
```

# The Three-Phase Workflow

DPmixGPD uses a **three-phase model-building paradigm** designed for clarity and control:

1. **Specification** (`compile_model_spec`): Validate inputs, NO NIMBLE code generation
2. **Bundle** (`build_nimble_bundle`): Generate NIMBLE code, compile sampler
3. **MCMC** (`run_mcmc_bundle_manual`): Execute posterior sampling

---

## Phase 1: Specification (Input Validation)

**Purpose**: Validate data, kernels, backends, and parameters WITHOUT generating NIMBLE code.

```{r spec-basic}
# Simulate data
set.seed(123)
y <- rgamma(100, shape = 2, rate = 1)

# Create specification
spec <- compile_model_spec(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = FALSE
)

# Inspect specification
print("Specification object class:", class(spec), "\n")
print("Kernel chosen:", spec$kernel, "\n")
print("Backend chosen:", spec$backend, "\n")
print("GPD enabled:", spec$GPD, "\n")
print("Sample size n:", length(spec$y), "\n")
```

### Specification for Different Model Variants

```{r spec-variants}
# 1. Unconditional DPmix (bulk only, no tail)
spec_unconditional <- compile_model_spec(y, kernel = "gamma", backend = "crp", GPD = FALSE)

# 2. Unconditional DPmixGPD (bulk + tail)
spec_gpd <- compile_model_spec(y, kernel = "gamma", backend = "crp", GPD = TRUE)

# 3. Conditional (with covariates)
X <- matrix(rnorm(100 * 2), nrow = 100, ncol = 2)
spec_conditional <- compile_model_spec(
  y = y,
  X = X,
  kernel = "lognormal",
  backend = "crp",
  GPD = FALSE
)

# 4. With custom MCMC settings
spec_custom <- compile_model_spec(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = FALSE,
  mcmc = list(niter = 2000, nburnin = 500, nchains = 2, thin = 2)
)

print("Created 4 specification objects successfully.\n")
```

---

## Phase 2: Bundle (NIMBLE Code Generation & Compilation)

**Purpose**: Generate NIMBLE model code, compile sampler, prepare for MCMC execution.

### Building from Specification

```{r bundle-from-spec}
# Method 1: From specification object
bundle <- build_nimble_bundle(spec)

print("Bundle object class:", class(bundle), "\n")
print("Bundle contains compiled sampler ready for MCMC.\n")
```

### Building Directly (Alternative)

```{r bundle-direct}
# Method 2: Direct call (recommended for quick workflows)
bundle_direct <- build_nimble_bundle(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = FALSE,
  mcmc = list(niter = 500, nburnin = 100, nchains = 1)
)

print("Direct bundle creation successful.\n")
```

### Inspecting Bundle Contents

```{r bundle-inspect}
# Bundle is an S3 object with structure
print("Bundle class:", class(bundle_direct), "\n")
print("Bundle contains:\n")
print(names(bundle_direct))

# Access key components
print("\nMCMC settings:\n")
print(bundle_direct$mcmc_settings)
```

---

## Phase 3: MCMC Execution

**Purpose**: Run posterior sampling from the compiled bundle.

### Basic MCMC Run

```{r mcmc-basic}
# Run MCMC
fit <- run_mcmc_bundle_manual(bundle_direct)

print("Fit object class:", class(fit), "\n")
print("MCMC execution complete. Posterior samples collected.\n")
```

### Accessing Posterior Samples

```{r mcmc-posterior}
# Posterior summary
print("\n--- POSTERIOR SUMMARY ---\n")
summary(fit)
```

### Diagnostic Plots

```{r mcmc-diagnostics}
# Trace plots
plot(fit, params = "alpha|beta")
```

---

## Complete Workflow: End-to-End Example

```{r workflow-complete}
# Generate realistic data
set.seed(789)
y_data <- rgamma(150, shape = 1.8, rate = 0.9)

# PHASE 1: Specification
spec_final <- compile_model_spec(
  y = y_data,
  kernel = "gamma",
  backend = "crp",
  GPD = FALSE,
  Kmax = 5
)

# PHASE 2: Bundle
bundle_final <- build_nimble_bundle(
  spec_final,
  mcmc = list(
    niter = 1000,
    nburnin = 200,
    nchains = 2,
    thin = 1
  )
)

# PHASE 3: MCMC
fit_final <- run_mcmc_bundle_manual(bundle_final)

print("\n=== THREE-PHASE WORKFLOW COMPLETE ===\n")
summary(fit_final)
```

---

## Backend Comparison: CRP vs Stick-Breaking

### CRP Backend

```{r backend-crp}
# Chinese Restaurant Process
bundle_crp <- build_nimble_bundle(
  y = y_data,
  kernel = "gamma",
  backend = "crp",
  Kmax = 5,
  mcmc = list(niter = 500, nburnin = 100, nchains = 1)
)

fit_crp <- run_mcmc_bundle_manual(bundle_crp)
print("CRP execution complete.\n")
```

### Stick-Breaking Backend

```{r backend-sb}
# Stick-Breaking Process
bundle_sb <- build_nimble_bundle(
  y = y_data,
  kernel = "gamma",
  backend = "sb",
  J = 5,
  mcmc = list(niter = 500, nburnin = 100, nchains = 1)
)

fit_sb <- run_mcmc_bundle_manual(bundle_sb)
print("SB execution complete.\n")
```

---

## Kernel Selection Guide

```{r kernel-guide}
kernels_available <- c("gamma", "lognormal", "normal", "laplace", "invgauss", "amoroso", "cauchy")

print("Available kernels:\n")
for (k in kernels_available) {
  print("  -", k, "\n")
}

print("\nChoose kernel based on:\n")
print("  gamma:     Right-skewed, positive support\n")
print("  lognormal: Log-transformed normality\n")
print("  normal:    Symmetric, unbounded\n")
print("  laplace:   Sharp peak, exponential tails\n")
print("  invgauss:  Positive, near-normal shape\n")
print("  amoroso:   Generalized, maximum flexibility\n")
print("  cauchy:    Heavy-tailed, rare cases\n")
```

---

## GPD Tail Augmentation

### Unconditional with GPD

```{r gpd-example}
# Data with tail behavior
y_tail <- c(rgamma(120, shape = 2, rate = 1), rgamma(30, shape = 0.8, rate = 0.1))

# Build with GPD
bundle_gpd <- build_nimble_bundle(
  y = y_tail,
  kernel = "gamma",
  backend = "crp",
  GPD = TRUE,
  mcmc = list(niter = 500, nburnin = 100, nchains = 1)
)

fit_gpd <- run_mcmc_bundle_manual(bundle_gpd)
print("\nGPD augmentation applied to tail region.\n")
summary(fit_gpd)
```

---

## Summary of Key Functions

| Phase | Function | Input | Output |
|-------|----------|-------|--------|
| **1. Spec** | `compile_model_spec()` | y, X (optional), kernel, backend, GPD | `dpmixgpd_spec` |
| **2. Bundle** | `build_nimble_bundle()` | spec or y, kernel, backend, GPD | `dpmixgpd_bundle` |
| **3. MCMC** | `run_mcmc_bundle_manual()` | bundle | `mixgpd_fit` |

---

## Common Parameter Settings

```{r param-reference}
print("=== Recommended MCMC Parameters ===\n")
print("Quick test:     niter=500,  nburnin=100, nchains=1\n")
print("Standard:       niter=2000, nburnin=500, nchains=2\n")
print("Production:     niter=5000, nburnin=1000, nchains=3\n")
print("\n=== Backend Parameters ===\n")
print("CRP: Kmax=3-5 (truncation for # components)\n")
print("SB:  J=3-5    (fixed # components)\n")
print("\n=== Kernel Selection ===\n")
print("Positive data:     gamma, lognormal, invgauss\n")
print("Any real data:     normal\n")
print("Symmetric tails:   laplace\n")
print("Extreme outliers:  cauchy\n")
```

---

## Next Steps

- Move to **vignette 4-5** for unconditional models (CRP vs SB backends)
- Move to **vignette 6-7** for tail modeling with GPD
- Move to **vignette 8-11** for conditional models with covariates
- Move to **vignette 12-15** for causal inference workflows
