---
title: "Kernels: What They Mean and When to Use Which"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kernels: What They Mean and When to Use Which}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "cache/04-kernels-",
  fig.width = 6,
  warning = FALSE,
  message = FALSE
)
library(DPmixGPD)
library(nimble)
use_cached_fit <- TRUE
.fit_path <- function(name) {
  path <- system.file("extdata", name, package = "DPmixGPD")
  if (path == "") path <- file.path("inst", "extdata", name)
  path
}
fit_small <- readRDS(.fit_path("fit_small.rds"))
library(ggplot2)
library(dplyr)
```

What you'll learn: how each kernel shapes the DP mixture (via the index `j`), what domains they cover, and how to pick the right kernel for a given dataset with a decision-style guide.

## Kernel registry and naming

DPmixGPD exposes a registry where each kernel family provides density, quantile, and RNG functions under consistent names (`dGammaMix`, `qGammaMix`, `rGammaMix`, etc.). The `j`-th component always shares the same parameter layout across kernels (location, scale, shape), and the components count is controlled by `J` only.

| Kernel | Support | Shape hints | Recommended use |
| --- | --- | --- | --- |
| `normal` | `(-Inf, Inf)` | location & scale | Symmetric data with light tails; fast mixing. |
| `lognormal` | `(0, Inf)` | log-scale mean & sd | Positive skew; natural for positive incomes or durations. |
| `gamma` | `(0, Inf)` | shape & scale | Heavy-tailed positive data, interpretable tail shape. |
| `Laplace` | `(-Inf, Inf)` | double exponential | Sharp center with heavier tails than normal. |
| `inverse Gaussian` | `(0, Inf)` | mean & shape | Right-skewed durations when variance grows cubically. |
| `Amoroso` | `(0, Inf)` | location, scale, shape | Flexible extreme skew with explicit tail shape control. |
| `Cauchy` | `(-Inf, Inf)` | location & scale | Very heavy tails; anchors tail behavior when GPD is off.

## Decision guide

1. Are your data strictly positive? If yes, stay in `(0, Inf)` kernels (`lognormal`, `gamma`, `amoroso`, `inverse Gaussian`).
2. Do you need a flexible tail shape within the kernel itself (before the GPD kick-in)? Choose `Amoroso` or `Cauchy` for wide tail control.
3. Need interpretable mean/variance for a symmetric central mass? Use `normal` or `Laplace` and rely on GPD for extremes.
4. When the dataset shows multiple modes, increase `J` and keep the kernel simple to avoid identifiability issues.

## Minimal kernel example

```{r kernel-example}
y <- sim_bulk_tail(n = 160, seed = 77)
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "amoroso",
  GPD = TRUE,
  J = 6,
  mcmc = list(niter = 200, nburnin = 50, thin = 2, nchains = 2, seed = c(1, 2))
)
if (use_cached_fit) {
  fit <- fit_small
} else {
  fit <- run_mcmc_bundle_manual(bundle)
}
```

## Diagnostic plot

```{r kernel-plot}
data.frame(y = y) |>
  ggplot(aes(x = y)) +
  geom_histogram(boundary = 0, bins = 30, fill = "coral", alpha = 0.6) +
  labs(title = "Data histogram showing bulk + tail", x = "y", y = "Count")
```

## Prediction example

```{r kernel-pred}
predicted <- predict(fit, type = "quantile", p = c(0.5, 0.9, 0.99))
predicted
```

## Kernel recommendations summary

- Use `lognormal`/`gamma` for positive, moderately heavy tails where standard priors suffice.
- Use `normal`/`Laplace` when symmetry is expected and tails do not dominate.
- Explore `Amoroso` when you want implicit tail shape control inside the bulk, remembering that a nontrivial threshold still keeps the GPD tail smooth.

## Session info

```{r session-info}
sessionInfo()
```
