---
title: "4. Unconditional DPmix with CRP Backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Unconditional DPmix CRP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
library(ggplot2)
library(gridExtra)
set.seed(123)
```

# Unconditional DPmix: Chinese Restaurant Process (CRP)

**Goal**: Estimate density of univariate outcome $y$ using **nonparametric Dirichlet Process mixture** with **Chinese Restaurant Process** backend.

**Model**: $y_i | G \sim \int K(y_i; \theta) dG(\theta)$ where $G \sim \text{DP}(\alpha, G_0)$

**Backend**: CRP with **truncation at $K_{\max}$ components**

---

## Data Setup

```{r data-setup}
# Generate from mixture of gamma distributions
set.seed(42)
component1 <- rgamma(60, shape = 2, rate = 1.5)
component2 <- rgamma(40, shape = 1, rate = 0.5)
y_mixed <- c(component1, component2)

cat("Sample size:", length(y_mixed), "\n")
cat("Mean:", mean(y_mixed), "\n")
cat("SD:", sd(y_mixed), "\n")
cat("Range:", range(y_mixed), "\n")

# Visualization
df_data <- data.frame(y = y_mixed)
p_raw <- ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, alpha = 0.6, fill = "steelblue") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Raw Data: Mixed Gamma Distribution", x = "y", y = "Density") +
  theme_minimal()

print(p_raw)
```

---

## Model Specification

```{r spec-crp}
spec_crp <- compile_model_spec(
  y = y_mixed,
  kernel = "gamma",           # Use gamma kernel
  backend = "crp",            # CRP backend
  GPD = FALSE,                # No tail augmentation
  Kmax = 15,                  # Max components truncation
  alpha = 1.0,                # DP concentration
  verbose = FALSE
)

cat("Specification created:\n")
cat("  Kernel:", spec_crp$kernel, "\n")
cat("  Backend:", spec_crp$backend, "\n")
cat("  Kmax:", spec_crp$Kmax, "\n")
cat("  GPD:", spec_crp$GPD, "\n")
```

---

## Bundle & MCMC Settings

```{r bundle-mcmc-crp}
bundle_crp <- build_nimble_bundle(
  spec_crp,
  mcmc = list(
    niter = 2000,           # Total iterations
    nburnin = 500,          # Burn-in
    nchains = 2,            # Parallel chains
    thin = 1                # No thinning
  )
)

cat("Bundle compiled with CRP sampler.\n")
cat("Ready for MCMC execution.\n")
```

---

## MCMC Execution

```{r mcmc-crp}
fit_crp <- run_mcmc_bundle_manual(bundle_crp)

cat("\n=== POSTERIOR SUMMARY (CRP Backend) ===\n")
summary(fit_crp)
```

---

## Posterior Diagnostics

### Trace Plots

```{r diag-trace}
# Trace plots for key parameters
plot(fit_crp, params = "alpha|beta")
```

### Effective Sample Size & Convergence

```{r diag-ess}
# ESS and R-hat from summary
cat("Check ESS > 1000 and Rhat < 1.1 for convergence.\n")
```

---

## Posterior Predictions

### Predictive Density

```{r pred-density}
# Generate prediction grid
y_grid <- seq(0, max(y_mixed) * 1.2, length.out = 200)

# Posterior predictive density
pred_densities <- predict(fit_crp, newdata = y_grid)

df_pred <- data.frame(
  y = y_grid,
  density = pred_densities
)

# Combine with data
p_pred <- ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, alpha = 0.4, fill = "steelblue") +
  geom_line(data = df_pred, aes(x = y, y = density), color = "red", linewidth = 1.2) +
  labs(title = "Data vs Posterior Predictive Density (CRP)", 
       x = "y", y = "Density") +
  theme_minimal()

print(p_pred)
```

### Quantile Predictions

```{r pred-quantiles}
# Posterior predictive quantiles
quantiles_pred <- predict(fit_crp, newdata = y_grid, type = "quantile", 
                          probs = c(0.05, 0.25, 0.5, 0.75, 0.95))

cat("Posterior Predictive Quantiles:\n")
print(head(quantiles_pred, 10))
```

---

## Mixture Component Analysis

### Number of Active Components

```{r comp-analysis}
# Extract number of components used
# (CRP backend infers K from cluster assignments)
cat("The CRP allows flexibility in # components.\n")
cat("Check MCMC samples for K (number of occupied tables).\n")
```

### Component Parameters

```{r comp-params}
# Estimated component parameters
# E.g., for gamma kernel: shape and rate per component
cat("Posterior mean component parameters available via coef().\n")
coef_vals <- coef(fit_crp)
print(head(coef_vals))
```

---

## Model Comparison: Different Kernels

### Gamma Kernel (Current)

```{r comp-gamma}
bundle_gamma <- build_nimble_bundle(
  y = y_mixed,
  kernel = "gamma",
  backend = "crp",
  Kmax = 15,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)
fit_gamma <- run_mcmc_bundle_manual(bundle_gamma)
summary(fit_gamma)
```

### Lognormal Kernel (Alternative)

```{r comp-lognormal}
bundle_lognormal <- build_nimble_bundle(
  y = y_mixed,
  kernel = "lognormal",
  backend = "crp",
  Kmax = 15,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)
fit_lognormal <- run_mcmc_bundle_manual(bundle_lognormal)
summary(fit_lognormal)
```

### Model Comparison via Predictions

```{r comp-summary}
# Compare fitted values
fitted_gamma <- fitted(fit_gamma)
fitted_lognormal <- fitted(fit_lognormal)

cat("\n=== KERNEL COMPARISON (CRP Backend) ===\n")
cat("Gamma fitted values (first 5):    ", head(fitted_gamma, 5), "\n")
cat("Lognormal fitted values (first 5):", head(fitted_lognormal, 5), "\n")
cat("\nUse summary() to compare convergence diagnostics.\n")
```

---

## Varying Truncation Level (Kmax)

```{r kmax-sensitivity}
kmax_values <- c(5, 10, 15, 20)

cat("Testing Kmax values:", paste(kmax_values, collapse = ", "), "\n")
cat("Recommendation: Start with Kmax=10-15 for most applications.\n")
cat("Monitor effective number of components via summary().\n")

# Demonstrate with one value
bundle_kmax <- build_nimble_bundle(
  y = y_mixed,
  kernel = "gamma",
  backend = "crp",
  Kmax = 10,
  mcmc = list(niter = 500, nburnin = 100, nchains = 1)
)
fit_kmax <- run_mcmc_bundle_manual(bundle_kmax)
summary(fit_kmax)
```

---

## Residual Analysis

```{r residuals}
# Extract residuals
residuals_vals <- residuals(fit_crp)

df_resid <- data.frame(residuals = residuals_vals)

p_resid <- ggplot(df_resid, aes(x = residuals)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, alpha = 0.6, fill = "coral") +
  geom_density(color = "darkred", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray") +
  labs(title = "Residual Distribution", x = "Residuals", y = "Density") +
  theme_minimal()

print(p_resid)

cat("Residual Summary:\n")
print(summary(residuals_vals))
```

---

## Key Takeaways

- **CRP Backend**: Flexible component allocation, ideal for unknown mixture complexity
- **Kmax Parameter**: Higher Kmax allows more components but increases computation
- **Kernel Choice**: Gamma suitable for positive, skewed data
- **Diagnostics**: Check convergence (Rhat, ESS) and posterior predictive fit
- **Next**: Compare with **Stick-Breaking (SB)** backend in vignette 5

