---
title: "4. Backends, Kernels, and Workflow Map"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{4. Backends, Kernels, and Workflow Map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  eval = FALSE,
  cache = TRUE,
  cache.path = "../.cache/vignettes/"
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
```

# Big picture

DPmixGPD has two orthogonal “dials” you turn when building models:

- **Backend** (how the mixture weights / clustering are represented)
  - **CRP**: Chinese Restaurant Process representation.
  - **SB**: stick-breaking truncation with a fixed maximum number of components.

- **Kernel / family** (what distribution models the bulk of the data)
  - Examples: normal, lognormal, gamma, inverse-Gaussian, Laplace, Cauchy, Amoroso, etc.

Optionally, you can also turn on:

- **GPD = TRUE/FALSE** to splice a Generalized Pareto tail beyond a threshold.

# What changes between CRP and SB?

Both backends target the same posterior over densities. The difference is *representation*:

- **CRP** learns a random number of occupied clusters (conceptually infinite), and you track allocations.
- **SB** uses a finite truncation level `Kmax` and learns stick-breaking weights.

Practical rule of thumb:

- **CRP** is convenient when you want adaptive complexity without choosing `Kmax`.
- **SB** is convenient when you want predictable memory/time and easy vectorization.

# What does the workflow look like?

DPmixGPD uses a consistent build → run → summarize loop:

1. **Build a bundle** using `build_nimble_bundle()` (or causal builders if you’re doing TE work).
2. **Run MCMC** using `run_mcmc_bundle_manual()`.
3. **Inspect and summarize** using `print()`, `summary()`, `plot()`.
4. **Predict** using `predict()` (and optionally `fitted()`).

```{r}
# Build
bundle <- build_nimble_bundle(
  y = rnorm(50),
  backend = "crp",
  kernel = "normal",
  GPD = FALSE,
  mcmc = list(niter = 200, nburnin = 100, thin = 1, nchains = 1, seed = 1)
)

# Run
fit <- run_mcmc_bundle_manual(bundle)

# Summarize
print(fit)
summary(fit)
plot(fit)
```

# Kernel support quick check

Use `kernel_support()` and the kernel registry helpers to confirm what is available.

```{r}
kernel_support()
```

# Where to go next

- **Available distributions**: see **v05** for the `d/p/q/r` functions and examples.
- **Basic build/compile/run**: see **v03**.
- **Unconditional / Conditional / Causal**: continue through v06+.
