---
title: "5. Unconditional DPmix with Stick-Breaking Backend"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{5. Unconditional DPmix with Stick-Breaking Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  cache = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(dplyr)
library(patchwork)
library(purrr)
set.seed(123)
```

# Unconditional DPmix: Stick-Breaking (SB) Backend

**Purpose**: Showcase the stick-breaking backend that uses a fixed number of mixture components (`components = J`) and contrasts it with the CRP backend from `v04`. The bulk kernel is again a lognormal so that the comparison is apples-to-apples.

---

## Data Setup

```{r data-setup}
# Load benchmark dataset
data("nc_pos200_k3")
y_mixed <- nc_pos200_k3$y

df_data <- data.frame(y = y_mixed)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y_mixed), mean(y_mixed), sd(y_mixed), min(y_mixed), max(y_mixed))
)

p_raw <- ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "darkorange", alpha = 0.7, color = "black") +
  geom_density(color = "darkred", linewidth = 1.2) +
  labs(title = "Raw Data: Mixed Gamma Distribution", x = "y", y = "Density") +
  theme_minimal()

grid.arrange(p_raw, ncol = 1)
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Summary of the SB Dataset") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Model Specification & Bundle

We build the stick-breaking mixture explicitly with `components = 5` so that there is room for weight decay while keeping the MCMC runtime manageable for a vignette.

```{r bundle-mcmc-sb}
bundle_sb <- build_nimble_bundle(
  y = y_mixed,
  kernel = "lognormal",
  backend = "sb",
  components = 5,            # fixed truncation level for SB
  GPD = FALSE,               # bulk-only scenario
  mcmc = list(
    niter = 300,
    nburnin = 60,
    nchains = 1,
    thin = 1
  )
)
```

---

## Running MCMC & Summary

```{r mcmc-run}
fit_sb <- run_mcmc_bundle_manual(bundle_sb)
```

```{r fit-summary}
summary(fit_sb)
```

---

## MCMC Diagnostics (S3 Plot Methods)

```{r diag-plots}
# Default diagnostics for the mixgpd fit
plots_fit <- plot(fit_sb, family = c("traceplot", "density"))
print(plots_fit)

 
``` 

Use `summary(fit_sb)` to inspect effective sample size, R-hat, and other convergence diagnostics; the `plot()` calls above show the trace/density pairs for both global and stick-breaking weight parameters.

---

## Stick-Breaking Weights & Component Activity

The stick-breaking weights are exposed via the `plot()` diagnostics above, and you can refer to the raw `fit_sb` object or its `summary()` output for exact posterior summaries of each component weight if your version exposes them.

---

## Posterior Predictions

```{r density-predict}
y_grid <- seq(0, max(y_mixed) * 1.3, length.out = 250)
pred_density <- predict(fit_sb, y = y_grid, type = "density")
plot(pred_density)
```

```{r quantile-predict}
quant_probs <- c(0.05, 0.25, 0.5, 0.75, 0.95)
pred_q <- predict(fit_sb, type = "quantile", index = quant_probs, interval = "credible")
plot(pred_q)
```

```{r mean-predict}
pred_mean <- predict(fit_sb, type = "mean")
plot(pred_mean)
```

---

## Bulk-only vs CRP Backends

```{r compare-backend}
bundle_crp_small <- build_nimble_bundle(
  y = y_mixed,
  kernel = "lognormal",
  backend = "crp",
  components = 5,
  GPD = FALSE,
  mcmc = list(niter = 300, nburnin = 60, nchains = 1)
)
fit_crp_small <- run_mcmc_bundle_manual(bundle_crp_small)

crp_quant <- predict(fit_crp_small, type = "quantile", index = quant_probs)
sb_quant <- predict(fit_sb, type = "quantile", index = quant_probs)

bind_rows(
  crp_quant$fit %>% mutate(model = "CRP"),
  sb_quant$fit %>% mutate(model = "SB")
) %>%
  select(model, index, estimate, lwr, upr) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(caption = "Quantile Comparison: CRP vs SB") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r compare-plots}
plot(crp_quant)
plot(sb_quant)
```

---

## Residuals & Fitted Values

```{r residuals}
fit_vals <- fitted(fit_sb)
plot(fit_vals)
```

---

## Takeaways

- **SB Backend**: Fixed `components` keeps label-switching manageable, and the diagnostic plots via the S3 `plot()` method show weight dynamics.
- **Predictions**: Posterior density, quantiles, and mean are all accessible via `predict()` and display credible bands for each summary.
- **Backend comparison**: The CRP fit (`v04`) and SB fit deliver similar central quantiles while SB offers more control over truncation.
- **Next**: Explore tail augmentation (`GPD = TRUE`) with the CRP backend in `v06` or SB backend in `v07`.

