---
title: "Causal Inference with DPmixGPD: CQTE End-to-End"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Inference with DPmixGPD: CQTE End-to-End}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "cache/05-causal-cqte-",
  warning = FALSE,
  message = FALSE,
  fig.width = 6
)
library(DPmixGPD)
library(nimble)
use_cached_fit <- TRUE
.fit_path <- function(name) {
  path <- system.file("extdata", name, package = "DPmixGPD")
  if (path == "") path <- file.path("inst", "extdata", name)
  path
}
fit_causal_con <- readRDS(.fit_path("fit_causal_con.rds"))
fit_causal_trt <- readRDS(.fit_path("fit_causal_trt.rds"))
fit_causal_meta <- readRDS(.fit_path("fit_causal_meta.rds"))
fit_causal_small <- list(
  ps_fit = NULL,
  outcome_fit = list(con = fit_causal_con, trt = fit_causal_trt),
  bundle = fit_causal_meta,
  call = NULL
)
class(fit_causal_small) <- "dpmixgpd_causal_fit"
library(ggplot2)
```

What you'll learn: how to encode causal assumptions, build a causal bundle, diagnose overlap, and compute conditional quantile treatment effects (CQTEs) that spotlight tail differences.

## 5.1 Setup and assumptions

- We model treated (`T=1`) and control (`T=0`) potential outcomes separately and adjust for covariates via a propensity score.
- Overlap (positivity) is essential: each arm needs support across the covariate space.
- Causality still rests on no unmeasured confounding; the DP mixture / GPD tail machinery simply learns flexible outcome distributions.

## 5.2 Data format

Requires:

- `y` outcome.
- `T` binary treatment indicator.
- `X` covariate matrix/data frame (columns `x1`, `x2`, `x3` in our examples).

## 5.3 Workflow overview

1. Propensity score stage (`T | X`).
2. Outcome model for `Y | X, T=0` and `Y | X, T=1`, each a DP mixture + optional GPD.
3. Predictions per arm at new covariate grids.
4. CQTE = quantile(trt) - quantile(con) for each probability and covariate slice.

## 5.4 Minimal causal example

```{r causal-example}
set.seed(101)
dat <- sim_causal_cqte(220)
J <- 6
bundle <- build_causal_bundle(
  y = dat$y,
  X = dat$X,
  T = dat$t,
  backend = c("sb", "sb"),
  kernel = c("normal", "normal"),
  GPD = c(FALSE, FALSE),
  J = c(J, J),
  mcmc_outcome = list(niter = 200, nburnin = 50, thin = 1, nchains = 2, seed = c(1, 2)),
  mcmc_ps = list(niter = 200, nburnin = 50, thin = 1, nchains = 2, seed = c(3, 4))
)
if (use_cached_fit) {
  cf <- fit_causal_small
} else {
  cf <- run_mcmc_causal(bundle, show_progress = FALSE)
}
print(cf)
summary(cf$outcome_fit$trt)
```

For vignette speed the cached fit uses a normal kernel with `GPD = FALSE`. In practice, switch
`kernel` and `GPD` to match your outcome support and tail goals.

## 5.5 Diagnostics

### Overlap / positivity

```{r overlap-plot, fig.cap="Propensity score distributions by arm (logistic GLM proxy)."}
ps_df <- data.frame(dat$X, T = dat$t)
ps_fit <- glm(T ~ x1 + x2 + x3, data = ps_df, family = binomial())
ps_df$propensity <- predict(ps_fit, type = "response")
ggplot(ps_df, aes(x = propensity, fill = factor(T, labels = c("con", "trt")))) +
  geom_density(alpha = 0.4) +
  labs(title = "Propensity score overlap", fill = "Arm")
```

### Trace plots (arm-specific mixture weights)

```{r tail-trace, fig.cap="Trace plots for tail_scale and tail_shape (treated arm)."}
plot(cf$outcome_fit$trt, family = "traceplot", params = c("alpha", "w[1]"))
```

## 5.6 CQTE estimation and plotting

```{r cqte-slices}
grid <- expand.grid(
  x1 = seq(-1.2, 1.2, length.out = 15),
  x2 = seq(-0.8, 0.8, length.out = 5),
  x3 = 0
)
taus <- c(.1, .5, .9, .95, .99)
pr_trt <- predict(cf$outcome_fit$trt, newdata = grid, type = "quantile", p = taus)
pr_con <- predict(cf$outcome_fit$con, newdata = grid, type = "quantile", p = taus)
cqte_df <- data.frame(
  x1 = rep(grid$x1, each = length(taus)),
  tau = rep(taus, times = nrow(grid)),
  cqte = c(pr_trt$fit - pr_con$fit)
)
slice_df <- cqte_df[cqte_df$tau %in% c(.5, .95, .99) & cqte_df$x1 %in% c(-1, 0, 1), ]
ggplot(slice_df, aes(x = x1, y = cqte, color = factor(tau))) +
  geom_line() +
  labs(title = "CQTE slices (selected taus)", color = "tau")
```

```{r cqte-surface, fig.cap="CQTE surface for tau=0.95 varying x1 / x2 at x3=0."}
surf_grid <- expand.grid(x1 = seq(-1, 1, length.out = 25), x2 = seq(-1, 1, length.out = 25), x3 = 0)
tau_surface <- 0.95
surf_trt <- predict(cf$outcome_fit$trt, newdata = surf_grid, type = "quantile", p = tau_surface)
surf_con <- predict(cf$outcome_fit$con, newdata = surf_grid, type = "quantile", p = tau_surface)
surf_grid$cqte <- as.numeric(surf_trt$fit - surf_con$fit)
ggplot(surf_grid, aes(x = x1, y = x2, fill = cqte)) +
  geom_raster() +
  scale_fill_viridis_c(option = "C") +
  labs(title = "CQTE surface (tau = 0.95)", fill = "CQTE")
```

```{r tail-focused}
tail_cqte <- predict(cf$outcome_fit$trt, newdata = data.frame(x1 = 0, x2 = 0, x3 = 0), type = "quantile", p = c(.95, .99, .995))
control_tail <- predict(cf$outcome_fit$con, newdata = data.frame(x1 = 0, x2 = 0, x3 = 0), type = "quantile", p = c(.95, .99, .995))
data.frame(
  tau = c(.95, .99, .995),
  cqte = tail_cqte$fit - control_tail$fit
)
```

## 5.7 Interpretation guidance

- Differences near the median (`tau ~ 0.5`) reflect central CFTE; tail differences (`tau >= 0.95`) highlight extreme treatment effects.
- Credible intervals widen where data are sparse???this is expected and flags caution, not failure.
- Try alternative kernels or toggle `GPD` per arm (e.g., `GPD = c(TRUE, FALSE)`) when shape estimates blow up.

## Summary table of CQTE

```{r cqte-table}
summary_points <- data.frame(
  x1 = c(-1, 0, 1),
  x2 = c(0, 0, 0),
  label = c("x1 = -1, x2 = 0", "x1 = 0, x2 = 0", "x1 = 1, x2 = 0"),
  cqte_median = NA_real_,
  cqte_tail = NA_real_
)
for (i in seq_len(nrow(summary_points))) {
  newdata <- data.frame(x1 = summary_points$x1[i], x2 = summary_points$x2[i], x3 = 0)
  med <- predict(cf$outcome_fit$trt, newdata = newdata, type = "quantile", p = 0.5)$fit
  med_con <- predict(cf$outcome_fit$con, newdata = newdata, type = "quantile", p = 0.5)$fit
  tail <- predict(cf$outcome_fit$trt, newdata = newdata, type = "quantile", p = 0.95)$fit
  tail_con <- predict(cf$outcome_fit$con, newdata = newdata, type = "quantile", p = 0.95)$fit
  summary_points$cqte_median[i] <- med - med_con
  summary_points$cqte_tail[i] <- tail - tail_con
}
knitr::kable(summary_points[, c("label", "cqte_median", "cqte_tail")], digits = 3)
```

## Session info

```{r session}
sessionInfo()
```
