---
title: "5. Unconditional DPmix with Stick-Breaking Backend"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{5. Unconditional DPmix SB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  cache = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
set.seed(123)
```

# Unconditional DPmix: Stick-Breaking (SB) Backend

**Goal**: Estimate density of univariate outcome $y$ using **nonparametric Dirichlet Process mixture** with **Stick-Breaking** backend.

**Model**: $y_i | G \sim \sum_{j=1}^{J} w_j K(y_i; \theta_j)$ where weights $w_j$ follow stick-breaking process

**Backend**: SB with **fixed $J$ components**

---

## Data Setup (Same as CRP)

```{r data-setup}
# Generate from mixture of gamma distributions
set.seed(42)
component1 <- rgamma(60, shape = 2, rate = 1.5)
component2 <- rgamma(40, shape = 1, rate = 0.5)
y_mixed <- c(component1, component2)

paste("Sample size:", length(y_mixed))
paste("Mean:", mean(y_mixed))
paste("SD:", sd(y_mixed))

# Visualization
df_data <- data.frame(y = y_mixed)
p_raw <- ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, alpha = 0.6, fill = "darkorange") +
  geom_density(color = "darkred", linewidth = 1) +
  labs(title = "Raw Data: Mixed Gamma Distribution", x = "y", y = "Density") +
  theme_minimal()

print(p_raw)
```

---

## Model Specification (SB Backend)

```{r spec-sb}
spec_sb <- compile_model_spec(
  y = y_mixed,
  kernel = "gamma",           # Use gamma kernel
  backend = "sb",             # Stick-Breaking backend
  GPD = FALSE,                # No tail augmentation
  components = 5              # FIXED number of components
)

print("Specification created:")
print(paste("  Kernel:", spec_sb$meta$kernel))
print(paste("  Backend:", spec_sb$meta$backend))
print(paste("  Components:", spec_sb$meta$components))
print(paste("  GPD:", spec_sb$meta$GPD))
```

### Key Difference: J vs Kmax

```{r spec-comparison}
print("=== CRP vs SB Parameters ===")
print("CRP (Vignette 4):")
print("  - Kmax: TRUNCATION level (max possible components)")
print("  - # components inferred from data (≤ Kmax)")
print("  - Flexible, data-adaptive")
print("")
print("SB (This Vignette):")
print("  - J: FIXED number of components")
print("  - Always uses exactly J components")
print("  - Simpler, faster (no label-switching)")
```

---

## Bundle & MCMC Settings

```{r bundle-mcmc-sb}
bundle_sb <- build_nimble_bundle(
  y = y_mixed,
  kernel = "gamma",
  backend = "sb",
  components = 5,
  mcmc = list(
    niter = 2000,           # Total iterations
    nburnin = 500,          # Burn-in
    nchains = 2,            # Parallel chains
    thin = 1                # No thinning
  )
)

message("Bundle compiled with SB sampler.")
message(paste("Fixed 5 components, no truncation variability."))
```

---

## MCMC Execution

```{r mcmc-sb}
fit_sb <- run_mcmc_bundle_manual(bundle_sb)

print("\n=== POSTERIOR SUMMARY (SB Backend) ===\n")
summary(fit_sb)
```

---

## Posterior Diagnostics

### Trace Plots

```{r diag-trace}
# Trace plots for key parameters
plot(fit_sb, params = "w_|alpha|beta")
```

### Effective Sample Size & Convergence

```{r diag-ess}
# ESS and R-hat from summary
print("Check ESS and Rhat for convergence.\n")
```

---

## Component Weights (Stick-Breaking)

### Weight Distribution

```{r weights-analysis}
# Extract stick-breaking weights for each component
# Weights w_j sum to 1, typically decreasing

print("Stick-breaking weight analysis:\n")
print("SB generates weights that sum to 1.\n")
print("Typically w_1 >= w_2 >= ... >= w_J\n")
print("Weights available from posterior samples.\n")
```

### Effective Number of Components

```{r eff-components}
# Count components with posterior mean weight > threshold
weight_threshold <- 0.05

print("Components with weight >", weight_threshold, "are 'active'.\n")
print("Remaining components absorb tail probability.\n")
```

---

## Posterior Predictions

### Predictive Density

```{r pred-density}
# Generate prediction grid
y_grid <- seq(0, max(y_mixed) * 1.2, length.out = 200)

# Posterior predictive density
pred_density <- predict(fit_sb, y = y_grid, type = "density")

# Use S3 plot method
plot(pred_density)
```

---

## CRP vs SB Backend Comparison

### Same Data, Both Backends

```{r comparison-both}
# CRP fit (from previous vignette)
bundle_crp <- build_nimble_bundle(
  y = y_mixed,
  kernel = "gamma",
  backend = "crp",
  Kmax = 5,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)
fit_crp <- run_mcmc_bundle_manual(bundle_crp)

# SB fit (current)
bundle_sb_quick <- build_nimble_bundle(
  y = y_mixed,
  kernel = "gamma",
  backend = "sb",
  J = 5,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)
fit_sb_quick <- run_mcmc_bundle_manual(bundle_sb_quick)

# Compare convergence diagnostics
print("\n=== BACKEND COMPARISON ===\n")
print("Use summary() to compare ESS and Rhat values.\n")
print("Both backends should produce similar fitted values.\n")
```

### Computational Efficiency

```{r comparison-speed}
print("\nComputational Considerations:\n")
print("CRP:\n")
print("  - Adaptive # components (Kmax truncation)\n")
print("  - Handles label-switching via NIMBLE\n")
print("  - Slower per iteration\n")
print("  - Better for unknown complexity\n\n")

print("SB:\n")
print("  - Fixed J components\n")
print("  - No label-switching overhead\n")
print("  - Faster per iteration\n")
print("  - Better when J known or for quick inference\n")
```

---

## Varying J (Number of Components)

```{r j-sensitivity}
j_values <- c(5, 10, 15, 20)

print("Testing J values:", paste(j_values, collapse = ", "), "\n")
print("Recommendation: Start with J=3-5 for mixture modeling.\n")
print("Adjust J based on data complexity and convergence diagnostics.\n")

# Demonstrate with one value
bundle_j <- build_nimble_bundle(
  y = y_mixed,
  kernel = "gamma",
  backend = "sb",
  J = 5,
  mcmc = list(niter = 500, nburnin = 100, nchains = 1)
)
fit_j <- run_mcmc_bundle_manual(bundle_j)
summary(fit_j)
```

---

## Residual Analysis

```{r residuals}
# Extract fitted values with diagnostics
fit_vals <- fitted(fit_sb_quick)

# Use S3 plot method for diagnostic plots
plot(fit_vals)

print("Residual Summary:\n")
print(summary(residuals(fit_sb_quick)))
```

---

## Mixture Component Details

### Component Parameters

```{r comp-details}
# Extract component parameters (shape, rate for gamma)
coef_sb <- coef(fit_sb_quick)
print("Component parameters (first 5 components):\n")
print(head(coef_sb, 5))
```

### Posterior Mean Components

```{r comp-visualization}
# Get component densities
print("Individual component contributions available.\n")
print("SB provides clean, fixed-J structure.\n")
```

---

## Key Takeaways

- **SB Backend**: Fixed components, simpler inference, faster computation
- **J Parameter**: Choose based on expected mixture complexity
- **Trade-offs**: SB trades flexibility (CRP) for computational efficiency
- **Model Comparison**: Use summary() to compare ESS, Rhat, and convergence
- **Next**: Add **GPD tail augmentation** in vignette 6

---

## When to Use SB vs CRP

```{r when-use}
print("=== DECISION GUIDE ===\n\n")

print("USE CRP (Vignette 4) IF:\n")
print("  ✓ Unknown mixture complexity\n")
print("  ✓ Want data-adaptive # components\n")
print("  ✓ Willing to pay computational cost\n")
print("  ✓ Small to medium sample size\n\n")

print("USE SB (This Vignette) IF:\n")
print("  ✓ Known or assumed # components\n")
print("  ✓ Need faster MCMC per iteration\n")
print("  ✓ Large dataset (SB scales better)\n")
print("  ✓ Prefer simpler, more stable inference\n")
```

