---
title: "Causal Extras: ATE, CATE, and Comparators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Causal Extras: ATE, CATE, and Comparators}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "cache/06-causal-",
  fig.width = 6,
  warning = FALSE,
  message = FALSE
)
library(DPmixGPD)
library(nimble)
use_cached_fit <- TRUE
.fit_path <- function(name) {
  path <- system.file("extdata", name, package = "DPmixGPD")
  if (path == "") path <- file.path("inst", "extdata", name)
  path
}
fit_causal_con <- readRDS(.fit_path("fit_causal_con.rds"))
fit_causal_trt <- readRDS(.fit_path("fit_causal_trt.rds"))
fit_causal_meta <- readRDS(.fit_path("fit_causal_meta.rds"))
fit_causal_small <- list(
  ps_fit = NULL,
  outcome_fit = list(con = fit_causal_con, trt = fit_causal_trt),
  bundle = fit_causal_meta,
  call = NULL
)
class(fit_causal_small) <- "dpmixgpd_causal_fit"
library(ggplot2)
library(dplyr)
```

What you'll learn: how CQTE fits into the broader causal toolkit, how to compare it with ATE/CATE estimands, and how to add simple baselines such as IPW quantiles.

## Estimand relationships

- CQTE subtracts conditional quantiles at levels `tau` for treated (`trt`) vs control (`con`) arms at the same covariate point.
- ATE averages the mean outcomes; CATE averages conditional means. CQTE instead tracks tail behavior that matter for extremes.
- If CQTE is flat across `tau`, the treatment effect does not depend on extremeness; rising curves point to tail dominance.

## Minimal causal-comparison example

```{r causal-extras-example}
set.seed(10)
data <- sim_causal_cqte(n = 200)
X <- data$X
bundle <- build_causal_bundle(
  y = data$y,
  X = X,
  T = data$t,
  backend = "sb",
  kernel = "normal",
  GPD = FALSE,
  J = 5,
  mcmc_outcome = list(niter = 200, nburnin = 50, thin = 3, nchains = 2, seed = c(1, 2)),
  mcmc_ps = list(niter = 200, nburnin = 50, thin = 3, nchains = 2, seed = c(3, 4))
)
if (use_cached_fit) {
  fit <- fit_causal_small
} else {
  fit <- run_mcmc_causal(bundle)
}
cq <- cqte(fit, probs = c(0.5, 0.9, 0.99), newdata = head(X, 3))
cq
```

The cached fit keeps `GPD = FALSE` for speed. Turn it on and adjust the kernel when you need tail-focused causal contrasts.

## Diagnostic plot: treatment densities

```{r causal-plot}
df_plot <- data.frame(y = data$y, arm = ifelse(data$t == 1, "trt", "con"))
ggplot(df_plot, aes(x = y, fill = arm)) +
  geom_density(alpha = 0.4) +
  labs(title = "Outcome densities by arm", x = "Outcome y", fill = "Arm")
```

## Comparison table and alternative baseline

| Estimand | Description |
| --- | --- |
| CQTE | Difference between treatment and control quantiles (`trt` minus `con`) at tail probability `tau`. |
| CATE | Mean difference conditional on covariates (special case of CQTE at `tau=0.5` when tails align). |
| ATE | Average of CATE across the empirical covariate distribution.

```{r ipw-baseline}
arm_tbl <- data.frame(y = data$y, treatment = ifelse(data$t == 1, "trt", "con"))
ipw_q <- arm_tbl |>
  group_by(treatment) |>
  summarize(median = quantile(y, 0.5), `90%` = quantile(y, 0.9))
ipw_q
```

## Prediction handoff

```{r causal-pred}
pred_grid <- data.frame(x1 = seq(-1, 1, length.out = 5), x2 = rep(0, 5), x3 = 0)
cq_pred <- cqte(fit, probs = c(0.9, 0.99), newdata = pred_grid)
cq_pred
```

## Session info

```{r session-info}
sessionInfo()
```
