---
title: "7. Unconditional DPmixGPD with Stick-Breaking Backend"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{7. Unconditional DPmixGPD SB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  cache = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(dplyr)
library(tibble)
set.seed(123)
```

# Unconditional DPmixGPD: Stick-Breaking (SB) Backend with Tail Augmentation

**Purpose**: Demonstrate the stick-breaking backend (`components = J`) while augmenting the extreme tail with a GPD. This mirrors the CRP+GPD pipeline (v06) but highlights the fixed-truncation behavior for bulk components.

---

## Data Setup

```{r data-setup}
# Tail-heavy data
data("nc_pos_tail200_k4")
y_tail <- nc_pos_tail200_k4$y

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y_tail), mean(y_tail), sd(y_tail), min(y_tail), max(y_tail))
)

df_data <- data.frame(y = y_tail)

p_raw <- ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "darkorange", alpha = 0.6, color = "black") +
  geom_density(color = "darkred", linewidth = 1) +
  labs(title = "Tail-Designed Data", x = "y", y = "Density") +
  theme_minimal()

grid.arrange(p_raw, ncol = 1)
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Tail Dataset Summaries") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Threshold Selection

```{r threshold}
thresholds <- quantile(y_tail, c(0.70, 0.75, 0.80, 0.85))
u_threshold <- thresholds["80%"]

ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "skyblue", alpha = 0.6) +
  geom_vline(xintercept = u_threshold, linetype = "dashed", color = "black") +
  labs(title = paste("Threshold at", round(u_threshold, 2)), x = "y", y = "Density") +
  theme_minimal()
```

---

## Model Specification & Bundle

```{r spec-bundle}
bundle_sb_gpd <- build_nimble_bundle(
  y = y_tail,
  kernel = "lognormal",
  backend = "sb",
  GPD = TRUE,
  components = 5,
  param_specs = list(
    gpd = list(
      threshold = list(mode = "fixed", value = u_threshold)
    )
  ),
  mcmc = list(
    niter = 300,
    nburnin = 60,
    nchains = 1,
    thin = 1
  )
)
```

---

## Running MCMC

```{r mcmc-run}
fit_sb_gpd <- run_mcmc_bundle_manual(bundle_sb_gpd)
summary(fit_sb_gpd)
```

---

## Posterior Predictions

```{r density-predict}
y_grid <- seq(0, max(y_tail) * 1.3, length.out = 300)
pred_density <- predict(fit_sb_gpd, y = y_grid, type = "density")
plot(pred_density)
```

```{r survival-predict}
y_surv <- seq(u_threshold, max(y_tail) * 1.1, length.out = 60)
pred_surv <- predict(fit_sb_gpd, y = y_surv, type = "survival")
plot(pred_surv)
```

```{r quantile-predict}
quant_probs <- c(0.90, 0.95, 0.99)
pred_quant <- predict(fit_sb_gpd, type = "quantile", index = quant_probs, interval = "credible")
plot(pred_quant)
```

---

## Tail vs Bulk Comparison

```{r bulk-comparison}
bundle_sb_bulk <- build_nimble_bundle(
  y = y_tail,
  kernel = "lognormal",
  backend = "sb",
  GPD = FALSE,
  components = 5,
  mcmc = list(niter = 300, nburnin = 60, nchains = 1)
)
fit_sb_bulk <- run_mcmc_bundle_manual(bundle_sb_bulk)

bulk_quant <- predict(fit_sb_bulk, type = "quantile", index = quant_probs)
t_quant <- predict(fit_sb_gpd, type = "quantile", index = quant_probs)

bind_rows(
  bulk_quant$fit %>% mutate(model = "Bulk-only"),
  t_quant$fit %>% mutate(model = "Bulk + GPD")
) %>%
  select(model, index, estimate, lwr, upr) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(caption = "Quantiles: Bulk-only vs GPD-augmented") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r comparison-plots}
plot(bulk_quant)
plot(t_quant)
```

---

## Residuals & Diagnostics

```{r residuals}
fit_vals <- fitted(fit_sb_gpd)
plot(fit_vals)
```

```{r diagnostics}
plot(fit_sb_gpd, family = c("traceplot", "density"))
plot(fit_sb_gpd, params = "w_", family = c("traceplot", "density"))
```

---

## Key Takeaways

- Stick-breaking truncation fixes the number of bulk components but still flexibly models the tail via GPD.
- Posterior `predict()` + `plot()` workflows visualize densities, survival probabilities, and extreme quantiles.
- Comparing bulk-only vs GPD-augmented quantiles reveals how tail augmentation shifts the 95–99% levels.
- Next: conditional DPmix (v08–v11) to explore covariate effects before moving to causal regimes.

