---
title: "7. Unconditional DPmixGPD with Stick-Breaking Backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{7. Unconditional DPmixGPD SB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
library(ggplot2)
set.seed(123)
```

# Unconditional DPmixGPD: Stick-Breaking (SB) Backend with Tail

**Goal**: Apply **Stick-Breaking backend** with **GPD tail augmentation** for bulk + tail modeling.

**Model**: 
- **Bulk** ($y \leq u$): Fixed-J stick-breaking mixture
- **Tail** ($y > u$): Generalized Pareto Distribution

---

## Data Setup

```{r data-setup}
set.seed(42)
bulk_comp1 <- rgamma(80, shape = 2, rate = 1.5)
bulk_comp2 <- rgamma(60, shape = 1, rate = 0.8)
bulk <- c(bulk_comp1, bulk_comp2)

tail_component <- DPmixGPD::rGpd(30, threshold = 0, scale = 1.5, shape = 0.3)
y_tail <- c(bulk, tail_component)

u_threshold <- quantile(y_tail, 0.8)

cat("Sample size:", length(y_tail), "\n")
cat("Bulk:", sum(y_tail <= u_threshold), "| Tail:", sum(y_tail > u_threshold), "\n")

df_raw <- data.frame(y = y_tail)
p_raw <- ggplot(df_raw, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, alpha = 0.6, fill = "darkorange") +
  geom_density(color = "darkred", linewidth = 1) +
  geom_vline(xintercept = u_threshold, linetype = "dashed", color = "gray") +
  labs(title = "Data with Tail", x = "y", y = "Density") +
  theme_minimal()

print(p_raw)
```

---

## Model Specification (SB + GPD)

```{r spec-sb-gpd}
spec_sb_gpd <- compile_model_spec(
  y = y_tail,
  kernel = "gamma",
  backend = "sb",           # Stick-Breaking backend
  GPD = TRUE,               # Enable tail
  threshold = u_threshold,
  J = 15,                   # Fixed components
  verbose = FALSE
)

cat("SB + GPD Specification:\n")
cat("  Backend (bulk): sb (J =", spec_sb_gpd$J, "components)\n")
cat("  Tail: GPD (threshold =", u_threshold, ")\n")
```

---

## Bundle & MCMC

```{r bundle-mcmc}
bundle_sb_gpd <- build_nimble_bundle(
  spec_sb_gpd,
  mcmc = list(
    niter = 2000,
    nburnin = 500,
    nchains = 2,
    thin = 1
  )
)

fit_sb_gpd <- run_mcmc_bundle_manual(bundle_sb_gpd)

cat("\n=== POSTERIOR SUMMARY (SB + GPD) ===\n")
summary(fit_sb_gpd)
```

---

## Posterior Predictions

### Full Distribution

```{r pred-full}
y_grid <- seq(0, max(y_tail) * 1.3, length.out = 300)
pred_dens <- predict(fit_sb_gpd, newdata = y_grid)

df_pred <- data.frame(y = y_grid, density = pred_dens)

p_pred <- ggplot(df_raw, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, alpha = 0.4, fill = "darkorange") +
  geom_line(data = df_pred, aes(y = density), color = "darkred", linewidth = 1.2) +
  geom_vline(xintercept = u_threshold, linetype = "dashed") +
  labs(title = "SB + GPD Posterior Predictive", x = "y", y = "Density") +
  theme_minimal()

print(p_pred)
```

---

## CRP vs SB for GPD Tail

```{r comparison-crp-sb}
# CRP + GPD (from vignette 6)
bundle_crp_gpd <- build_nimble_bundle(
  y = y_tail,
  kernel = "gamma",
  backend = "crp",
  GPD = TRUE,
  threshold = u_threshold,
  Kmax = 15,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)
fit_crp_gpd <- run_mcmc_bundle_manual(bundle_crp_gpd)

cat("\n=== CRP vs SB (with GPD Tail) ===\n")
```

---

## Residual Analysis

```{r residuals}
residuals_vals <- residuals(fit_sb_gpd)

p_resid <- ggplot(data.frame(r = residuals_vals), aes(x = r)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, alpha = 0.6, fill = "mediumpurple") +
  geom_density(color = "purple", linewidth = 1) +
  labs(title = "Residuals: SB + GPD", x = "Residuals", y = "Density") +
  theme_minimal()

print(p_resid)
```

---

## Key Takeaways

- **SB + GPD**: Fixed-component bulk with flexible tail
- **Speed**: Generally faster than CRP + GPD
- **Trade-off**: CRP more flexible; SB more efficient
- **Next**: Conditional models (vignettes 8-11)
