---
title: "Survival Outcomes: Censoring + Heavy Tails"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Survival Outcomes: Censoring + Heavy Tails}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "cache/07-survival-",
  fig.width = 6,
  warning = FALSE,
  message = FALSE
)
library(DPmixGPD)
library(nimble)
use_cached_fit <- TRUE
.fit_path <- function(name) {
  path <- system.file("extdata", name, package = "DPmixGPD")
  if (path == "") path <- file.path("inst", "extdata", name)
  path
}
fit_small <- readRDS(.fit_path("fit_small.rds"))
library(ggplot2)
```

What you'll learn: how to adapt DPmixGPD to survival-style data, what is currently available (bulk + GPD for tail times), and how to visualize survival curves even under right censoring.

## Survival data format

Use a data frame with columns `time`, `status` (1 for event, 0 for censor), and covariates. DPmixGPD currently treats all outcomes as positive durations and uses the conventional density + tail augmentation, so censoring is handled via manual likelihood reweighting outside the package (future work may integrate censored nodes inside Nimble).

## Minimal survival example

```{r survival-example}
data <- sim_survival_tail(n = 220, seed = 220)
y <- data$time
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "gamma",
  GPD = TRUE,
  J = 5,
  mcmc = list(niter = 200, nburnin = 50, thin = 2, nchains = 2, seed = c(1, 2))
)
if (use_cached_fit) {
  fit <- fit_small
} else {
  fit <- run_mcmc_bundle_manual(bundle)
}
```

## Diagnostic: survival density

```{r survival-density}
data.frame(time = y, status = factor(data$status, labels = c("censored", "event"))) |>
  ggplot(aes(x = time, fill = status)) +
  geom_density(alpha = 0.4) +
  labs(title = "Observed survival times (censored vs event)", x = "Time", fill = "Status")
```

## Prediction: survival curves via tail extrapolation

```{r survival-pred}
tau <- seq(0.01, 0.99, length.out = 40)
quantiles <- predict(fit, type = "quantile", p = tau)
surv_curve <- data.frame(time = as.numeric(quantiles$fit), survival = 1 - tau)
surv_curve
```

We treat the quantiles as inverse survival: the higher the quantile, the lower the survival probability, which mirrors how extreme durations dominate upper tails.
## Roadmap and assumptions

- **Overlap assumption:** the model must see enough events (status = 1) to learn tail parameters; too few will make the GPD estimate unstable.
- **Censoring note:** handle censoring outside DPmixGPD by imputing truncated observations or reweighting the data before bundling.
- **Diagnostic tip:** compare `predict(..., type = "quantile", p = c(.9, .99))` with Kaplan-Meier estimates to monitor tail stability.

## Session info

```{r session-info}
sessionInfo()
```
