---
title: "Troubleshooting and Common Errors"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Troubleshooting and Common Errors}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "cache/09-troubleshooting-",
  warning = FALSE,
  message = FALSE,
  fig.width = 6
)
library(DPmixGPD)
library(nimble)
use_cached_fit <- TRUE
.fit_path <- function(name) {
  path <- system.file("extdata", name, package = "DPmixGPD")
  if (path == "") path <- file.path("inst", "extdata", name)
  path
}
fit_small <- readRDS(.fit_path("fit_small.rds"))
library(ggplot2)
```

What you'll learn: how to interpret common warnings/errors, when to check dimensions, and how to keep posterior tails stable.

## Not fully initialized

This note from NIMBLE is normal on complex models; it flags variables not touched yet. Always run `fit$model$initializeInfo()` to confirm there are no truly missing nodes. The short example below logs the message and shows how to inspect:

```{r init-note}
bundle <- build_nimble_bundle(y = abs(rnorm(40)) + 0.1, backend = "sb", kernel = "normal", J = 5)
if (use_cached_fit) {
  fit <- fit_small
} else {
  fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE)
}
fit$bundle$code  # ensures model objects exist
```

When the warning persists, check `bundle$dimensions` and explicit initial values in `bundle$inits` to ensure every latent vector has compatible length.

## Dimension mismatch

Mixing data with different numbers of covariates quickly raises dimension errors. Always confirm `ncol(newdata) == ncol(bundle$design$data$X)` before calling `predict()`.

```{r dimension-check}
bad_X <- matrix(1, nrow = 5, ncol = 1)
stopifnot(ncol(bad_X) == ncol(bundle$dimensions$X %||% matrix()))
```

## `'object "v" not found'` and stick variables

This class of errors arises when the `components` truncation is too small or `epsilon` drives many weight draws to zero. Remedy: increase `components` (your \(J\)) and raise `epsilon` slightly. The following histogram shows posterior weight spread:

```{r weight-hist, fig.cap="Posterior weights that survived truncation."}
ws <- draw_w <- DPmixGPD:::.extract_draws_matrix(fit, drop_v = TRUE)
weight_cols <- grep("^w\\[[0-9]+\\]$", colnames(draw_w), value = TRUE)
weight_df <- data.frame(weight = c(draw_w[, weight_cols]))
ggplot(weight_df, aes(x = weight)) +
  geom_histogram(bins = 25, fill = "#1f77b4", color = "white") +
  labs(title = "Posterior weight distribution (some components may collapse)")
```

## Slow mixing

If trace plots show drift, lengthen `nburnin`, increase `thin`, or reduce `niter` under raw CPU pressure. Try the `plot(fit)` diagnostics above for thresholds and tail shape, and use `plot(fit, family = "traceplot", params = "alpha")` to monitor concentration.

## Tail instability

When tail shape/scale oscillate, the usual remedies are:

- **Prior sensitivity:** tighten `tail_shape` priors (e.g., `param_specs = list(tail_shape = list(mode = "fixed", value = 0.2))`).
- **More exceedances:** raise `threshold` by setting a higher initial guess or increasing `J`.
- **GPD flag:** temporarily set `GPD = FALSE` to ensure your bulk is stable before reintroducing the tail.

## Decision chart

```{r decision-chart}
symptoms <- data.frame(
  symptom = c("Dimension mismatch", "Not fully initialized", "Tail shape blows up", "Slow mixing"),
  cause = c("newdata matrix columns differ", "latent nodes not set before compile", "too few tail exceedances", "thin/nchains too small"),
  fix = c("match columns/newdata", "provide explicit inits", "increase components/J or seed", "increase nburnin/thin or reduce autocorr")
)
knitr::kable(symptoms)
```

## Prediction example

```{r predictive-snapshot}
quant <- predict(fit, type = "quantile", p = c(.5, .9, .99))
quant
```

## Session info

```{r session}
sessionInfo()
```
