---
title: "9. Conditional DPmix with Stick-Breaking Backend"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{9. Conditional DPmix SB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  cache = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(dplyr)
library(tibble)
set.seed(123)
```

# Conditional DPmix: Stick-Breaking Backend

**Purpose**: Replace the CRP backend with stick-breaking truncation while keeping the covariate-dependent bulk structure. This demonstrates how fixed `components` interplay with covariates.

---

## Data Setup

```{r data-setup}
data("nc_posX100_p3_k2")
y <- nc_posX100_p3_k2$y
X <- as.matrix(nc_posX100_p3_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

ggplot(data.frame(y = y, x1 = X[, 1]), aes(x = x1, y = y)) +
  geom_point(alpha = 0.6, color = "darkorange") +
  geom_smooth(method = "loess", color = "navy", fill = NA) +
  labs(title = "y vs X1 (SB)", x = "X1", y = "y") +
  theme_minimal()
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Conditional Dataset Summary (SB)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Model Specification

```{r spec}
bundle_sb_cond <- build_nimble_bundle(
  y = y,
  X = X,
  kernel = "gamma",
  backend = "sb",
  GPD = FALSE,
  components = 5,
  mcmc = list(
    niter = 1000,
    nburnin = 250,
    nchains = 2,
    thin = 1
  )
)
```

---

## Running MCMC

```{r mcmc}
fit_sb_cond <- run_mcmc_bundle_manual(bundle_sb_cond)
summary(fit_sb_cond)
```

---

## Conditional Predictive Density

```{r cond-predict}
X_new <- expand.grid(x1 = seq(-2, 2, length.out = 3), x2 = c(-1, 1), x3 = 0)
y_grid <- seq(-1, 8, length.out = 200)

densities <- lapply(seq_len(nrow(X_new)), function(i) {
  pred <- predict(fit_sb_cond, x = as.matrix(X_new[i, ]), y = y_grid, type = "density")
  data.frame(
    y = y_grid,
    density = pred$fit[[1]],
    label = paste0("x1=", round(X_new[i, "x1"], 1), ", x2=", X_new[i, "x2"])
  )
})

df_cond <- bind_rows(densities)

ggplot(df_cond, aes(x = y, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "SB Conditional Predictive Densities", x = "y", y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Quantile Drift with Covariates

```{r quantile-drift}
X_eval <- cbind(x1 = seq(-2, 2, length.out = 5), x2 = 0, x3 = 0)
quant_probs <- c(0.25, 0.5, 0.75)

quant_df <- lapply(seq_len(nrow(X_eval)), function(i) {
  pred <- predict(fit_sb_cond, x = X_eval[i, , drop = FALSE], type = "quantile", index = quant_probs)
  data.frame(
    x1 = X_eval[i, "x1"],
    quant = quant_probs,
    estimate = pred$fit$estimate
  )
})

bind_rows(quant_df) %>%
  ggplot(aes(x = x1, y = estimate, color = factor(quant))) +
  geom_line(linewidth = 1) +
  labs(title = "SB Conditional Quantiles vs x1", x = "x1", y = "y", color = "Quantile") +
  theme_minimal()
```

---

## Residuals & Diagnostics

```{r residuals}
plot(fitted(fit_sb_cond))
```

```{r diagnostics}
plot(fit_sb_cond, family = c("traceplot", "density"))
```

---

## Takeaways

- Stick-breaking component count is fixed but still supports covariate-dependent mixtures.
- `predict(..., type = "density")` returns group-specific densities for each `X`.
- `predict(..., type = "quantile")` shows how median and 75% quantiles shift with `x1`.
- Diagnoses rely on the same S3 `plot()`/`fitted()` pipeline available in other vignettes.
