---
title: "Developer Guide: Adding a Kernel / Extending the Engine"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Developer Guide: Adding a Kernel / Extending the Engine}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE,
  cache.path = "cache/10-developer-",
  fig.width = 6,
  warning = FALSE,
  message = FALSE
)
library(DPmixGPD)
library(nimble)
use_cached_fit <- TRUE
.fit_path <- function(name) {
  path <- system.file("extdata", name, package = "DPmixGPD")
  if (path == "") path <- file.path("inst", "extdata", name)
  path
}
fit_small <- readRDS(.fit_path("fit_small.rds"))
library(ggplot2)
```

What you'll learn: how to register a new kernel, the Nimble hooks needed, and the minimal regression suite for keeping the engine stable.

## Kernel registry steps

1. Call `init_kernel_registry()` to inspect current labels and slot names.
2. Each entry needs `density`, `quantile`, `rng`, and `nimble` stubs following the naming pattern `d<Name>Mix`, `q<Name>Mix`, `r<Name>Mix`, `d<Name>`. The mixture index `j` is always the component number.
3. Add the kernel to `get_kernel_registry()` in `R/00-kernel-registry.R`, providing:
   - `support`: e.g., `(0, Inf)` or `(-Inf, Inf)`,
   - `params`: names of location/scale/shape,
   - `nimble`: the generated Nimble code snippet for the kernel density.

## Nimble density/rng hooks

```{r kernel-template, eval=FALSE}
# Dummy skeleton (copy into R/0-base-kernels.R and register)
nimbleFunction(
  run = function(y = double(1), mu = double(0), sigma = double(0)) {
    returnType(double(0))
    dnorm(y, mu, sigma, log = 1)
  }
)
```

Register the kernel using `init_kernel_registry()` for the lazy-loaded metadata; the backend generator picks up names automatically from `get_kernel_registry()`.

## Adding a kernel entry

```{r register-template, eval=FALSE}
init_kernel_registry("custom", list(
  density = "dCustomMix",
  quantile = "qCustomMix",
  rng = "rCustomMix",
  support = "(0, Inf)",
  params = c("mu", "sigma")
))
```

## Prediction sanity check (one snippet)

```{r developer-pred}
y <- sim_bulk_tail(n = 80, seed = 123)
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "lognormal",
  GPD = FALSE,
  J = 4,
  mcmc = list(niter = 200, nburnin = 50, thin = 2, nchains = 2, seed = c(1, 2))
)
if (use_cached_fit) {
  fit <- fit_small
} else {
  fit <- run_mcmc_bundle_manual(bundle)
}
predict(fit, type = "quantile", p = c(0.5, 0.9))
```

## Diagnostic plot

```{r developer-plot}
data.frame(y = y) |>
  ggplot(aes(x = y)) +
  geom_freqpoly(color = "steelblue") +
  labs(title = "Developer-mode data distribution", x = "y", y = "Count")
```

## Diagnostics to keep in the test checklist

1. `build_nimble_bundle()` with explicit `components`/`GPD` values for each kernel.
2. `run_mcmc_bundle_manual()` for at least 500 iterations on `seed`-locked data.
3. `predict(..., type = "quantile"/"density"/"draws")` to ensure a shared interface.
4. `validate_fit()` (or any internal helper) to assert thresholds and tail shapes obey the `GPD` flag.
5. Pre-run plotting scripts such as `plot(fit)` to verify mixture weights are accessible via `j` components.

## Session info

```{r session-info}
sessionInfo()
```
