---
title: "10. Conditional DPmixGPD with CRP Backend"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{10. Conditional DPmixGPD CRP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  cache = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(dplyr)
library(tibble)
set.seed(123)
```

# Conditional DPmixGPD: CRP Backend with Tail Augmentation

**Purpose**: Combine conditional modeling with GPD tail augmentation so each covariate slice inherits both mixture bulk and tail behavior. This extends the unconditional GPD (v06) and conditional DP (v08).

---

## Data Setup

```{r data-setup}
data("nc_pos_tailX200_p3_k4")
y <- nc_pos_tailX200_p3_k4$y
X <- as.matrix(nc_pos_tailX200_p3_k4$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

ggplot(data.frame(y = y, x1 = X[, 1]), aes(x = x1, y = y)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_smooth(method = "loess", color = "steelblue", fill = NA) +
  labs(title = "Outcome vs X1 (Tail dataset)", x = "X1", y = "y") +
  theme_minimal()
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Conditional Tail Dataset Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Threshold Selection

```{r threshold}
u_threshold <- quantile(y, 0.85)

ggplot(data.frame(y = y), aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "magenta", alpha = 0.6) +
  geom_vline(xintercept = u_threshold, linetype = "dashed", color = "black") +
  labs(title = paste("Threshold at", signif(u_threshold, 3)), x = "y", y = "Density") +
  theme_minimal()
```

---

## Model Specification & Bundle

```{r bundle}
bundle_cond_gpd <- build_nimble_bundle(
  y = y,
  X = X,
  kernel = "lognormal",
  backend = "crp",
  GPD = TRUE,
  components = 5,
  param_specs = list(
    gpd = list(
      threshold = list(mode = "fixed", value = u_threshold)
    )
  ),
  mcmc = list(
    niter = 1000,
    nburnin = 250,
    nchains = 2,
    thin = 1
  )
)
```

---

## Running MCMC

```{r mcmc-run}
fit_cond_gpd <- run_mcmc_bundle_manual(bundle_cond_gpd)
summary(fit_cond_gpd)
```

---

## Conditional Tail-aware Predictions

```{r density-predict}
X_new <- matrix(c(-1, 0, 0, 0, 0, 0, 1, 1, 0), ncol = 3, byrow = TRUE)
y_grid <- seq(0, max(y) * 1.2, length.out = 200)

df_pred <- lapply(seq_len(nrow(X_new)), function(i) {
  pred <- predict(fit_cond_gpd, x = X_new[i, , drop = FALSE], y = y_grid, type = "density")
  data.frame(
    y = y_grid,
    density = pred$fit[[1]],
    label = paste("x1=", X_new[i, 1], ", x2=", X_new[i, 2], sep = "")
  )
})

bind_rows(df_pred) %>%
  ggplot(aes(x = y, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Conditional Density with GPD Tail", x = "y", y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Tail Quantiles vs Covariates

```{r quantiles}
X_grid <- cbind(x1 = seq(-1, 1, length.out = 5), x2 = 0, x3 = 0)
quant_probs <- c(0.90, 0.95)

quant_df <- lapply(seq_len(nrow(X_grid)), function(i) {
  pred <- predict(fit_cond_gpd, x = X_grid[i, , drop = FALSE], type = "quantile", index = quant_probs)
  data.frame(
    x1 = X_grid[i, "x1"],
    quantile = quant_probs,
    estimate = pred$fit$estimate
  )
})

bind_rows(quant_df) %>%
  ggplot(aes(x = x1, y = estimate, color = factor(quantile))) +
  geom_line(linewidth = 1) +
  labs(title = "Tail Quantiles vs x1 (CRP)", x = "x1", y = "Quantile", color = "Probability") +
  theme_minimal()
```

---

## Residuals & Diagnostics

```{r residuals}
plot(fitted(fit_cond_gpd))
```

```{r diagnostics}
plot(fit_cond_gpd, family = c("traceplot", "density"))
```

---

## Takeaways

- Conditional DPmix with a GPD tail lets extreme quantiles vary with covariates.
- The CRP backend samples the bulk and tail jointly while thresholding at the 85th percentile.
- `predict()` + `plot()` remain the main tools for densities, survival curves, and quantiles; residual diagnostics check fit quality.
- Next: Mirror this workflow with the SB backend in `v11`.
