---
title: "11. Conditional DPmixGPD with Stick-Breaking Backend"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{11. Conditional DPmixGPD SB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
library(ggplot2)
set.seed(123)
```

# Conditional DPmixGPD: Stick-Breaking Backend with Covariates & Tail

**Goal**: Apply **SB backend** with **covariates** and **GPD tail** for fixed-component conditional modeling.

---

## Data Setup

```{r data-setup}
set.seed(42)
n <- 150

X <- cbind(
  x1 = rnorm(n, mean = 0, sd = 1),
  x2 = rnorm(n, mean = 0, sd = 1)
)

mean_shift_1 <- 1 + 0.5 * X[, 1]
mean_shift_2 <- 3 + 0.2 * X[, 1]

component_1 <- rgamma(75, shape = 2, rate = 1.5) + mean_shift_1[1:75]
component_2 <- rgamma(75, shape = 1, rate = 0.8) + mean_shift_2[76:150]

tail_component <- DPmixGPD::rGpd(20, threshold = 0, scale = 1, shape = 0.3)
y_tail <- c(component_1, component_2, tail_component)

X_tail <- rbind(X, X[1:20, ])
u_threshold <- quantile(y_tail, 0.8)

cat("Sample size:", length(y_tail), "\n")
```

---

## Model Specification

```{r spec}
spec_cond_sb_gpd <- compile_model_spec(
  y = y_tail,
  X = X_tail,
  kernel = "gamma",
  backend = "sb",            # SB backend
  GPD = TRUE,
  threshold = u_threshold,
  J = 15,
  verbose = FALSE
)

cat("Conditional SB + GPD Specification created.\n")
```

---

## Bundle & MCMC

```{r bundle-mcmc}
bundle_cond_sb_gpd <- build_nimble_bundle(
  spec_cond_sb_gpd,
  mcmc = list(niter = 2000, nburnin = 500, nchains = 2, thin = 1)
)

fit_cond_sb_gpd <- run_mcmc_bundle_manual(bundle_cond_sb_gpd)

summary(fit_cond_sb_gpd)
```

---

## Predictions

```{r pred}
X_new <- cbind(x1 = c(-1, 0, 1), x2 = 0)
y_grid <- seq(-2, 15, length.out = 150)

densities_list <- list()
for (i in 1:nrow(X_new)) {
  dens <- predict(fit_cond_sb_gpd, newdata = y_grid, X = X_new[i, , drop = FALSE])
  densities_list[[i]] <- data.frame(
    y = y_grid,
    density = dens,
    group = paste("X1=", round(X_new[i, 1], 1), sep = "")
  )
}

df_pred <- do.call(rbind, densities_list)

ggplot(df_pred, aes(x = y, y = density, color = group)) +
  geom_line(linewidth = 1) +
  labs(title = "Conditional Density with Tail (SB + GPD)", x = "y", y = "Density") +
  theme_minimal()
```

---

## Key Takeaways

- **SB + Covariates + Tail**: Fixed components with flexible tail
- **Efficiency**: Faster than CRP for large data
- **Next**: Causal models in vignettes 12-15
