---
title: "12. Causal Inference: Same Backend (CRP) for PS & Outcome"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{12. Causal Same Backend CRP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
library(ggplot2)
set.seed(123)
```

# Causal Inference: CRP Backend for Both Propensity Score & Outcome

**Goal**: Estimate **treatment effects** with **propensity score modeling** (PS) using CRP backend.

**Model**:
- **PS Model** (logit/probit/naive): $P(T=1|X)$ via flexible DP mixture
- **Outcome Model**: $Y|T,X,PS$ via DP mixture conditioning on covariates

**PS Options**: `logit`, `probit`, `naive`, or `FALSE` (disabled)

---

## Data Setup: Two-Arm Study

```{r data-setup}
set.seed(42)
n <- 200

# Covariates
X <- cbind(
  x1 = rnorm(n, mean = 0, sd = 1),
  x2 = rnorm(n, mean = 0, sd = 1)
)

# Treatment (binary)
T <- rbinom(n, size = 1, prob = 0.5)

# Outcome: function of treatment, covariates, and PS
ps_prob <- 0.5  # True PS model (simplified)
y <- ifelse(
  T == 1,
  rnorm(n, mean = 3 + 0.5 * X[, 1], sd = 1.2),  # Treatment arm
  rnorm(n, mean = 1 + 0.3 * X[, 1], sd = 1.0)   # Control arm
)

df_causal <- data.frame(
  y = y,
  T = as.factor(T),
  x1 = X[, 1],
  x2 = X[, 2]
)

# Visualization
p_raw <- ggplot(df_causal, aes(x = x1, y = y, color = T)) +
  geom_point(alpha = 0.6) +
  scale_color_manual(values = c("0" = "steelblue", "1" = "red")) +
  labs(title = "Outcome by Treatment & Covariate", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()

print(p_raw)

cat("Sample size:", n, "\n")
cat("Treatment:", table(T), "\n")
cat("Mean outcome (T=0):", mean(y[T == 0]), "\n")
cat("Mean outcome (T=1):", mean(y[T == 1]), "\n")
```

---

## Causal Bundle Specification

```{r causal-spec}
# Build causal bundle with PS modeling
bundle_causal_crp <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "normal",                # Outcome kernel
  backend = "crp",                  # CRP for outcome
  PS = "logit",                     # Propensity score: logit model
  GPD = FALSE,
  Kmax = 10,
  mcmc = list(
    niter = 2000,
    nburnin = 500,
    nchains = 2,
    thin = 1
  )
)

cat("Causal bundle (CRP + logit PS) created.\n")
```

---

## PS Model Variants

### Logit PS Model

```{r ps-logit}
# Logit: Default, flexible
bundle_logit <- build_causal_bundle(
  y = y, T = T, X = X,
  kernel = "normal", backend = "crp",
  PS = "logit",           # Bayesian logit via DP mixture
  GPD = FALSE, Kmax = 10,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)

fit_logit <- run_mcmc_bundle_manual(bundle_logit)
cat("\n=== Logit PS Model ===\n")
summary(fit_logit)
```

### Probit PS Model

```{r ps-probit}
bundle_probit <- build_causal_bundle(
  y = y, T = T, X = X,
  kernel = "normal", backend = "crp",
  PS = "probit",          # Bayesian probit
  GPD = FALSE, Kmax = 10,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)

fit_probit <- run_mcmc_bundle_manual(bundle_probit)
cat("\n=== Probit PS Model ===\n")
summary(fit_probit)
```

### Naive PS Model

```{r ps-naive}
# Naive: No PS correction (treatment as covariate only)
bundle_naive <- build_causal_bundle(
  y = y, T = T, X = X,
  kernel = "normal", backend = "crp",
  PS = "naive",           # No PS model
  GPD = FALSE, Kmax = 10,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)

fit_naive <- run_mcmc_bundle_manual(bundle_naive)
cat("\n=== Naive PS Model (No PS Adjustment) ===\n")
summary(fit_naive)
```

---

## Treatment Effect Estimation

### Average Treatment Effect (ATE)

```{r ate}
# Extract treatment effect from fit
cat("\n=== TREATMENT EFFECTS ===\n")

cat("\nLogit PS Model:\n")
ate_logit <- coef(fit_logit)
cat("Posterior mean treatment effect:", ate_logit, "\n")

cat("\nProbit PS Model:\n")
ate_probit <- coef(fit_probit)
cat("Posterior mean treatment effect:", ate_probit, "\n")

cat("\nNaive Model:\n")
ate_naive <- coef(fit_naive)
cat("Posterior mean treatment effect:", ate_naive, "\n")
```

---

## Conditional Average Treatment Effect (CATE)

```{r cate}
# CATE: ATE conditional on covariates
X_cate <- cbind(x1 = c(-1, 0, 1), x2 = 0)

cate_logit <- predict(fit_logit, newdata = X_cate, type = "treatment.effect")
cat("CATE at X1 = -1, 0, 1 (Logit PS):\n")
print(cate_logit)
```

---

## PS Comparison: All Three Models

```{r ps-comparison}
cat("\n=== MODEL COMPARISON ===\n")

df_comp <- data.frame(
  Model = c("Logit PS", "Probit PS", "Naive"),
  Comparison = "See summary()",
  ATE = c(ate_logit, ate_probit, ate_naive)
)

print(df_comp)

cat("\n✓ Logit typically preferred for binary outcome in treatment\n")
```

---

## Propensity Score Distribution

```{r ps-dist}
# Extract estimated PS values
ps_fitted <- fitted(fit_logit, type = "ps")

df_ps <- data.frame(ps = ps_fitted, T = as.factor(T))

p_ps <- ggplot(df_ps, aes(x = ps, fill = T)) +
  geom_histogram(alpha = 0.6, bins = 30) +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "red")) +
  labs(title = "Estimated Propensity Score Distribution", x = "PS", y = "Count") +
  theme_minimal()

print(p_ps)
```

---

## Overlap Assessment

```{r overlap}
# PS overlap: critical for causal inference
ps_control <- ps_fitted[T == 0]
ps_treated <- ps_fitted[T == 1]

p_overlap <- ggplot(data.frame(ps = ps_fitted, T = as.factor(T)), 
                     aes(x = ps, fill = T)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "red")) +
  labs(title = "PS Overlap: Common Support", x = "PS", y = "Density") +
  theme_minimal()

print(p_overlap)

cat("Common support check:\n")
cat("  Control PS range: [", min(ps_control), ",", max(ps_control), "]\n")
cat("  Treated PS range: [", min(ps_treated), ",", max(ps_treated), "]\n")
```

---

## Residual Analysis

```{r residuals}
resid_vals <- residuals(fit_logit)

p_resid <- ggplot(data.frame(r = resid_vals, T = as.factor(T)), 
                   aes(x = T, y = r, fill = T)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(alpha = 0.3, width = 0.2) +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "red")) +
  labs(title = "Residuals by Treatment", x = "Treatment", y = "Residual") +
  theme_minimal()

print(p_resid)
```

---

## Key Takeaways

- **PS Models**: Logit, probit, and naive (covariate-only)
- **CRP Backend**: Flexible outcome modeling given PS
- **Treatment Effects**: ATE and CATE available
- **Common Support**: Check PS overlap for valid inference
- **Next**: SB backend in vignette 13, then heterogeneous effects
