---
title: "12. Causal Inference: Same Backend (CRP) - Normal Kernel"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{12. Causal Inference: Same Backend (CRP) - Normal Kernel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  cache = FALSE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(kableExtra)
library(dplyr)
library(tibble)
set.seed(123)
```

# Causal Inference: Same Backend (CRP) - Normal Kernel {#causal-crp-same-backend}

This vignette fits two CRP-based causal models using the same kernel (normal):

- Model A: bulk-only outcome models (GPD = FALSE)
- Model B: GPD-augmented outcome models (GPD = TRUE)

Each model demonstrates the full causal S3 workflow: bundle, fit, predict, plot, ATE, and QTE.

---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- causal_alt_real500_p4_k2$y
T <- causal_alt_real500_p4_k2$T
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Causal Dataset Summary", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")

x_eval <- X[1:40, , drop = FALSE]
y_eval <- y[1:40]
u_threshold <- as.numeric(stats::quantile(y, 0.8, names = FALSE))
```

```{r data-scatter}
df_causal <- data.frame(y = y, T = as.factor(T), x1 = X[, 1], x2 = X[, 2])

p_scatter <- ggplot(df_causal, aes(x = x1, y = y, color = T)) +
  geom_point(alpha = 0.5) +
  labs(title = "Outcome vs X1 by Treatment", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()

p_scatter
```

---

## Model A: CRP Bulk-only (Normal)

```{r bundle-crp-bulk}
bundle_crp_bulk <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "normal",
  backend = "crp",
  PS = "logit",
  GPD = FALSE,
  J = 5,
  mcmc_outcome = list(niter = 40, nburnin = 10, nchains = 1, thin = 1, seed = 1),
  mcmc_ps = list(niter = 40, nburnin = 10, nchains = 1, thin = 1, seed = 1)
)

bundle_crp_bulk
summary(bundle_crp_bulk)
```

```{r fit-crp-bulk}
fit_crp_bulk <- run_mcmc_causal(bundle_crp_bulk)
summary(fit_crp_bulk)
params(fit_crp_bulk)
```

```{r plot-fit-crp-bulk}
plot(fit_crp_bulk)
```

```{r predict-mean-crp-bulk}
pred_mean_bulk <- predict(fit_crp_bulk, x = x_eval, type = "mean",
                          interval = "credible", nsim_mean = 100)
plot(pred_mean_bulk)
```

```{r predict-quantile-crp-bulk}
pred_q_bulk <- predict(fit_crp_bulk, x = x_eval, type = "quantile",
                       p = 0.5, interval = "credible")
plot(pred_q_bulk)
```

```{r predict-density-crp-bulk}
pred_d_bulk <- predict(fit_crp_bulk, x = x_eval, y = y_eval,
                       type = "density", interval = "credible")
plot(pred_d_bulk)
```

```{r predict-prob-crp-bulk}
pred_p_bulk <- predict(fit_crp_bulk, x = x_eval, y = y_eval,
                       type = "prob", interval = "credible")
plot(pred_p_bulk)
```

```{r ate-crp-bulk}
ate_bulk <- ate(fit_crp_bulk, newdata = x_eval,
                interval = "credible", nsim_mean = 100)
plot(ate_bulk)
```

```{r qte-crp-bulk}
qte_bulk <- qte(fit_crp_bulk, probs = c(0.25, 0.5, 0.75),
                newdata = x_eval, interval = "credible")
plot(qte_bulk)
```

---

## Model B: CRP with GPD Tail (Normal)

```{r bundle-crp-gpd}
param_specs_gpd <- list(
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "lognormal",
      args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
    )
  )
)

bundle_crp_gpd <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "normal",
  backend = "crp",
  PS = "logit",
  GPD = TRUE,
  J = 5,
  param_specs = param_specs_gpd,
  mcmc_outcome = list(niter = 40, nburnin = 10, nchains = 1, thin = 1, seed = 2),
  mcmc_ps = list(niter = 40, nburnin = 10, nchains = 1, thin = 1, seed = 2)
)

bundle_crp_gpd
summary(bundle_crp_gpd)
```

```{r fit-crp-gpd}
fit_crp_gpd <- run_mcmc_causal(bundle_crp_gpd)
summary(fit_crp_gpd)
params(fit_crp_gpd)
```

```{r plot-fit-crp-gpd}
plot(fit_crp_gpd)
```

```{r predict-mean-crp-gpd}
pred_mean_gpd <- predict(fit_crp_gpd, x = x_eval, type = "mean",
                         interval = "credible", nsim_mean = 100)
plot(pred_mean_gpd)
```

```{r predict-quantile-crp-gpd}
pred_q_gpd <- predict(fit_crp_gpd, x = x_eval, type = "quantile",
                      p = 0.5, interval = "credible")
plot(pred_q_gpd)
```

```{r predict-density-crp-gpd}
pred_d_gpd <- predict(fit_crp_gpd, x = x_eval, y = y_eval,
                      type = "density", interval = "credible")
plot(pred_d_gpd)
```

```{r predict-prob-crp-gpd}
pred_p_gpd <- predict(fit_crp_gpd, x = x_eval, y = y_eval,
                      type = "prob", interval = "credible")
plot(pred_p_gpd)
```

```{r ate-crp-gpd}
ate_gpd <- ate(fit_crp_gpd, newdata = x_eval,
               interval = "credible", nsim_mean = 100)
plot(ate_gpd)
```

```{r qte-crp-gpd}
qte_gpd <- qte(fit_crp_gpd, probs = c(0.25, 0.5, 0.75),
               newdata = x_eval, interval = "credible")
plot(qte_gpd)
```
