---
title: "13. Causal Inference: Same Backend (SB) for PS & Outcome"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{13. Causal Same Backend SB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  cache = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(dplyr)
library(tibble)
library(patchwork)
set.seed(123)
```

# Causal Inference: Same Backend (SB) for PS & Outcome

**Purpose**: Repeat the causal workflow with a stick-breaking backend for both arms, highlighting fixed truncation and its effect on CQTE.

---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- causal_alt_real500_p4_k2$y
T <- causal_alt_real500_p4_k2$T
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

df_causal <- data.frame(y = y, T = as.factor(T), x1 = X[, 1], x2 = X[, 2])

p_scatter <- ggplot(df_causal, aes(x = x1, y = y, color = T)) +
  geom_point(alpha = 0.5) +
  labs(title = "Outcomes by X1 & Treatment (SB)", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()

grid.arrange(p_scatter, ncol = 1)
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Causal Dataset Summary (SB)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Bundle & MCMC

```{r bundle}
bundle_causal_sb <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = c("normal", "normal"),
  backend = c("sb", "sb"),
  PS = "logit",
  GPD = c(FALSE, FALSE),
  J = c(5, 5),
  mcmc = list(
    niter = 300,
    nburnin = 60,
    nchains = 1,
    thin = 1
  )
)

fit_causal_sb <- run_mcmc_causal(bundle_causal_sb)
summary(fit_causal_sb)
```

---

## Propensity Score Overlap

```{r ps}
ps_vals <- fitted(fit_causal_sb, type = "ps")

df_ps <- data.frame(ps = ps_vals, T = as.factor(T))

ggplot(df_ps, aes(x = ps, fill = T)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "firebrick")) +
  labs(title = "Propensity Scores for SB Causal Fit", x = "PS", y = "Count") +
  theme_minimal()
```

---

## CQTE

```{r cqte}
X_ref <- matrix(colMeans(X), nrow = 1)
tau_grid <- seq(0.05, 0.95, length.out = 10)
pred_tau <- predict(fit_causal_sb, newdata = X_ref, type = "quantile", p = tau_grid)

q_trt <- as.numeric(pred_tau$trt$fit)
q_con <- as.numeric(pred_tau$con$fit)

qte_tbl <- tibble(
  tau = tau_grid,
  cqte = q_trt - q_con
)

ggplot(qte_tbl, aes(x = tau, y = cqte)) +
  geom_line(color = "darkgreen", linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  labs(title = "CQTE (SB backend)", x = "τ", y = "Q1 - Q0") +
  theme_minimal()
```

---

## Density Comparison

```{r density-diff}
y_grid <- seq(min(y), max(y), length.out = 200)
pred_density <- predict(fit_causal_sb, newdata = X_ref, y = y_grid, type = "density")

df_density <- data.frame(
  y = y_grid,
  control = pred_density$con$fit[[1]],
  treated = pred_density$trt$fit[[1]]
)

ggplot(df_density, aes(x = y)) +
  geom_line(aes(y = control, color = "Control")) +
  geom_line(aes(y = treated, color = "Treated")) +
  scale_color_manual(values = c("Control" = "steelblue", "Treated" = "firebrick")) +
  labs(title = "Posterior Densities by Arm (SB)", y = "Density", color = "Arm") +
  theme_minimal()
```

---

## Takeaways

- Stick-breaking truncation fixes component count while PS logit allows flexible weighting between arms.
- CQTE curves still show heterogeneity and can be compared with the CRP results in `v12`.
- Use `fitted(..., type = "ps")` + `predict()` for full PS-outcome diagnostics before trusting CQTE.
- Next: switch to mixed backend scenarios (v14, v15) to highlight CRP/SB asymmetry.

