---
title: "13. Causal Inference: Same Backend (SB) for PS & Outcome"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{13. Causal Same Backend SB}
  %\VignetteEngine{kmitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
library(ggplot2)
set.seed(123)
```

# Causal Inference: SB Backend for Both Propensity Score & Outcome

**Goal**: Apply **Stick-Breaking backend** for both PS and outcome modeling in causal inference.

---

## Data Setup

```{r data-setup}
set.seed(42)
n <- 200

X <- cbind(
  x1 = rnorm(n, mean = 0, sd = 1),
  x2 = rnorm(n, mean = 0, sd = 1)
)

T <- rbinom(n, size = 1, prob = 0.5)

y <- ifelse(
  T == 1,
  rnorm(n, mean = 3 + 0.5 * X[, 1], sd = 1.2),
  rnorm(n, mean = 1 + 0.3 * X[, 1], sd = 1.0)
)

cat("Sample size:", n, "\n")
```

---

## Causal Bundle with SB Backend

```{r causal-sb}
bundle_causal_sb <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "normal",
  backend = "sb",           # Stick-Breaking
  PS = "logit",
  GPD = FALSE,
  J = 10,                   # Fixed components
  mcmc = list(
    niter = 2000,
    nburnin = 500,
    nchains = 2,
    thin = 1
  )
)

fit_causal_sb <- run_mcmc_bundle_manual(bundle_causal_sb)

cat("\n=== CAUSAL SB MODEL ===\n")
summary(fit_causal_sb)
```

---

## Treatment Effect Comparison: CRP vs SB

```{r comparison}
# CRP (from vignette 12)
bundle_crp_logit <- build_causal_bundle(
  y = y, T = T, X = X,
  kernel = "normal", backend = "crp",
  PS = "logit", GPD = FALSE, Kmax = 10,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)
fit_crp_logit <- run_mcmc_bundle_manual(bundle_crp_logit)

ate_crp <- coef(fit_crp_logit)
ate_sb <- coef(fit_causal_sb)

cat("\n=== TREATMENT EFFECT COMPARISON ===\n")
cat("CRP ATE:", ate_crp, "\n")
cat("SB ATE: ", ate_sb, "\n")
cat("Difference:", abs(ate_crp - ate_sb), "\n")

```

---

## Propensity Score

```{r ps}
ps_fitted <- fitted(fit_causal_sb, type = "ps")

df_ps <- data.frame(ps = ps_fitted, T = as.factor(T))

ggplot(df_ps, aes(x = ps, fill = T)) +
  geom_histogram(alpha = 0.6, bins = 30) +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "red")) +
  labs(title = "PS Distribution (SB Backend)", x = "PS", y = "Count") +
  theme_minimal()
```

---

## Key Takeaways

- **SB Causal Models**: Fixed-component approach for faster inference
- **Same PS Options**: logit, probit, naive available
- **CRP vs SB Trade-off**: Applies to causal setting too
- **Next**: Different backends for PS vs outcome in vignettes 14-15
