---
title: "14. Causal Inference: Different Backends per Arm (Bulk-only)"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{14. Causal Different Backends CRP}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  cache = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(dplyr)
library(tibble)
library(patchwork)
set.seed(123)
```

# Causal Inference: Different Backends per Arm (Bulk-only)

**Purpose**: Show how outcome backends can differ across arms (CRP vs SB) while PS is shared. This vignette fits both combinations and compares ATE/CQTE across settings, keeping `GPD = FALSE`.

---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- causal_alt_real500_p4_k2$y
T <- causal_alt_real500_p4_k2$T
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

ggplot(data.frame(y = y, x1 = X[, 1], T = as.factor(T)), aes(x = x1, y = y, color = T)) +
  geom_point(alpha = 0.5) +
  labs(title = "Outcome vs X1 by Treatment", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Causal Dataset Summary") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Backend Combinations

```{r bundles}
# Treated: SB, Control: CRP
bundle_sb_crp <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = c("normal", "normal"),
  backend = c("sb", "crp"),
  PS = "logit",
  GPD = c(FALSE, FALSE),
  J = c(6, 6),
  mcmc = list(niter = 2000, nburnin = 500, nchains = 2, thin = 1)
)

# Treated: CRP, Control: SB
bundle_crp_sb <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = c("normal", "normal"),
  backend = c("crp", "sb"),
  PS = "logit",
  GPD = c(FALSE, FALSE),
  J = c(6, 6),
  mcmc = list(niter = 2000, nburnin = 500, nchains = 2, thin = 1)
)
```

---

## MCMC Fits & ATE Comparison

```{r mcmc-fits}
fit_sb_crp <- run_mcmc_causal(bundle_sb_crp)
fit_crp_sb <- run_mcmc_causal(bundle_crp_sb)

ate_sb_crp <- coef(fit_sb_crp)
ate_crp_sb <- coef(fit_crp_sb)

ave_tbl <- tibble(
  model = c("Treated=SB, Control=CRP", "Treated=CRP, Control=SB"),
  ATE = c(ate_sb_crp, ate_crp_sb)
)

ggplot(ave_tbl, aes(x = model, y = ATE)) +
  geom_col(fill = c("mediumseagreen", "steelblue")) +
  labs(title = "ATE under Different Backends", x = "Backend Combination", y = "ATE") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## PS Overlap (SB/CRP mixture)

```{r ps-overlap}
ps_vals <- fitted(fit_sb_crp, type = "ps")
df_ps <- data.frame(ps = ps_vals, T = as.factor(T))

ggplot(df_ps, aes(x = ps, fill = T)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "firebrick")) +
  labs(title = "Propensity Score Overlap (shared PS)", x = "PS", y = "Density") +
  theme_minimal()
```

---

## CQTE Across Backend Choices

```{r cqte-diff}
X_ref <- matrix(colMeans(X), nrow = 1)
tau_grid <- seq(0.05, 0.95, length.out = 10)

pred_sb_crp <- predict(fit_sb_crp, newdata = X_ref, type = "quantile", p = tau_grid)
pred_crp_sb <- predict(fit_crp_sb, newdata = X_ref, type = "quantile", p = tau_grid)

qte_sb_crp <- as.numeric(pred_sb_crp$trt$fit - pred_sb_crp$con$fit)
qte_crp_sb <- as.numeric(pred_crp_sb$trt$fit - pred_crp_sb$con$fit)

qte_df <- tibble(
  tau = rep(tau_grid, 2),
  cqte = c(qte_sb_crp, qte_crp_sb),
  model = rep(c("SB/CRP", "CRP/SB"), each = length(tau_grid))
)

ggplot(qte_df, aes(x = tau, y = cqte, color = model)) +
  geom_line(linewidth = 1.1) +
  labs(title = "CQTE Comparison by Backend Combo", x = "τ", y = "Q1 − Q0") +
  theme_minimal()
```

---

## Takeaways

- Different backend mixtures demonstrate how truncation style influences extreme quantiles even when PS is shared.
- `predict()` returns paired `trt`/`con` outputs so subtracting them yields CQTE curves for any $\tau$.
- Use the same PS diagnostics and `coef()` summaries across fits to ensure proper overlap before trusting treatment effects.
- Next: add GPD tails to these mixed-backend workflows (`v15`).
