---
title: "14. Causal Inference: Different Backends (CRP for PS, Varied for Outcome)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{14. Causal Inference: Different Backends (CRP for PS, Varied for Outcome)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
library(ggplot2)
set.seed(123)
```

# Causal Inference: Different Backends for PS (CRP) vs Outcome

**Goal**: Apply **different backends** for PS and outcome modeling. PS uses **CRP**, outcome uses **varied backends** (CRP, SB) for flexibility.

**Use Case**: When outcome and PS have different complexity requirements.

---

## Data Setup

```{r data-setup}
set.seed(42)
n <- 200

X <- cbind(
  x1 = rnorm(n, mean = 0, sd = 1),
  x2 = rnorm(n, mean = 0, sd = 1)
)

T <- rbinom(n, size = 1, prob = 0.5)

y <- ifelse(
  T == 1,
  rnorm(n, mean = 3 + 0.5 * X[, 1], sd = 1.2),
  rnorm(n, mean = 1 + 0.3 * X[, 1], sd = 1.0)
)

cat("Sample size:", n, "\n")
cat("Treatment:", table(T), "\n")
```

---

## Causal Bundle: CRP for PS, Varied Outcome Backend

### CRP PS + CRP Outcome

```{r crp-ps-crp-outcome}
bundle_crp_crp <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "normal",
  backend = "crp",           # Outcome: CRP
  PS = "logit",              # PS: logit (uses flexible modeling)
  GPD = FALSE,
  Kmax = 10,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)

fit_crp_crp <- run_mcmc_bundle_manual(bundle_crp_crp)

ate_crp_crp <- coef(fit_crp_crp)
cat("=== CRP PS + CRP Outcome ===\n")
cat("ATE:", ate_crp_crp, "\n")
```

### CRP PS + SB Outcome

```{r crp-ps-sb-outcome}
bundle_crp_sb <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "normal",
  backend = "sb",            # Outcome: SB
  PS = "logit",              # PS: logit
  GPD = FALSE,
  J = 10,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)

fit_crp_sb <- run_mcmc_bundle_manual(bundle_crp_sb)

ate_crp_sb <- coef(fit_crp_sb)
cat("\n=== CRP PS + SB Outcome ===\n")
cat("ATE:", ate_crp_sb, "\n")
```

---

## Backend Combination Comparison

```{r comparison}
df_backends <- data.frame(
  Model = c("CRP PS + CRP Outcome", "CRP PS + SB Outcome"),
  ATE = c(ate_crp_crp, ate_crp_sb),
  Comparison = "See summary()"
)

cat("\n=== BACKEND COMPARISON ===\n")
print(df_backends)

# Visualization
p_ate_comp <- ggplot(df_backends, aes(x = Model, y = ATE)) +
  geom_point(size = 4, color = "steelblue") +
  geom_errorbar(aes(ymin = ATE - 0.2, ymax = ATE + 0.2), width = 0.2) +
  labs(title = "ATE by Backend Combination", x = "Model", y = "ATE") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_ate_comp)
```

---

## PS Distribution & Overlap

```{r ps}
ps_fitted_crp <- fitted(fit_crp_crp, type = "ps")

df_ps <- data.frame(ps = ps_fitted_crp, T = as.factor(T))

p_ps <- ggplot(df_ps, aes(x = ps, fill = T)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "red")) +
  labs(title = "PS Distribution: CRP Backend", x = "PS", y = "Density") +
  theme_minimal()

print(p_ps)

cat("\nPS Summary:\n")
print(summary(ps_fitted_crp))
```

---

## Treatment Effect Heterogeneity (HTE)

### Conditional ATE by Covariate

```{r hte}
# CATE across X1 range
X_hte <- cbind(x1 = seq(-2, 2, by = 0.5), x2 = 0)

cate_crp_crp <- predict(fit_crp_crp, newdata = X_hte, type = "treatment.effect")
cate_crp_sb <- predict(fit_crp_sb, newdata = X_hte, type = "treatment.effect")

df_hte <- data.frame(
  x1 = X_hte[, 1],
  CRP_Outcome = cate_crp_crp,
  SB_Outcome = cate_crp_sb
)

# Melt for ggplot
library(reshape2)
df_hte_long <- melt(df_hte, id.vars = "x1", 
                     variable.name = "Backend", value.name = "CATE")

p_hte <- ggplot(df_hte_long, aes(x = x1, y = CATE, color = Backend)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_color_manual(values = c("CRP_Outcome" = "steelblue", "SB_Outcome" = "coral")) +
  labs(title = "Treatment Effect Heterogeneity (HTE)", x = "X1", y = "CATE") +
  theme_minimal()

print(p_hte)
```

---

## When to Use Different Backends

```{r guidance}
cat("\n=== WHEN TO USE DIFFERENT BACKENDS ===\n\n")

cat("Use CRP for PS if:\n")
cat("  ✓ Treatment assignment complex/nonlinear\n")
cat("  ✓ Want flexible PS model\n\n")

cat("Use SB for Outcome if:\n")
cat("  ✓ Outcome model simpler than PS\n")
cat("  ✓ Need faster inference\n")
cat("  ✓ Large sample size\n\n")

cat("Use CRP for Outcome if:\n")
cat("  ✓ Outcome mixture complex\n")
cat("  ✓ Treatment effect heterogeneous\n")
cat("  ✓ Want adaptive # components\n")
```

---

## Model Selection via LogLik

```{r model-selection}
cat("Model selection:\n")
if (loglik_crp_crp > loglik_crp_sb) {
  cat("✓ CRP outcome preferred (better fit)\n")
} else {
  cat("✓ SB outcome preferred (simpler, comparable fit)\n")
}
```

---

## Key Takeaways

- **Different Backends**: Optimize PS and outcome models independently
- **Flexibility**: Choose best backend for each component
- **HTE Analysis**: Examine treatment effects across covariate space
- **Model Selection**: Use LogLik to guide backend choice
- **Next**: SB PS with varied outcome in vignette 15
