---
title: "15. Causal Inference: Different Backends (SB for PS, Varied for Outcome)"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{15. Causal Different Backends SB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
set.seed(123)
```

# Causal Inference: Different Backends for PS (SB) vs Outcome

**Goal**: Apply **Stick-Breaking for PS** with **varied outcome backends** for fast, efficient causal inference.

**Use Case**: When PS model is simpler, outcome requires more flexibility (or vice versa).

---

## Data Setup

```{r data-setup}
set.seed(42)
n <- 200

X <- cbind(
  x1 = rnorm(n, mean = 0, sd = 1),
  x2 = rnorm(n, mean = 0, sd = 1)
)

T <- rbinom(n, size = 1, prob = 0.5)

y <- ifelse(
  T == 1,
  rnorm(n, mean = 3 + 0.5 * X[, 1], sd = 1.2),
  rnorm(n, mean = 1 + 0.3 * X[, 1], sd = 1.0)
)

print("Sample size:", n, "\n")
```

---

## Causal Models: SB PS with Varied Outcome Backends

### SB PS + CRP Outcome

```{r sb-ps-crp-outcome}
bundle_sb_crp <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "normal",
  backend = "crp",           # Outcome: CRP (flexible)
  PS = "logit",              # PS: via SB (fixed components)
  GPD = FALSE,
  Kmax = 5,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)

fit_sb_crp <- run_mcmc_bundle_manual(bundle_sb_crp)

ate_sb_crp <- coef(fit_sb_crp)
print("=== SB PS + CRP Outcome ===\n")
print("ATE:", ate_sb_crp, "\n")
```

### SB PS + SB Outcome

```{r sb-ps-sb-outcome}
bundle_sb_sb <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "normal",
  backend = "sb",            # Outcome: SB (fixed)
  PS = "logit",              # PS: via SB
  GPD = FALSE,
  J = 5,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)

fit_sb_sb <- run_mcmc_bundle_manual(bundle_sb_sb)

ate_sb_sb <- coef(fit_sb_sb)
print("\n=== SB PS + SB Outcome ===\n")
print("ATE:", ate_sb_sb, "\n")
```

---

## Comprehensive Backend Comparison (All 4 Combinations)

```{r all-combinations}
# CRP PS + CRP Outcome (from vignette 14)
bundle_crp_crp <- build_causal_bundle(
  y = y, T = T, X = X,
  kernel = "normal", backend = "crp",
  PS = "logit", GPD = FALSE, Kmax = 5,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)
fit_crp_crp <- run_mcmc_bundle_manual(bundle_crp_crp)
ate_crp_crp <- coef(fit_crp_crp)
# CRP PS + SB Outcome (from vignette 14)
bundle_crp_sb <- build_causal_bundle(
  y = y, T = T, X = X,
  kernel = "normal", backend = "sb",
  PS = "logit", GPD = FALSE, J = 5,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)
fit_crp_sb <- run_mcmc_bundle_manual(bundle_crp_sb)
ate_crp_sb <- coef(fit_crp_sb)
# Summary table
df_all <- data.frame(
  PS_Backend = rep(c("CRP", "SB"), each = 2),
  Outcome_Backend = rep(c("CRP", "SB"), times = 2),
  ATE = c(ate_crp_crp, ate_crp_sb, ate_sb_crp, ate_sb_sb),
  Comparison = "See summary()"
)

print("\n=== ALL 4 BACKEND COMBINATIONS ===\n")
print(df_all)

# Visualization
p_all_ate <- ggplot(df_all, aes(x = paste(PS_Backend, Outcome_Backend, sep="+"), 
                                y = ATE, fill = PS_Backend)) +
  geom_col(alpha = 0.7) +
  scale_fill_manual(values = c("CRP" = "steelblue", "SB" = "coral")) +
  labs(title = "ATE Across Backend Combinations", x = "Backend Combo", y = "ATE") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(p_all_ate)
```

---

## PS Distribution: SB Backend

```{r ps}
ps_fitted_sb <- fitted(fit_sb_crp, type = "ps")

df_ps <- data.frame(ps = ps_fitted_sb, T = as.factor(T))

p_ps <- ggplot(df_ps, aes(x = ps, fill = T)) +
  geom_histogram(alpha = 0.6, bins = 30) +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "red")) +
  labs(title = "PS Distribution: SB Backend", x = "PS", y = "Count") +
  theme_minimal()

print(p_ps)
```

---

## Treatment Effect Heterogeneity: All Combinations

```{r hte}
X_hte <- cbind(x1 = seq(-2, 2, by = 0.5), x2 = 0)

# Extract CATE from all fits
cate_crp_crp <- predict(fit_crp_crp, newdata = X_hte, type = "treatment.effect")
cate_crp_sb <- predict(fit_crp_sb, newdata = X_hte, type = "treatment.effect")
cate_sb_crp <- predict(fit_sb_crp, newdata = X_hte, type = "treatment.effect")
cate_sb_sb <- predict(fit_sb_sb, newdata = X_hte, type = "treatment.effect")

df_hte <- data.frame(
  x1 = X_hte[, 1],
  `CRP+CRP` = cate_crp_crp,
  `CRP+SB` = cate_crp_sb,
  `SB+CRP` = cate_sb_crp,
  `SB+SB` = cate_sb_sb,
  check.names = FALSE
)

# Melt for ggplot
df_hte_long <- reshape2::melt(df_hte, id.vars = "x1",
                               variable.name = "Combo", value.name = "CATE")

p_hte <- ggplot(df_hte_long, aes(x = x1, y = CATE, color = Combo)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  labs(title = "CATE: All Backend Combinations", x = "X1", y = "CATE") +
  theme_minimal() +
  theme(legend.position = "bottom")

print(p_hte)
```

---

## PS Variant Comparison: Logit vs Probit

```{r ps-variants}
# SB PS + CRP Outcome with Probit PS
bundle_probit <- build_causal_bundle(
  y = y, T = T, X = X,
  kernel = "normal", backend = "crp",
  PS = "probit",           # Probit instead of logit
  GPD = FALSE, Kmax = 5,
  mcmc = list(niter = 1000, nburnin = 200, nchains = 1)
)
fit_probit <- run_mcmc_bundle_manual(bundle_probit)

ate_probit <- coef(fit_probit)
print("\n=== PS MODEL VARIANT COMPARISON ===\n")
print("Logit PS + CRP Outcome ATE: ", ate_sb_crp, "\n")
print("Probit PS + CRP Outcome ATE:", ate_probit, "\n")
print("Difference:", abs(ate_sb_crp - ate_probit), "\n")
```

---

## Computational Efficiency

```{r efficiency}
print("\n=== COMPUTATIONAL TRADE-OFFS ===\n\n")

print("CRP + CRP:\n")
print("  Pros: Most flexible both PS and outcome\n")
print("  Cons: Slowest\n\n")

print("CRP + SB:\n")
print("  Pros: Flexible PS, efficient outcome\n")
print("  Cons: Medium speed\n\n")

print("SB + CRP:\n")
print("  Pros: Efficient PS, flexible outcome\n")
print("  Cons: Medium speed\n\n")

print("SB + SB:\n")
print("  Pros: Fastest, simplest\n")
print("  Cons: Least flexible\n")
```

---

## Recommendation Algorithm

```{r recommendation}
print("\n=== CHOOSING BACKENDS ===\n\n")

print("Step 1: Check data complexity\n")
print("  - Sample size?\n")
print("  - Treatment assignment complexity?\n")
print("  - Outcome distribution shape?\n\n")

print("Step 2: Fit models with different backends\n")
print("  - Check convergence (Rhat, ESS)\n")
print("  - Compare treatment effects\n\n")

print("Step 3: Select based on:\n")
print("  ✓ Computational budget\n")
print("  ✓ Convergence diagnostics\n")
print("  ✓ Stability across runs\n")
```

---

## Final Summary: All Vignettes

```{r summary}
print("\n=== 15-VIGNETTE ROADMAP ===\n\n")

print("v01: Introduction - Overview & three-phase workflow\n")
print("v02: Distributions - All 7 kernels + GPD\n")
print("v03: Basic workflow - Spec → Bundle → MCMC\n\n")

print("v04-v05: Unconditional DPmix\n")
print("  v04: CRP backend\n")
print("  v05: SB backend\n\n")

print("v06-v07: Unconditional DPmixGPD\n")
print("  v06: CRP + GPD tail\n")
print("  v07: SB + GPD tail\n\n")

print("v08-v09: Conditional DPmix\n")
print("  v08: CRP + covariates\n")
print("  v09: SB + covariates\n\n")

print("v10-v11: Conditional DPmixGPD\n")
print("  v10: CRP + covariates + GPD tail\n")
print("  v11: SB + covariates + GPD tail\n\n")

print("v12-v13: Causal Same Backend\n")
print("  v12: CRP for both PS & outcome\n")
print("  v13: SB for both PS & outcome\n\n")

print("v14-v15: Causal Different Backends\n")
print("  v14: CRP PS, varied outcome\n")
print("  v15: SB PS, varied outcome (YOU ARE HERE)\n\n")

print("✓ COMPLETE 15-VIGNETTE TUTORIAL SERIES\n")
```

---

## Key Takeaways: Entire Series

- **Flexibility**: DPmixGPD supports unconditional, conditional, and causal models
- **Backends**: CRP (adaptive) vs SB (fixed) for different needs
- **Tail Modeling**: GPD augmentation captures extreme values
- **PS Methods**: logit, probit, naive for robust causal inference
- **Treatment Effects**: ATE and CATE available across all models
- **Model Selection**: Use LogLik, diagnostics, and subject-matter knowledge

