---
title: "15. Causal Inference: Different Backends per Arm (GPD Augmentation)"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{15. Causal Different Backends SB}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  cache = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(dplyr)
library(tibble)
library(patchwork)
set.seed(123)
```

# Causal Inference: Different Backends per Arm (GPD Augmentation)

**Purpose**: Repeat the mixed-backend experiment but augment both arms with GPD tails to focus on extreme quantiles and CQTE behavior under different backend asymmetries.

---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- causal_alt_real500_p4_k2$y
T <- causal_alt_real500_p4_k2$T
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

ggplot(data.frame(y = y, x1 = X[, 1], T = as.factor(T)), aes(x = x1, y = y, color = T)) +
  geom_point(alpha = 0.5) +
  labs(title = "Outcome vs X1 by Treatment", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Causal Dataset Summary with Tails") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

---

## Backend + Tail Specifications

```{r bundles}
# Treated: SB, Control: CRP
bundle_sb_crp <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = c("lognormal", "lognormal"),
  backend = c("sb", "crp"),
  PS = "logit",
  GPD = c(TRUE, TRUE),
  J = c(5, 5),
  mcmc = list(niter = 1000, nburnin = 250, nchains = 2, thin = 1)
)

# Treated: CRP, Control: SB
bundle_crp_sb <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = c("lognormal", "lognormal"),
  backend = c("crp", "sb"),
  PS = "logit",
  GPD = c(TRUE, TRUE),
  J = c(5, 5),
  mcmc = list(niter = 1000, nburnin = 250, nchains = 2, thin = 1)
)
```

---

## MCMC Execution & Summary

```{r mcmc}
fit_sb_crp <- run_mcmc_causal(bundle_sb_crp)
fit_crp_sb <- run_mcmc_causal(bundle_crp_sb)

summary(fit_sb_crp)
summary(fit_crp_sb)
```

---

## Propensity Scores

```{r ps}
ps_vals <- fitted(fit_sb_crp, type = "ps")
df_ps <- data.frame(ps = ps_vals, T = as.factor(T))

ggplot(df_ps, aes(x = ps, fill = T)) +
  geom_density(alpha = 0.5) +
  scale_fill_manual(values = c("0" = "steelblue", "1" = "firebrick")) +
  labs(title = "Propensity Scores with GPD Mix", x = "PS", y = "Density") +
  theme_minimal()
```

---

## CQTE with Tails

```{r cqte}
X_ref <- matrix(colMeans(X), nrow = 1)
tau_grid <- seq(0.90, 0.99, length.out = 8)

pred_sb_crp <- predict(fit_sb_crp, newdata = X_ref, type = "quantile", p = tau_grid)
pred_crp_sb <- predict(fit_crp_sb, newdata = X_ref, type = "quantile", p = tau_grid)

qte_sb_crp <- as.numeric(pred_sb_crp$trt$fit - pred_sb_crp$con$fit)
qte_crp_sb <- as.numeric(pred_crp_sb$trt$fit - pred_crp_sb$con$fit)

qte_df <- tibble(
  tau = rep(tau_grid, 2),
  cqte = c(qte_sb_crp, qte_crp_sb),
  combo = rep(c("SB/CRP", "CRP/SB"), each = length(tau_grid))
)

ggplot(qte_df, aes(x = tau, y = cqte, color = combo)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Tail CQTE under GPD Augmentation", x = "τ", y = "Q1 − Q0") +
  theme_minimal()
```

---

## Density Tail Comparison

```{r density-tail}
y_grid <- seq(quantile(y, 0.7), max(y), length.out = 200)
pred_density_sb_crp <- predict(fit_sb_crp, newdata = X_ref, y = y_grid, type = "density")
pred_density_crp_sb <- predict(fit_crp_sb, newdata = X_ref, y = y_grid, type = "density")

density_sb_crp <- pred_density_sb_crp$con$fit[[1]] + pred_density_sb_crp$trt$fit[[1]]
density_crp_sb <- pred_density_crp_sb$con$fit[[1]] + pred_density_crp_sb$trt$fit[[1]]

df_density <- tibble(
  y = rep(y_grid, 2),
  density = c(density_sb_crp, density_crp_sb),
  combo = rep(c("SB/CRP", "CRP/SB"), each = length(y_grid))
)

ggplot(df_density, aes(x = y, y = density, color = combo)) +
  geom_line(linewidth = 1.2) +
  labs(title = "Tail Density Comparison (sum of arms)", y = "Density", x = "y") +
  theme_minimal()
```

---

## Takeaways

- GPD tails amplify differences between backend choices in the far right tail.
- CQTE at extreme τ (0.90-0.99) reveals how tail modeling interacts with stick-breaking vs CRP mixtures.
- Reuse the same PS diagnostics and `predict()`/`coef()` pipelines from earlier vignettes to maintain comparability across settings.
