---
title: "Backends, Kernels, and Workflow Map"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = '../Examples/legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(DPmixGPD)

# Ensure package is fully loaded for data access
if (!require("devtools", quietly = TRUE)) {
  library(devtools)
}
pkg_root <- rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
  devtools::load_all(path = pkg_root, quiet = TRUE)

# Precomputed results for fast website builds
source("_legacy-setup.R")
if (!exists("safe_plot", mode = "function")) {
  safe_plot <- function(x, ...) {
    obj <- tryCatch(plot(x, ...), error = function(e) e)
    if (inherits(obj, "ggplot")) {
      try(print(obj), silent = TRUE)
    } else if (is.list(obj)) {
      for (elt in obj) if (inherits(elt, "ggplot")) try(print(elt), silent = TRUE)
    }
    invisible(obj)
  }
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (basic, unconditional, conditional, and causal)..


# Big picture

DPmixGPD has two orthogonal dials you turn when building models:

- **Backend** (how the mixture weights / clustering are represented)
  - **CRP**: Chinese Restaurant Process representation.
  - **SB**: stick-breaking truncation with a fixed number of components.

- **Kernel / family** (what distribution models the bulk of the data)
  - Examples: normal, lognormal, gamma, inverse-Gaussian, Laplace, Cauchy, Amoroso, etc.

Optionally, you can also turn on:

- **GPD = TRUE/FALSE** to splice a Generalized Pareto tail beyond a threshold.

# What changes between CRP and SB?

Both backends target the same posterior over densities. The difference is representation:

- **CRP** learns a random number of occupied clusters within a finite `components` cap.
- **SB** uses the same finite `components` cap and learns stick-breaking weights.

Practical rule of thumb:

- **CRP** is convenient when you want adaptive complexity while still using a finite `components` cap.
- **SB** is convenient when you want predictable memory/time and easy vectorization.

# What does the workflow look like?

DPmixGPD uses a consistent build -> run -> summarize loop:

1. **Build a bundle** using `DPmixGPD::build_nimble_bundle()` (or causal builders if you are doing TE work).
2. **Run MCMC** using `run_mcmc_bundle_manual()`.
3. **Inspect and summarize** using `print()`, `summary()`, `plot()`.
4. **Predict** using `predict()`. For conditional (covariate) models, `fitted()` is also available.

```{r, cache=FALSE}
# Build
bundle <- DPmixGPD::build_nimble_bundle(
  y = rnorm(50),
  backend = "crp",
  kernel = "normal",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)

# Run
fit <- load_or_fit("quickstart-backends-and-workflow-fit", run_mcmc_bundle_manual(bundle))

# Summarize
print(fit)
summary(fit)
safe_plot(fit)
```

# Kernel support quick check

Use `kernel_support_table()` and the kernel registry helpers to confirm what is available.

```{r}
kernel_support_table()
```

# Where to go next

- **Available distributions**: see **QuickStart introduction** for the `d/p/q/r` functions and examples.
- **Basic build/compile/run**: see **QuickStart basic-model-compile-run**.
- **Unconditional / Conditional / Causal**: continue through ex06+.

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [basic-model-compile-run](basic-model-compile-run.html)
- Next: [ex05-unconditional-DPmix-CRP](../Examples/ex05-unconditional-DPmix-CRP.html)
- Workflow index: [Roadmap](../roadmap.html)
- Practical entry: [Cookbook](../cookbook.html)
<!-- workflow-nav:end -->
















