---
title: "Start Here"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = '../Examples/legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy


# Ensure package is fully loaded for data access
if (!require("devtools", quietly = TRUE)) {
  library(devtools)
}
pkg_root <- rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
  devtools::load_all(path = pkg_root, quiet = TRUE)

# Precomputed results for fast website builds
source("_legacy-setup.R")
if (!exists("safe_plot", mode = "function")) {
  safe_plot <- function(x, ...) {
    obj <- tryCatch(plot(x, ...), error = function(e) e)
    if (inherits(obj, "ggplot")) {
      try(print(obj), silent = TRUE)
    } else if (is.list(obj)) {
      for (elt in obj) if (inherits(elt, "ggplot")) try(print(elt), silent = TRUE)
    }
    invisible(obj)
  }
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (basic, unconditional, conditional, and causal)..


## Quick start

This vignette gives a minimal, fully working workflow for an **unconditional** model and a **conditional** model. Everything uses short MCMC runs so the vignette renders quickly.

---

## Unconditional Model (CRP, bulk-only)

```{r start-uncond-data}
data("nc_pos200_k3", package = "DPmixGPD")
y <- nc_pos200_k3$y
```

```{r start-uncond-bundle, cache=FALSE}
bundle_uncond <- DPmixGPD::build_nimble_bundle(
  y = y,
  backend = "crp",
  kernel = "gamma",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)
```

```{r start-uncond-fit, cache=FALSE}
fit_uncond <- load_or_fit("quickstart-start-here-fit_uncond", quiet_mcmc(run_mcmc_bundle_manual(bundle_uncond, show_progress = FALSE)))
summary(fit_uncond)
```

```{r start-uncond-predict, cache=FALSE}
pred_q <- predict(fit_uncond, type = "quantile", p = c(0.5, 0.9), interval = "credible")
head(pred_q$fit)
safe_plot(pred_q)
```

---

## Conditional Model (SB, bulk-only)

```{r start-cond-data}
data("nc_posX100_p3_k2", package = "DPmixGPD")
yc <- nc_posX100_p3_k2$y
X <- as.matrix(nc_posX100_p3_k2$X)
```

```{r start-cond-bundle, cache=FALSE}
bundle_cond <- DPmixGPD::build_nimble_bundle(
  y = yc,
  X = X,
  backend = "sb",
  kernel = "lognormal",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)
```

```{r start-cond-fit, cache=FALSE}
fit_cond <- load_or_fit("quickstart-start-here-fit_cond", quiet_mcmc(run_mcmc_bundle_manual(bundle_cond, show_progress = FALSE)))
summary(fit_cond)
```

```{r start-cond-predict, cache=FALSE}
x_new <- X[1:20, , drop = FALSE]
pred_mean <- predict(fit_cond, x = x_new, type = "mean", interval = "credible", nsim_mean = 200)
head(pred_mean$fit)
safe_plot(pred_mean)
```

---

## Useful S3 Methods

```{r start-s3}
params(fit_uncond)
safe_plot(fit_uncond, family = c("traceplot", "running"))
```

---

## Next Steps

- `QuickStart/introduction`: package overview and terminology
- `Examples/ex05`: full three-phase workflow (spec -> bundle -> MCMC)
- `Examples/ex06-ex13`: unconditional/conditional models with and without GPD
- `Examples/ex14-ex19`: causal workflows

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: none (start of workflow series)
- Next: [introduction](introduction.html)
- Workflow index: [Roadmap](../roadmap.html)
- Practical entry: [Examples](../Examples/index.html)
<!-- workflow-nav:end -->

















