---
title: "Normal"
pkgdown:
  slug: "kernels/kernel-normal"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\\VignetteIndexEntry{Normal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
source("_load_pkg.R")
```

## Normal {#v02-normal}


### Normal mixture kernel

A normal component is parameterized by \(\mu\in\mathbb{R}\) and \(\sigma>0\):
\[
f(y\mid\mu,\sigma)=\frac{1}{\sqrt{2\pi}\sigma}\exp\!\left(-\frac{(y-\mu)^2}{2\sigma^2}\right).
\]

A finite normal mixture with \(J\) components has density
\[
f(y)=\sum_{j=1}^J w_j\,\mathcal{N}(y\mid\mu_j,\sigma_j^2),
\qquad \sum_{j=1}^J w_j=1.
\]

**Parameter mapping (math \(\rightarrow\) code):** \(\mu_j\to\) `mean[j]`, \(\sigma_j\to\) `sd[j]`, \(w_j\to\) `w[j]`.

### Normal mixture with GPD tail

In the spliced version, the bulk distribution is the normal mixture below \(u\), and a GPD tail is attached above \(u\). The tail parameters are \(\sigma>0\) and \(\xi\), applied to exceedances \(x-u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD (mixture kernel)

```{r normal-mix}
grid <- seq(-4, 5, length.out = 400)
normal_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), mean = c(-1, 0.5, 2), sd = c(2, 0.6, 1.1)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), mean = c(-1.2, 0.3, 1.5), sd = c(0.9, 0.7, 1.0)),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), mean = c(-0.5, 2, 2.5), sd = c(0.7, 0.6, 1.2))
)

example <- normal_sets[[1]]
```

```{r normal-mix-d}
dNormMix(0, w = example$w, mean = example$mean, sd = example$sd)
dNormMix(0, w = example$w, mean = example$mean, sd = example$sd, log = TRUE)
```

```{r normal-mix-p}
pNormMix(0, w = example$w, mean = example$mean, sd = example$sd)
pNormMix(0, w = example$w, mean = example$mean, sd = example$sd, lower.tail = FALSE)
pNormMix(0, w = example$w, mean = example$mean, sd = example$sd, log.p = TRUE)
```

```{r normal-mix-q}
q_vec(qNormMix, c(0.25, 0.5, 0.75), w = example$w, mean = example$mean,
      sd = example$sd)
q_vec(qNormMix, c(0.25, 0.5, 0.75), w = example$w, mean = example$mean,
      sd = example$sd, lower.tail = FALSE)
q_vec(qNormMix, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      mean = example$mean, sd = example$sd, log.p = TRUE)
```

```{r normal-mix-r}
draw_many(rNormMix, list(w = example$w, mean = example$mean, sd = example$sd))
```

```{r normal-mix-plot}
df_norm <- do.call(rbind, lapply(normal_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dNormMix, list(w = ps$w, mean = ps$mean, sd = ps$sd)), label = ps$label)
}))

ggplot(df_norm, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Normal mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

Spliced density uses the same mixture for $x<u$ and attaches $f_{GPD}$ above $u$ with continuity. CDF combines mixture CDF up to $u$ and GPD exceedance beyond $u$; quantiles invert this spliced CDF; RNG draws bulk vs tail by the CDF mass at $u$.

```{r normal-gpd}
normal_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.4), mean = c(-1, 2), sd = c(0.5, 0.8), threshold = 1.8, tail_scale = 0.4, tail_shape = 0.25),
  list(label = "Mix B", w = c(0.5, 0.5), mean = c(0, 1), sd = c(0.6, 0.6), threshold = 1.5, tail_scale = 0.3, tail_shape = 0.2),
  list(label = "Mix C", w = c(0.7, 0.3), mean = c(0.5, 2.5), sd = c(0.4, 1.0), threshold = 2.0, tail_scale = 0.5, tail_shape = 0.15)
)

example <- normal_gpd_sets[[1]]
```

```{r normal-gpd-d}
dNormMixGpd(2, w = example$w, mean = example$mean, sd = example$sd, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r normal-gpd-p}
pNormMixGpd(2, w = example$w, mean = example$mean, sd = example$sd, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r normal-gpd-q}
q_vec(qNormMixGpd, c(0.5, 0.9), w = example$w, mean = example$mean, sd = example$sd, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r normal-gpd-r}
draw_many(rNormMixGpd, example)
```

```{r normal-gpd-plot}
df_norm_gpd <- do.call(rbind, lapply(normal_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dNormMixGpd, list(w = ps$w, mean = ps$mean, sd = ps$sd, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_norm_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Normal mixtures with GPD tail (different thresholds)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```


