---
title: "Legacy: Kernel reference: Amoroso"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Legacy: Kernel reference: Amoroso}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
library(DPmixGPD)
```

> **Legacy note:** This kernel reference is retained for historical context alongside the extended legacy materials.

The **Amoroso** kernel is parameterized by (`a`, `theta`, `alpha`, `beta`).

## Exported mixture helpers

- `dAmorosoMix(x, a, theta, alpha, beta, w, log = FALSE)`
- `pAmorosoMix(q, a, theta, alpha, beta, w, lower.tail = TRUE, log.p = FALSE)`
- `qAmorosoMix(p, a, theta, alpha, beta, w, lower.tail = TRUE, log.p = FALSE)`
- `rAmorosoMix(n, a, theta, alpha, beta, w)`

For the full catalog (including CRP utilities and GPD splicing), see:

- [Legacy distributions](legacy/v02-available-distributions.html)

## Using Amoroso in a model

```{r}
y <- abs(stats::rnorm(50)) + 0.1
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel  = "amoroso",
  GPD     = TRUE,
  components = 6
)
```


