---
title: "Legacy: Kernel reference: Cauchy"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Legacy: Kernel reference: Cauchy}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
library(DPmixGPD)
```

> **Legacy note:** This kernel reference is retained for historical context alongside the extended legacy materials.

The **Cauchy** kernel is parameterized by `location` and `scale`.

## Exported mixture helpers

- `dCauchyMix(x, location, scale, w, log = FALSE)`
- `pCauchyMix(q, location, scale, w, lower.tail = TRUE, log.p = FALSE)`
- `qCauchyMix(p, location, scale, w, lower.tail = TRUE, log.p = FALSE)`
- `rCauchyMix(n, location, scale, w)`

For the full catalog (including CRP utilities and GPD splicing), see:

- [Legacy distributions](legacy/v02-available-distributions.html)

## Using Cauchy in a model

```{r}
y <- abs(stats::rnorm(50)) + 0.1
bundle <- build_nimble_bundle(
  y = y,
  backend = "crp",
  kernel  = "cauchy",
  GPD     = FALSE,
  components = 6
)
```


