---
title: "Legacy: 2. Available Distributions and Related Functions"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Legacy: 2. Available Distributions and Related Functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  tidy = FALSE,
  width = 80
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
set.seed(42)
```

> **Legacy note:** This page is preserved for historical context and extended detail. It predates the streamlined official vignettes and may include longer runs or exploratory material.


```{r helpers}
# Helpers
q_vec <- function(fn, probs, ...) vapply(probs, function(p) fn(p, ...), numeric(1))
density_curve <- function(grid, fn, args) {
  vapply(grid, function(x) do.call(fn, c(list(x = x), args)), numeric(1))
}

draw_many <- function(fn, args, n_draws = 5) {
  if (!is.null(args$label)) args$label <- NULL
  vapply(seq_len(n_draws), function(i) do.call(fn, c(list(n = 1), args)), numeric(1))

}
```

## Mathematical definitions and conventions

This vignette documents the probability distributions exported by **DPmixGPD** and the corresponding `d/p/q/r` functions used throughout the package. For a continuous random variable \(Y\) with parameter scalars \(\boldsymbol{\theta}\), we use the standard quartet:

- **Density (PDF)**: \(f(y\mid\boldsymbol{\theta})\), implemented by `d*()`.
- **Distribution (CDF)**: \(F(y\mid\boldsymbol{\theta})=\Pr(Y\le y)\), implemented by `p*()`.
- **Quantile**: \(Q(p\mid\boldsymbol{\theta}) = \inf\{y: F(y\mid\boldsymbol{\theta})\ge p\}\), implemented by `q*()`.
- **Random generation**: \(Y = Q(U\mid\boldsymbol{\theta})\) with \(U\sim \text{Unif}(0,1)\), implemented by `r*()`.

All `d*()` functions optionally return log-densities via `log = TRUE`, i.e., \(\log f(y\mid\boldsymbol{\theta})\). This is numerically convenient because likelihoods multiply densities but add log-densities:
\[
\log\Big\{\prod_{i=1}^n f(y_i\mid\boldsymbol{\theta})\Big\}=\sum_{i=1}^n \log f(y_i\mid\boldsymbol{\theta}).
\]

All `p*()` functions support `lower.tail` and `log.p`:
\[
\Pr(Y>y)=1-F(y),\qquad \log F(y)\ \text{or}\ \log\{1-F(y)\}.
\]

The `d*()`, `p*()`, and `q*()` functions accept vector inputs for their first argument and evaluate elementwise; `r*()` supports `n > 1` (including `n = 0`).

```{r vectorized-dqrp}
w <- c(0.60, 0.40)
shape <- c(2.0, 5.0)
scale <- c(1.0, 2.0)

dGammaMix(seq(0.5, 2.0, length.out = 5), w = w, shape = shape, scale = scale)
qGammaMix(c(0.1, 0.5, 0.9), w = w, shape = shape, scale = scale)
rGammaMix(5, w = w, shape = shape, scale = scale)
```

### Finite mixtures

Several kernels are available as **finite mixtures**. Let \(f_j(\cdot\mid\boldsymbol{\theta}_j)\) denote the \(j\)th component density and let \(\boldsymbol{w}=(w_1,\ldots,w_J)\) be mixture weights with \(w_j\ge 0\) and \(\sum_{j=1}^J w_j=1\). The mixture density and CDF are

\[
f(y)=\sum_{j=1}^J w_j f_j(y\mid\boldsymbol{\theta}_j),
\qquad
F(y)=\sum_{j=1}^J w_j F_j(y\mid\boldsymbol{\theta}_j).
\]

Random generation proceeds by sampling a latent component index \(Z\sim\text{Categorical}(\boldsymbol{w})\), then sampling \(Y\mid Z=j\sim f_j(\cdot\mid\boldsymbol{\theta}_j)\). Mixture quantiles typically require numerical inversion of the mixture CDF.

### GPD tails and splicing

The **Generalized Pareto Distribution (GPD)** is used for tail modeling relative to a threshold \(u\). In the spliced kernels (the `*Gpd` functions), a bulk kernel is used below \(u\) and a GPD tail is attached above \(u\), with parameters \((u,\sigma,\xi)\). In code, these are represented by `threshold = u`, `tail_scale = \sigma`, and `tail_shape = \xi` (the standalone GPD functions use `scale` and `shape` for \(\sigma\) and \(\xi\)).




## Introduction with GPD kernel


### Generalized Pareto distribution (standalone tail)

The standalone GPD functions `dGpd`, `pGpd`, `qGpd`, and `rGpd` implement the distribution of \(X\) **above a threshold** \(u\). For \(x\ge u\), the density is

\[
f(x;u,\sigma,\xi)=\frac{1}{\sigma}\left(1+\xi\frac{x-u}{\sigma}\right)^{-1/\xi-1},
\quad \sigma>0,
\quad 1+\xi\frac{x-u}{\sigma}>0.
\]

The CDF is
\[
F(x)=1-\left(1+\xi\frac{x-u}{\sigma}\right)^{-1/\xi},
\quad x\ge u,
\]
and the quantile is
\[
Q(p)=u+\frac{\sigma}{\xi}\Big[(1-p)^{-\xi}-1\Big],\quad p\in(0,1),\ \xi\ne 0,
\]
with the exponential-limit case obtained as \(\xi\to 0\).

**Parameter mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `scale`, \(\xi\to\) `shape`. Log-densities are returned when `log = TRUE`.


We show the DPmixGPD kernels in pairs: bulk-only and GPD-augmented. A GPD tail above threshold $u$ uses

$$f_{GPD}(x; u, \sigma, \xi) = \frac{1}{\sigma} \left(1 + \xi \frac{x-u}{\sigma}\right)^{-1/\xi - 1} \ \text{for } x \ge u.$$

The standalone GPD tail distribution is available via `dGpd`, `pGpd`, `qGpd`, and `rGpd` functions.


```{r gpd-density}
dGpd(1.8, threshold = 1.5, scale = 0.5, shape = 0.2)
dGpd(1.8, threshold = 1.5, scale = 0.5, shape = 0.2, log = TRUE)
```

```{r gpd-cdf}
pGpd(1.8, threshold = 1.5, scale = 0.5, shape = 0.2)
pGpd(1.8, threshold = 1.5, scale = 0.5, shape = 0.2, lower.tail = FALSE)
pGpd(1.8, threshold = 1.5, scale = 0.5, shape = 0.2, log.p = TRUE)
```

```{r gpd-quantile}
q_vec(qGpd, c(0.25, 0.5, 0.75), threshold = 1.5, scale = 0.5, shape = 0.2)
q_vec(qGpd, c(0.25, 0.5, 0.75), threshold = 1.5, scale = 0.5, shape = 0.2,
      lower.tail = FALSE)
q_vec(qGpd, c(log(0.25), log(0.5), log(0.75)), threshold = 1.5, scale = 0.5,
      shape = 0.2, log.p = TRUE)
```

```{r gpd-random}
draw_many(rGpd, list(threshold = 1.5, scale = 0.5, shape = 0.2))
```

```{r gpd-plot}
grid <- seq(-4, 15, length.out = 500)
gpd_sets <- list(
  list(label = "GPD A", threshold = 1.5, tail_scale = 0.5, tail_shape = 0.2),
  list(label = "GPD B", threshold = 2.0, tail_scale = 0.6, tail_shape = 0.15),
  list(label = "GPD C", threshold = 1.0, tail_scale = 0.4, tail_shape = 0.25)
)

df_gpd <- do.call(rbind, lapply(gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dGpd, list(threshold = ps$threshold, scale = ps$tail_scale, shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Standalone GPD tails", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

Each subsection below defines the available $d/p/q/r$ functions, prints example outputs for fixed parameters, and overlays density curves for **three parameter sets** with clear legends. The same parameter sets are reused within the bulk and GPD variants of a section.

## Normal


### Normal mixture kernel

A normal component is parameterized by \(\mu\in\mathbb{R}\) and \(\sigma>0\):
\[
f(y\mid\mu,\sigma)=\frac{1}{\sqrt{2\pi}\sigma}\exp\!\left(-\frac{(y-\mu)^2}{2\sigma^2}\right).
\]

A finite normal mixture with \(J\) components has density
\[
f(y)=\sum_{j=1}^J w_j\,\mathcal{N}(y\mid\mu_j,\sigma_j^2),
\qquad \sum_{j=1}^J w_j=1.
\]

**Parameter mapping (math \(\rightarrow\) code):** \(\mu_j\to\) `mean[j]`, \(\sigma_j\to\) `sd[j]`, \(w_j\to\) `w[j]`.

### Normal mixture with GPD tail

In the spliced version, the bulk distribution is the normal mixture below \(u\), and a GPD tail is attached above \(u\). The tail parameters are \(\sigma>0\) and \(\xi\), applied to exceedances \(x-u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD (mixture kernel)

```{r normal-mix}
grid <- seq(-4, 5, length.out = 400)
normal_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), mean = c(-1, 0.5, 2), sd = c(2, 0.6, 1.1)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), mean = c(-1.2, 0.3, 1.5), sd = c(0.9, 0.7, 1.0)),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), mean = c(-0.5, 2, 2.5), sd = c(0.7, 0.6, 1.2))
)

example <- normal_sets[[1]]
```

```{r normal-mix-d}
dNormMix(0, w = example$w, mean = example$mean, sd = example$sd)
dNormMix(0, w = example$w, mean = example$mean, sd = example$sd, log = TRUE)
```

```{r normal-mix-p}
pNormMix(0, w = example$w, mean = example$mean, sd = example$sd)
pNormMix(0, w = example$w, mean = example$mean, sd = example$sd, lower.tail = FALSE)
pNormMix(0, w = example$w, mean = example$mean, sd = example$sd, log.p = TRUE)
```

```{r normal-mix-q}
q_vec(qNormMix, c(0.25, 0.5, 0.75), w = example$w, mean = example$mean,
      sd = example$sd)
q_vec(qNormMix, c(0.25, 0.5, 0.75), w = example$w, mean = example$mean,
      sd = example$sd, lower.tail = FALSE)
q_vec(qNormMix, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      mean = example$mean, sd = example$sd, log.p = TRUE)
```

```{r normal-mix-r}
draw_many(rNormMix, list(w = example$w, mean = example$mean, sd = example$sd))
```

```{r normal-mix-plot}
df_norm <- do.call(rbind, lapply(normal_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dNormMix, list(w = ps$w, mean = ps$mean, sd = ps$sd)), label = ps$label)
}))

ggplot(df_norm, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Normal mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

Spliced density uses the same mixture for $x<u$ and attaches $f_{GPD}$ above $u$ with continuity. CDF combines mixture CDF up to $u$ and GPD exceedance beyond $u$; quantiles invert this spliced CDF; RNG draws bulk vs tail by the CDF mass at $u$.

```{r normal-gpd}
normal_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.4), mean = c(-1, 2), sd = c(0.5, 0.8), threshold = 1.8, tail_scale = 0.4, tail_shape = 0.25),
  list(label = "Mix B", w = c(0.5, 0.5), mean = c(0, 1), sd = c(0.6, 0.6), threshold = 1.5, tail_scale = 0.3, tail_shape = 0.2),
  list(label = "Mix C", w = c(0.7, 0.3), mean = c(0.5, 2.5), sd = c(0.4, 1.0), threshold = 2.0, tail_scale = 0.5, tail_shape = 0.15)
)

example <- normal_gpd_sets[[1]]
```

```{r normal-gpd-d}
dNormMixGpd(2, w = example$w, mean = example$mean, sd = example$sd, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r normal-gpd-p}
pNormMixGpd(2, w = example$w, mean = example$mean, sd = example$sd, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r normal-gpd-q}
q_vec(qNormMixGpd, c(0.5, 0.9), w = example$w, mean = example$mean, sd = example$sd, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r normal-gpd-r}
draw_many(rNormMixGpd, example)
```

```{r normal-gpd-plot}
df_norm_gpd <- do.call(rbind, lapply(normal_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dNormMixGpd, list(w = ps$w, mean = ps$mean, sd = ps$sd, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_norm_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Normal mixtures with GPD tail (different thresholds)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```


## Gamma


### Gamma mixture kernel

A gamma component with shape \(\alpha>0\) and scale \(\beta>0\) has density
\[
f(y\mid \alpha,\beta)
=
\frac{1}{\Gamma(\alpha)\,\beta^\alpha}\,y^{\alpha-1}\exp\!\left(-\frac{y}{\beta}\right),
\qquad y>0.
\]

A finite gamma mixture with \(J\) components is
\[
f(y)=\sum_{j=1}^J w_j\,f_j(y\mid \alpha_j,\beta_j),
\qquad w_j\ge 0,\ \sum_{j=1}^J w_j=1.
\]

**Parameter mapping (math \(\rightarrow\) code):** \(\alpha\to\) `shape`, \(\beta\to\) `scale`, and weights \(w_j\to\) `w[j]`.

The `*MixGpd` variant uses the same splicing idea: gamma mixture below \(u\) and a GPD tail above \(u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD (mixture kernel)

```{r gamma-mix}
grid <- seq(0, 10, length.out = 400)
gamma_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), shape = c(2.0, 5.0, 9.0), scale = c(1.0, 0.6, 0.3)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), shape = c(1.5, 4.0, 7.0), scale = c(1.2, 0.7, 0.35)),
  list(label = "Mix C", w = c(0.4, 0.3, 0.3), shape = c(1.2, 3.5, 6.0), scale = c(1.4, 0.75, 0.4))
)
example <- gamma_sets[[1]]

dens <- do.call(rbind, lapply(gamma_sets, function(s) {
  data.frame(
    x = grid,
    density = dGammaMix(grid, w = s$w, shape = s$shape, scale = s$scale),
    label = s$label
  )
}))
```

```{r gamma-mix-d}
dGammaMix(2, w = example$w, shape = example$shape, scale = example$scale)
dGammaMix(2, w = example$w, shape = example$shape, scale = example$scale, log = TRUE)
```

```{r gamma-mix-p}
pGammaMix(2, w = example$w, shape = example$shape, scale = example$scale)
pGammaMix(2, w = example$w, shape = example$shape, scale = example$scale, lower.tail = FALSE)
pGammaMix(2, w = example$w, shape = example$shape, scale = example$scale, log.p = TRUE)
```

```{r gamma-mix-q}
qGammaMix(0.95, w = example$w, shape = example$shape, scale = example$scale)
qGammaMix(0.95, w = example$w, shape = example$shape, scale = example$scale, lower.tail = FALSE)
```

```{r gamma-mix-r}
df_gamma <- do.call(rbind, lapply(gamma_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dGammaMix, list(w = ps$w, shape = ps$shape, scale = ps$scale)), label = ps$label)
}))

ggplot(df_norm, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Gamma mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### Gamma mixture with GPD tail

```{r gamma-gpd}
u <- 6
tail_scale <- 1.0
tail_shape <- 0.2
```

```{r gamma-gpd-d}
dGammaMixGpd(6.5, w = example$w, shape = example$shape, scale = example$scale,
             threshold = u, tail_scale = tail_scale, tail_shape = tail_shape)
dGammaMixGpd(6.5, w = example$w, shape = example$shape, scale = example$scale,
             threshold = u, tail_scale = tail_scale, tail_shape = tail_shape, log = TRUE)
```

```{r gamma-gpd-p}
pGammaMixGpd(6.5, w = example$w, shape = example$shape, scale = example$scale,
             threshold = u, tail_scale = tail_scale, tail_shape = tail_shape)
pGammaMixGpd(6.5, w = example$w, shape = example$shape, scale = example$scale,
             threshold = u, tail_scale = tail_scale, tail_shape = tail_shape, lower.tail = FALSE)
```

```{r gamma-gpd-q}
qGammaMixGpd(0.95, w = example$w, shape = example$shape, scale = example$scale,
             threshold = u, tail_scale = tail_scale, tail_shape = tail_shape)
```

```{r gamma-gpd-r}
gamma_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.4), shape = c(2.0, 5.0), scale = c(1.0, 0.6), threshold = 6.0, tail_scale = 1.0, tail_shape = 0.2),
  list(label = "Mix B", w = c(0.5, 0.5), shape = c(1.5, 4.0), scale = c(1.2, 0.7), threshold = 5.5, tail_scale = 0.8, tail_shape = 0.15),
  list(label = "Mix C", w = c(0.7, 0.3), shape = c(1.2, 3.5), scale = c(1.4, 0.75), threshold = 6.5, tail_scale = 1.2, tail_shape = 0.25)
)



df_gamma_gpd <- do.call(rbind, lapply(gamma_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dGammaMixGpd, list(w = ps$w, shape = ps$shape, scale = ps$scale, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_norm_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Gamma mixtures with GPD tail (different thresholds)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```
## Lognormal


### Lognormal mixture kernel

A lognormal component is defined by \(\log Y \sim \mathcal{N}(\mu,\sigma^2)\), i.e.,
\[
f(y\mid\mu,\sigma)=\frac{1}{y\,\sigma\sqrt{2\pi}}\exp\!\left(-\frac{(\log y-\mu)^2}{2\sigma^2}\right),\quad y>0.
\]

A finite lognormal mixture has density
\[
f(y)=\sum_{j=1}^J w_j\,\text{Lognormal}(y\mid\mu_j,\sigma_j^2),\quad y>0.
\]

**Parameter mapping (math \(\rightarrow\) code):** \(\mu_j\to\) `meanlog[j]`, \(\sigma_j\to\) `sdlog[j]`, \(w_j\to\) `w[j]`.

### Lognormal mixture with GPD tail

The `*MixGpd` variant uses the same splicing idea: lognormal mixture below \(u\), GPD tail above \(u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD (mixture kernel)

```{r lognormal-mix}
grid <- seq(0, 8, length.out = 400)
logn_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), meanlog = c(0.0, 0.3, 0.6), sdlog = c(0.4, 0.5, 0.6)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), meanlog = c(0.1, 0.4, 0.7), sdlog = c(0.35, 0.45, 0.55)),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), meanlog = c(0.2, 0.5, 2), sdlog = c(0.3, 0.4, 0.5))
)

example <- logn_sets[[1]]
```

```{r lognormal-mix-d}
dLognormalMix(1, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog)
dLognormalMix(1, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, log = TRUE)
```

```{r lognormal-mix-p}
pLognormalMix(1, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog)
pLognormalMix(1, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, lower.tail = FALSE)
pLognormalMix(1, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, log.p = TRUE)
```

```{r lognormal-mix-q}
q_vec(qLognormalMix, c(0.25, 0.5, 0.75), w = example$w,
      meanlog = example$meanlog, sdlog = example$sdlog)
q_vec(qLognormalMix, c(0.25, 0.5, 0.75), w = example$w,
      meanlog = example$meanlog, sdlog = example$sdlog, lower.tail = FALSE)
q_vec(qLognormalMix, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      meanlog = example$meanlog, sdlog = example$sdlog, log.p = TRUE)
```

```{r lognormal-mix-r}
draw_many(rLognormalMix, list(w = example$w, meanlog = example$meanlog, sdlog = example$sdlog))
```

```{r lognormal-mix-plot}
df_logn <- do.call(rbind, lapply(logn_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dLognormalMix, list(w = ps$w, meanlog = ps$meanlog, sdlog = ps$sdlog)), label = ps$label)
}))

ggplot(df_logn, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Lognormal mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

```{r lognormal-gpd}
logn_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.4), meanlog = c(0, 1), sdlog = c(0.3, 0.5), threshold = 2.5, tail_scale = 0.5, tail_shape = 0.2),
  list(label = "Mix B", w = c(0.5, 0.5), meanlog = c(0.5, 1.2), sdlog = c(0.4, 0.4), threshold = 2.0, tail_scale = 0.4, tail_shape = 0.15),
  list(label = "Mix C", w = c(0.7, 0.3), meanlog = c(0.8, 1.5), sdlog = c(0.25, 0.6), threshold = 3.0, tail_scale = 0.6, tail_shape = 0.18)
)

example <- logn_gpd_sets[[1]]
```

```{r lognormal-gpd-d}
dLognormalMixGpd(2.5, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r lognormal-gpd-p}
pLognormalMixGpd(2.5, w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r lognormal-gpd-q}
q_vec(qLognormalMixGpd, c(0.5, 0.9), w = example$w, meanlog = example$meanlog, sdlog = example$sdlog, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r lognormal-gpd-r}
draw_many(rLognormalMixGpd, example)
```

```{r lognormal-gpd-plot}
df_logn_gpd <- do.call(rbind, lapply(logn_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dLognormalMixGpd, list(w = ps$w, meanlog = ps$meanlog, sdlog = ps$sdlog, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_logn_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Lognormal mixtures with GPD tail (different thresholds)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

## Laplace


### Laplace mixture kernel

A Laplace (double-exponential) component with location \(\ell\in\mathbb{R}\) and scale \(b>0\) has density
\[
f(y\mid \ell,b)=\frac{1}{2b}\exp\!\left(-\frac{|y-\ell|}{b}\right),\quad y\in\mathbb{R}.
\]

A finite Laplace mixture is
\[
f(y)=\sum_{j=1}^J w_j\,\text{Laplace}(y\mid \ell_j,b_j).
\]

**Parameter mapping (math \(\rightarrow\) code):** \(\ell_j\to\) `location[j]`, \(b_j\to\) `scale[j]`, \(w_j\to\) `w[j]`.

### Laplace mixture with GPD tail

The spliced kernel uses the Laplace mixture below \(u\) and a GPD tail above \(u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD (mixture kernel)

Laplace mixture density $f(x)=\sum_j w_j\,\text{Laplace}(x\mid \ell_j, b_j)$.

```{r laplace-mix}
grid <- seq(-4, 4, length.out = 400)
lap_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), location = c(0.0, 1.0, -2), scale = c(1.0, 0.9, 1.1)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), location = c(0.2, 1.2, -0.5), scale = c(0.9, 2, 1.0)),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), location = c(-0.2, 2, -1.0), scale = c(1.1, 0.95, 1.05))
)

example <- lap_sets[[1]]
```

```{r laplace-mix-d}
dLaplaceMix(0, w = example$w, location = example$location, scale = example$scale)
dLaplaceMix(0, w = example$w, location = example$location, scale = example$scale, log = TRUE)
```

```{r laplace-mix-p}
pLaplaceMix(0, w = example$w, location = example$location, scale = example$scale)
pLaplaceMix(0, w = example$w, location = example$location, scale = example$scale, lower.tail = FALSE)
pLaplaceMix(0, w = example$w, location = example$location, scale = example$scale, log.p = TRUE)
```

```{r laplace-mix-q}
q_vec(qLaplaceMix, c(0.25, 0.5, 0.75), w = example$w,
      location = example$location, scale = example$scale)
q_vec(qLaplaceMix, c(0.25, 0.5, 0.75), w = example$w,
      location = example$location, scale = example$scale, lower.tail = FALSE)
q_vec(qLaplaceMix, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      location = example$location, scale = example$scale, log.p = TRUE)
```

```{r laplace-mix-r}
draw_many(rLaplaceMix, list(w = example$w, location = example$location, scale = example$scale))
```

```{r laplace-mix-plot}
df_lap <- do.call(rbind, lapply(lap_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dLaplaceMix, list(w = ps$w, location = ps$location, scale = ps$scale)), label = ps$label)
}))

ggplot(df_lap, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Laplace mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

```{r laplace-gpd}
lap_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.4), location = c(-0.5, 1), scale = c(0.4, 0.6), threshold = 1.5, tail_scale = 0.35, tail_shape = 0.3),
  list(label = "Mix B", w = c(0.5, 0.5), location = c(0, 0.8), scale = c(0.5, 0.5), threshold = 1.2, tail_scale = 0.3, tail_shape = 0.25),
  list(label = "Mix C", w = c(0.7, 0.3), location = c(0.2, 1.5), scale = c(0.35, 0.7), threshold = 1.8, tail_scale = 0.4, tail_shape = 0.22)
)

example <- lap_gpd_sets[[1]]
```

```{r laplace-gpd-d}
dLaplaceMixGpd(1, w = example$w, location = example$location, scale = example$scale, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r laplace-gpd-p}
pLaplaceMixGpd(1, w = example$w, location = example$location, scale = example$scale, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r laplace-gpd-q}
q_vec(qLaplaceMixGpd, c(0.5, 0.9), w = example$w, location = example$location, scale = example$scale, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r laplace-gpd-r}
draw_many(rLaplaceMixGpd, example)
```

```{r laplace-gpd-plot}
df_lap_gpd <- do.call(rbind, lapply(lap_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dLaplaceMixGpd, list(w = ps$w, location = ps$location, scale = ps$scale, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_lap_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Laplace mixtures with GPD tail (different thresholds)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

## Inverse Gaussian (mixture kernel)


### Inverse Gaussian mixture kernel

An inverse Gaussian component with mean \(\mu>0\) and shape \(\lambda>0\) has density
\[
f(y\mid \mu,\lambda)
=
\left(\frac{\lambda}{2\pi y^3}\right)^{1/2}
\exp\!\left(-\frac{\lambda (y-\mu)^2}{2\mu^2 y}\right),
\quad y>0.
\]

A finite inverse Gaussian mixture is
\[
f(y)=\sum_{j=1}^J w_j\,\text{IG}(y\mid \mu_j,\lambda_j),\quad y>0.
\]

**Parameter mapping (math \(\rightarrow\) code):** \(\mu_j\to\) `mean[j]`, \(\lambda_j\to\) `shape[j]`, \(w_j\to\) `w[j]`.

### Inverse Gaussian mixture with GPD tail

The spliced variant attaches a GPD tail above \(u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD

```{r invgauss-mix}
grid <- seq(0.1, 6, length.out = 400)
ig_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), mean = c(1.0, 1.5, 2.0), shape = c(2.0, 3.0, 4.0)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), mean = c(1.1, 1.6, 2.2), shape = c(2.2, 3.2, 4.2)),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), mean = c(0.9, 1.4, 2.1), shape = c(1.8, 2.8, 3.8))
)

example <- ig_sets[[1]]
```

```{r invgauss-mix-d}
dInvGaussMix(1, w = example$w, mean = example$mean, shape = example$shape)
dInvGaussMix(1.5, w = example$w, mean = example$mean, shape = example$shape, log = TRUE)
```

```{r invgauss-mix-p}
pInvGaussMix(1.5, w = example$w, mean = example$mean, shape = example$shape)
pInvGaussMix(1.5, w = example$w, mean = example$mean, shape = example$shape, lower.tail = FALSE)
pInvGaussMix(1.5, w = example$w, mean = example$mean, shape = example$shape, log.p = TRUE)
```

```{r invgauss-mix-q}
q_vec(qInvGaussMix, c(0.25, 0.5, 0.75), w = example$w, mean = example$mean,
      shape = example$shape)
q_vec(qInvGaussMix, c(0.25, 0.5, 0.75), w = example$w, mean = example$mean,
      shape = example$shape, lower.tail = FALSE)
q_vec(qInvGaussMix, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      mean = example$mean, shape = example$shape, log.p = TRUE)
```

```{r invgauss-mix-r}
draw_many(rInvGaussMix, list(w = example$w, mean = example$mean, shape = example$shape))
```

```{r invgauss-mix-plot}
df_ig <- do.call(rbind, lapply(ig_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dInvGaussMix, list(w = ps$w, mean = ps$mean, shape = ps$shape)), label = ps$label)
}))

ggplot(df_ig, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Inverse Gaussian mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

```{r invgauss-gpd}
ig_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.4), mean = c(1, 2.5), shape = c(2, 3), threshold = 2.5, tail_scale = 0.5, tail_shape = 0.25),
  list(label = "Mix B", w = c(0.5, 0.5), mean = c(1.5, 2), shape = c(2.5, 2.5), threshold = 2.0, tail_scale = 0.4, tail_shape = 0.2),
  list(label = "Mix C", w = c(0.7, 0.3), mean = c(1.2, 3), shape = c(3, 2), threshold = 3.0, tail_scale = 0.6, tail_shape = 0.18)
)

example <- ig_gpd_sets[[1]]
```

```{r invgauss-gpd-d}
dInvGaussMixGpd(2.5, w = example$w, mean = example$mean, shape = example$shape, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r invgauss-gpd-p}
pInvGaussMixGpd(2.5, w = example$w, mean = example$mean, shape = example$shape, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r invgauss-gpd-q}
q_vec(qInvGaussMixGpd, c(0.5, 0.9), w = example$w, mean = example$mean, shape = example$shape, threshold = example$threshold, tail_scale = example$tail_scale, tail_shape = example$tail_shape)
```

```{r invgauss-gpd-r}
draw_many(rInvGaussMixGpd, example)
```

```{r invgauss-gpd-plot}
df_ig_gpd <- do.call(rbind, lapply(ig_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dInvGaussMixGpd, list(w = ps$w, mean = ps$mean, shape = ps$shape, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_ig_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Inverse Gaussian mixtures with GPD tail", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

## Amoroso (bulk kernel and mixture)


### Amoroso mixture kernel

An Amoroso component with location \(a\in\mathbb{R}\), scale \(\theta\neq 0\), and shapes \(\alpha>0\), \(\beta>0\) has density (for \(y>a\))
\[
f(y\mid a,\theta,\alpha,\beta)
=
\frac{1}{\Gamma(\alpha)}
\left|\frac{\beta}{\theta}\right|
\left(\frac{y-a}{\theta}\right)^{\alpha\beta-1}
\exp\!\left[-\left(\frac{y-a}{\theta}\right)^{\beta}\right].
\]

A finite Amoroso mixture is
\[
f(y)=\sum_{j=1}^J w_j\,f_j(y\mid a_j,\theta_j,\alpha_j,\beta_j).
\]

**Parameter mapping (math \(\rightarrow\) code):** \(a_j\to\) `loc[j]`, \(\theta_j\to\) `scale[j]`, \(\alpha_j\to\) `shape1[j]`, \(\beta_j\to\) `shape2[j]`, \(w_j\to\) `w[j]`.

### Amoroso mixture with GPD tail

The spliced kernel attaches a GPD tail above \(u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD

```{r amoroso-mix}
grid <- seq(0.01, 6, length.out = 400)
amor_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), loc = c(0.5, 0.5, 0.5), scale = c(1.0, 1.3, 1.6), shape1 = c(2.5, 3.0, 4.0), shape2 = c(1.2, 1.2, 1.2)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), loc = c(0.4, 0.6, 0.6), scale = c(1.1, 1.2, 1.5), shape1 = c(2.2, 2.8, 3.8), shape2 = c(1.1, 1.2, 1.3)),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), loc = c(0.6, 0.6, 0.6), scale = c(0.9, 1.1, 1.4), shape1 = c(2.0, 2.6, 3.5), shape2 = c(1.0, 1.1, 1.2))
)

example <- amor_sets[[1]]
```

```{r amoroso-mix-d}
dAmorosoMix(2, w = example$w, loc = example$loc, scale = example$scale,
            shape1 = example$shape1, shape2 = example$shape2)
dAmorosoMix(2, w = example$w, loc = example$loc, scale = example$scale,
            shape1 = example$shape1, shape2 = example$shape2, log = TRUE)
```

```{r amoroso-mix-p}
pAmorosoMix(2, w = example$w, loc = example$loc, scale = example$scale,
            shape1 = example$shape1, shape2 = example$shape2)
pAmorosoMix(2, w = example$w, loc = example$loc, scale = example$scale,
            shape1 = example$shape1, shape2 = example$shape2, lower.tail = FALSE)
pAmorosoMix(2, w = example$w, loc = example$loc, scale = example$scale,
            shape1 = example$shape1, shape2 = example$shape2, log.p = TRUE)
```

```{r amoroso-mix-q}
q_vec(qAmorosoMix, c(0.25, 0.5, 0.75), w = example$w, loc = example$loc,
      scale = example$scale, shape1 = example$shape1, shape2 = example$shape2)
q_vec(qAmorosoMix, c(0.25, 0.5, 0.75), w = example$w, loc = example$loc,
      scale = example$scale, shape1 = example$shape1, shape2 = example$shape2,
      lower.tail = FALSE)
q_vec(qAmorosoMix, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      loc = example$loc, scale = example$scale, shape1 = example$shape1,
      shape2 = example$shape2, log.p = TRUE)
```

```{r amoroso-mix-r}
draw_many(rAmorosoMix, list(w = example$w, loc = example$loc,
                            scale = example$scale, shape1 = example$shape1,
                            shape2 = example$shape2))
```

```{r amoroso-mix-plot}
df_amor <- do.call(rbind, lapply(amor_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dAmorosoMix, list(w = ps$w, loc = ps$loc, scale = ps$scale, shape1 = ps$shape1, shape2 = ps$shape2)), label = ps$label)
}))

ggplot(df_amor, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Amoroso mixtures (bulk)", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

```{r amoroso-gpd}
amor_gpd_sets <- list(
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), loc = c(0.5, 0.5, 0.5), scale = c(1.0, 1.3, 1.6), shape1 = c(2.5, 3.0, 4.0), shape2 = c(1.2, 1.2, 1.2), threshold = 2.8, tail_scale = 0.4, tail_shape = 0.2),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), loc = c(0.4, 0.6, 0.6), scale = c(1.1, 1.2, 1.5), shape1 = c(2.2, 2.8, 3.8), shape2 = c(1.1, 1.2, 1.3), threshold = 3.0, tail_scale = 0.35, tail_shape = 0.18),
  list(label = "Mix C", w = c(0.4, 0.35, 0.25), loc = c(0.6, 0.6, 0.6), scale = c(0.9, 1.1, 1.4), shape1 = c(2.0, 2.6, 3.5), shape2 = c(1.0, 1.1, 1.2), threshold = 2.5, tail_scale = 0.45, tail_shape = 0.22)
)
example <- amor_gpd_sets[[1]]
```

```{r amoroso-gpd-d}
dAmorosoMixGpd(3, w = example$w, loc = example$loc, scale = example$scale,
               shape1 = example$shape1, shape2 = example$shape2,
               threshold = example$threshold, tail_scale = example$tail_scale,
               tail_shape = example$tail_shape)
dAmorosoMixGpd(3, w = example$w, loc = example$loc, scale = example$scale,
               shape1 = example$shape1, shape2 = example$shape2,
               threshold = example$threshold, tail_scale = example$tail_scale,
               tail_shape = example$tail_shape, log = TRUE)
```

```{r amoroso-gpd-p}
pAmorosoMixGpd(3, w = example$w, loc = example$loc, scale = example$scale,
               shape1 = example$shape1, shape2 = example$shape2,
               threshold = example$threshold, tail_scale = example$tail_scale,
               tail_shape = example$tail_shape)
pAmorosoMixGpd(3, w = example$w, loc = example$loc, scale = example$scale,
               shape1 = example$shape1, shape2 = example$shape2,
               threshold = example$threshold, tail_scale = example$tail_scale,
               tail_shape = example$tail_shape, lower.tail = FALSE)
pAmorosoMixGpd(3, w = example$w, loc = example$loc, scale = example$scale,
               shape1 = example$shape1, shape2 = example$shape2,
               threshold = example$threshold, tail_scale = example$tail_scale,
               tail_shape = example$tail_shape, log.p = TRUE)
```

```{r amoroso-gpd-q}
q_vec(qAmorosoMixGpd, c(0.25, 0.5, 0.75), w = example$w, loc = example$loc,
      scale = example$scale, shape1 = example$shape1, shape2 = example$shape2,
      threshold = example$threshold, tail_scale = example$tail_scale,
      tail_shape = example$tail_shape)
q_vec(qAmorosoMixGpd, c(0.25, 0.5, 0.75), w = example$w, loc = example$loc,
      scale = example$scale, shape1 = example$shape1, shape2 = example$shape2,
      threshold = example$threshold, tail_scale = example$tail_scale,
      tail_shape = example$tail_shape, lower.tail = FALSE)
q_vec(qAmorosoMixGpd, c(log(0.25), log(0.5), log(0.75)), w = example$w,
      loc = example$loc, scale = example$scale, shape1 = example$shape1,
      shape2 = example$shape2, threshold = example$threshold,
      tail_scale = example$tail_scale, tail_shape = example$tail_shape,
      log.p = TRUE)
```

```{r amoroso-gpd-r}
draw_many(rAmorosoMixGpd, example)
```

```{r amoroso-gpd-plot}
df_amor_gpd <- do.call(rbind, lapply(amor_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dAmorosoMixGpd, list(w = ps$w, loc = ps$loc, scale = ps$scale, shape1 = ps$shape1, shape2 = ps$shape2, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_amor_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Amoroso mixtures with GPD tail", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

## Inverse Gaussian (base kernel)


### Inverse Gaussian base kernel

This section documents the same inverse Gaussian density as above, but for a **single** component rather than a mixture:
\[
f(y\mid \mu,\lambda)
=
\left(\frac{\lambda}{2\pi y^3}\right)^{1/2}
\exp\!\left(-\frac{\lambda (y-\mu)^2}{2\mu^2 y}\right),
\quad y>0.
\]

**Parameter mapping (math \(\rightarrow\) code):** \(\mu\to\) `mean`, \(\lambda\to\) `shape`.

### Inverse Gaussian base with GPD tail

The `dInvGaussGpd`, `pInvGaussGpd`, `qInvGaussGpd`, and `rInvGaussGpd` functions splice the base inverse Gaussian below \(u\) with a GPD tail above \(u\).

**Tail mapping (math \(\rightarrow\) code):** \(u\to\) `threshold`, \(\sigma\to\) `tail_scale`, \(\xi\to\) `tail_shape`.


### Without GPD

```{r invgauss-base}
grid <- seq(0.1, 6, length.out = 400)
ig_base_sets <- list(
  list(label = "Base A", mean = 1.2, shape = 3.0),
  list(label = "Base B", mean = 1.5, shape = 4.0),
  list(label = "Base C", mean = 1.0, shape = 2.5)
)

example <- ig_base_sets[[1]]
```

```{r invgauss-base-d}
dInvGauss(1, mean = example$mean, shape = example$shape)
dInvGauss(1, mean = example$mean, shape = example$shape, log = TRUE)
```

```{r invgauss-base-p}
pInvGauss(1, mean = example$mean, shape = example$shape)
pInvGauss(1, mean = example$mean, shape = example$shape, lower.tail = FALSE)
pInvGauss(1, mean = example$mean, shape = example$shape, log.p = TRUE)
```

```{r invgauss-base-q}
q_vec(qInvGauss, c(0.25, 0.5, 0.75), mean = example$mean,
      shape = example$shape)
q_vec(qInvGauss, c(0.25, 0.5, 0.75), mean = example$mean,
      shape = example$shape, lower.tail = FALSE)
q_vec(qInvGauss, c(log(0.25), log(0.5), log(0.75)), mean = example$mean,
      shape = example$shape, log.p = TRUE)
```

```{r invgauss-base-r}
draw_many(rInvGauss, list(mean = example$mean, shape = example$shape))
```

```{r invgauss-base-plot}
df_ig_base <- do.call(rbind, lapply(ig_base_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dInvGauss, list(mean = ps$mean, shape = ps$shape)), label = ps$label)
}))

ggplot(df_ig_base, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Inverse Gaussian base kernels", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

### With GPD tail

```{r invgauss-base-gpd}
ig_base_gpd_sets <- list(
  list(label = "Base A", mean = 1.5, shape = 2, threshold = 2.0, tail_scale = 0.4, tail_shape = 0.2),
  list(label = "Base B", mean = 1.2, shape = 2.5, threshold = 1.8, tail_scale = 0.35, tail_shape = 0.18),
  list(label = "Base C", mean = 1.8, shape = 3, threshold = 2.5, tail_scale = 0.5, tail_shape = 0.22)
)
example <- ig_base_gpd_sets[[1]]
```

```{r invgauss-base-gpd-d}
dInvGaussGpd(2, mean = example$mean, shape = example$shape,
             threshold = example$threshold, tail_scale = example$tail_scale,
             tail_shape = example$tail_shape)
dInvGaussGpd(2, mean = example$mean, shape = example$shape,
             threshold = example$threshold, tail_scale = example$tail_scale,
             tail_shape = example$tail_shape, log = TRUE)
```

```{r invgauss-base-gpd-p}
pInvGaussGpd(2, mean = example$mean, shape = example$shape,
             threshold = example$threshold, tail_scale = example$tail_scale,
             tail_shape = example$tail_shape)
pInvGaussGpd(2, mean = example$mean, shape = example$shape,
             threshold = example$threshold, tail_scale = example$tail_scale,
             tail_shape = example$tail_shape, lower.tail = FALSE)
pInvGaussGpd(2, mean = example$mean, shape = example$shape,
             threshold = example$threshold, tail_scale = example$tail_scale,
             tail_shape = example$tail_shape, log.p = TRUE)
```

```{r invgauss-base-gpd-q}
q_vec(qInvGaussGpd, c(0.25, 0.5, 0.75), mean = example$mean,
      shape = example$shape, threshold = example$threshold,
      tail_scale = example$tail_scale, tail_shape = example$tail_shape)
q_vec(qInvGaussGpd, c(0.25, 0.5, 0.75), mean = example$mean,
      shape = example$shape, threshold = example$threshold,
      tail_scale = example$tail_scale, tail_shape = example$tail_shape,
      lower.tail = FALSE)
q_vec(qInvGaussGpd, c(log(0.25), log(0.5), log(0.75)), mean = example$mean,
      shape = example$shape, threshold = example$threshold,
      tail_scale = example$tail_scale, tail_shape = example$tail_shape,
      log.p = TRUE)
```

```{r invgauss-base-gpd-r}
draw_many(rInvGaussGpd, example)
```

```{r invgauss-base-gpd-plot}
df_ig_base_gpd <- do.call(rbind, lapply(ig_base_gpd_sets, function(ps) {
  data.frame(x = grid, density = density_curve(grid, dInvGaussGpd, list(mean = ps$mean, shape = ps$shape, threshold = ps$threshold, tail_scale = ps$tail_scale, tail_shape = ps$tail_shape)), label = ps$label)
}))

ggplot(df_ig_base_gpd, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Inverse Gaussian base with GPD tail", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

## Cauchy


### Cauchy base and mixtures

A Cauchy distribution with location \(x_0\in\mathbb{R}\) and scale \(\gamma>0\) has density
\[
f(y\mid x_0,\gamma)=\frac{1}{\pi\gamma\left[1+\left(\frac{y-x_0}{\gamma}\right)^2\right]},
\quad y\in\mathbb{R}.
\]

A finite Cauchy mixture is
\[
f(y)=\sum_{j=1}^J w_j\,\text{Cauchy}(y\mid x_{0j},\gamma_j).
\]

**Parameter mapping (math \(\rightarrow\) code):** \(x_0\to\) `location`, \(\gamma\to\) `scale`; in mixtures \(x_{0j}\to\) `location[j]`, \(\gamma_j\to\) `scale[j]`, \(w_j\to\) `w[j]`.


### Without GPD (base and mixture)

```{r cauchy-base}
grid <- seq(-8, 8, length.out = 400)
cauchy_sets <- list(
  list(label = "Base", location = 0, scale = 1),
  list(label = "Mix A", w = c(0.6, 0.3, 0.1), location = c(-1, 0, 1), scale = c(1.0, 1.2, 2)),
  list(label = "Mix B", w = c(0.5, 0.3, 0.2), location = c(-1.5, 0.5, 1.5), scale = c(1.1, 1.0, 0.9))
)

base_par <- cauchy_sets[[1]]
mix_par1 <- cauchy_sets[[2]]
```

```{r cauchy-d}
dCauchy(0, location = base_par$location, scale = base_par$scale)
dCauchy(0, location = base_par$location, scale = base_par$scale, log = TRUE)
```

```{r cauchy-p}
pCauchy(0, location = base_par$location, scale = base_par$scale)
pCauchy(0, location = base_par$location, scale = base_par$scale, lower.tail = FALSE)
pCauchy(0, location = base_par$location, scale = base_par$scale, log.p = TRUE)
```

```{r cauchy-q}
q_vec(qCauchy, c(0.25, 0.5, 0.75), location = base_par$location,
      scale = base_par$scale)
q_vec(qCauchy, c(0.25, 0.5, 0.75), location = base_par$location,
      scale = base_par$scale, lower.tail = FALSE)
q_vec(qCauchy, c(log(0.25), log(0.5), log(0.75)), location = base_par$location,
      scale = base_par$scale, log.p = TRUE)
```

```{r cauchy-r}
draw_many(rCauchy, list(location = base_par$location, scale = base_par$scale))
```

```{r cauchy-mix-d}
dCauchyMix(0, w = mix_par1$w, location = mix_par1$location, scale = mix_par1$scale)
dCauchyMix(0, w = mix_par1$w, location = mix_par1$location, scale = mix_par1$scale, log = TRUE)
```

```{r cauchy-mix-r}
draw_many(rCauchyMix, list(w = mix_par1$w, location = mix_par1$location, scale = mix_par1$scale))
```

```{r cauchy-plot}
df_cauchy <- rbind(
  data.frame(x = grid, density = density_curve(grid, dCauchy, list(location = base_par$location, scale = base_par$scale)), label = "Base"),
  data.frame(x = grid, density = density_curve(grid, dCauchyMix, list(w = mix_par1$w, location = mix_par1$location, scale = mix_par1$scale)), label = "Mix A"),
  data.frame(x = grid, density = density_curve(grid, dCauchyMix, list(w = cauchy_sets[[3]]$w, location = cauchy_sets[[3]]$location, scale = cauchy_sets[[3]]$scale)), label = "Mix B")
)

ggplot(df_cauchy, aes(x = x, y = density, color = label)) +
  geom_line(linewidth = 1) +
  labs(title = "Cauchy base and mixtures", x = "x", y = "density") +
  theme_minimal() + theme(legend.position = "top")
```

## Summary: Distribution & Backend Reference

This section provides a compact reference for the distributions supported by **DPmixGPD** and the corresponding user-facing function families.  
The **CRP backend** uses base (single-kernel) functions, while the **SB backend** uses mixture (multi-component) functions. The “Type” column indicates whether each backend expects **scalar** parameters (single value per parameter) or **vectors** indexed by mixture component.


```{r dist-backend-summary, echo=FALSE, message=FALSE, warning=FALSE}
df <- data.frame(
  Distribution = c(
    "Normal","Gamma","Lognormal","Laplace","Inverse Gaussian","Amoroso","Cauchy",
    "GPD",
    "Normal + GPD","Gamma + GPD","Lognormal + GPD","Laplace + GPD","InvGauss + GPD","Amoroso + GPD"
  ),
  Parameters = c(
    "$\\mu,\\,\\sigma$",
    "$\\alpha,\\,\\beta$",
    "$\\mu,\\,\\sigma$",
    "$\\ell,\\,b$",
    "$\\mu,\\,\\lambda$",
    "$a,\\,\\theta,\\,\\alpha,\\,\\beta$",
    "$x_0,\\,\\gamma$",
    "$u,\\,\\sigma,\\,\\xi$",
    "$\\mu,\\,\\sigma,\\,u,\\,\\sigma,\\,\\xi$",
    "$\\alpha,\\,\\beta,\\,u,\\,\\sigma,\\,\\xi$",
    "$\\mu,\\,\\sigma,\\,u,\\,\\sigma,\\,\\xi$",
    "$\\ell,\\,b,\\,u,\\,\\sigma,\\,\\xi$",
    "$\\mu,\\,\\lambda,\\,u,\\,\\sigma,\\,\\xi$",
    "$a,\\,\\theta,\\,\\alpha,\\,\\beta,\\,u,\\,\\sigma,\\,\\xi$"
  ),
  `Arguments` = c(
    "`mean`, `sd`",
    "`shape`, `scale`",
    "`meanlog`, `sdlog`",
    "`location`, `scale`",
    "`mean`, `shape`",
    "`loc`, `scale`, `shape1`, `shape2`",
    "`location`, `scale`",
    "`threshold`, `scale`, `shape`",
    "`mean`, `sd`, `threshold`, `tail_scale`, `tail_shape`",
    "`shape`, `scale`, `threshold`, `tail_scale`, `tail_shape`",
    "`meanlog`, `sdlog`, `threshold`, `tail_scale`, `tail_shape`",
    "`location`, `scale`, `threshold`, `tail_scale`, `tail_shape`",
    "`mean`, `shape`, `threshold`, `tail_scale`, `tail_shape`",
    "`loc`, `scale`, `shape1`, `shape2`, `threshold`, `tail_scale`, `tail_shape`"
  ),
  `Type` = c(
    "scalars","scalars","scalars","scalars","scalars","scalars","scalars",
    "scalars",
    "scalars","scalars","scalars","scalars","scalars","scalars"
  ),
  `Function` = c(
    "`dnorm()`, `pnorm()`, `qnorm()`, `rnorm()`",
    "`dgamma()`, `pgamma()`, `qgamma()`, `rgamma()`",
    "`dlnorm()`, `plnorm()`, `qlnorm()`, `rlnorm()`",
    "`ddexp()`, `pdexp()`, `qdexp()`, `rdexp()`",
    "`dinvgauss()`, `pinvgauss()`, `qinvgauss()`, `rinvgauss()`",
    "`damoroso()`, `pamoroso()`, `qamoroso()`, `ramoroso()`",
    "`dcauchy()`, `pcauchy()`, `qcauchy()`, `rcauchy()`",
    "`dGpd()`, `pGpd()`, `qGpd()`, `rGpd()`",
    "`dNormGpd()`, `pNormGpd()`, `qNormGpd()`, `rNormGpd()`",
    "`dGammaGpd()`, `pGammaGpd()`, `qGammaGpd()`, `rGammaGpd()`",
    "`dLognormalGpd()`, `pLognormalGpd()`, `qLognormalGpd()`, `rLognormalGpd()`",
    "`dLaplaceGpd()`, `pLaplaceGpd()`, `qLaplaceGpd()`, `rLaplaceGpd()`",
    "`dInvGaussGpd()`, `pInvGaussGpd()`, `qInvGaussGpd()`, `rInvGaussGpd()`",
    "`dAmorosoGpd()`, `pAmorosoGpd()`, `qAmorosoGpd()`, `rAmorosoGpd()`"
  ),
  `Type.` = c(
    "vectors","vectors","vectors","vectors","vectors","vectors","vectors",
    "NA",
    "vectors","vectors + scalars","vectors + scalars","vectors + scalars","vectors + scalars","vectors + scalars"
  ),
  `Function.` = c(
    "`dNormMix()`, `pNormMix()`, `qNormMix()`, `rNormMix()`",
    "`dGammaMix()`, `pGammaMix()`, `qGammaMix()`, `rGammaMix()`",
    "`dLognormalMix()`, `pLognormalMix()`, `qLognormalMix()`, `rLognormalMix()`",
    "`dLaplaceMix()`, `pLaplaceMix()`, `qLaplaceMix()`, `rLaplaceMix()`",
    "`dInvGaussMix()`, `pInvGaussMix()`, `qInvGaussMix()`, `rInvGaussMix()`",
    "`dAmorosoMix()`, `pAmorosoMix()`, `qAmorosoMix()`, `rAmorosoMix()`",
    "`dCauchyMix()`, `pCauchyMix()`, `qCauchyMix()`, `rCauchyMix()`",
    "NA",
    "`dNormMixGpd()`, `pNormMixGpd()`, `qNormMixGpd()`, `rNormMixGpd()`",
    "`dGammaMixGpd()`, `pGammaMixGpd()`, `qGammaMixGpd()`, `rGammaMixGpd()`",
    "`dLognormalMixGpd()`, `pLognormalMixGpd()`, `qLognormalMixGpd()`, `rLognormalMixGpd()`",
    "`dLaplaceMixGpd()`, `pLaplaceMixGpd()`, `qLaplaceMixGpd()`, `rLaplaceMixGpd()`",
    "`dInvGaussMixGpd()`, `pInvGaussMixGpd()`, `qInvGaussMixGpd()`, `rInvGaussMixGpd()`",
    "`dAmorosoMixGpd()`, `pAmorosoMixGpd()`, `qAmorosoMixGpd()`, `rAmorosoMixGpd()`"
  ),
  check.names = FALSE
)

# Rename SB columns to match header labels cleanly
names(df)[4:7] <- c("Type", "Function", "Type", "Function")

knitr::kable(
  df,
  format = "html",
  escape = FALSE,
  align = rep("c", 7)
) |>
  kableExtra::kable_styling(full_width = FALSE, position = "center") |>
  kableExtra::add_header_above(c(" " = 3, "CRP" = 2, "SB" = 2)) |>
  kableExtra::row_spec(0, bold = TRUE)
```

### Notes 

- **CRP Backend** uses base (single-kernel) functions; parameters are scalar.

- **SB Backend** uses mixture/spliced mixture functions; component-specific parameters are typically vectors indexed by component \(j\); mixture weights `w` are required as vector.

- For **GPD-only**, SB entries are **NA** because there is no mixture-only GPD wrapper in the SB family.





