---
title: "Legacy: 10. Conditional DPmix with CRP Backend"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{Legacy: 10. Conditional DPmix with CRP Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(dplyr)
library(tibble)
set.seed(123)
```

> **Legacy note:** This page is preserved for historical context and extended detail. It predates the streamlined official vignettes and may include longer runs or exploratory material.

# Conditional DPmix: CRP Backend with Covariates

**Purpose**: Show how the CRP backend can model $y | X$ via a covariate-dependent Dirichlet Process mixture. This vignette parallels `v04` but includes $X$ so we can explore conditional predictions.

---

## Data Setup

```{r data-setup}
data("nc_posX100_p3_k2")
y <- nc_posX100_p3_k2$y
X <- as.matrix(nc_posX100_p3_k2$X)
if (is.null(colnames(X))) {
  colnames(X) <- paste0("x", seq_len(ncol(X)))
}

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

df_cov <- data.frame(y = y, x1 = X[, 1], x2 = X[, 2])

p_cov <- ggplot(df_cov, aes(x = x1, y = y)) +
  geom_point(alpha = 0.6, color = "steelblue") +
  geom_smooth(method = "loess", color = "firebrick", fill = NA) +
  labs(title = "y vs X1 (loess smoother)", x = "X1", y = "y") +
  theme_minimal()

grid.arrange(p_cov, ncol = 1)
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Summary of Conditional Dataset", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

---

## Model Specification & Bundle

```{r spec-bundle}
bundle_cond_normal <- build_nimble_bundle(
  y = y,
  X = X,
  kernel = "normal",
  backend = "crp",
  GPD = FALSE,
  components = 5,
  mcmc = list(
    niter = 60,
    nburnin = 10,
    nchains = 2,
    thin = 1
  )
)

bundle_cond_amoroso <- build_nimble_bundle(
  y = y,
  X = X,
  kernel = "amoroso",
  backend = "crp",
  GPD = FALSE,
  components = 5,
  mcmc = list(
    niter = 60,
    nburnin = 10,
    nchains = 1,
    thin = 1
  )
)
```

---

## Running MCMC

```{r mcmc-run}
fit_cond_normal <- run_mcmc_bundle_manual(bundle_cond_normal)
fit_cond_amoroso <- run_mcmc_bundle_manual(bundle_cond_amoroso)
summary(fit_cond_normal)
summary(fit_cond_amoroso)
```

```{r params-cond-crp}
params_cond <- params(fit_cond_normal)
params_cond
```

---

## Conditional Predictions

```{r cond-predict}
X_new <- expand.grid(
  x1 = seq(-2, 2, length.out = 3),
  x2 = c(0, 1),
  x3 = 0
)
colnames(X_new) <- colnames(X)

y_grid <- seq(-1, 10, length.out = 200)
densities_normal <- lapply(seq_len(nrow(X_new)), function(i) {
  pred <- predict(fit_cond_normal, x = as.matrix(X_new[i, , drop = FALSE]), y = y_grid, type = "density")
  data.frame(
    y = pred$fit$y,
    density = pred$fit$density,
    group = paste0("x1=", round(X_new[i, "x1"], 1), ", x2=", X_new[i, "x2"]),
    model = "Normal"
  )
})

densities_amoroso <- lapply(seq_len(nrow(X_new)), function(i) {
  pred <- predict(fit_cond_amoroso, x = as.matrix(X_new[i, , drop = FALSE]), y = y_grid, type = "density")
  data.frame(
    y = pred$fit$y,
    density = pred$fit$density,
    group = paste0("x1=", round(X_new[i, "x1"], 1), ", x2=", X_new[i, "x2"]),
    model = "Amoroso (shape1=1)"
  )
})

df_dens <- bind_rows(densities_normal, densities_amoroso)

ggplot(df_dens, aes(x = y, y = density, color = group)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ model) +
  labs(title = "Conditional Predictive Densities", x = "y", y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Covariate Effect on Conditional Quantiles

```{r quantile-effect}
X_grid <- cbind(
  x1 = seq(-2, 2, length.out = 5),
  x2 = 0,
  x3 = 0
)
colnames(X_grid) <- colnames(X)

quant_probs <- c(0.25, 0.5, 0.75)
pred_q_normal <- predict(fit_cond_normal, x = as.matrix(X_grid), type = "quantile", index = quant_probs)
pred_q_amoroso <- predict(fit_cond_amoroso, x = as.matrix(X_grid), type = "quantile", index = quant_probs)

quant_df_normal <- pred_q_normal$fit
quant_df_normal$x1 <- X_grid[quant_df_normal$id, "x1"]
quant_df_normal$model <- "Normal"

quant_df_amoroso <- pred_q_amoroso$fit
quant_df_amoroso$x1 <- X_grid[quant_df_amoroso$id, "x1"]
quant_df_amoroso$model <- "Amoroso (shape1=1)"

bind_rows(quant_df_normal, quant_df_amoroso) %>%
  ggplot(aes(x = x1, y = estimate, color = factor(index), group = index)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ model) +
  labs(title = "Conditional Quantiles vs x1 (x2=0)", x = "x1", y = "y", color = "Quantile") +
  theme_minimal()
```

---

## Residuals & Diagnostics

```{r residuals}
fit_vals <- fitted(fit_cond_normal)
plot(fit_vals)
```

```{r diagnostics}
plot(fit_cond_normal, family = c("traceplot", "autocorrelation", "geweke"))
plot(fit_cond_amoroso, family = c("density", "running", "caterpillar"))
```

---

## Takeaways

- Covariate-informed DP mixtures predict outcome distributions that shift with `x1` (and other covariates).
- Use `predict(..., type = "density")` to visualize conditional densities and `type = "quantile"` for posterior-mean location shifts.
- Diagnostics (`plot(fit_cond_normal)`, `fitted(fit_cond_normal)`) ensure the chains mix before relying on predictions.
- Next vignette extends the same idea to the SB backend before adding tails.







