---
title: "ex01. Unconditional DPM (CRP Backend)"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = 'legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
max_workers <- local({
  n_phys <- suppressWarnings(tryCatch(parallel::detectCores(logical = FALSE), error = function(e) NA_integer_))
  n_log <- suppressWarnings(tryCatch(parallel::detectCores(logical = TRUE), error = function(e) NA_integer_))
  n <- if (is.finite(n_phys) && !is.na(n_phys) && n_phys >= 1L) n_phys else n_log
  if (!is.finite(n) || is.na(n) || n < 1L) n <- 1L
  as.integer(min(10L, n))
})
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, workers = max_workers)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_chains = TRUE, workers = max_workers, z_update_every = 2)

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(CausalMixGPD)

# Precomputed results for fast website builds
source("_legacy-setup.R")
max_workers <- min(max_workers, CMGPD_MAX_WORKERS)
```


> **Website workflow note.**
> This page reflects the current exported API and recommended wrapper-first usage.
> Last updated: **2026-02-19**.
>
> For the full package narrative, see the main package vignettes (basic, unconditional, conditional, and causal).


# Unconditional DPmix: Chinese Restaurant Process (CRP)

## What you’ll learn

- Fit a **bulk-only** one-arm model with `dpmix()` using the CRP backend.
- Use the package’s **S3 surface** (`summary()`, `plot()`, `predict()`) to diagnose and extract posterior predictive summaries.

## When to use this template

- You have a single outcome \(y\) and want a flexible density model without covariates \(X\).
- You do not need tail splicing (`GPD = FALSE`).
- You want adaptive clustering behavior via CRP (within a finite `components` cap).

## Goal (in one sentence)

Estimate the density of a univariate outcome \(y\) using a **Dirichlet-process mixture** with a **Chinese Restaurant Process** backend.

**Model**: $y_i | G \sim \int K(y_i; \theta) dG(\theta)$ where $G \sim \text{DP}(\alpha, G_0)$

**Backend**: CRP with **truncation at max components**

---

## Data Setup

```{r data-setup}
# Load pre-generated dataset: 200 observations from mixture of 3 gamma components
data("nc_pos200_k3")
y_mixed <- nc_pos200_k3$y

paste("Sample size:", length(y_mixed))
paste("Mean:", mean(y_mixed))
paste("SD:", sd(y_mixed))
paste("Range:", paste(range(y_mixed), collapse = " to "))

# Visualization
df_data <- data.frame(y = y_mixed)
p_raw <- ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, alpha = 0.6, fill = "steelblue",color = "black") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Raw Data: Mixed Gamma Distribution", x = "y", y = "Density") +
  theme_minimal()

print(p_raw)
```

---

## Model Specification & Bundle

This page keeps the direct `bundle()` workflow because it is still useful for showing how the unconditional CRP specification is assembled before fitting.

```{r mcmc-crp, cache = FALSE}
bundle_crp <- bundle(
  y = y_mixed,
  kernel = "laplace",
  backend = "crp",
  GPD = FALSE,
  components = 3,
  alpha_random = TRUE,
  mcmc = mcmc
)
```

### Summary of MCMC Bundle
```{r}
summary(bundle_crp)
```

### Running MCMC
```{r mcmc-crp-fit, results = "hide", cache = FALSE}
fit_crp <- load_or_fit("ex01-unconditional-dpm-crp-fit_crp", dpmix(bundle_crp))
```

### Summary of Fitted MCMC model

```{r}
summary(fit_crp)
```

```{r params-crp}
params_crp <- params(fit_crp)
params_crp
```

---

## MCMC Diagnostics Plots


```{r diag-trace}
# Trace plots for key parameters
plot(fit_crp, params = "alpha", family = c("traceplot", "density", "geweke"))
```


---

## Posterior Predictions

### Predictive Density

```{r pred-density}
# Generate prediction grid
y_grid <- seq(0, max(y_mixed) * 1.2, length.out = 200)

# Posterior predictive density
pred_density <- predict(fit_crp, y = y_grid, type = "density")

# Use S3 plot method
plot(pred_density)
```

### Quantile Predictions

```{r pred-quantiles}
# Posterior predictive quantiles with credible intervals
quantiles_pred <- predict(fit_crp, type = "quantile", p = c(0.05, 0.25, 0.5, 0.75, 0.95),
                          interval = "credible")

# Use S3 plot method
plot(quantiles_pred)
```

---

## Varying Truncation Level (components)

```{r components-sensitivity, cache = FALSE}
# Demonstrate with one value
bundle_components <- bundle(
  y = y_mixed,
  kernel = "laplace",
  backend = "crp",
  components = 5,
  mcmc = mcmc
)
fit_components <- load_or_fit("ex01-unconditional-dpm-crp-fit_components", dpmix(bundle_components))
```


```{r}
summary(fit_components)
```

---

For unconditional models, use `predict()` for posterior summaries; `fitted()` and `residuals()` are not supported.

---

## Model Comparison: Different Kernels

### Laplace Kernel (Current)

```{r comp-laplace, cache = FALSE}
bundle_laplace <- bundle(
  y = y_mixed,
  kernel = "laplace",
  backend = "crp",
  components = 5,
  mcmc = mcmc
)
fit_laplace <- load_or_fit("ex01-unconditional-dpm-crp-fit_laplace", dpmix(bundle_laplace))
```


```{r}
summary(fit_laplace)
```

### Amoroso Kernel (Alternative)

```{r comp-amoroso, cache = FALSE}
bundle_amoroso <- bundle(
  y = y_mixed,
  kernel = "amoroso",
  backend = "crp",
  components = 5,
  mcmc = mcmc
)
fit_amoroso <- load_or_fit("ex01-unconditional-dpm-crp-fit_amoroso", dpmix(bundle_amoroso))
```


```{r}
summary(fit_amoroso)
```

### Model Comparison via Predictions

```{r comp-summary, cache = FALSE}
# Compare models using predict() (fitted/residuals not supported for unconditional)
pred_laplace <- predict(fit_laplace, type = "mean", level = 0.9, interval = "credible")
pred_amoroso <- predict(fit_amoroso, type = "mean", level = 0.9, interval = "credible")
```

```{r fig.height=6, fig.width=10, cache = FALSE}
# For unconditional models, fitted() is not supported; use predict() for comparisons (see pred_laplace, pred_amoroso above).
```



## Key Takeaways

- **CRP Backend**: Flexible cluster assignment, ideal for unknown mixture complexity
- **components Parameter**: A higher `components` cap allows more represented mixture components but increases computation
- **Kernel Choice**: Laplace and Amoroso illustrate two different bulk-kernel choices for the same unconditional CRP workflow
- **Diagnostics**: Check convergence (Rhat, ESS) and posterior predictive fit
- **Next**: Compare with **Stick-Breaking (SB)** backend in ex02

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [backends-and-workflow](../start/backends-and-workflow.html)
- Next: [ex02-unconditional-dpm-sb](ex02-unconditional-dpm-sb.html)
- Workflow index: [Roadmap](../start/roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [Model Umbrella](../advanced/model-umbrella.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->



