---
title: "ex02. Unconditional DPM (SB Backend)"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = 'legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, workers = 2)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_chains = TRUE, workers = 2, z_update_every = 2)

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(CausalMixGPD)

# Precomputed results for fast website builds
source("_legacy-setup.R")
```


> **Website workflow note.**
> This page reflects the current exported API and recommended wrapper-first usage.
> Last updated: **2026-02-19**.
>
> For the full package narrative, see the main package vignettes (basic, unconditional, conditional, and causal).


# Unconditional DPmix: Stick-Breaking (SB) Backend

**Purpose**: Showcase the stick-breaking backend that uses a fixed number of mixture components (`components = J`) and contrast it with the CRP backend from `start/backends-and-workflow`. In this vignette we fit **two** bulk kernels (**Gamma** and **Cauchy**) on the same dataset to highlight how heavier tails in the bulk can change the fit even without a GPD tail.

## What you’ll learn

- How the **stick-breaking (SB)** backend differs from CRP in practice (fixed truncation via `components = J`).
- How to use `bundle()` + `dpmix()` + `predict()`/`plot()` for an **unconditional bulk-only** DP mixture.
- How **bulk kernel choice** (e.g., Gamma vs Cauchy) can change tail behavior even when `GPD = FALSE`.

## When to use this template

- You want a **predictable runtime** mixture workflow with a fixed upper bound on components.
- You mainly care about bulk fit, density estimation, and posterior predictive summaries.

## Next steps

- If extremes matter, extend to `dpmgpd()` with `GPD = TRUE` (see `ex03`/`ex04`).

---

## Data Setup

```{r data-setup, fig.width=8, fig.height=6}
# Generate a reproducible mock dataset instead of relying on saved cache files.
set.seed(10102)
n <- 200
weights <- c(0.4, 0.35, 0.25)
params <- list(
  list(shape = 2.0, rate = 1.5),
  list(shape = 4.0, rate = 1.0),
  list(shape = 7.0, rate = 0.7)
)
component <- sample(seq_along(weights), size = n, replace = TRUE, prob = weights)
y_mixed <- numeric(n)
for (k in seq_along(params)) {
  idx <- which(component == k)
  if (length(idx) > 0) {
    y_mixed[idx] <- rgamma(length(idx), shape = params[[k]]$shape, rate = params[[k]]$rate)
  }
}

df_data <- data.frame(y = y_mixed)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y_mixed), mean(y_mixed), sd(y_mixed), min(y_mixed), max(y_mixed))
)

p_raw <- ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, fill = "darkorange", alpha = 0.7, color = "black") +
  geom_density(color = "darkred", linewidth = 1.2) +
  labs(title = "Raw Data: Mixed Gamma Distribution", x = "y", y = "Density") +
  theme_minimal()

print(p_raw)
```

```{r data-table, echo=FALSE}
summary_tbl
```

---

## Model Specification & Bundle

We build the stick-breaking mixture explicitly with `components = 5` so that there is room for weight decay while keeping the MCMC runtime manageable for a vignette. We then refit the same SB model with a **Cauchy** kernel for comparison.

```{r bundle-mcmc-sb, cache = FALSE}
# --- Gamma kernel ---
bundle_sb_gamma <- bundle(
  x = y_mixed,
  kernel = "gamma",
  backend = "sb",
  components = 5,            # fixed truncation level for SB
  GPD = FALSE,               # bulk-only scenario
  mcmc = mcmc
)

# --- Cauchy kernel ---
bundle_sb_cauchy <- bundle(
  x = y_mixed,
  kernel = "cauchy",
  backend = "sb",
  components = 5,
  GPD = FALSE,
  mcmc = mcmc
)
```

---

## Running MCMC & Summary

```{r mcmc-run, cache = FALSE}
fit_sb_gamma <- load_or_fit("ex02-unconditional-dpm-sb-fit_sb_gamma", dpmix(bundle_sb_gamma))
fit_sb_cauchy <- load_or_fit("ex02-unconditional-dpm-sb-fit_sb_cauchy", dpmix(bundle_sb_cauchy))
```

```{r fit-summary}
summary(fit_sb_gamma)
summary(fit_sb_cauchy)
```

```{r params-sb}
params_gamma <- params(fit_sb_gamma)
params_gamma
```

---

## MCMC Diagnostics (S3 Plot Methods)

```{r diag-plots}
# Default diagnostics for each fit
plot(fit_sb_gamma, family = c("traceplot", "autocorrelation", "running"))
plot(fit_sb_cauchy, family = c("density", "geweke", "caterpillar"))
``` 

Use `summary(fit_sb_gamma)` and `summary(fit_sb_cauchy)` to inspect effective sample size, R-hat, and other convergence diagnostics; the `plot()` calls above show the trace/density pairs for both global and stick-breaking weight parameters.

---

## Stick-Breaking Weights & Component Activity

The stick-breaking weights are exposed via the `plot()` diagnostics above, and you can refer to the raw `fit_sb_gamma` / `fit_sb_cauchy` objects (or their `summary()` output) for posterior summaries of each component weight.

---

## Posterior Predictions

```{r density-predict}
y_grid <- seq(min(y_mixed), max(y_mixed) * 1.3, length.out = 250)

pred_density_gamma <- predict(fit_sb_gamma, y = y_grid, type = "density")
pred_density_cauchy <- predict(fit_sb_cauchy, y = y_grid, type = "density")

plot(pred_density_gamma)
plot(pred_density_cauchy)
```

```{r quantile-predict}
quant_probs <- c(0.05, 0.25, 0.5, 0.75, 0.95)

pred_q_gamma <- predict(fit_sb_gamma, type = "quantile", p = quant_probs, interval = "credible")
pred_q_cauchy <- predict(fit_sb_cauchy, type = "quantile", p = quant_probs, interval = "credible")

plot(pred_q_gamma)
plot(pred_q_cauchy)
```

```{r mean-predict}
pred_mean_gamma <- predict(fit_sb_gamma, type = "mean")
pred_mean_cauchy <- predict(fit_sb_cauchy, type = "mean")

plot(pred_mean_gamma)
plot(pred_mean_cauchy)
```

---

## Bulk-only vs CRP Backends

```{r compare-backend, cache = FALSE}
bundle_crp_small <- bundle(
  x = y_mixed,
  kernel = "gamma",
  backend = "crp",
  components = 5,
  GPD = FALSE,
  mcmc = mcmc
)
fit_crp_small <- load_or_fit("ex02-unconditional-dpm-sb-fit_crp_small", dpmix(bundle_crp_small))

crp_quant <- predict(fit_crp_small, type = "quantile", p = quant_probs, interval = "credible")
sb_quant <- predict(fit_sb_gamma, type = "quantile", p = quant_probs, interval = "credible")

bind_rows(
  crp_quant$fit %>% mutate(model = "CRP"),
  sb_quant$fit %>% mutate(model = "SB")
) %>%
  select(any_of(c("model", "index", "estimate", "lower", "upper")))
```

```{r compare-plots}
plot(crp_quant)
plot(sb_quant)
```

---

For unconditional models, use `predict()` only; `fitted()` and `residuals()` are not supported.

---

## Takeaways

- **SB Backend**: Fixed `components` keeps label-switching manageable, and the diagnostic plots via the S3 `plot()` method show weight dynamics.
- **Kernels matter**: Gamma vs Cauchy can behave very differently in the shoulders/tails, even without a GPD tail.
- **Predictions**: Posterior density, posterior-mean quantiles, and mean (via predictive sampling) are accessible via `predict()` with pointwise 95% credible intervals.
- **Backend comparison**: The CRP fit (`start/backends-and-workflow`) and the SB-gamma fit deliver similar central quantiles while SB offers more control over truncation.
- **Next**: Explore tail augmentation (`GPD = TRUE`) with the CRP backend in `ex03` or SB backend in `ex04`.

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex01-unconditional-dpm-crp](ex01-unconditional-dpm-crp.html)
- Next: [ex03-unconditional-dpmgpd-crp](ex03-unconditional-dpmgpd-crp.html)
- Workflow index: [Roadmap](../start/roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [Model Umbrella](../advanced/model-umbrella.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->





