---
title: "ex06. Unconditional DPmix with CRP Backend"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = 'legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Required for %>% and kable_styling
library(magrittr)
library(kableExtra)
library(patchwork)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy
# Ensure package is fully loaded for data access
if (!require("devtools", quietly = TRUE)) {
  library(devtools)
}
pkg_root <- rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
  devtools::load_all(path = pkg_root, quiet = TRUE)
library(DPmixGPD)

# Precomputed results for fast website builds
source("_legacy-setup.R")
if (!exists("safe_plot", mode = "function")) {
  safe_plot <- function(x, ...) {
    obj <- tryCatch(plot(x, ...), error = function(e) e)
    if (inherits(obj, "ggplot")) {
      try(print(obj), silent = TRUE)
    } else if (is.list(obj)) {
      for (elt in obj) if (inherits(elt, "ggplot")) try(print(elt), silent = TRUE)
    }
    invisible(obj)
  }
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (basic, unconditional, conditional, and causal)..


# Unconditional DPmix: Chinese Restaurant Process (CRP)

**Goal**: Estimate density of univariate outcome $y$ using **nonparametric Dirichlet Process mixture** with **Chinese Restaurant Process** backend.

**Model**: $y_i | G \sim \int K(y_i; \theta) dG(\theta)$ where $G \sim \text{DP}(\alpha, G_0)$

**Backend**: CRP with **truncation at max components**

---

## Data Setup

```{r data-setup}
# Load pre-generated dataset: 200 observations from mixture of 3 gamma components
data("nc_pos200_k3", package = "DPmixGPD")
y_mixed <- nc_pos200_k3$y

paste("Sample size:", length(y_mixed))
paste("Mean:", mean(y_mixed))
paste("SD:", sd(y_mixed))
paste("Range:", paste(range(y_mixed), collapse = " to "))

# Visualization
df_data <- data.frame(y = y_mixed)
p_raw <- ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, alpha = 0.6, fill = "steelblue",color = "black") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Raw Data: Mixed Gamma Distribution", x = "y", y = "Density") +
  theme_minimal()

print(p_raw)
```

---

## Model Specification & Bundle

We'll use the `build_nimble_bundle` function directly which handles both specification and bundle creation.

```{r bundle-mcmc-crp, cache=FALSE}
bundle_crp <- DPmixGPD::build_nimble_bundle(
  y = y_mixed,
  kernel = "laplace",         # Use laplace kernel
  backend = "crp",            # CRP backend
  GPD = FALSE,                # No tail augmentation
  components = 3,             # Minimal for testing
  alpha_random = TRUE,        # Random DP concentration
  mcmc = mcmc
)

```

---

### Building MCMC bundle

```{r mcmc-crp, cache=FALSE}
bundle_crp <- DPmixGPD::build_nimble_bundle(
  y_mixed,
  kernel = "laplace",
  backend = "crp",
  GPD = FALSE,
  components = 3,
  alpha_random = TRUE,
  mcmc = mcmc
)
```

### Summary of MCMC Bundle
```{r}
summary(bundle_crp)
```

### Running MCMC
```{r mcmc-crp-fit, results = "hide", cache=FALSE}
fit_crp <- load_or_fit("ex06-unconditional-DPmix-CRP-fit_crp", run_mcmc_bundle_manual(bundle_crp))
```

### Summary of Fitted MCMC model

```{r}
summary(fit_crp)
```

```{r params-crp}
params_crp <- params(fit_crp)
params_crp
```

---

## MCMC Diagnostics Plots


```{r diag-trace}
# Trace plots for key parameters
safe_plot(fit_crp, params = "alpha", family = c("traceplot", "density", "geweke"))
```


---

## Posterior Predictions

### Predictive Density

```{r pred-density}
# Generate prediction grid
y_grid <- seq(0, max(y_mixed) * 1.2, length.out = 200)

# Posterior predictive density
pred_density <- predict(fit_crp, y = y_grid, type = "density")

# Use S3 plot method
safe_plot(pred_density)
```

### Quantile Predictions

```{r pred-quantiles}
# Posterior predictive quantiles with credible intervals
quantiles_pred <- predict(fit_crp, type = "quantile", 
                          index = c(0.05, 0.25, 0.5, 0.75, 0.95),
                          interval = "credible")

# Display table
quantiles_pred$fit %>%
 kbl(caption = "Posterior Predictive Quantiles with Credible Intervals",
   align = "c",
                  digits = 3) %>%
 kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "center")

# Use S3 plot method
safe_plot(quantiles_pred)
```

---

## Varying Truncation Level (components)

```{r components-sensitivity, cache=FALSE}
# Demonstrate with one value
bundle_components <- DPmixGPD::build_nimble_bundle(
  y = y_mixed,
  kernel = "laplace",
  backend = "crp",
  components = 5,
  mcmc = mcmc
)
fit_components <- load_or_fit("ex06-unconditional-DPmix-CRP-fit_components", run_mcmc_bundle_manual(bundle_components))
```


```{r}
summary(fit_components)
```

---

For unconditional models, use `predict()` for posterior summaries; `fitted()` and `residuals()` are not supported.

---

## Model Comparison: Different Kernels

### Laplace Kernel (Current)

```{r comp-laplace, cache=FALSE}
bundle_laplace <- DPmixGPD::build_nimble_bundle(
  y = y_mixed,
  kernel = "laplace",
  backend = "crp",
  components = 5,
  mcmc = mcmc
)
fit_laplace <- load_or_fit("ex06-unconditional-DPmix-CRP-fit_laplace", run_mcmc_bundle_manual(bundle_laplace))
```


```{r}
summary(fit_laplace)
```

### Amoroso Kernel (Alternative)

```{r comp-amoroso, cache=FALSE}
bundle_amoroso <- DPmixGPD::build_nimble_bundle(
  y = y_mixed,
  kernel = "amoroso",
  backend = "crp",
  components = 5,
  mcmc = mcmc
)
fit_amoroso <- load_or_fit("ex06-unconditional-DPmix-CRP-fit_amoroso", run_mcmc_bundle_manual(bundle_amoroso))
```


```{r}
summary(fit_amoroso)
```

### Model Comparison via Predictions

```{r comp-summary, cache=FALSE}
# Compare models using predict() (fitted/residuals not supported for unconditional)
pred_laplace <- predict(fit_laplace, type = "mean", cred.level = 0.9, interval = "credible")
pred_amoroso <- predict(fit_amoroso, type = "mean", cred.level = 0.9, interval = "credible")
```

```{r fig.height=6, fig.width=10, cache=FALSE}
# For unconditional models, fitted() is not supported; use predict() for comparisons (see pred_laplace, pred_amoroso above).
```



## Key Takeaways

- **CRP Backend**: Flexible component allocation, ideal for unknown mixture complexity
- **components Parameter**: Higher components allows more components but increases computation
- **Kernel Choice**: Gamma suitable for positive, skewed data
- **Diagnostics**: Check convergence (Rhat, ESS) and posterior predictive fit
- **Next**: Compare with **Stick-Breaking (SB)** backend in ex07

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex05-unconditional-DPmix-CRP](ex05-unconditional-DPmix-CRP.html)
- Next: [ex07-unconditional-DPmix-SB](ex07-unconditional-DPmix-SB.html)
- Workflow index: [Roadmap](../roadmap.html)
- Practical entry: [Examples](../Examples/index.html)
<!-- workflow-nav:end -->


















