---
title: "ex07. Conditional DPMGPD (CRP Backend)"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache=TRUE,
  cache.path = 'legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(DPmixGPD)

# Ensure package is fully loaded for data access
if (!require("devtools", quietly = TRUE)) {
  library(devtools)
}
pkg_root <- rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
  devtools::load_all(path = pkg_root, quiet = TRUE)

# Precomputed results for fast website builds
source("_legacy-setup.R")
if (!exists("safe_plot", mode = "function")) {
  safe_plot <- function(x, ...) {
    obj <- tryCatch(plot(x, ...), error = function(e) e)
    if (inherits(obj, "ggplot")) {
      try(print(obj), silent = TRUE)
    } else if (is.list(obj)) {
      for (elt in obj) if (inherits(elt, "ggplot")) try(print(elt), silent = TRUE)
    }
    invisible(obj)
  }
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (basic, unconditional, conditional, and causal)..


# Conditional DPmixGPD: CRP Backend with Tail Augmentation

**Purpose**: Combine conditional modeling with GPD tail augmentation so each covariate slice inherits both mixture bulk and tail behavior. This extends the unconditional GPD (ex03) and conditional DP (ex05).

---

## Data Setup

```{r data-setup}
data("nc_posX100_p5_k4", package = "DPmixGPD")
y <- nc_posX100_p5_k4$y
X <- as.matrix(nc_posX100_p5_k4$X)
if (is.null(colnames(X))) {
  colnames(X) <- paste0("x", seq_len(ncol(X)))
}

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

ggplot(data.frame(y = y, x1 = X[, 1]), aes(x = x1, y = y)) +
  geom_point(alpha = 0.5, color = "darkgreen") +
  geom_smooth(method = "loess", color = "steelblue", fill = NA) +
  labs(title = "Outcome vs X1 (Tail dataset)", x = "X1", y = "y") +
  theme_minimal()
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Conditional Tail Dataset Summary", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

---

## Threshold Selection

```{r threshold}
u_threshold <- quantile(y, 0.85)

ggplot(data.frame(y = y), aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "magenta", alpha = 0.6, color = "black") +
  geom_vline(xintercept = u_threshold, linetype = "dashed", color = "black") +
  labs(title = paste("Threshold at", signif(u_threshold, 3)), x = "y", y = "Density") +
  theme_minimal()
```

---

## Model Specification & Bundle

```{r bundle, cache=TRUE}
bundle_cond_gpd_lognormal <- DPmixGPD::build_nimble_bundle(
  y = y,
  X = X,
  kernel = "lognormal",
  backend = "crp",
  GPD = TRUE,
  components = 5,
  param_specs = list(
    gpd = list(
      threshold = list(mode = "link", link = "exp")
    )
  ),
  mcmc = mcmc
)

bundle_cond_gpd_normal <- DPmixGPD::build_nimble_bundle(
  y = y,
  X = X,
  kernel = "normal",
  backend = "crp",
  GPD = TRUE,
  components = 5,
  param_specs = list(
    gpd = list(
      threshold = list(mode = "link", link = "exp")
    )
  ),
  mcmc = mcmc
)
```

---

## Running MCMC

```{r mcmc-run, cache=TRUE}
fit_cond_gpd_lognormal <- load_or_fit("ex07-conditional-dpmgpd-crp-fit_cond_gpd_lognormal", run_mcmc_bundle_manual(bundle_cond_gpd_lognormal))
fit_cond_gpd_normal <- load_or_fit("ex07-conditional-dpmgpd-crp-fit_cond_gpd_normal", run_mcmc_bundle_manual(bundle_cond_gpd_normal))
summary(fit_cond_gpd_lognormal)
summary(fit_cond_gpd_normal)
```

```{r params-cond-gpd-crp}
params_cond_gpd <- params(fit_cond_gpd_lognormal)
params_cond_gpd
```

---

## Conditional Tail-aware Predictions

```{r density-predict}
X_new <- rbind(
  c(-1, 0, 0, 0, 0),
  c(0, 0, 0, 0, 0),
  c(1, 1, 0, 0, 0)
)
colnames(X_new) <- colnames(X)
y_grid <- seq(0, max(y) * 1.2, length.out = 200)

df_pred_lognormal <- lapply(seq_len(nrow(X_new)), function(i) {
  pred <- predict(fit_cond_gpd_lognormal, x = as.matrix(X_new[i, , drop = FALSE]), y = y_grid, type = "density")
  data.frame(
    y = pred$fit$y,
    density = pred$fit$density,
    label = paste("x1=", X_new[i, 1], ", x2=", X_new[i, 2], sep = ""),
    model = "Lognormal"
  )
})

df_pred_normal <- lapply(seq_len(nrow(X_new)), function(i) {
  pred <- predict(fit_cond_gpd_normal, x = as.matrix(X_new[i, , drop = FALSE]), y = y_grid, type = "density")
  data.frame(
    y = pred$fit$y,
    density = pred$fit$density,
    label = paste("x1=", X_new[i, 1], ", x2=", X_new[i, 2], sep = ""),
    model = "Normal"
  )
})

bind_rows(df_pred_lognormal, df_pred_normal) %>%
  ggplot(aes(x = y, y = density, color = label)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ model) +
  labs(title = "Conditional Density with GPD Tail", x = "y", y = "Density") +
  theme_minimal() +
  theme(legend.position = "bottom")
```

---

## Tail Quantiles vs Covariates

```{r quantiles}
X_grid <- cbind(x1 = seq(-1, 1, length.out = 5), x2 = 0, x3 = 0, x4 = 0, x5 = 0)
colnames(X_grid) <- colnames(X)
quant_probs <- c(0.90, 0.95)

pred_q_lognormal <- predict(fit_cond_gpd_lognormal, x = as.matrix(X_grid), type = "quantile", index = quant_probs)
pred_q_normal <- predict(fit_cond_gpd_normal, x = as.matrix(X_grid), type = "quantile", index = quant_probs)

quant_df_lognormal <- pred_q_lognormal$fit
quant_df_lognormal$x1 <- X_grid[quant_df_lognormal$id, "x1"]
quant_df_lognormal$model <- "Lognormal"

quant_df_normal <- pred_q_normal$fit
quant_df_normal$x1 <- X_grid[quant_df_normal$id, "x1"]
quant_df_normal$model <- "Normal"

bind_rows(quant_df_lognormal, quant_df_normal) %>%
  ggplot(aes(x = x1, y = estimate, color = factor(index), group = index)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~ model) +
  labs(title = "Tail Quantiles vs x1 (CRP)", x = "x1", y = "Quantile", color = "Probability") +
  theme_minimal()
```

---

## Residuals & Diagnostics

```{r residuals, cache=TRUE}
safe_plot(fitted(fit_cond_gpd_lognormal))
```

```{r diagnostics}
safe_plot(fit_cond_gpd_lognormal, family = c("traceplot", "density", "autocorrelation"))
safe_plot(fit_cond_gpd_normal, family = c("running", "geweke", "caterpillar"))
```

---

## Takeaways

- Conditional DPmix with a GPD tail lets posterior-mean extreme quantiles vary with covariates.
- The CRP backend samples the bulk and tail jointly while thresholding at the 85th percentile.
- `predict()` + `plot()` remain the main tools for densities, survival curves, and quantiles; residual diagnostics check fit quality.
- Next: Mirror this workflow with the SB backend in `ex08`.

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex06-conditional-dpm-sb](ex06-conditional-dpm-sb.html)
- Next: [ex08-conditional-dpmgpd-sb](ex08-conditional-dpmgpd-sb.html)
- Workflow index: [Roadmap](../start/roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->





















