---
title: "ex09. Causal (No X, CRP Backend) - Gamma Kernel"
execute:
  freeze: true
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = 'legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
max_workers <- local({
  n_phys <- suppressWarnings(tryCatch(parallel::detectCores(logical = FALSE), error = function(e) NA_integer_))
  n_log <- suppressWarnings(tryCatch(parallel::detectCores(logical = TRUE), error = function(e) NA_integer_))
  n <- if (is.finite(n_phys) && !is.na(n_phys) && n_phys >= 1L) n_phys else n_log
  if (!is.finite(n) || is.na(n) || n < 1L) n <- 1L
  as.integer(min(10L, n))
})
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, parallel_arms = TRUE, workers = max_workers)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_arms = TRUE, workers = max_workers, z_update_every = 2)

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(CausalMixGPD)

# Ensure package is fully loaded for data access

# Precomputed results for fast website builds
source("_legacy-setup.R")
max_workers <- min(max_workers, CMGPD_MAX_WORKERS)
causal_run_mcmc <- list(show_progress = FALSE, parallel_arms = TRUE, workers = max_workers, timing = TRUE)
causal_run_mcmc_gpd <- list(show_progress = FALSE, parallel_arms = TRUE, workers = max_workers, timing = TRUE)
components_ex09 <- if (isTRUE(FAST)) 2L else 6L
bulk_mcmc_ex09 <- if (isTRUE(FAST)) {
  list(niter = 80, nburnin = 20, nchains = 1, thin = 1, seed = 1)
} else {
  list(niter = 300, nburnin = 80, nchains = 1, thin = 1, seed = 1)
}
gpd_mcmc_ex09 <- if (isTRUE(FAST)) {
  list(niter = 40, nburnin = 10, nchains = 1, thin = 1, seed = 2)
} else {
  list(niter = 300, nburnin = 80, nchains = 1, thin = 1, seed = 2)
}
predict_workers_ex09 <- max_workers
nsim_mean_ex09 <- if (isTRUE(FAST)) 50L else 150L
tag_crp_bulk <- if (isTRUE(FAST)) "ex09-causal-no-x-crp-fit_crp_bulk_fast" else "ex09-causal-no-x-crp-fit_crp_bulk"
tag_crp_gpd <- if (isTRUE(FAST)) "ex09-causal-no-x-crp-fit_crp_gpd_fast" else "ex09-causal-no-x-crp-fit_crp_gpd"
```


> **Website workflow note.**
> This page reflects the current exported API and recommended wrapper-first usage.
> Last updated: **2026-02-19**.
>
> For the full package narrative, see the main package vignettes (basic, unconditional, conditional, and causal).


# Causal Inference: No Covariates (CRP)

## What you’ll learn

- How causal fits are structured when **no covariates** are available (`X = NULL`).
- How to interpret and report **marginal estimands**: `ate()`, `att()`, `qte()`, `qtt()`.
- How tail augmentation (`GPD = TRUE`) changes effect behavior in the extremes.

## Key limitation (important)

Because `X = NULL`, conditional estimands `cate()` and `cqte()` are unavailable. This is a feature: it prevents interpreting conditional effects without covariates.

## What this page does

This vignette fits **two no-X causal models** where the treatment arms are modeled independently using unconditional outcome models:

- Model A: CRP bulk-only (GPD = FALSE)
- Model B: CRP with GPD tail (GPD = TRUE)

---

## Data Setup (No X)

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- abs(causal_alt_real500_p4_k2$y) + 0.01
A <- causal_alt_real500_p4_k2$A

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  print()

u_threshold <- as.numeric(stats::quantile(y, 0.8, names = FALSE))
y_eval <- y[seq_len(if (isTRUE(FAST)) 20L else 40L)]
```

---

## Model A: CRP Bulk-only (Gamma)

```{r bundle-crp-bulk, cache = FALSE}
bundle_crp_bulk <- bundle(
  y = y,
  treat = A,
  X = NULL,
  kernel = "gamma",
  backend = "crp",
  PS = FALSE,
  GPD = FALSE,
  components = components_ex09,
  mcmc_outcome = bulk_mcmc_ex09
)

bundle_crp_bulk
```

```{r fit-crp-bulk, cache = FALSE}
fit_crp_bulk <- load_or_fit(
  tag_crp_bulk,
  quiet_mcmc(dpmix.causal(bundle_crp_bulk, mcmc = causal_run_mcmc))
)
summary(fit_crp_bulk)
```

```{r predict-mean-crp-bulk}
pred_mean_bulk <- predict(fit_crp_bulk, type = "mean", interval = "credible", nsim_mean = nsim_mean_ex09, workers = predict_workers_ex09)
head(pred_mean_bulk)
plot(pred_mean_bulk)
```

```{r predict-quantile-crp-bulk}
pred_q_bulk <- predict(fit_crp_bulk, type = "quantile", p = 0.5, interval = "credible", workers = predict_workers_ex09)
head(pred_q_bulk)
plot(pred_q_bulk)
```

```{r predict-density-crp-bulk}
pred_d_bulk <- predict(fit_crp_bulk, y = y_eval, type = "density", interval = "credible", workers = predict_workers_ex09)
head(pred_d_bulk)
plot(pred_d_bulk)
```

```{r predict-survival-crp-bulk}
pred_surv_bulk <- predict(fit_crp_bulk, y = y_eval, type = "survival", interval = "credible", workers = predict_workers_ex09)
head(pred_surv_bulk)
plot(pred_surv_bulk)
```

```{r ate-crp-bulk}
ate_bulk <- ate(fit_crp_bulk, interval = "credible", nsim_mean = nsim_mean_ex09)
head(ate_bulk)
plot(ate_bulk)
```

```{r qte-crp-bulk}
qte_bulk <- qte(fit_crp_bulk, probs = c(0.25, 0.5, 0.75), interval = "credible")
head(qte_bulk)
plot(qte_bulk)
```

---

## Model B: CRP with GPD Tail (Gamma)

```{r bundle-crp-gpd, cache = FALSE}
param_specs_gpd <- list(
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "lognormal",
      args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
    )
  )
)

bundle_crp_gpd <- bundle(
  y = y,
  treat = A,
  X = NULL,
  kernel = "gamma",
  backend = "crp",
  PS = FALSE,
  GPD = TRUE,
  components = components_ex09,
  param_specs = param_specs_gpd,
  mcmc_outcome = gpd_mcmc_ex09
)

bundle_crp_gpd
```

```{r fit-crp-gpd, cache = FALSE}
fit_crp_gpd <- load_or_fit(
  tag_crp_gpd,
  quiet_mcmc(dpmgpd.causal(bundle_crp_gpd, mcmc = causal_run_mcmc_gpd))
)
summary(fit_crp_gpd)
```

```{r predict-mean-crp-gpd}
pred_mean_gpd <- predict(fit_crp_gpd, type = "mean", interval = "credible", nsim_mean = nsim_mean_ex09, workers = predict_workers_ex09)
head(pred_mean_gpd)
plot(pred_mean_gpd)
```

```{r predict-quantile-crp-gpd}
pred_q_gpd <- predict(fit_crp_gpd, type = "quantile", p = 0.5, interval = "credible", workers = predict_workers_ex09)
head(pred_q_gpd)
plot(pred_q_gpd)
```

```{r predict-density-crp-gpd}
pred_d_gpd <- predict(fit_crp_gpd, y = y_eval, type = "density", interval = "credible", workers = predict_workers_ex09)
head(pred_d_gpd)
plot(pred_d_gpd)
```

```{r predict-survival-crp-gpd}
pred_surv_gpd <- predict(fit_crp_gpd, y = y_eval, type = "survival", interval = "credible", workers = predict_workers_ex09)
head(pred_surv_gpd)
plot(pred_surv_gpd)
```

```{r ate-crp-gpd}
ate_gpd <- ate(fit_crp_gpd, interval = "credible", nsim_mean = nsim_mean_ex09)
head(ate_gpd)
plot(ate_gpd)
```

```{r qte-crp-gpd}
qte_gpd <- qte(fit_crp_gpd, probs = c(0.25, 0.5, 0.75), interval = "credible")
head(qte_gpd)
plot(qte_gpd)
```

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex08-conditional-dpmgpd-sb](ex08-conditional-dpmgpd-sb.html)
- Next: [ex10-causal-x-no-ps-sb](ex10-causal-x-no-ps-sb.html)
- Workflow index: [Roadmap](../start/roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [03 Causal Inference Objects](../advanced/theory/03-causal-inference-objects.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->
