---
title: "ex11. Causal (Same backend, CRP Backend) - Laplace Kernel"
execute:
  freeze: true
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = 'legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
max_workers <- local({
  n_phys <- suppressWarnings(tryCatch(parallel::detectCores(logical = FALSE), error = function(e) NA_integer_))
  n_log <- suppressWarnings(tryCatch(parallel::detectCores(logical = TRUE), error = function(e) NA_integer_))
  n <- if (is.finite(n_phys) && !is.na(n_phys) && n_phys >= 1L) n_phys else n_log
  if (!is.finite(n) || is.na(n) || n < 1L) n <- 1L
  as.integer(min(10L, n))
})
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, parallel_arms = TRUE, workers = max_workers)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_arms = TRUE, workers = max_workers, z_update_every = 2)

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(CausalMixGPD)

# Ensure package is fully loaded for data access

# Precomputed results for fast website builds
source("_legacy-setup.R")
max_workers <- min(max_workers, CMGPD_MAX_WORKERS)
causal_run_mcmc_bulk <- list(show_progress = FALSE, parallel_arms = TRUE, workers = max_workers, timing = TRUE)
causal_run_mcmc_gpd <- list(
  show_progress = FALSE,
  parallel_arms = if (isTRUE(FAST)) FALSE else TRUE,
  workers = if (isTRUE(FAST)) 1L else max_workers,
  timing = TRUE
)
components_ex11 <- if (isTRUE(FAST)) 3L else 5L
bulk_mcmc_ex11 <- if (isTRUE(FAST)) {
  list(niter = 120, nburnin = 30, nchains = 1, thin = 1, seed = 1)
} else {
  list(niter = 400, nburnin = 100, nchains = 1, thin = 2, seed = 1)
}
gpd_mcmc_ex11 <- if (isTRUE(FAST)) {
  list(niter = 80, nburnin = 20, nchains = 1, thin = 1, seed = 2)
} else {
  list(niter = 400, nburnin = 100, nchains = 1, thin = 2, seed = 2)
}
mcmc_ps_bulk_ex11 <- bulk_mcmc_ex11
mcmc_ps_gpd_ex11 <- gpd_mcmc_ex11
predict_workers_ex11 <- if (isTRUE(FAST)) 1L else max_workers
nsim_mean_ex11 <- if (isTRUE(FAST)) 50L else 100L
plot_family_bulk_ex11 <- if (isTRUE(FAST)) "traceplot" else c("traceplot", "autocorrelation", "running")
plot_family_gpd_ex11 <- if (isTRUE(FAST)) "density" else c("density", "geweke", "caterpillar")
tag_crp_bulk_ex11 <- if (isTRUE(FAST)) "ex11-causal-same-backend-crp-fit_crp_bulk_laplace_fast" else "ex11-causal-same-backend-crp-fit_crp_bulk_laplace"
tag_crp_gpd_ex11 <- if (isTRUE(FAST)) "ex11-causal-same-backend-crp-fit_crp_gpd_laplace_fast" else "ex11-causal-same-backend-crp-fit_crp_gpd_laplace"
```

> **Website workflow note.**
> This page reflects the current exported API and recommended wrapper-first usage.
> Last updated: **2026-02-19**.
>
> For the full package narrative, see the main package vignettes (basic, unconditional, conditional, and causal).

# Causal Inference: Same Backend (CRP) - Laplace Kernel {#causal-crp-same-backend}

This vignette fits two CRP-based causal models using the same kernel (Laplace):

- Model A: bulk-only outcome models (`GPD = FALSE`)
- Model B: GPD-augmented outcome models (`GPD = TRUE`)

Each model demonstrates the standard causal S3 workflow: bundle, fit, predict, plot, ATE, and QTE.

## What you'll learn

- How to hold backend and kernel fixed and isolate the impact of tail augmentation.
- How tail augmentation can change upper-quantile behavior even when mean effects look similar.
- How to run the standard causal post-fit workflow for multiple estimands.

## When to use this template

- You want a clean sensitivity check focused on tail modeling rather than backend or kernel changes.
- You want predictable runtime and an explicit complexity knob (`components`) inside a causal workflow.

## Next steps

- Compare this page with the SB version to study how tail augmentation behaves under a different backend.

---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- causal_alt_real500_p4_k2$y
A <- causal_alt_real500_p4_k2$A
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl

x_eval <- X[seq_len(if (isTRUE(FAST)) 20L else 40L), , drop = FALSE]
y_eval <- y[seq_len(if (isTRUE(FAST)) 20L else 40L)]
u_threshold <- as.numeric(stats::quantile(y, 0.8, names = FALSE))
```

```{r data-scatter}
df_causal <- data.frame(y = y, A = as.factor(A), x1 = X[, 1], x2 = X[, 2])

p_scatter <- ggplot(df_causal, aes(x = x1, y = y, color = A)) +
  geom_point(alpha = 0.5) +
  labs(title = "Outcome vs X1 by Treatment", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()

p_scatter
```

---

## Model A: CRP Bulk-only (Laplace)

```{r bundle-crp-bulk, cache = FALSE}
bundle_crp_bulk <- bundle(
  y = y,
  treat = A,
  X = X,
  kernel = "laplace",
  backend = "crp",
  PS = "logit",
  GPD = FALSE,
  components = components_ex11,
  mcmc_outcome = bulk_mcmc_ex11,
  mcmc_ps = mcmc_ps_bulk_ex11
)

bundle_crp_bulk
summary(bundle_crp_bulk)
```

```{r fit-crp-bulk, cache = FALSE}
fit_crp_bulk <- load_or_fit(
  tag_crp_bulk_ex11,
  quiet_mcmc(dpmix.causal(bundle_crp_bulk, mcmc = causal_run_mcmc_bulk))
)
summary(fit_crp_bulk)
params(fit_crp_bulk)
```

```{r plot-fit-crp-bulk}
plot(fit_crp_bulk, family = plot_family_bulk_ex11)
```

```{r predict-mean-crp-bulk}
pred_mean_bulk <- predict(fit_crp_bulk, newdata = x_eval, type = "mean",
                          interval = "credible", nsim_mean = nsim_mean_ex11, workers = predict_workers_ex11)
head(pred_mean_bulk)
plot(pred_mean_bulk)
```

```{r predict-quantile-crp-bulk}
pred_q_bulk <- predict(fit_crp_bulk, newdata = x_eval, type = "quantile",
                       p = 0.5, interval = "credible", workers = predict_workers_ex11)
head(pred_q_bulk)
plot(pred_q_bulk)
```

```{r predict-density-crp-bulk}
pred_d_bulk <- predict(fit_crp_bulk, newdata = x_eval, y = y_eval,
                       type = "density", interval = "credible", workers = predict_workers_ex11)
head(pred_d_bulk)
plot(pred_d_bulk)
```

```{r predict-survival-crp-bulk}
pred_surv_bulk <- predict(fit_crp_bulk, newdata = x_eval, y = y_eval,
                          type = "survival", interval = "credible", workers = predict_workers_ex11)
head(pred_surv_bulk)
plot(pred_surv_bulk)
```

```{r ate-crp-bulk}
ate_bulk <- ate(fit_crp_bulk, interval = "credible", nsim_mean = nsim_mean_ex11)
head(ate_bulk)
plot(ate_bulk)
```

```{r qte-crp-bulk}
qte_bulk <- qte(fit_crp_bulk, probs = c(0.25, 0.5, 0.75), interval = "credible")
head(qte_bulk)
plot(qte_bulk)
```

---

## Model B: CRP with GPD Tail (Laplace)

```{r bundle-crp-gpd, cache = FALSE}
param_specs_gpd <- list(
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "lognormal",
      args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
    )
  )
)

bundle_crp_gpd <- bundle(
  y = y,
  treat = A,
  X = X,
  kernel = "laplace",
  backend = "crp",
  PS = "logit",
  GPD = TRUE,
  components = components_ex11,
  param_specs = param_specs_gpd,
  mcmc_outcome = gpd_mcmc_ex11,
  mcmc_ps = mcmc_ps_gpd_ex11
)

bundle_crp_gpd
summary(bundle_crp_gpd)
```

```{r fit-crp-gpd, cache = FALSE}
fit_crp_gpd <- load_or_fit(
  tag_crp_gpd_ex11,
  quiet_mcmc(dpmgpd.causal(bundle_crp_gpd, mcmc = causal_run_mcmc_gpd))
)
summary(fit_crp_gpd)
params(fit_crp_gpd)
```

```{r plot-fit-crp-gpd}
plot(fit_crp_gpd, family = plot_family_gpd_ex11)
```

```{r predict-mean-crp-gpd}
pred_mean_gpd <- predict(fit_crp_gpd, newdata = x_eval, type = "mean",
                         interval = "credible", nsim_mean = nsim_mean_ex11, workers = predict_workers_ex11)
head(pred_mean_gpd)
plot(pred_mean_gpd)
```

```{r predict-quantile-crp-gpd}
pred_q_gpd <- predict(fit_crp_gpd, newdata = x_eval, type = "quantile",
                      p = 0.5, interval = "credible", workers = predict_workers_ex11)
head(pred_q_gpd)
plot(pred_q_gpd)
```

```{r predict-density-crp-gpd}
pred_d_gpd <- predict(fit_crp_gpd, newdata = x_eval, y = y_eval,
                      type = "density", interval = "credible", workers = predict_workers_ex11)
head(pred_d_gpd)
plot(pred_d_gpd)
```

```{r predict-survival-crp-gpd}
pred_surv_gpd <- predict(fit_crp_gpd, newdata = x_eval, y = y_eval,
                         type = "survival", interval = "credible", workers = predict_workers_ex11)
head(pred_surv_gpd)
plot(pred_surv_gpd)
```

```{r ate-crp-gpd}
ate_gpd <- ate(fit_crp_gpd, interval = "credible", nsim_mean = nsim_mean_ex11)
head(ate_gpd)
plot(ate_gpd)
```

```{r qte-crp-gpd}
qte_gpd <- qte(fit_crp_gpd, probs = c(0.25, 0.5, 0.75), interval = "credible")
head(qte_gpd)
plot(qte_gpd)
```

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex10-causal-x-no-ps-sb](ex10-causal-x-no-ps-sb.html)
- Next: [ex12-causal-same-backend-sb](ex12-causal-same-backend-sb.html)
- Workflow index: [Roadmap](../start/roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [03 Causal Inference Objects](../advanced/theory/03-causal-inference-objects.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->
