---
title: "ex11. Causal (Same backend, CRP Backend) - Cauchy/Laplace"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = 'legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, workers = 2)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_arms = TRUE, workers = 2, z_update_every = 2)

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(CausalMixGPD)

# Ensure package is fully loaded for data access

# Precomputed results for fast website builds
source("_legacy-setup.R")
```


> **Website workflow note.**
> This page reflects the current exported API and recommended wrapper-first usage.
> Last updated: **2026-02-19**.
>
> For the full package narrative, see the main package vignettes (basic, unconditional, conditional, and causal).


# Causal Inference: Mixed Backends (CRP+SB) {#causal-mixed-backends}

This vignette fits two causal models using different backends and kernels:

- Model A: CRP bulk-only (Cauchy)
- Model B: SB with GPD tail (Laplace)

Each model demonstrates the full causal S3 workflow: bundle, fit, predict, plot, ATE, and QTE.

## What you’ll learn

- How backend choice (CRP vs SB) and tail choice (`GPD = FALSE/TRUE`) affect predictive uncertainty and downstream causal estimands.
- How to keep comparisons interpretable by using the same post-fit S3 surface across models (`summary()`, `predict()`, `plot()`).
- How to read both mean and quantile effect summaries and treat them as model-based functionals.
## When to use this template

- You want an apples-to-apples comparison of causal models that differ by **backend** and/or **tail augmentation**.
- You want a compact example of the full causal S3 workflow in one place.
## Next steps

- Use the “same-backend” causal examples to isolate tail effects from backend effects.
---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- causal_alt_real500_p4_k2$y
A <- causal_alt_real500_p4_k2$A
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl

x_eval <- X[1:40, , drop = FALSE]
y_eval <- y[1:40]
u_threshold <- as.numeric(stats::quantile(y, 0.8, names = FALSE))
```

```{r data-scatter}
df_causal <- data.frame(y = y, A = as.factor(A), x1 = X[, 1], x2 = X[, 2])

p_scatter <- ggplot(df_causal, aes(x = x1, y = y, color = A)) +
  geom_point(alpha = 0.5) +
  labs(title = "Outcome vs X1 by Treatment", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()

p_scatter
```

---

## Model A: CRP Bulk-only (Cauchy)

```{r bundle-crp-bulk, cache = FALSE}
bundle_crp_bulk <- bundle(
  x = y,
  treat = A,
  X = X,
  kernel = "cauchy",
  backend = "crp",
  PS = "logit",
  GPD = FALSE,
  components = 5,
  mcmc_outcome = list(niter = 400, nburnin = 100, nchains = 1, thin = 2, seed = 1),
  mcmc_ps = list(niter = 400, nburnin = 100, nchains = 1, thin = 2, seed = 1)
)

bundle_crp_bulk
summary(bundle_crp_bulk)
```

```{r fit-crp-bulk, cache = FALSE}
fit_crp_bulk <- load_or_fit(
  "ex11-causal-same-backend-crp-fit_crp_bulk",
  quiet_mcmc(dpmix.causal(bundle_crp_bulk))
)
summary(fit_crp_bulk)
params(fit_crp_bulk)
```

```{r plot-fit-crp-bulk}
plot(fit_crp_bulk, family = c("traceplot", "autocorrelation", "running"))
```

```{r predict-mean-crp-bulk}
pred_mean_bulk <- predict(fit_crp_bulk, x = x_eval, type = "mean",
                          interval = "credible", nsim_mean = 100)
head(pred_mean_bulk)
plot(pred_mean_bulk)
```

```{r predict-quantile-crp-bulk}
pred_q_bulk <- predict(fit_crp_bulk, x = x_eval, type = "quantile",
                       p = 0.5, interval = "credible")
head(pred_q_bulk)
plot(pred_q_bulk)
```

```{r predict-density-crp-bulk}
pred_d_bulk <- predict(fit_crp_bulk, x = x_eval, y = y_eval,
                       type = "density", interval = "credible")
head(pred_d_bulk)
plot(pred_d_bulk)
```

```{r predict-survival-crp-bulk}
pred_surv_bulk <- predict(fit_crp_bulk, x = x_eval, y = y_eval,
                          type = "survival", interval = "credible")
head(pred_surv_bulk)
plot(pred_surv_bulk)
```

```{r ate-crp-bulk}
ate_bulk <- ate(fit_crp_bulk, interval = "credible", nsim_mean = 100)
head(ate_bulk)
plot(ate_bulk)
```

```{r qte-crp-bulk}
qte_bulk <- qte(fit_crp_bulk, probs = c(0.25, 0.5, 0.75), interval = "credible")
head(qte_bulk)
plot(qte_bulk)
```

---

## Model B: SB with GPD Tail (Laplace)

```{r bundle-crp-gpd, cache = FALSE}
param_specs_gpd <- list(
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "lognormal",
      args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
    )
  )
)

bundle_sb_gpd <- bundle(
  x = y,
  treat = A,
  X = X,
  kernel = "laplace",
  backend = "sb",
  PS = "logit",
  GPD = TRUE,
  components = 5,
  param_specs = param_specs_gpd,
  mcmc_outcome = list(niter = 400, nburnin = 100, nchains = 1, thin = 2, seed = 2),
  mcmc_ps = list(niter = 400, nburnin = 100, nchains = 1, thin = 2, seed = 2)
)

bundle_sb_gpd
summary(bundle_sb_gpd)
```

```{r fit-sb-gpd, cache = FALSE}
fit_sb_gpd <- load_or_fit(
  "ex11-causal-same-backend-crp-fit_sb_gpd",
  quiet_mcmc(dpmgpd.causal(bundle_sb_gpd))
)
summary(fit_sb_gpd)
params(fit_sb_gpd)
```

```{r plot-fit-sb-gpd}
plot(fit_sb_gpd, family = c("density", "geweke", "caterpillar"))
```

```{r predict-mean-sb-gpd}
pred_mean_gpd <- predict(fit_sb_gpd, x = x_eval, type = "mean",
                         interval = "credible", nsim_mean = 100)
head(pred_mean_gpd)
plot(pred_mean_gpd)
```

```{r predict-quantile-sb-gpd}
pred_q_gpd <- predict(fit_sb_gpd, x = x_eval, type = "quantile",
                      p = 0.5, interval = "credible")
head(pred_q_gpd)
plot(pred_q_gpd)
```

```{r predict-density-sb-gpd}
pred_d_gpd <- predict(fit_sb_gpd, x = x_eval, y = y_eval,
                      type = "density", interval = "credible")
head(pred_d_gpd)
plot(pred_d_gpd)
```

```{r predict-survival-sb-gpd}
pred_surv_gpd <- predict(fit_sb_gpd, x = x_eval, y = y_eval,
                         type = "survival", interval = "credible")
head(pred_surv_gpd)
plot(pred_surv_gpd)
```

```{r cate-sb-gpd}
cate_gpd <- cate(fit_sb_gpd, newdata = x_eval,
                interval = "credible", nsim_mean = 100)
head(format_cate_table(cate_gpd))
plot(cate_gpd)
```

```{r cqte-sb-gpd}
cqte_gpd <- cqte(fit_sb_gpd, probs = c(0.25, 0.5, 0.75),
                newdata = x_eval, interval = "credible")
head(format_cqte_table(cqte_gpd))
plot(cqte_gpd)
```

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex10-causal-x-no-ps-sb](ex10-causal-x-no-ps-sb.html)
- Next: [ex12-causal-same-backend-sb](ex12-causal-same-backend-sb.html)
- Workflow index: [Roadmap](../start/roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [03 Causal Inference Objects](../advanced/theory/03-causal-inference-objects.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->


