---
title: "ex11. Causal (Same backend, CRP Backend) - Cauchy/Laplace"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache=TRUE,
  cache.path = 'legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(DPmixGPD)

# Ensure package is fully loaded for data access
if (!require("devtools", quietly = TRUE)) {
  library(devtools)
}
pkg_root <- rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
  devtools::load_all(path = pkg_root, quiet = TRUE)

# Precomputed results for fast website builds
source("_legacy-setup.R")
if (!exists("safe_plot", mode = "function")) {
  safe_plot <- function(x, ...) {
    obj <- tryCatch(plot(x, ...), error = function(e) e)
    if (inherits(obj, "ggplot")) {
      try(print(obj), silent = TRUE)
    } else if (is.list(obj)) {
      for (elt in obj) if (inherits(elt, "ggplot")) try(print(elt), silent = TRUE)
    }
    invisible(obj)
  }
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (basic, unconditional, conditional, and causal)..


# Causal Inference: Mixed Backends (CRP+SB) {#causal-mixed-backends}

This vignette fits two causal models using different backends and kernels:

- Model A: CRP bulk-only (Cauchy)
- Model B: SB with GPD tail (Laplace)

Each model demonstrates the full causal S3 workflow: bundle, fit, predict, plot, ATE, and QTE.

---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2", package = "DPmixGPD")
y <- causal_alt_real500_p4_k2$y
T <- causal_alt_real500_p4_k2$T
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Causal Dataset Summary", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")

x_eval <- X[1:40, , drop = FALSE]
y_eval <- y[1:40]
u_threshold <- as.numeric(stats::quantile(y, 0.8, names = FALSE))
```

```{r data-scatter}
df_causal <- data.frame(y = y, T = as.factor(T), x1 = X[, 1], x2 = X[, 2])

p_scatter <- ggplot(df_causal, aes(x = x1, y = y, color = T)) +
  geom_point(alpha = 0.5) +
  labs(title = "Outcome vs X1 by Treatment", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()

p_scatter
```

---

## Model A: CRP Bulk-only (Cauchy)

```{r bundle-crp-bulk, cache=TRUE}
bundle_crp_bulk <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "cauchy",
  backend = "crp",
  PS = "logit",
  GPD = FALSE,
  components = 5,
  mcmc_outcome = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 1),
  mcmc_ps = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 1)
)

bundle_crp_bulk
summary(bundle_crp_bulk)
```

```{r fit-crp-bulk, cache=TRUE}
fit_crp_bulk <- quiet_mcmc(run_mcmc_causal(bundle_crp_bulk))
summary(fit_crp_bulk)
params(fit_crp_bulk)
```

```{r plot-fit-crp-bulk}
safe_plot(fit_crp_bulk, family = c("traceplot", "autocorrelation", "running"))
```

```{r predict-mean-crp-bulk}
pred_mean_bulk <- predict(fit_crp_bulk, x = x_eval, type = "mean",
                          interval = "credible", nsim_mean = 100)
head(pred_mean_bulk)
safe_plot(pred_mean_bulk)
```

```{r predict-quantile-crp-bulk}
pred_q_bulk <- predict(fit_crp_bulk, x = x_eval, type = "quantile",
                       p = 0.5, interval = "credible")
head(pred_q_bulk)
safe_plot(pred_q_bulk)
```

```{r predict-density-crp-bulk}
pred_d_bulk <- predict(fit_crp_bulk, x = x_eval, y = y_eval,
                       type = "density", interval = "credible")
head(pred_d_bulk)
safe_plot(pred_d_bulk)
```

```{r predict-survival-crp-bulk}
pred_surv_bulk <- predict(fit_crp_bulk, x = x_eval, y = y_eval,
                          type = "survival", interval = "credible")
head(pred_surv_bulk)
safe_plot(pred_surv_bulk)
```

```{r ate-crp-bulk}
ate_bulk <- ate(fit_crp_bulk, interval = "credible", nsim_mean = 100)
head(ate_bulk)
safe_plot(ate_bulk)
```

```{r qte-crp-bulk}
qte_bulk <- qte(fit_crp_bulk, probs = c(0.25, 0.5, 0.75), interval = "credible")
head(qte_bulk)
safe_plot(qte_bulk)
```

---

## Model B: SB with GPD Tail (Laplace)

```{r bundle-crp-gpd, cache=TRUE}
param_specs_gpd <- list(
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "lognormal",
      args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
    )
  )
)

bundle_sb_gpd <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "laplace",
  backend = "sb",
  PS = "logit",
  GPD = TRUE,
  components = 5,
  param_specs = param_specs_gpd,
  mcmc_outcome = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 2),
  mcmc_ps = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 2)
)

bundle_sb_gpd
summary(bundle_sb_gpd)
```

```{r fit-sb-gpd, cache=TRUE}
fit_sb_gpd <- quiet_mcmc(run_mcmc_causal(bundle_sb_gpd))
summary(fit_sb_gpd)
params(fit_sb_gpd)
```

```{r plot-fit-sb-gpd}
safe_plot(fit_sb_gpd, family = c("density", "geweke", "caterpillar"))
```

```{r predict-mean-sb-gpd}
pred_mean_gpd <- predict(fit_sb_gpd, x = x_eval, type = "mean",
                         interval = "credible", nsim_mean = 100)
head(pred_mean_gpd)
safe_plot(pred_mean_gpd)
```

```{r predict-quantile-sb-gpd}
pred_q_gpd <- predict(fit_sb_gpd, x = x_eval, type = "quantile",
                      p = 0.5, interval = "credible")
head(pred_q_gpd)
safe_plot(pred_q_gpd)
```

```{r predict-density-sb-gpd}
pred_d_gpd <- predict(fit_sb_gpd, x = x_eval, y = y_eval,
                      type = "density", interval = "credible")
head(pred_d_gpd)
safe_plot(pred_d_gpd)
```

```{r predict-survival-sb-gpd}
pred_surv_gpd <- predict(fit_sb_gpd, x = x_eval, y = y_eval,
                         type = "survival", interval = "credible")
head(pred_surv_gpd)
safe_plot(pred_surv_gpd)
```

```{r ate-sb-gpd}
ate_gpd <- cate(fit_sb_gpd, newdata = x_eval,
               interval = "credible", nsim_mean = 100)
head(ate_gpd)
safe_plot(ate_gpd)
```

```{r qte-sb-gpd}
qte_gpd <- cqte(fit_sb_gpd, probs = c(0.25, 0.5, 0.75),
               newdata = x_eval, interval = "credible")
head(qte_gpd)
safe_plot(qte_gpd)
```

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex10-causal-x-no-ps-sb](ex10-causal-x-no-ps-sb.html)
- Next: [ex12-causal-same-backend-sb](ex12-causal-same-backend-sb.html)
- Workflow index: [Roadmap](../start/roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->





















