---
title: "ex13. Causal (Different backends, CRP Backend) - Cauchy Kernel"
execute:
  freeze: true
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = 'legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
max_workers <- local({
  n_phys <- suppressWarnings(tryCatch(parallel::detectCores(logical = FALSE), error = function(e) NA_integer_))
  n_log <- suppressWarnings(tryCatch(parallel::detectCores(logical = TRUE), error = function(e) NA_integer_))
  n <- if (is.finite(n_phys) && !is.na(n_phys) && n_phys >= 1L) n_phys else n_log
  if (!is.finite(n) || is.na(n) || n < 1L) n <- 1L
  as.integer(min(10L, n))
})
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, parallel_arms = TRUE, workers = max_workers)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_arms = TRUE, workers = max_workers, z_update_every = 2)

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(CausalMixGPD)

# Ensure package is fully loaded for data access

# Precomputed results for fast website builds
source("_legacy-setup.R")
max_workers <- min(max_workers, CMGPD_MAX_WORKERS)
causal_run_mcmc <- list(show_progress = FALSE, parallel_arms = TRUE, workers = max_workers, timing = TRUE)
components_ex13 <- if (isTRUE(FAST)) c(3L, 3L) else c(5L, 5L)
bulk_mcmc_ex13 <- if (isTRUE(FAST)) {
  list(niter = 120, nburnin = 30, nchains = 1, thin = 1, seed = 1)
} else {
  list(niter = 400, nburnin = 100, nchains = 1, thin = 2, seed = 1)
}
bulk_mcmc_ex13_b <- if (isTRUE(FAST)) {
  list(niter = 120, nburnin = 30, nchains = 1, thin = 1, seed = 2)
} else {
  list(niter = 400, nburnin = 100, nchains = 1, thin = 2, seed = 2)
}
mcmc_ps_ex13 <- bulk_mcmc_ex13
mcmc_ps_ex13_b <- bulk_mcmc_ex13_b
predict_workers_ex13 <- max_workers
nsim_mean_ex13 <- if (isTRUE(FAST)) 50L else 100L
rmean_cutoff_ex13 <- 10
plot_family_sb_crp_ex13 <- if (isTRUE(FAST)) "traceplot" else c("traceplot", "autocorrelation", "running")
plot_family_crp_sb_ex13 <- if (isTRUE(FAST)) "density" else c("density", "geweke", "caterpillar")
tag_sb_crp_ex13 <- if (isTRUE(FAST)) "ex13-causal-different-backends-crp-fit_sb_crp_fast" else "ex13-causal-different-backends-crp-fit_sb_crp"
tag_crp_sb_ex13 <- if (isTRUE(FAST)) "ex13-causal-different-backends-crp-fit_crp_sb_fast" else "ex13-causal-different-backends-crp-fit_crp_sb"
```


> **Website workflow note.**
> This page reflects the current exported API and recommended wrapper-first usage.
> Last updated: **2026-02-19**.
>
> For the full package narrative, see the main package vignettes (basic, unconditional, conditional, and causal).


# Causal Inference: Mixed Backends (Bulk-only) - Cauchy Kernel {#causal-different-backends-bulk}

This vignette fits two mixed-backend causal models with a shared Cauchy kernel:

- Model A: Treated = SB, Control = CRP
- Model B: Treated = CRP, Control = SB

## What you’ll learn

- How to specify **arm-specific backends** (treated vs control) within one causal bundle.
- How to interpret causal predictions/effects when the two potential-outcome models have different partition structures.
- How to keep the comparison focused on backend differences by using a shared kernel and `GPD = FALSE`.

## When to use this template

- You suspect treated and control outcomes differ in ways that make one backend preferable for one arm.
- You want to study how much causal estimands change under **backend asymmetry** while keeping tail modeling off.

## Next steps

- Turn on tail augmentation (mixed-backend + `GPD = TRUE`) when extremes plausibly drive effect heterogeneity.
- Because the Cauchy kernel does not have a finite mean, this page uses median and upper-quantile predictions, plus restricted-mean causal effects.
---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- causal_alt_real500_p4_k2$y
A <- causal_alt_real500_p4_k2$A
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl

x_eval <- X[seq_len(if (isTRUE(FAST)) 20L else 40L), , drop = FALSE]
y_eval <- y[seq_len(if (isTRUE(FAST)) 20L else 40L)]
```

```{r data-scatter}
df_causal <- data.frame(y = y, A = as.factor(A), x1 = X[, 1])

p_scatter <- ggplot(df_causal, aes(x = x1, y = y, color = A)) +
  geom_point(alpha = 0.5) +
  labs(title = "Outcome vs X1", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()

p_scatter
```

---

## Model A: Treated SB, Control CRP (Cauchy)

```{r bundle-sb-crp, cache = FALSE}
bundle_sb_crp <- bundle(
  y = y,
  treat = A,
  X = X,
  kernel = c("cauchy", "cauchy"),
  backend = c("sb", "crp"),
  PS = "logit",
  GPD = c(FALSE, FALSE),
  components = components_ex13,
  mcmc_outcome = bulk_mcmc_ex13,
  mcmc_ps = mcmc_ps_ex13
)

bundle_sb_crp
summary(bundle_sb_crp)
```

```{r fit-sb-crp, cache = FALSE}
fit_sb_crp <- load_or_fit(
  tag_sb_crp_ex13,
  quiet_mcmc(dpmix.causal(bundle_sb_crp, mcmc = causal_run_mcmc))
)
summary(fit_sb_crp)
params(fit_sb_crp)
```

```{r plot-fit-sb-crp}
plot(fit_sb_crp, family = plot_family_sb_crp_ex13)
```

```{r predict-mean-sb-crp}
pred_median_sb_crp <- predict(
  fit_sb_crp,
  newdata =x_eval,
  type = "quantile",
  p = 0.5,
  interval = "credible",
  workers = predict_workers_ex13
)
plot(pred_median_sb_crp)
```

```{r predict-quantile-sb-crp}
pred_q_sb_crp <- predict(fit_sb_crp, newdata =x_eval, type = "quantile",
                         p = 0.9, interval = "credible", workers = predict_workers_ex13)
plot(pred_q_sb_crp)
```

```{r predict-density-sb-crp}
pred_d_sb_crp <- predict(fit_sb_crp, newdata =x_eval, y = y_eval,
                         type = "density", interval = "credible", workers = predict_workers_ex13)
plot(pred_d_sb_crp)
```

```{r predict-survival-sb-crp}
pred_surv_sb_crp <- predict(fit_sb_crp, newdata =x_eval, y = y_eval,
                            type = "survival", interval = "credible", workers = predict_workers_ex13)
plot(pred_surv_sb_crp)
```

```{r ate-sb-crp}
att_sb_crp <- att(
  fit_sb_crp,
  type = "rmean",
  cutoff = rmean_cutoff_ex13,
  interval = "credible",
  nsim_mean = nsim_mean_ex13
)
plot(att_sb_crp)
```

```{r qte-sb-crp}
qtt_sb_crp <- qtt(fit_sb_crp, probs = c(0.25, 0.5, 0.75), interval = "credible")
plot(qtt_sb_crp)
```

---

## Model B: Treated CRP, Control SB (Cauchy)

```{r bundle-crp-sb, cache = FALSE}
bundle_crp_sb <- bundle(
  y = y,
  treat = A,
  X = X,
  kernel = c("cauchy", "cauchy"),
  backend = c("crp", "sb"),
  PS = "logit",
  GPD = c(FALSE, FALSE),
  components = components_ex13,
  mcmc_outcome = bulk_mcmc_ex13_b,
  mcmc_ps = mcmc_ps_ex13_b
)

bundle_crp_sb
summary(bundle_crp_sb)
```

```{r fit-crp-sb, cache = FALSE}
fit_crp_sb <- load_or_fit(
  tag_crp_sb_ex13,
  quiet_mcmc(dpmix.causal(bundle_crp_sb, mcmc = causal_run_mcmc))
)
summary(fit_crp_sb)
params(fit_crp_sb)
```

```{r plot-fit-crp-sb}
plot(fit_crp_sb, family = plot_family_crp_sb_ex13)
```

```{r predict-mean-crp-sb}
pred_median_crp_sb <- predict(
  fit_crp_sb,
  newdata =x_eval,
  type = "quantile",
  p = 0.5,
  interval = "credible",
  workers = predict_workers_ex13
)
plot(pred_median_crp_sb)
```

```{r predict-quantile-crp-sb}
pred_q_crp_sb <- predict(fit_crp_sb, newdata =x_eval, type = "quantile",
                         p = 0.9, interval = "credible", workers = predict_workers_ex13)
plot(pred_q_crp_sb)
```

```{r predict-density-crp-sb}
pred_d_crp_sb <- predict(fit_crp_sb, newdata =x_eval, y = y_eval,
                         type = "density", interval = "credible", workers = predict_workers_ex13)
plot(pred_d_crp_sb)
```

```{r predict-survival-crp-sb}
pred_surv_crp_sb <- predict(fit_crp_sb, newdata =x_eval, y = y_eval,
                            type = "survival", interval = "credible", workers = predict_workers_ex13)
plot(pred_surv_crp_sb)
```

```{r cate-crp-sb}
cate_crp_sb <- cate(
  fit_crp_sb,
  type = "rmean",
  cutoff = rmean_cutoff_ex13,
  newdata = x_eval,
  interval = "credible",
  nsim_mean = nsim_mean_ex13
)
head(format_cate_table(cate_crp_sb))
plot(cate_crp_sb)
```

```{r cqte-crp-sb}
cqte_crp_sb <- cqte(fit_crp_sb, probs = c(0.25, 0.5, 0.75),
                   newdata = x_eval, interval = "credible")
head(format_cqte_table(cqte_crp_sb))
plot(cqte_crp_sb)
```

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex12-causal-same-backend-sb](ex12-causal-same-backend-sb.html)
- Next: [ex14-causal-different-backends-sb](ex14-causal-different-backends-sb.html)
- Workflow index: [Roadmap](../start/roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [03 Causal Inference Objects](../advanced/theory/03-causal-inference-objects.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->
