---
title: "ex15. Clustering with GPD Tail (Weights + Covariates)"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = 'legacy-cache-v2/'
)

set.seed(1)

FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, workers = 2)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_chains = TRUE, workers = 2, z_update_every = 2)

mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(CausalMixGPD)
source("_legacy-setup.R")
```

> **Website workflow note.**
> This page reflects the current exported clustering API built around `dpmgpd.cluster()`, `predict()`, `summary()`, and `plot()`.
> Last updated: **2026-03-19**.
>
> For the package-level theory and longer narrative, see the manuscript and the clustering discussion in the main package articles.

# Clustering with a Spliced GPD Tail and Covariate-Dependent Weights

**Purpose**: Fit a tail-aware clustering model in which covariates affect **cluster prevalence** through mixture weights (`type = "weights"`), while the component likelihood is a **spliced bulk + GPD tail** model.

---

## Data Setup

```{r data-setup}
data("nc_posX100_p5_k4")
dat <- data.frame(y = nc_posX100_p5_k4$y, nc_posX100_p5_k4$X)

train_id <- seq_len(70)
train_dat <- dat[train_id, , drop = FALSE]
test_dat <- dat[-train_id, , drop = FALSE]

summary_tbl <- tibble(
  split = c("train", "test"),
  n = c(nrow(train_dat), nrow(test_dat)),
  y_mean = c(mean(train_dat$y), mean(test_dat$y)),
  y_sd = c(stats::sd(train_dat$y), stats::sd(test_dat$y)),
  y_max = c(max(train_dat$y), max(test_dat$y))
)

summary_tbl %>%
  mutate(across(c(y_mean, y_sd, y_max), ~ signif(.x, 4))) %>%
  kable(caption = "Train/test split for the spliced clustering workflow", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

```{r data-scatter}
ggplot(train_dat, aes(x = x1, y = y, color = x2)) +
  geom_point(alpha = 0.65) +
  labs(
    title = "Training sample for spliced clustering",
    subtitle = "Covariates drive mixture weights via type = 'weights'",
    x = "x1",
    y = "y",
    color = "x2"
  ) +
  theme_minimal()
```

---

## Fit a Spliced Clustering Model

```{r fit-cluster, cache=FALSE}
fit_cluster_gpd <- load_or_fit(
  "ex15-clustering-dpmgpd-weights-fit_cluster_gpd",
  dpmgpd.cluster(
    y ~ x1 + x2 + x3 + x4 + x5,
    data = train_dat,
    kernel = "lognormal",
    type = "weights",
    components = 10,
    mcmc = mcmc
  )
)

summary(fit_cluster_gpd, vars = c("y", "x1", "x2", "x3"))
```

This clustering configuration uses the formula/data-frame interface directly. Because `type = "weights"`, covariates enter the **gating model** for cluster membership, while the spliced bulk-tail likelihood captures upper-tail differences inside each component.

---

## Training Labels and Posterior Similarity Matrix

```{r predict-train}
cluster_psm_gpd <- predict(fit_cluster_gpd, type = "psm")
cluster_train_gpd <- predict(fit_cluster_gpd, type = "label", return_scores = TRUE)

train_summary_gpd <- summary(
  cluster_train_gpd,
  vars = c("y", "x1", "x2", "x3"),
  top_n = 5
)

train_sizes_tbl <- tibble(
  cluster = paste0("C", names(train_summary_gpd$cluster_sizes)),
  n = as.integer(train_summary_gpd$cluster_sizes)
)

train_sizes_tbl %>%
  kable(caption = "Representative training cluster sizes", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

```{r psm-plot}
plot(cluster_psm_gpd)
```

```{r train-summary-plot}
plot(cluster_train_gpd, type = "summary", top_n = 5)
```

The PSM summarizes posterior co-clustering probabilities on the training sample, while the training label object gives the Dahl representative partition together with assignment scores and cluster-level summaries.

---

## Predict Labels for New Observations

```{r predict-newdata}
cluster_new_gpd <- predict(
  fit_cluster_gpd,
  newdata = test_dat,
  type = "label",
  return_scores = TRUE
)

new_summary_gpd <- summary(
  cluster_new_gpd,
  vars = c("y", "x1", "x2", "x3"),
  top_n = 5
)

new_sizes_tbl <- tibble(
  cluster = paste0("C", names(new_summary_gpd$cluster_sizes)),
  n = as.integer(new_summary_gpd$cluster_sizes)
)

new_sizes_tbl %>%
  kable(caption = "Predicted cluster sizes for the held-out sample", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

```{r new-certainty-plot}
plot(cluster_new_gpd, type = "certainty")
```

```{r fit-size-plot}
plot(fit_cluster_gpd, which = "sizes", top_n = 5)
```

`predict(..., newdata = ..., type = "label")` maps held-out observations into the space of the representative training clusters. The returned score matrix can be summarized either numerically or with the built-in certainty plot above.

---

## Takeaways

- `dpmgpd.cluster()` is the current wrapper for **tail-aware** clustering.
- `type = "weights"` is appropriate when covariates should change **cluster prevalence** rather than cluster-specific kernel parameters.
- The standard post-fit workflow is `predict(..., type = "psm")`, `predict(..., type = "label")`, then `summary()`/`plot()` on the returned cluster objects.
- The next page contrasts this with a **bulk-only** clustering model where covariates enter through **parameter links** instead of weights.

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex14-causal-different-backends-sb](ex14-causal-different-backends-sb.html)
- Next: [ex16-clustering-dpm-param](ex16-clustering-dpm-param.html)
- Workflow index: [Roadmap](../start/roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->
