---
title: "ex15. Causal Inference: X Without PS (CRP+SB) - InvGauss/Amoroso"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = 'legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(DPmixGPD)
library(tibble)
library(dplyr)

# Ensure package is fully loaded for data access
if (!require("devtools", quietly = TRUE)) {
  library(devtools)
}
pkg_root <- rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
  devtools::load_all(path = pkg_root, quiet = TRUE)

# Precomputed results for fast website builds
source("_legacy-setup.R")
if (!exists("safe_plot", mode = "function")) {
  safe_plot <- function(x, ...) {
    obj <- tryCatch(plot(x, ...), error = function(e) e)
    if (inherits(obj, "ggplot")) {
      try(print(obj), silent = TRUE)
    } else if (is.list(obj)) {
      for (elt in obj) if (inherits(elt, "ggplot")) try(print(elt), silent = TRUE)
    }
    invisible(obj)
  }
}
source("_legacy-setup.R")
if (!exists("safe_plot", mode = "function")) {
  safe_plot <- function(x, ...) {
    obj <- tryCatch(plot(x, ...), error = function(e) e)
    if (inherits(obj, "ggplot")) {
      try(print(obj), silent = TRUE)
    } else if (is.list(obj)) {
      for (elt in obj) if (inherits(elt, "ggplot")) try(print(elt), silent = TRUE)
    }
    invisible(obj)
  }
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (basic, unconditional, conditional, and causal)..


# Causal Inference: X Without PS (CRP+SB)

This vignette uses **covariates X** but disables PS estimation. Outcome models are conditional on X only.

- Model A: CRP with GPD tail (InvGauss)
- Model B: SB bulk-only (Amoroso)

---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2", package = "DPmixGPD")
y <- abs(causal_alt_real500_p4_k2$y) + 0.01
T <- causal_alt_real500_p4_k2$T
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  print()

x_eval <- X[1:40, , drop = FALSE]
y_eval <- y[1:40]
u_threshold <- as.numeric(stats::quantile(y, 0.8, names = FALSE))
```

---

## Model A: CRP with GPD Tail (InvGauss)

```{r bundle-crp-gpd, cache=FALSE}
param_specs_gpd <- list(
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "lognormal",
      args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
    )
  )
)

bundle_crp_gpd <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "invgauss",
  backend = "crp",
  PS = FALSE,
  GPD = TRUE,
  components = 6,
  param_specs = param_specs_gpd,
  mcmc_outcome = list(niter = 300, nburnin = 80, nchains = 1, thin = 1, seed = 3)
)

bundle_crp_gpd
```

```{r fit-crp-gpd, cache=FALSE}
fit_crp_gpd <- quiet_mcmc(run_mcmc_causal(bundle_crp_gpd))
summary(fit_crp_gpd)
```

```{r predict-mean-crp-gpd}
pred_mean_gpd <- predict(fit_crp_gpd, x = x_eval, type = "mean", interval = "credible", nsim_mean = 150)
head(pred_mean_gpd)
safe_plot(pred_mean_gpd)
```

```{r predict-quantile-crp-gpd}
pred_q_gpd <- predict(fit_crp_gpd, x = x_eval, type = "quantile", p = 0.5, interval = "credible")
head(pred_q_gpd)
safe_plot(pred_q_gpd)
```

```{r predict-density-crp-gpd}
pred_d_gpd <- predict(fit_crp_gpd, x = x_eval, y = y_eval, type = "density", interval = "credible")
head(pred_d_gpd)
safe_plot(pred_d_gpd)
```

```{r predict-survival-crp-gpd}
pred_surv_gpd <- predict(fit_crp_gpd, x = x_eval, y = y_eval, type = "survival", interval = "credible")
head(pred_surv_gpd)
safe_plot(pred_surv_gpd)
```

```{r ate-crp-gpd}
ate_gpd <- ate(fit_crp_gpd, newdata = x_eval, interval = "credible", nsim_mean = 150)
head(ate_gpd)
safe_plot(ate_gpd)
```

```{r qte-crp-gpd}
qte_gpd <- qte(fit_crp_gpd, probs = c(0.25, 0.5, 0.75), newdata = x_eval, interval = "credible")
head(qte_gpd)
safe_plot(qte_gpd)
```

---

## Model B: SB Bulk-only (Amoroso)

```{r bundle-sb-bulk, cache=FALSE}
bundle_sb_bulk <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "amoroso",
  backend = "sb",
  PS = FALSE,
  GPD = FALSE,
  components = 6,
  mcmc_outcome = list(niter = 300, nburnin = 80, nchains = 1, thin = 1, seed = 4)
)

bundle_sb_bulk
```

```{r fit-sb-bulk, cache=FALSE}
fit_sb_bulk <- quiet_mcmc(run_mcmc_causal(bundle_sb_bulk))
summary(fit_sb_bulk)
```

```{r predict-mean-sb-bulk}
pred_mean_bulk <- predict(fit_sb_bulk, x = x_eval, type = "mean", interval = "credible", nsim_mean = 150)
head(pred_mean_bulk)
safe_plot(pred_mean_bulk)
```

```{r predict-quantile-sb-bulk}
pred_q_bulk <- predict(fit_sb_bulk, x = x_eval, type = "quantile", p = 0.5, interval = "credible")
head(pred_q_bulk)
safe_plot(pred_q_bulk)
```

```{r predict-density-sb-bulk}
pred_d_bulk <- predict(fit_sb_bulk, x = x_eval, y = y_eval, type = "density", interval = "credible")
head(pred_d_bulk)
safe_plot(pred_d_bulk)
```

```{r predict-survival-sb-bulk}
pred_surv_bulk <- predict(fit_sb_bulk, x = x_eval, y = y_eval, type = "survival", interval = "credible")
head(pred_surv_bulk)
safe_plot(pred_surv_bulk)
```

```{r ate-sb-bulk}
ate_bulk <- ate(fit_sb_bulk, newdata = x_eval, interval = "credible", nsim_mean = 150)
head(ate_bulk)
safe_plot(ate_bulk)
```

```{r qte-sb-bulk}
qte_bulk <- qte(fit_sb_bulk, probs = c(0.25, 0.5, 0.75), newdata = x_eval, interval = "credible")
head(qte_bulk)
safe_plot(qte_bulk)
```

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex14-causal-no-x-CRP](ex14-causal-no-x-CRP.html)
- Next: [ex16-causal-same-backend-CRP](ex16-causal-same-backend-CRP.html)
- Workflow index: [Roadmap](../roadmap.html)
- Practical entry: [Cookbook](../cookbook.html)
<!-- workflow-nav:end -->
















