---
title: "ex16. Clustering with Regular Kernel (Parameter Links + Covariates)"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = 'legacy-cache-v2/'
)

set.seed(1)

FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, workers = 2)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_chains = TRUE, workers = 2, z_update_every = 2)

mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(CausalMixGPD)
source("_legacy-setup.R")
```

> **Website workflow note.**
> This page reflects the current exported clustering API built around `dpmix.cluster()`, `predict()`, `summary()`, and `plot()`.
> Last updated: **2026-03-19**.
>
> For the package-level theory and longer narrative, see the manuscript and the clustering discussion in the main package articles.

# Clustering with a Regular Kernel and Covariate-Dependent Parameter Links

**Purpose**: Fit a bulk-only clustering model in which covariates affect **component-specific kernel parameters** (`type = "param"`), while mixture weights remain global.

## What you’ll learn

- How `dpmix.cluster()` differs from outcome modeling: the goal is **partitioning / labeling**, not density prediction alone.\n- What `type = \"param\"` means: covariates shift **within-cluster kernel parameters** while mixture weights remain global.\n- How to use `predict(..., type = \"label\" | \"psm\")` plus `summary()`/`plot()` for cluster interpretation and stability.

## When to use this template

- You want covariates to explain **how clusters behave internally** (component parameters), not just reweight cluster membership.\n- You need train/test labeling where new points map back to a representative training partition.

## Next steps

- Compare against the `type = \"weights\"` clustering mode (`ex15`) when covariates should drive **cluster membership** rather than component parameters.

---

## Data Setup

```{r data-setup}
data("nc_posX100_p3_k2")
dat <- data.frame(y = nc_posX100_p3_k2$y, nc_posX100_p3_k2$X)

train_id <- seq_len(70)
train_dat <- dat[train_id, , drop = FALSE]
test_dat <- dat[-train_id, , drop = FALSE]

summary_tbl <- tibble(
  split = c("train", "test"),
  n = c(nrow(train_dat), nrow(test_dat)),
  y_mean = c(mean(train_dat$y), mean(test_dat$y)),
  y_sd = c(stats::sd(train_dat$y), stats::sd(test_dat$y)),
  x1_mean = c(mean(train_dat$x1), mean(test_dat$x1))
)

summary_tbl
```

```{r data-scatter}
ggplot(train_dat, aes(x = x1, y = y, color = x2)) +
  geom_point(alpha = 0.65) +
  labs(
    title = "Training sample for bulk-only clustering",
    subtitle = "Covariates drive kernel parameters via type = 'param'",
    x = "x1",
    y = "y",
    color = "x2"
  ) +
  theme_minimal()
```

---

## Fit a Bulk-Only Clustering Model

```{r fit-cluster, cache=FALSE}
fit_cluster_param <- load_or_fit(
  "ex16-clustering-dpm-param-fit_cluster_param",
  dpmix.cluster(
    y ~ x1 + x2 + x3,
    data = train_dat,
    kernel = "normal",
    type = "param",
    components = 8,
    mcmc = mcmc
  )
)

summary(fit_cluster_param, vars = c("y", "x1", "x2"))
```

Here the formula still defines the full covariate structure, but `type = "param"` means the model uses those predictors to shift component-specific kernel parameters rather than the cluster weights. This matches the parameter-link clustering formulation described in the manuscript.

---

## Training Labels and Cluster Summaries

```{r predict-train}
cluster_psm_param <- predict(fit_cluster_param, type = "psm")
cluster_train_param <- predict(fit_cluster_param, type = "label", return_scores = TRUE)

fit_summary_param <- summary(
  fit_cluster_param,
  vars = c("y", "x1", "x2"),
  top_n = 5
)

train_sizes_tbl <- tibble(
  cluster = paste0("C", names(fit_summary_param$cluster_sizes)),
  n = as.integer(fit_summary_param$cluster_sizes)
)

train_sizes_tbl
```

```{r fit-summary-plot}
plot(fit_cluster_param, which = "summary", top_n = 5)
```

```{r train-size-plot}
plot(cluster_train_param, type = "sizes", top_n = 5)
```

```{r psm-plot}
plot(cluster_psm_param)
```

The fit-level and label-level plotting methods expose the same representative partition from different angles: size summaries, response summaries, and the full posterior similarity matrix.

---

## Predict Labels for Held-Out Data

```{r predict-newdata}
cluster_new_param <- predict(
  fit_cluster_param,
  newdata = test_dat,
  type = "label",
  return_scores = TRUE
)

new_summary_param <- summary(
  cluster_new_param,
  vars = c("y", "x1", "x2"),
  top_n = 5
)

new_sizes_tbl <- tibble(
  cluster = paste0("C", names(new_summary_param$cluster_sizes)),
  n = as.integer(new_summary_param$cluster_sizes)
)

new_sizes_tbl
```

```{r new-summary-plot}
plot(cluster_new_param, type = "summary", top_n = 5)
```

```{r new-certainty-plot}
plot(cluster_new_param, type = "certainty")
```

For newdata, the labels are returned in the space of the representative training clusters, which makes train/test comparison straightforward without exposing component-label switching from the sampler.

---

## Takeaways

- `dpmix.cluster()` is the current wrapper for **bulk-only** clustering.
- `type = "param"` is appropriate when covariates should alter **within-cluster regression/parameter structure** while keeping global mixture weights.
- The same three-step workflow applies as in the spliced example: fit, `predict(..., type = "psm" | "label")`, then `summary()` and `plot()`.
- This closes the examples sequence with the two main clustering modes now exposed by the package website.

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [ex15-clustering-dpmgpd-weights](ex15-clustering-dpmgpd-weights.html)
- Next: [troubleshooting](../start/troubleshooting.html)
- Workflow index: [Roadmap](../start/roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [02 Clustering Extension](../advanced/theory/02-clustering-extension.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->





