---
<<<<<<< HEAD
<<<<<<< HEAD
title: "Kernel reference: Amoroso"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kernel reference: Amoroso}
=======
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
title: "Legacy: Kernel reference: Lognormal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Legacy: Kernel reference: Lognormal}
<<<<<<< HEAD
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<<<<<<< HEAD
<<<<<<< HEAD
```{r legacy-setup, include=FALSE}
source(file.path(dirname(knitr::current_input(dir = TRUE)), "_legacy-setup.R"))
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(DPmixGPD)

# Precomputed results for fast website builds
PRECOMP_DIR <- file.path("articles", "legacy-precomputed")
if (!dir.exists(PRECOMP_DIR)) dir.create(PRECOMP_DIR, recursive = TRUE, showWarnings = FALSE)

.precomp_path <- function(tag) file.path(PRECOMP_DIR, paste0(tag, ".rds"))

load_or_fit <- function(tag, expr) {
  path <- .precomp_path(tag)
  if (isTRUE(FAST) && file.exists(path)) {
    return(readRDS(path))
  }
  # If FAST but cache missing, run the expression with fast MCMC and save.
  out <- eval(substitute(expr), envir = parent.frame())
  # Save only if it looks like a fit object
  try(saveRDS(out, path), silent = TRUE)
  out
}


quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == 'windows') 'NUL' else '/dev/null'
  out <- NULL
  utils::capture.output(
    out <- eval(substitute(expr), envir = parent.frame()),
    file = nullfile
  )
  out
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).


The **Amoroso** kernel is parameterized by (`a`, `theta`, `alpha`, `beta`).

## Exported mixture helpers

- `dAmorosoMix(x, a, theta, alpha, beta, w, log = FALSE)`
- `pAmorosoMix(q, a, theta, alpha, beta, w, lower.tail = TRUE, log.p = FALSE)`
- `qAmorosoMix(p, a, theta, alpha, beta, w, lower.tail = TRUE, log.p = FALSE)`
- `rAmorosoMix(n, a, theta, alpha, beta, w)`

For the full catalog (including CRP utilities and GPD splicing), see:

- [Available Distributions](v02-available-distributions.html)

## Using Amoroso in a model
=======
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
library(DPmixGPD)
```

> **Legacy note:** This kernel reference is retained for historical context alongside the extended legacy materials.

The **Lognormal** kernel is parameterized by `meanlog` and `sdlog` (i.e., `log(y)` is Normal).

## Exported mixture helpers

- `dLognormalMix(x, meanlog, sdlog, w, log = FALSE)`
- `pLognormalMix(q, meanlog, sdlog, w, lower.tail = TRUE, log.p = FALSE)`
- `qLognormalMix(p, meanlog, sdlog, w, lower.tail = TRUE, log.p = FALSE)`
- `rLognormalMix(n, meanlog, sdlog, w)`

For the full catalog (including CRP utilities and GPD splicing), see:

- [Legacy distributions](legacy/v02-available-distributions.html)

## Using Lognormal in a model
<<<<<<< HEAD
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090

```{r}
y <- abs(stats::rnorm(50)) + 0.1
bundle <- build_nimble_bundle(
  y = y,
<<<<<<< HEAD
<<<<<<< HEAD
  backend = "sb",
  kernel  = "amoroso",
=======
  backend = "crp",
  kernel  = "lognormal",
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
  backend = "crp",
  kernel  = "lognormal",
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  GPD     = TRUE,
  components = 6
)
```


<<<<<<< HEAD
<<<<<<< HEAD
=======

>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======

>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
