---
title: "Kernel reference: Lognormal"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Kernel reference: Lognormal}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(DPmixGPD)

# Precomputed results for fast website builds
source("_legacy-setup.R")
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).


The **Lognormal** kernel is parameterized by `meanlog` and `sdlog` (i.e., `log(y)` is Normal).

## Exported mixture helpers

- `dLognormalMix(x, meanlog, sdlog, w, log = FALSE)`
- `pLognormalMix(q, meanlog, sdlog, w, lower.tail = TRUE, log.p = FALSE)`
- `qLognormalMix(p, meanlog, sdlog, w, lower.tail = TRUE, log.p = FALSE)`
- `rLognormalMix(n, meanlog, sdlog, w)`

For the full catalog (including CRP utilities and GPD splicing), see:

- [Available Distributions](v02-available-distributions.html)

## Using Lognormal in a model

```{r}
y <- abs(stats::rnorm(50)) + 0.1
bundle <- build_nimble_bundle(
  y = y,
  backend = "crp",
  kernel  = "lognormal",
  GPD     = TRUE,
  components = 6
)
```



