---
title: "Legacy: Kernel reference: Gamma"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Legacy: Kernel reference: Gamma}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = TRUE
)
library(DPmixGPD)
```

> **Legacy note:** This kernel reference is retained for historical context alongside the extended legacy materials.

The **Gamma** kernel is parameterized by `shape` and `rate`.

## Exported mixture helpers

- `dGammaMix(x, shape, rate, w, log = FALSE)`
- `pGammaMix(q, shape, rate, w, lower.tail = TRUE, log.p = FALSE)`
- `qGammaMix(p, shape, rate, w, lower.tail = TRUE, log.p = FALSE)`
- `rGammaMix(n, shape, rate, w)`

For the full catalog (including CRP utilities and GPD splicing), see:

- [Legacy distributions](legacy/v02-available-distributions.html)

## Using Gamma in a model

```{r}
y <- abs(stats::rnorm(50)) + 0.1
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel  = "gamma",
  GPD     = TRUE,
  components = 6
)
```



