---
<<<<<<< HEAD
<<<<<<< HEAD
title: "0. Start Here"
=======
title: "Legacy: 0. Getting Started"
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
title: "Legacy: 0. Getting Started"
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
<<<<<<< HEAD
<<<<<<< HEAD
  %\VignetteIndexEntry{0. Start Here}
=======
  %\VignetteIndexEntry{Legacy: 0. Getting Started}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
  %\VignetteIndexEntry{Legacy: 0. Getting Started}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<<<<<<< HEAD
<<<<<<< HEAD
```{r legacy-setup, include=FALSE}
source(file.path(dirname(knitr::current_input(dir = TRUE)), "_legacy-setup.R"))
```


=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
<<<<<<< HEAD
<<<<<<< HEAD
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(nimble)
library(DPmixGPD)

# Precomputed results for fast website builds
PRECOMP_DIR <- file.path("articles", "legacy-precomputed")
if (!dir.exists(PRECOMP_DIR)) dir.create(PRECOMP_DIR, recursive = TRUE, showWarnings = FALSE)

.precomp_path <- function(tag) file.path(PRECOMP_DIR, paste0(tag, ".rds"))

load_or_fit <- function(tag, expr) {
  path <- .precomp_path(tag)
  if (isTRUE(FAST) && file.exists(path)) {
    return(readRDS(path))
  }
  # If FAST but cache missing, run the expression with fast MCMC and save.
  out <- eval(substitute(expr), envir = parent.frame())
  # Save only if it looks like a fit object
  try(saveRDS(out, path), silent = TRUE)
  out
}


quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == 'windows') 'NUL' else '/dev/null'
=======
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  eval = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
set.seed(1)

quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == "windows") "NUL" else "/dev/null"
<<<<<<< HEAD
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  out <- NULL
  utils::capture.output(
    out <- eval(substitute(expr), envir = parent.frame()),
    file = nullfile
  )
  out
}
```

<<<<<<< HEAD
<<<<<<< HEAD

> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).


# Start Here
=======
> **Legacy note:** This page is preserved for historical context and extended detail. It predates the streamlined official vignettes and may include longer runs or exploratory material.

# Legacy Overview {#legacy-overview}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
> **Legacy note:** This page is preserved for historical context and extended detail. It predates the streamlined official vignettes and may include longer runs or exploratory material.

# Legacy Overview {#legacy-overview}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090

This vignette gives a minimal, fully working workflow for an **unconditional** model and a **conditional** model. Everything uses short MCMC runs so the vignette renders quickly.

---

## Unconditional Model (CRP, bulk-only)

```{r start-uncond-data}
data("nc_pos200_k3")
y <- nc_pos200_k3$y
```

```{r start-uncond-bundle}
bundle_uncond <- build_nimble_bundle(
  y = y,
  backend = "crp",
  kernel = "gamma",
  GPD = FALSE,
  components = 5,
<<<<<<< HEAD
<<<<<<< HEAD
  mcmc = mcmc
=======
  mcmc = list(niter = 200, nburnin = 50, thin = 1, nchains = 1, seed = 1)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
  mcmc = list(niter = 200, nburnin = 50, thin = 1, nchains = 1, seed = 1)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
)
```

```{r start-uncond-fit}
<<<<<<< HEAD
<<<<<<< HEAD
fit_uncond <- load_or_fit("v00-start-here-fit_uncond", quiet_mcmc(run_mcmc_bundle_manual(bundle_uncond, show_progress = FALSE)))
=======
fit_uncond <- quiet_mcmc(run_mcmc_bundle_manual(bundle_uncond, show_progress = FALSE))
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
fit_uncond <- quiet_mcmc(run_mcmc_bundle_manual(bundle_uncond, show_progress = FALSE))
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
summary(fit_uncond)
```

```{r start-uncond-predict, cache=FALSE}
pred_q <- predict(fit_uncond, type = "quantile", index = c(0.5, 0.9), interval = "credible")
head(pred_q$fit)
plot(pred_q)
```

---

## Conditional Model (SB, bulk-only)

```{r start-cond-data}
data("nc_posX100_p3_k2")
yc <- nc_posX100_p3_k2$y
X <- as.matrix(nc_posX100_p3_k2$X)
```

```{r start-cond-bundle}
bundle_cond <- build_nimble_bundle(
  y = yc,
  X = X,
  backend = "sb",
  kernel = "lognormal",
  GPD = FALSE,
  components = 5,
<<<<<<< HEAD
<<<<<<< HEAD
  mcmc = mcmc
=======
  mcmc = list(niter = 250, nburnin = 50, thin = 1, nchains = 1, seed = 2)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
  mcmc = list(niter = 250, nburnin = 50, thin = 1, nchains = 1, seed = 2)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
)
```

```{r start-cond-fit}
<<<<<<< HEAD
<<<<<<< HEAD
fit_cond <- load_or_fit("v00-start-here-fit_cond", quiet_mcmc(run_mcmc_bundle_manual(bundle_cond, show_progress = FALSE)))
=======
fit_cond <- quiet_mcmc(run_mcmc_bundle_manual(bundle_cond, show_progress = FALSE))
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
fit_cond <- quiet_mcmc(run_mcmc_bundle_manual(bundle_cond, show_progress = FALSE))
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
summary(fit_cond)
```

```{r start-cond-predict, cache=FALSE}
x_new <- X[1:20, , drop = FALSE]
pred_mean <- predict(fit_cond, x = x_new, type = "mean", interval = "credible", nsim_mean = 200)
head(pred_mean$fit)
plot(pred_mean)
```

---

## Useful S3 Methods

```{r start-s3}
params(fit_uncond)
plot(fit_uncond, family = c("traceplot", "running"))
```

---

## Next Steps

- `vignette 1`: package overview and terminology
- `vignette 5`: full three-phase workflow (spec ? bundle ? MCMC)
- `vignette 6-13`: unconditional/conditional models with and without GPD
- `vignette 14-19`: causal workflows



