---
<<<<<<< HEAD
title: "1. Introduction to DPmixGPD"
=======
title: "Legacy: 1. Introduction to DPmixGPD"
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
<<<<<<< HEAD
  %\VignetteIndexEntry{1. Introduction to DPmixGPD}
=======
  %\VignetteIndexEntry{Legacy: 1. Introduction to DPmixGPD}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<<<<<<< HEAD
```{r legacy-setup, include=FALSE}
source(file.path(dirname(knitr::current_input(dir = TRUE)), "_legacy-setup.R"))
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(nimble)
library(DPmixGPD)

# Precomputed results for fast website builds
PRECOMP_DIR <- file.path("articles", "legacy-precomputed")
if (!dir.exists(PRECOMP_DIR)) dir.create(PRECOMP_DIR, recursive = TRUE, showWarnings = FALSE)

.precomp_path <- function(tag) file.path(PRECOMP_DIR, paste0(tag, ".rds"))

load_or_fit <- function(tag, expr) {
  path <- .precomp_path(tag)
  if (isTRUE(FAST) && file.exists(path)) {
    return(readRDS(path))
  }
  # If FAST but cache missing, run the expression with fast MCMC and save.
  out <- eval(substitute(expr), envir = parent.frame())
  # Save only if it looks like a fit object
  try(saveRDS(out, path), silent = TRUE)
  out
}


quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == 'windows') 'NUL' else '/dev/null'
  out <- NULL
  utils::capture.output(
    out <- eval(substitute(expr), envir = parent.frame()),
    file = nullfile
  )
  out
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).

=======
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 7,
  fig.height = 5,
  eval = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
```

> **Legacy note:** This page is preserved for historical context and extended detail. It predates the streamlined official vignettes and may include longer runs or exploratory material.
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090

# DPmixGPD: Bayesian Nonparametric Mixture Models with GPD Tails

## Overview

**DPmixGPD** is an R package for Bayesian nonparametric Dirichlet Process mixture models with flexible kernels and optional Generalized Pareto Distribution (GPD) tail augmentation.

### Key Features

- **Flexible mixture kernels**: normal, gamma, lognormal, laplace, invgauss, amoroso, cauchy
- **Two MCMC backends**: Stick-Breaking (SB) and Chinese Restaurant Process (CRP)
- **Optional GPD tail**: Enhance tail probability estimation
- **Conditional modeling**: Add covariates via parameter linking
- **Causal workflows**: Propensity score adjustment with multiple PS strategies
- **Comprehensive prediction**: density, survival, quantile, sample, mean

---

## Model Structure

### Single-Outcome Model

**Bulk**: Dirichlet Process mixture
$$Y_i \sim \sum_{k=1}^{\infty} w_k f_k(y_i; \theta_k)$$

**Tail (optional)**: GPD beyond threshold
$$P(Y > u | Y > u) \sim \text{GPD}(\text{scale}, \text{shape})$$

### Causal Model

Two outcome arms with optional propensity score (PS) adjustment:
- **Control** ($T=0$): $Y_i^{(0)} \sim \text{DPM}_{\text{con}}(X_i, PS_i)$
- **Treated** ($T=1$): $Y_i^{(1)} \sim \text{DPM}_{\text{trt}}(X_i, PS_i)$

---

## MCMC Backends

### Stick-Breaking (SB)
- Fixed number of mixture components `components`
- Exact truncation at the chosen `components`
- Faster per-iteration computation
- Use when computational cost is primary concern

### Chinese Restaurant Process (CRP)
- Finite `components` cap for sampling
- Adaptively discovers number of occupied clusters within that cap
- More flexible cluster allocation
- Default for exploratory analysis

---

## Propensity Score (PS) Models

**Three PS strategies** (causal workflows only):

1. **`PS="logit"`** (default): Bayesian logistic regression
2. **`PS="probit"`**: Bayesian probit regression
3. **`PS="naive"`**: Gaussian naive Bayes classifier
4. **`PS=FALSE`**: No PS; outcomes depend on $X$ only

When PS is estimated, scores are injected into outcome model covariates for confounding adjustment.

---

## Workflow Overview

### Direct Model Building

```r
bundle <- build_nimble_bundle(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = TRUE,
  components = 5,
<<<<<<< HEAD
  mcmc = mcmc
=======
  mcmc = list(niter = 1000, nburnin = 100, nchains = 1)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
)
fit <- run_mcmc_bundle_manual(bundle)
```

---

## Quick Start

```{r intro-quickstart}
# Load packaged dataset
data("nc_pos200_k3")
y <- nc_pos200_k3$y

# Build unconditional model: bulk-only, SB backend
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "gamma",
  GPD = FALSE,
  components = 5,
<<<<<<< HEAD
  mcmc = mcmc
=======
  mcmc = list(niter = 200, nburnin = 10, nchains = 1)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
)

# Print bundle
print(bundle)
```

---

## Outline of This Vignette Series

1. **Introduction** (this vignette)
2. **Available Distributions**: dqrp functions, 10k samples, visualizations
3. **Basic Model Workflow**: compile → bundle → MCMC
4. **Unconditional DPmix (CRP)**: bulk-only, no covariates
5. **Unconditional DPmix (SB)**: bulk-only, no covariates
6. **Unconditional DPmixGPD (CRP)**: bulk+tail, no covariates
7. **Unconditional DPmixGPD (SB)**: bulk+tail, no covariates
8. **Conditional DPmix (CRP)**: bulk-only, with covariates
9. **Conditional DPmix (SB)**: bulk-only, with covariates
10. **Conditional DPmixGPD (CRP)**: bulk+tail, with covariates
11. **Conditional DPmixGPD (SB)**: bulk+tail, with covariates
12. **Causal (Same Backend CRP)**: PS + treatment effects, CRP
13. **Causal (Same Backend SB)**: PS + treatment effects, SB
14. **Causal (Different Backends CRP)**: Per-arm backends, CRP
15. **Causal (Different Backends SB)**: Per-arm backends, SB

---

## Dependencies

```{r intro-deps}
packageVersion("DPmixGPD")
packageVersion("nimble")
packageVersion("ggplot2")
```

Each vignette demonstrates complete workflows with reproducible examples.






