---
title: "3. Basic Workflow: Model Specification, Bundle, & MCMC"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{3. Basic Workflow: Model Specification, Bundle, & MCMC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r legacy-setup, include=FALSE}
source(file.path(dirname(knitr::current_input(dir = TRUE)), "_legacy-setup.R"))
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(nimble)
library(DPmixGPD)

# Precomputed results for fast website builds
PRECOMP_DIR <- file.path("articles", "legacy-precomputed")
if (!dir.exists(PRECOMP_DIR)) dir.create(PRECOMP_DIR, recursive = TRUE, showWarnings = FALSE)

.precomp_path <- function(tag) file.path(PRECOMP_DIR, paste0(tag, ".rds"))

load_or_fit <- function(tag, expr) {
  path <- .precomp_path(tag)
  if (isTRUE(FAST) && file.exists(path)) {
    return(readRDS(path))
  }
  # If FAST but cache missing, run the expression with fast MCMC and save.
  out <- eval(substitute(expr), envir = parent.frame())
  # Save only if it looks like a fit object
  try(saveRDS(out, path), silent = TRUE)
  out
}


quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == 'windows') 'NUL' else '/dev/null'
  out <- NULL
  utils::capture.output(
    out <- eval(substitute(expr), envir = parent.frame()),
    file = nullfile
  )
  out
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).


# Workflow Overview

DPmixGPD uses a direct two-step workflow:

1. **Bundle** (`build_nimble_bundle`): Generate NIMBLE code and compile the sampler.
2. **MCMC** (`run_mcmc_bundle_manual`): Execute posterior sampling.

---

## Phase 1: Bundle (NIMBLE Code Generation & Compilation)

**Purpose**: Generate NIMBLE model code, compile sampler, prepare for MCMC execution.

### Building Directly

```{r bundle-direct}
# Load packaged data
data("nc_pos200_k3")
y <- nc_pos200_k3$y

# Direct call
bundle_direct <- build_nimble_bundle(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)

print("Direct bundle creation successful.\n")
```

### Inspecting Bundle Contents

```{r bundle-inspect}
# Bundle is an S3 object with structure
cat("Bundle class:", class(bundle_direct), "\n")
cat("Bundle contains:\n")
print(names(bundle_direct))

# Access key components
cat("\nMCMC settings:\n")
print(bundle_direct$mcmc_settings)
```

---

## Phase 3: MCMC Execution

**Purpose**: Run posterior sampling from the compiled bundle.

### Basic MCMC Run

```{r mcmc-basic}
# Run MCMC
fit <- load_or_fit("v03-basic-model-compile-run-fit", run_mcmc_bundle_manual(bundle_direct, show_progress = FALSE))

cat("Fit object class:", class(fit), "\n")
cat("MCMC execution complete. Posterior samples collected.\n")
```

### Accessing Posterior Samples

```{r mcmc-posterior}
# Posterior summary
print("\n--- POSTERIOR SUMMARY ---\n")
summary(fit)

# Posterior mean parameters in original form
params_fit <- params(fit)
params_fit
```

### Diagnostic Plots

```{r mcmc-diagnostics}
# Trace plots
plot(fit, params = "alpha|beta", family = c("traceplot", "running", "autocorrelation"))
```

---

## Complete Workflow: End-to-End Example

```{r workflow-complete}
# Load packaged data
data("nc_pos200_k3")
y_data <- nc_pos200_k3$y

# PHASE 1: Bundle
bundle_final <- build_nimble_bundle(
  y = y_data,
  kernel = "gamma",
  backend = "crp",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)

# PHASE 2: MCMC
fit_final <- load_or_fit("v03-basic-model-compile-run-fit_final", run_mcmc_bundle_manual(bundle_final, show_progress = FALSE))

print("\n=== THREE-PHASE WORKFLOW COMPLETE ===\n")
summary(fit_final)
```

---

## Backend Comparison: CRP vs Stick-Breaking

### CRP Backend

```{r backend-crp}
# Chinese Restaurant Process
bundle_crp <- build_nimble_bundle(
  y = y_data,
  kernel = "gamma",
  backend = "crp",
  components = 5,
  mcmc = mcmc
)

fit_crp <- load_or_fit("v03-basic-model-compile-run-fit_crp", run_mcmc_bundle_manual(bundle_crp, show_progress = FALSE))
print("CRP execution complete.\n")
```

### Stick-Breaking Backend

```{r backend-sb}
# Stick-Breaking Process
bundle_sb <- build_nimble_bundle(
  y = y_data,
  kernel = "gamma",
  backend = "sb",
  components = 5,
  mcmc = mcmc
)

fit_sb <- load_or_fit("v03-basic-model-compile-run-fit_sb", run_mcmc_bundle_manual(bundle_sb, show_progress = FALSE))
print("SB execution complete.\n")
```

---

## Kernel Selection Guide

```{r kernel-guide}
kernels_available <- c("gamma", "lognormal", "normal", "laplace", "invgauss", "amoroso", "cauchy")

cat("Available kernels:\n")
for (k in kernels_available) {
  cat("  -", k, "\n")
}

print("\nChoose kernel based on:\n")
print("  gamma:     Right-skewed, positive support\n")
print("  lognormal: Log-transformed normality\n")
print("  normal:    Symmetric, unbounded\n")
print("  laplace:   Sharp peak, exponential tails\n")
print("  invgauss:  Positive, near-normal shape\n")
print("  amoroso:   Generalized, maximum flexibility\n")
print("  cauchy:    Heavy-tailed, rare cases\n")
```

---

## GPD Tail Augmentation

### Unconditional with GPD

```{r gpd-example, eval=FALSE}
# Data with tail behavior
data("nc_pos_tail200_k4")
y_tail <- nc_pos_tail200_k4$y

# Build with GPD
bundle_gpd <- build_nimble_bundle(
  y = y_tail,
  kernel = "gamma",
  backend = "sb",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit_gpd <- load_or_fit("v03-basic-model-compile-run-fit_gpd", run_mcmc_bundle_manual(bundle_gpd, show_progress = FALSE))
print("\nGPD augmentation applied to tail region.\n")
summary(fit_gpd)
```

---

## Summary of Key Functions

| Phase | Function | Input | Output |
|-------|----------|-------|--------|
| **1. Bundle** | `build_nimble_bundle()` | y, X (optional), kernel, backend, GPD, components | `dpmixgpd_bundle` |
| **2. MCMC** | `run_mcmc_bundle_manual()` | bundle | `mixgpd_fit` |

---

## Common Parameter Settings

```{r param-reference}
print("=== Recommended MCMC Parameters ===\n")
print("Quick test:     niter=500,  nburnin=100, nchains=1\n")
print("Standard:       niter=1000, nburnin=250, nchains=2\n")
print("Production:     niter=1000, nburnin=250, nchains=3\n")
print("\n=== Backend Parameters ===\n")
print("Use components=3-5 for both backends in this implementation.\n")
print("\n=== Kernel Selection ===\n")
print("Positive data:     gamma, lognormal, invgauss\n")
print("Any real data:     normal\n")
print("Symmetric tails:   laplace\n")
print("Extreme outliers:  cauchy\n")
```

---

## Next Steps

- Move to **vignette 6-7** for unconditional models (CRP vs SB backends)
- Move to **vignette 8-9** for tail modeling with GPD
- Move to **vignette 10-13** for conditional models with covariates
- Move to **vignette 14-19** for causal inference workflows








