---
<<<<<<< HEAD
title: "4. Backends, Kernels, and Workflow Map"
=======
title: "Legacy: 4. Backends, Kernels, and Workflow Map"
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
<<<<<<< HEAD
  %\VignetteIndexEntry{4. Backends, Kernels, and Workflow Map}
=======
  %\VignetteIndexEntry{Legacy: 4. Backends, Kernels, and Workflow Map}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<<<<<<< HEAD
```{r legacy-setup, include=FALSE}
source(file.path(dirname(knitr::current_input(dir = TRUE)), "_legacy-setup.R"))
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
=======
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
<<<<<<< HEAD
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(nimble)
library(DPmixGPD)

# Precomputed results for fast website builds
PRECOMP_DIR <- file.path("articles", "legacy-precomputed")
if (!dir.exists(PRECOMP_DIR)) dir.create(PRECOMP_DIR, recursive = TRUE, showWarnings = FALSE)

.precomp_path <- function(tag) file.path(PRECOMP_DIR, paste0(tag, ".rds"))

load_or_fit <- function(tag, expr) {
  path <- .precomp_path(tag)
  if (isTRUE(FAST) && file.exists(path)) {
    return(readRDS(path))
  }
  # If FAST but cache missing, run the expression with fast MCMC and save.
  out <- eval(substitute(expr), envir = parent.frame())
  # Save only if it looks like a fit object
  try(saveRDS(out, path), silent = TRUE)
  out
}


quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == 'windows') 'NUL' else '/dev/null'
  out <- NULL
  utils::capture.output(
    out <- eval(substitute(expr), envir = parent.frame()),
    file = nullfile
  )
  out
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).

=======
  eval = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
```

> **Legacy note:** This page is preserved for historical context and extended detail. It predates the streamlined official vignettes and may include longer runs or exploratory material.
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090

# Big picture

DPmixGPD has two orthogonal dials you turn when building models:

- **Backend** (how the mixture weights / clustering are represented)
  - **CRP**: Chinese Restaurant Process representation.
  - **SB**: stick-breaking truncation with a fixed number of components.

- **Kernel / family** (what distribution models the bulk of the data)
  - Examples: normal, lognormal, gamma, inverse-Gaussian, Laplace, Cauchy, Amoroso, etc.

Optionally, you can also turn on:

- **GPD = TRUE/FALSE** to splice a Generalized Pareto tail beyond a threshold.

# What changes between CRP and SB?

Both backends target the same posterior over densities. The difference is representation:

- **CRP** learns a random number of occupied clusters within a finite `components` cap.
- **SB** uses the same finite `components` cap and learns stick-breaking weights.

Practical rule of thumb:

- **CRP** is convenient when you want adaptive complexity while still using a finite `components` cap.
- **SB** is convenient when you want predictable memory/time and easy vectorization.

# What does the workflow look like?

DPmixGPD uses a consistent build -> run -> summarize loop:

1. **Build a bundle** using `build_nimble_bundle()` (or causal builders if you are doing TE work).
2. **Run MCMC** using `run_mcmc_bundle_manual()`.
3. **Inspect and summarize** using `print()`, `summary()`, `plot()`.
4. **Predict** using `predict()` (and optionally `fitted()`).

```{r}
# Build
bundle <- build_nimble_bundle(
  y = rnorm(50),
  backend = "crp",
  kernel = "normal",
  GPD = FALSE,
  components = 5,
<<<<<<< HEAD
  mcmc = mcmc
)

# Run
fit <- load_or_fit("v04-backends-and-workflow-fit", run_mcmc_bundle_manual(bundle))
=======
  mcmc = list(niter = 200, nburnin = 100, thin = 1, nchains = 1, seed = 1)
)

# Run
fit <- run_mcmc_bundle_manual(bundle)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090

# Summarize
print(fit)
summary(fit)
plot(fit)
```

# Kernel support quick check

Use `kernel_support_table()` and the kernel registry helpers to confirm what is available.

```{r}
kernel_support_table()
```

# Where to go next

- **Available distributions**: see **v02** for the `d/p/q/r` functions and examples.
- **Basic build/compile/run**: see **v03**.
- **Unconditional / Conditional / Causal**: continue through v06+.


