---
title: "Legacy: 5. Unconditional DPmix (CRP Backend)"
output:
  html_document:
    toc: true
    toc_depth: 3
    theme: flatly
    highlight: tango
vignette: >
  %\VignetteIndexEntry{Legacy: 5. Unconditional DPmix (CRP Backend)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  fig.width = 8,
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE,
  cache = FALSE
)
library(nimble)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(gridExtra)
library(kableExtra)
library(patchwork)
set.seed(123)
```

> **Legacy note:** This page is preserved for historical context and extended detail. It predates the streamlined official vignettes and may include longer runs or exploratory material.

# Overview

This vignette demonstrates an **unconditional DPmix model** using the **CRP backend** on a positive dataset.
We use a Laplace kernel for the bulk distribution and no GPD tail augmentation.

# Data Setup

```{r data-setup}
# Load pre-generated dataset: 200 observations from mixture of 3 gamma components
data(nc_pos200_k3)
y_mixed <- nc_pos200_k3$y

paste("Sample size:", length(y_mixed))
paste("Mean:", mean(y_mixed))
paste("SD:", sd(y_mixed))
paste("Range:", paste(range(y_mixed), collapse = " to "))

# Visualization
df_data <- data.frame(y = y_mixed)
p_raw <- ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 30, alpha = 0.6,
                 fill = "steelblue", color = "black") +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Raw Data: Mixed Gamma Distribution", x = "y", y = "Density") +
  theme_minimal()

print(p_raw)
```

# Build Bundle (CRP)

```{r bundle-mcmc-crp}
bundle_crp <- build_nimble_bundle(
  y = y_mixed,
  kernel = "laplace",         # Use laplace kernel
  backend = "crp",            # CRP backend
  GPD = FALSE,                # No tail augmentation
  components = 3,             # Minimal for testing
  alpha_random = TRUE,        # Random DP concentration
  mcmc = list(
    niter = 50,             # Minimal for testing
    nburnin = 10,           # Minimal burnin
    nchains = 2,            # Two chains for diagnostics
    thin = 1                # No thinning
  )
)
```

# Run MCMC (Longer Run)

```{r mcmc-crp}
bundle_crp <- build_nimble_bundle(
  y_mixed,
  kernel = "laplace",
  backend = "crp",
  GPD = FALSE,
  components = 3,
  alpha_random = TRUE,
  mcmc = list(niter = 1500, nburnin = 250, nchains = 2, thin = 1)
)
```

```{r}
summary(bundle_crp)
```

```{r mcmc-crp-fit, results = "hide"}
fit_crp <- run_mcmc_bundle_manual(bundle_crp)
```

```{r}
summary(fit_crp)
```

# Posterior Parameters

```{r params-crp}
params_crp <- params(fit_crp)
params_crp
```

# Diagnostics

```{r diag-trace}
# Trace plots for key parameters
plot(fit_crp, params = "alpha", family = c("traceplot", "density", "geweke"))
```

# Posterior Predictive Density

```{r pred-density}
# Generate prediction grid
y_grid <- seq(0, max(y_mixed) * 1.2, length.out = 200)

# Posterior predictive density
pred_density <- predict(fit_crp, y = y_grid, type = "density")

# Use S3 plot method
plot(pred_density)
```
