---
title: "8. Unconditional DPmixGPD with CRP Backend"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{8. Unconditional DPmixGPD with CRP Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(DPmixGPD)

# Precomputed results for fast website builds
PRECOMP_DIR <- file.path("articles", "legacy-precomputed")
if (!dir.exists(PRECOMP_DIR)) dir.create(PRECOMP_DIR, recursive = TRUE, showWarnings = FALSE)

.precomp_path <- function(tag) file.path(PRECOMP_DIR, paste0(tag, ".rds"))

load_or_fit <- function(tag, expr) {
  path <- .precomp_path(tag)
  if (isTRUE(FAST) && file.exists(path)) {
    return(readRDS(path))
  }
  # If FAST but cache missing, run the expression with fast MCMC and save.
  out <- eval(substitute(expr), envir = parent.frame())
  # Save only if it looks like a fit object
  try(saveRDS(out, path), silent = TRUE)
  out
}


quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == 'windows') 'NUL' else '/dev/null'
  out <- NULL
  utils::capture.output(
    out <- eval(substitute(expr), envir = parent.frame()),
    file = nullfile
  )
  out
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).


# Unconditional DPmixGPD: CRP Backend with Tail Augmentation

**Purpose**: Model the **bulk** of the distribution with a DP mixture (the "kernel" or bulk family) and let a **GPD tail** capture extreme values beyond a threshold. The CRP backend handles partitioning and we toggle `GPD = TRUE` to augment the tail.

---

## Data Setup

```{r data-setup}
# Load data with an obvious tail
data("nc_pos_tail200_k4")
y_tail <- nc_pos_tail200_k4$y
y_tail <- y_tail[is.finite(y_tail) & y_tail > 0]

# Summaries and table
summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y_tail), mean(y_tail), sd(y_tail), min(y_tail), max(y_tail))
)

df_data <- data.frame(y = y_tail)

ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "steelblue", color = "black", alpha = 0.6) +
  geom_density(color = "red", linewidth = 1) +
  labs(title = "Raw Data: Mix of Bulk and Tail", x = "y", y = "Density") +
  theme_minimal()
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Summary of the Tail Dataset", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

---

## Threshold & Tail Partition

```{r threshold-selection}
thresholds <- quantile(y_tail, c(0.70, 0.75, 0.80, 0.85, 0.90))
u_threshold <- thresholds["80%"]

df_tail <- tibble(
  y = y_tail,
  region = ifelse(y_tail > u_threshold, "Tail", "Bulk")
)

ggplot(df_tail, aes(x = y, fill = region)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, alpha = 0.7, color = "black") +
  geom_vline(xintercept = u_threshold, linetype = "dashed", color = "black") +
  scale_fill_manual(values = c("Bulk" = "steelblue", "Tail" = "red")) +
  labs(title = "Threshold Partition", x = "y", y = "Density") +
  theme_minimal()
```

---

## Model Specification & Bundle

This mirrors the direct `build_nimble_bundle()` workflow in `v04`, but with **tail augmentation** turned on (`GPD = TRUE`). Here we use an **Inverse Gaussian** bulk kernel and estimate the GPD threshold with a lognormal prior centered at the empirical 80th percentile.

```{r bundle-mcmc-crp-gpd}
bundle_gpd <- build_nimble_bundle(
  y = y_tail,
  kernel = "invgauss",
  backend = "crp",
  GPD = TRUE,
  components = 5,
  alpha_random = TRUE,
  param_specs = list(
    gpd = list(
      threshold = list(
        mode = "dist",
        dist = "lognormal",
        args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
      )
    )
  ),
  mcmc = mcmc
)
```

---

## Running MCMC & Diagnostics

```{r mcmc-run}
fit_gpd <- load_or_fit("v08-unconditional-DPmixGPD-CRP-fit_gpd", run_mcmc_bundle_manual(bundle_gpd))
```

```{r fit-summary}
summary(fit_gpd)
```

```{r params-gpd-crp}
params_gpd <- params(fit_gpd)
params_gpd
```

```{r fit-plots}
# S3 plot methods highlight trace + density diagnostics
plot(fit_gpd, family = c("traceplot", "density", "running"))
plot(fit_gpd, params = "alpha", family = c("traceplot", "autocorrelation", "geweke"))
```

---

## Posterior Predictions: Bulk and Tail

We use the **S3 `predict()` objects** and their dedicated `plot()` methods to visualize densities, survival curves, and posterior-mean quantiles (i.e., averages of q(p | θ) over draws).

```{r density-predict}
y_min <- max(min(y_tail), .Machine$double.eps)
y_grid <- seq(y_min, max(y_tail) * 1.3, length.out = 300)
pred_density <- predict(fit_gpd, y = y_grid, type = "density")
plot(pred_density)
```

```{r survival-predict}
y_surv <- seq(max(u_threshold, y_min), max(y_tail) * 1.1, length.out = 60)
pred_surv <- predict(fit_gpd, y = y_surv, type = "survival")
plot(pred_surv)
```

```{r quantile-predict}
quantile_probs <- c(0.90, 0.95, 0.99)
pred_quantiles <- predict(fit_gpd, type = "quantile", index = quantile_probs, interval = "credible")
plot(pred_quantiles)
```

---

## Bulk-Only Baseline Comparison

To demonstrate the value of tail augmentation, we fit a bulk-only lognormal DP mixture and compare posterior summaries against the InvGauss+GPD fit.

```{r bulk-only-fit}
bundle_bulk_only <- build_nimble_bundle(
  y = y_tail,
  kernel = "lognormal",
  backend = "crp",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)
fit_bulk_only <- load_or_fit("v08-unconditional-DPmixGPD-CRP-fit_bulk_only", run_mcmc_bundle_manual(bundle_bulk_only))
```

```{r compare-quantiles}
bulk_quantiles <- predict(fit_bulk_only, type = "quantile", index = quantile_probs)
tail_quantiles <- predict(fit_gpd, type = "quantile", index = quantile_probs)

compare_tbl <- bind_rows(
  bulk_quantiles$fit %>% mutate(model = "Bulk only"),
  tail_quantiles$fit %>% mutate(model = "Bulk + GPD")
) %>%
  select(any_of(c("model", "index", "estimate", "lwr", "upr", "lower", "upper")))

compare_tbl %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(caption = "Posterior-Mean Quantiles for Bulk-Only vs GPD Models", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

```{r quantile-plot}
plot(bulk_quantiles)
plot(tail_quantiles)
```

---

## Extreme Value Diagnostics & Residuals

```{r return-levels}
probs <- c(0.995, 0.99, 0.975)
return_levels <- predict(fit_gpd, type = "quantile", index = probs)

return_levels$fit %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(caption = "Extreme Quantile Estimates (Posterior Mean and Credible Intervals)", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

```{r residuals}
fit_vals <- fitted(fit_gpd)
plot(fit_vals)
```

---

## Threshold Sensitivity

```{r threshold-sensitivity}
sensitivity_fits <- lapply(thresholds, function(u) {
  bundle <- build_nimble_bundle(
    y = y_tail,
    kernel = "invgauss",
    backend = "crp",
    GPD = TRUE,
    param_specs = list(
      gpd = list(
        threshold = list(mode = "fixed", value = u)
      )
    ),
    components = 5,
    mcmc = mcmc
  )
  fit <- load_or_fit("v08-unconditional-DPmixGPD-CRP-fit", run_mcmc_bundle_manual(bundle))
  list(threshold = u, fit = fit)
})

sensitivity_tbl <- bind_rows(lapply(sensitivity_fits, function(entry) {
  quant <- predict(entry$fit, type = "quantile", index = 0.99, interval = "credible")
  qfit <- as.data.frame(quant$fit)
  qfit$threshold <- entry$threshold

  # Keep the usual columns if present
  keep <- intersect(names(qfit), c("threshold", "index", "estimate", "lower", "upper", "lwr", "upr"))
  qfit <- qfit[, keep, drop = FALSE]

  # Standardize to one set of CI column names
  if (!"lower" %in% names(qfit) && "lwr" %in% names(qfit)) qfit$lower <- qfit$lwr
  if (!"upper" %in% names(qfit) && "upr" %in% names(qfit)) qfit$upper <- qfit$upr

  # If estimate isn't present for some reason, fall back to the first numeric column
  if (!"estimate" %in% names(qfit)) {
    num_cols <- names(qfit)[vapply(qfit, is.numeric, logical(1))]
    if (length(num_cols) > 0) qfit$estimate <- qfit[[num_cols[1]]]
  }

  out <- data.frame(
    threshold = qfit$threshold[1],
    q_99 = qfit$estimate[1],
    q_99_lwr = if ("lower" %in% names(qfit)) qfit$lower[1] else NA_real_,
    q_99_upr = if ("upper" %in% names(qfit)) qfit$upper[1] else NA_real_
  )
  tibble::as_tibble(out)
}))

sensitivity_tbl %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(caption = "Threshold Sensitivity: 99th Quantile", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

---

## Key Takeaways

- **Bulk + Tail**: DPmix explains the bulk, and GPD captures the extremes.
- **Threshold**: 0.8 quantile is our running default, but sensitivity checks guide the choice.
- **S3 workflows**: `predict()` and its `plot()` method make it easy to compare density, survival, and quantiles.
- **Comparison**: Bulk-only fits under-estimate extreme quantiles without a GPD component.
- **Next**: Explore Stick-Breaking (`v07`) for another backend with tail augmentation.





