---
title: "9. Unconditional DPmixGPD with Stick-Breaking Backend"
output: 
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{9. Unconditional DPmixGPD with Stick-Breaking Backend}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r legacy-setup, include=FALSE}
source(file.path(dirname(knitr::current_input(dir = TRUE)), "_legacy-setup.R"))
```


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(nimble)
library(DPmixGPD)

# Precomputed results for fast website builds
PRECOMP_DIR <- file.path("articles", "legacy-precomputed")
if (!dir.exists(PRECOMP_DIR)) dir.create(PRECOMP_DIR, recursive = TRUE, showWarnings = FALSE)

.precomp_path <- function(tag) file.path(PRECOMP_DIR, paste0(tag, ".rds"))

load_or_fit <- function(tag, expr) {
  path <- .precomp_path(tag)
  if (isTRUE(FAST) && file.exists(path)) {
    return(readRDS(path))
  }
  # If FAST but cache missing, run the expression with fast MCMC and save.
  out <- eval(substitute(expr), envir = parent.frame())
  # Save only if it looks like a fit object
  try(saveRDS(out, path), silent = TRUE)
  out
}


quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == 'windows') 'NUL' else '/dev/null'
  out <- NULL
  utils::capture.output(
    out <- eval(substitute(expr), envir = parent.frame()),
    file = nullfile
  )
  out
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).


# Unconditional DPmixGPD: Stick-Breaking (SB) Backend with Tail Augmentation

**Purpose**: Demonstrate the stick-breaking backend (`components = J`) while augmenting the extreme tail with a GPD. This mirrors the CRP+GPD pipeline (v06) but highlights the fixed-truncation behavior for bulk components.

---

## Data Setup

```{r data-setup}
# Tail-heavy data
data("nc_pos_tail200_k4")
y_tail <- nc_pos_tail200_k4$y

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y_tail), mean(y_tail), sd(y_tail), min(y_tail), max(y_tail))
)

df_data <- data.frame(y = y_tail)

p_raw <- ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "darkorange", alpha = 0.6, color = "black") +
  geom_density(color = "darkred", linewidth = 1) +
  labs(title = "Tail-Designed Data", x = "y", y = "Density") +
  theme_minimal()

grid.arrange(p_raw, ncol = 1)
```

```{r data-table, echo=FALSE}
summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Tail Dataset Summaries", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

---

## Threshold Selection

```{r threshold}
thresholds <- quantile(y_tail, c(0.70, 0.75, 0.80, 0.85))
u_threshold <- thresholds["80%"]

ggplot(df_data, aes(x = y)) +
  geom_histogram(aes(y = after_stat(density)), bins = 40, fill = "skyblue", alpha = 0.6, color = "black") +
  geom_vline(xintercept = u_threshold, linetype = "dashed", color = "black") +
  labs(title = paste("Threshold at", round(u_threshold, 2)), x = "y", y = "Density") +
  theme_minimal()
```

---

## Model Specification & Bundle

This follows the same structure as the SB bulk-only vignette (`v05`): build a bundle with `build_nimble_bundle()`, run MCMC, then use the S3 `predict()` + `plot()` helpers. Compared with `v06` (CRP+GPD), here we keep the **stick-breaking** backend and use a **Gamma** bulk kernel with a lognormal threshold prior, then contrast it with a bulk-only **Laplace** fit.

```{r spec-bundle}
bundle_sb_gpd <- build_nimble_bundle(
  y = y_tail,
  kernel = "gamma",
  backend = "sb",
  GPD = TRUE,
  components = 5,
  param_specs = list(
    gpd = list(
      threshold = list(
        mode = "dist",
        dist = "lognormal",
        args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
      )
    )
  ),
  mcmc = mcmc
)
```

---

## Running MCMC

```{r mcmc-run}
fit_sb_gpd <- load_or_fit("v09-unconditional-DPmixGPD-SB-fit_sb_gpd", run_mcmc_bundle_manual(bundle_sb_gpd))
summary(fit_sb_gpd)
```

```{r params-sb-gpd}
params_sb_gpd <- params(fit_sb_gpd)
params_sb_gpd
```

---

## Posterior Predictions

```{r density-predict}
y_grid <- seq(0, max(y_tail) * 1.3, length.out = 300)
pred_density <- predict(fit_sb_gpd, y = y_grid, type = "density")
plot(pred_density)
```

```{r survival-predict}
y_surv <- seq(u_threshold, max(y_tail) * 1.1, length.out = 60)
pred_surv <- predict(fit_sb_gpd, y = y_surv, type = "survival")
plot(pred_surv)
```

```{r quantile-predict}
quant_probs <- c(0.90, 0.95, 0.99)
pred_quant <- predict(fit_sb_gpd, type = "quantile", index = quant_probs, interval = "credible")
plot(pred_quant)
```

---

## Tail vs Bulk Comparison

```{r bulk-comparison}
bundle_sb_bulk <- build_nimble_bundle(
  y = y_tail,
  kernel = "laplace",
  backend = "sb",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)
fit_sb_bulk <- load_or_fit("v09-unconditional-DPmixGPD-SB-fit_sb_bulk", run_mcmc_bundle_manual(bundle_sb_bulk))

bulk_quant <- predict(fit_sb_bulk, type = "quantile", index = quant_probs)
t_quant <- predict(fit_sb_gpd, type = "quantile", index = quant_probs)

bind_rows(
  bulk_quant$fit %>% mutate(model = "Bulk-only"),
  t_quant$fit %>% mutate(model = "Bulk + GPD")
) %>%
  select(any_of(c("model", "index", "estimate", "lwr", "upr", "lower", "upper"))) %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(caption = "Quantiles: Bulk-only vs GPD-augmented", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```

```{r comparison-plots}
plot(bulk_quant)
plot(t_quant)
```

---

## Residuals & Diagnostics

```{r residuals}
fit_vals <- fitted(fit_sb_gpd)
plot(fit_vals)
```

```{r diagnostics}
plot(fit_sb_gpd, family = c("histogram", "autocorrelation", "running"))
```

---

## Key Takeaways

- Stick-breaking truncation fixes the number of bulk components but still flexibly models the tail via GPD.
- Posterior `predict()` + `plot()` workflows visualize densities, survival probabilities, and posterior-mean extreme quantiles.
- Comparing bulk-only vs GPD-augmented quantiles reveals how tail augmentation shifts the 95–99% levels.
- Next: conditional DPmix (v08–v11) to explore covariate effects before moving to causal regimes.





