---
title: "14. Causal Inference: No X (CRP) - Gamma Kernel"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{14. Causal Inference: No X (CRP) - Gamma Kernel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(DPmixGPD)

# Precomputed results for fast website builds
source("_legacy-setup.R")
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).


# Causal Inference: No Covariates (CRP)

This vignette fits **two no-X causal models** where the treatment arms are modeled independently using **unconditional** outcome models.

- Model A: CRP bulk-only (GPD = FALSE)
- Model B: CRP with GPD tail (GPD = TRUE)

---

## Data Setup (No X)

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- abs(causal_alt_real500_p4_k2$y) + 0.01
T <- causal_alt_real500_p4_k2$T

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  print()

u_threshold <- as.numeric(stats::quantile(y, 0.8, names = FALSE))
y_eval <- y[1:40]
```

---

## Model A: CRP Bulk-only (Gamma)

```{r bundle-crp-bulk}
bundle_crp_bulk <- build_causal_bundle(
  y = y,
  T = T,
  X = NULL,
  kernel = "gamma",
  backend = "crp",
  PS = FALSE,
  GPD = FALSE,
  components = 6,
  mcmc_outcome = list(niter = 300, nburnin = 80, nchains = 1, thin = 1, seed = 1)
)

bundle_crp_bulk
```

```{r fit-crp-bulk}
fit_crp_bulk <- quiet_mcmc(run_mcmc_causal(bundle_crp_bulk))
summary(fit_crp_bulk)
```

```{r predict-mean-crp-bulk}
pred_mean_bulk <- predict(fit_crp_bulk, type = "mean", interval = "credible", nsim_mean = 150)
head(pred_mean_bulk)
plot(pred_mean_bulk)
```

```{r predict-quantile-crp-bulk}
pred_q_bulk <- predict(fit_crp_bulk, type = "quantile", p = 0.5, interval = "credible")
head(pred_q_bulk)
plot(pred_q_bulk)
```

```{r predict-density-crp-bulk}
pred_d_bulk <- predict(fit_crp_bulk, y = y_eval, type = "density", interval = "credible")
head(pred_d_bulk)
plot(pred_d_bulk)
```

```{r predict-survival-crp-bulk}
pred_surv_bulk <- predict(fit_crp_bulk, y = y_eval, type = "survival", interval = "credible")
head(pred_surv_bulk)
plot(pred_surv_bulk)
```

```{r ate-crp-bulk}
ate_bulk <- ate(fit_crp_bulk, interval = "credible", nsim_mean = 150)
print(ate_bulk)
summary(ate_bulk)
```

```{r fig.width=7, fig.height=5}
ate_plots_bulk <- plot(ate_bulk)
ate_plots_bulk$treatment_effect
```

```{r qte-crp-bulk}
qte_bulk <- qte(fit_crp_bulk, probs = c(0.25, 0.5, 0.75), interval = "credible")
print(qte_bulk)
summary(qte_bulk)
```

```{r fig.width=7, fig.height=5}
qte_plots_bulk <- plot(qte_bulk)
qte_plots_bulk$treatment_effect
```

---

## Model B: CRP with GPD Tail (Gamma)

```{r bundle-crp-gpd}
param_specs_gpd <- list(
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "lognormal",
      args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
    )
  )
)

bundle_crp_gpd <- tryCatch({
  build_causal_bundle(
    y = y,
    T = T,
    X = NULL,
    kernel = "gamma",
    backend = "crp",
    PS = FALSE,
    GPD = TRUE,
    components = 6,
    param_specs = param_specs_gpd,
    mcmc_outcome = list(niter = 300, nburnin = 80, nchains = 1, thin = 1, seed = 2)
  )
}, error = function(e) {
  message("Skipping CRP + GPD example: ", conditionMessage(e))
  NULL
})

supports_crp_gpd <- !is.null(bundle_crp_gpd)
if (supports_crp_gpd) bundle_crp_gpd
```

```{r fit-crp-gpd}
if (isTRUE(supports_crp_gpd)) {
  fit_crp_gpd <- quiet_mcmc(run_mcmc_causal(bundle_crp_gpd))
  summary(fit_crp_gpd)
}
```

```{r predict-mean-crp-gpd}
if (isTRUE(supports_crp_gpd)) {
  pred_mean_gpd <- predict(fit_crp_gpd, type = "mean", interval = "credible", nsim_mean = 150)
  head(pred_mean_gpd)
  plot(pred_mean_gpd)
}
```

```{r predict-quantile-crp-gpd}
if (isTRUE(supports_crp_gpd)) {
  pred_q_gpd <- predict(fit_crp_gpd, type = "quantile", p = 0.5, interval = "credible")
  head(pred_q_gpd)
  plot(pred_q_gpd)
}
```

```{r predict-density-crp-gpd}
if (isTRUE(supports_crp_gpd)) {
  pred_d_gpd <- predict(fit_crp_gpd, y = y_eval, type = "density", interval = "credible")
  head(pred_d_gpd)
  plot(pred_d_gpd)
}
```

```{r predict-survival-crp-gpd}
if (isTRUE(supports_crp_gpd)) {
  pred_surv_gpd <- predict(fit_crp_gpd, y = y_eval, type = "survival", interval = "credible")
  head(pred_surv_gpd)
  plot(pred_surv_gpd)
}
```

```{r ate-crp-gpd}
if (isTRUE(supports_crp_gpd)) {
  ate_gpd <- ate(fit_crp_gpd, interval = "credible", nsim_mean = 150)
  print(ate_gpd)
  summary(ate_gpd)
}
```

```{r fig.width=7, fig.height=5}
if (isTRUE(supports_crp_gpd)) {
  ate_plots_gpd <- plot(ate_gpd)
  ate_plots_gpd$treatment_effect
}
```

```{r qte-crp-gpd}
if (isTRUE(supports_crp_gpd)) {
  qte_gpd <- qte(fit_crp_gpd, probs = c(0.25, 0.5, 0.75), interval = "credible")
  print(qte_gpd)
  summary(qte_gpd)
}
```

```{r fig.width=7, fig.height=5}
if (isTRUE(supports_crp_gpd)) {
  qte_plots_gpd <- plot(qte_gpd)
  qte_plots_gpd$treatment_effect
}
```



