---
<<<<<<< HEAD
<<<<<<< HEAD
title: "16. Causal Inference: Mixed Backends (CRP+SB) - Cauchy/Laplace"
=======
title: "Legacy: 16. Causal Inference: Mixed Backends (CRP+SB) - Cauchy/Laplace"
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
title: "Legacy: 16. Causal Inference: Mixed Backends (CRP+SB) - Cauchy/Laplace"
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
<<<<<<< HEAD
<<<<<<< HEAD
  %\VignetteIndexEntry{16. Causal Inference: Mixed Backends (CRP+SB) - Cauchy/Laplace}
=======
  %\VignetteIndexEntry{Legacy: 16. Causal Inference: Mixed Backends (CRP+SB) - Cauchy/Laplace}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
  %\VignetteIndexEntry{Legacy: 16. Causal Inference: Mixed Backends (CRP+SB) - Cauchy/Laplace}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<<<<<<< HEAD
<<<<<<< HEAD
```{r legacy-setup, include=FALSE}
source(file.path(dirname(knitr::current_input(dir = TRUE)), "_legacy-setup.R"))
```


=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
<<<<<<< HEAD
<<<<<<< HEAD
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(DPmixGPD)
library(nimble)

# Precomputed results for fast website builds
PRECOMP_DIR <- file.path("articles", "legacy-precomputed")
if (!dir.exists(PRECOMP_DIR)) dir.create(PRECOMP_DIR, recursive = TRUE, showWarnings = FALSE)

.precomp_path <- function(tag) file.path(PRECOMP_DIR, paste0(tag, ".rds"))

load_or_fit <- function(tag, expr) {
  path <- .precomp_path(tag)
  if (isTRUE(FAST) && file.exists(path)) {
    return(readRDS(path))
  }
  # If FAST but cache missing, run the expression with fast MCMC and save.
  out <- eval(substitute(expr), envir = parent.frame())
  # Save only if it looks like a fit object
  try(saveRDS(out, path), silent = TRUE)
  out
}


quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == 'windows') 'NUL' else '/dev/null'
=======
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  fig.height = 6,
  message = FALSE,
  warning = FALSE,
  eval = TRUE
)
library(DPmixGPD)
if (requireNamespace("devtools", quietly = TRUE)) devtools::load_all(quiet = TRUE)
library(ggplot2)
library(kableExtra)
library(dplyr)
library(tibble)
set.seed(123)

quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == "windows") "NUL" else "/dev/null"
<<<<<<< HEAD
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  out <- NULL
  utils::capture.output(
    out <- eval(substitute(expr), envir = parent.frame()),
    file = nullfile
  )
  out
}
```

<<<<<<< HEAD
<<<<<<< HEAD

> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).

=======
> **Legacy note:** This page is preserved for historical context and extended detail. It predates the streamlined official vignettes and may include longer runs or exploratory material.
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
> **Legacy note:** This page is preserved for historical context and extended detail. It predates the streamlined official vignettes and may include longer runs or exploratory material.
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090

# Causal Inference: Mixed Backends (CRP+SB) {#causal-mixed-backends}

This vignette fits two causal models using different backends and kernels:

- Model A: CRP bulk-only (Cauchy)
- Model B: SB with GPD tail (Laplace)

Each model demonstrates the full causal S3 workflow: bundle, fit, predict, plot, ATE, and QTE.

---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- causal_alt_real500_p4_k2$y
T <- causal_alt_real500_p4_k2$T
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Causal Dataset Summary", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")

x_eval <- X[1:40, , drop = FALSE]
y_eval <- y[1:40]
u_threshold <- as.numeric(stats::quantile(y, 0.8, names = FALSE))
```

```{r data-scatter}
df_causal <- data.frame(y = y, T = as.factor(T), x1 = X[, 1], x2 = X[, 2])

p_scatter <- ggplot(df_causal, aes(x = x1, y = y, color = T)) +
  geom_point(alpha = 0.5) +
  labs(title = "Outcome vs X1 by Treatment", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()

p_scatter
```

---

## Model A: CRP Bulk-only (Cauchy)

```{r bundle-crp-bulk}
bundle_crp_bulk <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "cauchy",
  backend = "crp",
  PS = "logit",
  GPD = FALSE,
  components = 5,
  mcmc_outcome = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 1),
  mcmc_ps = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 1)
)

bundle_crp_bulk
summary(bundle_crp_bulk)
```

```{r fit-crp-bulk}
fit_crp_bulk <- quiet_mcmc(run_mcmc_causal(bundle_crp_bulk))
summary(fit_crp_bulk)
params(fit_crp_bulk)
```

```{r plot-fit-crp-bulk}
plot(fit_crp_bulk, family = c("traceplot", "autocorrelation", "running"))
```

```{r predict-mean-crp-bulk}
pred_mean_bulk <- predict(fit_crp_bulk, x = x_eval, type = "mean",
                          interval = "credible", nsim_mean = 100)
head(pred_mean_bulk)
plot(pred_mean_bulk)
```

```{r predict-quantile-crp-bulk}
pred_q_bulk <- predict(fit_crp_bulk, x = x_eval, type = "quantile",
                       p = 0.5, interval = "credible")
head(pred_q_bulk)
plot(pred_q_bulk)
```

```{r predict-density-crp-bulk}
pred_d_bulk <- predict(fit_crp_bulk, x = x_eval, y = y_eval,
                       type = "density", interval = "credible")
head(pred_d_bulk)
plot(pred_d_bulk)
```

```{r predict-survival-crp-bulk}
pred_surv_bulk <- predict(fit_crp_bulk, x = x_eval, y = y_eval,
                          type = "survival", interval = "credible")
head(pred_surv_bulk)
plot(pred_surv_bulk)
```

```{r ate-crp-bulk}
ate_bulk <- ate(fit_crp_bulk, newdata = x_eval,
                interval = "credible", nsim_mean = 100)
head(ate_bulk)
plot(ate_bulk)
```

```{r qte-crp-bulk}
qte_bulk <- qte(fit_crp_bulk, probs = c(0.25, 0.5, 0.75),
                newdata = x_eval, interval = "credible")
head(qte_bulk)
plot(qte_bulk)
```

---

## Model B: SB with GPD Tail (Laplace)

```{r bundle-crp-gpd}
param_specs_gpd <- list(
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "lognormal",
      args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
    )
  )
)

bundle_sb_gpd <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "laplace",
  backend = "sb",
  PS = "logit",
  GPD = TRUE,
  components = 5,
  param_specs = param_specs_gpd,
  mcmc_outcome = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 2),
  mcmc_ps = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 2)
)

bundle_sb_gpd
summary(bundle_sb_gpd)
```

```{r fit-sb-gpd}
fit_sb_gpd <- quiet_mcmc(run_mcmc_causal(bundle_sb_gpd))
summary(fit_sb_gpd)
params(fit_sb_gpd)
```

```{r plot-fit-sb-gpd}
plot(fit_sb_gpd, family = c("density", "geweke", "caterpillar"))
```

```{r predict-mean-sb-gpd}
pred_mean_gpd <- predict(fit_sb_gpd, x = x_eval, type = "mean",
                         interval = "credible", nsim_mean = 100)
head(pred_mean_gpd)
plot(pred_mean_gpd)
```

```{r predict-quantile-sb-gpd}
pred_q_gpd <- predict(fit_sb_gpd, x = x_eval, type = "quantile",
                      p = 0.5, interval = "credible")
head(pred_q_gpd)
plot(pred_q_gpd)
```

```{r predict-density-sb-gpd}
pred_d_gpd <- predict(fit_sb_gpd, x = x_eval, y = y_eval,
                      type = "density", interval = "credible")
head(pred_d_gpd)
plot(pred_d_gpd)
```

```{r predict-survival-sb-gpd}
pred_surv_gpd <- predict(fit_sb_gpd, x = x_eval, y = y_eval,
                         type = "survival", interval = "credible")
head(pred_surv_gpd)
plot(pred_surv_gpd)
```

```{r ate-sb-gpd}
ate_gpd <- ate(fit_sb_gpd, newdata = x_eval,
               interval = "credible", nsim_mean = 100)
head(ate_gpd)
plot(ate_gpd)
```

```{r qte-sb-gpd}
qte_gpd <- qte(fit_sb_gpd, probs = c(0.25, 0.5, 0.75),
               newdata = x_eval, interval = "credible")
head(qte_gpd)
plot(qte_gpd)
```






