---
title: "17. Causal Inference: Same Backend (SB) - Laplace Kernel"
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
  %\VignetteIndexEntry{17. Causal Inference: Same Backend (SB) - Laplace Kernel}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(DPmixGPD)

# Precomputed results for fast website builds
source("_legacy-setup.R")
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).


# Causal Inference: Same Backend (SB) - Laplace Kernel {#causal-sb-same-backend}

This vignette fits two SB-based causal models using the same kernel (Laplace):

- Model A: bulk-only outcome models (GPD = FALSE)
- Model B: GPD-augmented outcome models (GPD = TRUE)

---

## Data Setup

```{r data-setup}
data("causal_alt_real500_p4_k2")
y <- causal_alt_real500_p4_k2$y
T <- causal_alt_real500_p4_k2$T
X <- as.matrix(causal_alt_real500_p4_k2$X)

summary_tbl <- tibble(
  statistic = c("N", "Mean", "SD", "Min", "Max"),
  value = c(length(y), mean(y), sd(y), min(y), max(y))
)

summary_tbl %>%
  mutate(value = signif(value, 4)) %>%
  kable(caption = "Causal Dataset Summary", align = "c") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")

x_eval <- X[1:40, , drop = FALSE]
y_eval <- y[1:40]
u_threshold <- as.numeric(stats::quantile(y, 0.8, names = FALSE))
```

```{r data-scatter}
df_causal <- data.frame(y = y, T = as.factor(T), x1 = X[, 1], x2 = X[, 2])

p_scatter <- ggplot(df_causal, aes(x = x1, y = y, color = T)) +
  geom_point(alpha = 0.5) +
  labs(title = "Outcome vs X1 by Treatment", x = "X1", y = "y", color = "Treatment") +
  theme_minimal()

p_scatter
```

---

## Model A: SB Bulk-only (Laplace)

```{r bundle-sb-bulk}
bundle_sb_bulk <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "laplace",
  backend = "sb",
  PS = "logit",
  GPD = FALSE,
  components = 5,
  mcmc_outcome = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 1),
  mcmc_ps = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 1)
)

bundle_sb_bulk
summary(bundle_sb_bulk)
```

```{r fit-sb-bulk}
fit_sb_bulk <- run_mcmc_causal(bundle_sb_bulk)
summary(fit_sb_bulk)
params(fit_sb_bulk)
```

```{r plot-fit-sb-bulk}
plot(fit_sb_bulk, family = c("traceplot", "autocorrelation", "running"))
```

```{r predict-mean-sb-bulk}
pred_mean_bulk <- predict(fit_sb_bulk, x = x_eval, type = "mean",
                          interval = "hpd", nsim_mean = 100)
plot(pred_mean_bulk)
```

```{r predict-quantile-sb-bulk}
pred_q_bulk <- predict(fit_sb_bulk, x = x_eval, type = "quantile",
                       p = 0.5, interval = "hpd")
plot(pred_q_bulk)
```

```{r predict-density-sb-bulk}
pred_d_bulk <- predict(fit_sb_bulk, x = x_eval, y = y_eval,
                       type = "density", interval = "hpd")
plot(pred_d_bulk)
```

```{r predict-survival-sb-bulk}
pred_surv_bulk <- predict(fit_sb_bulk, x = x_eval, y = y_eval,
                          type = "survival", interval = "hpd")
plot(pred_surv_bulk)
```

```{r ate-sb-bulk}
ate_bulk <- ate(fit_sb_bulk, newdata = x_eval,
                interval = "hpd", nsim_mean = 100)
print(ate_bulk)
summary(ate_bulk)
```

```{r fig.width=7, fig.height=5}
ate_plots_bulk <- plot(ate_bulk)
ate_plots_bulk$treatment_effect
```

```{r qte-sb-bulk}
qte_bulk <- qte(fit_sb_bulk, probs = c(0.25, 0.5, 0.75),
                newdata = x_eval, interval = "hpd")
print(qte_bulk)
summary(qte_bulk)
```

```{r fig.width=7, fig.height=5}
qte_plots_bulk <- plot(qte_bulk)
qte_plots_bulk$treatment_effect
```

---

## Model B: SB with GPD Tail (Laplace)

```{r bundle-sb-gpd}
param_specs_gpd <- list(
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "lognormal",
      args = list(meanlog = log(max(u_threshold, .Machine$double.eps)), sdlog = 0.25)
    )
  )
)

bundle_sb_gpd <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = "laplace",
  backend = "sb",
  PS = "logit",
  GPD = TRUE,
  components = 5,
  param_specs = param_specs_gpd,
  mcmc_outcome = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 2),
  mcmc_ps = list(niter = 2500, nburnin = 500, nchains = 1, thin = 1, seed = 2)
)

bundle_sb_gpd
summary(bundle_sb_gpd)
```

```{r fit-sb-gpd}
fit_sb_gpd <- run_mcmc_causal(bundle_sb_gpd)
summary(fit_sb_gpd)
params(fit_sb_gpd)
```

```{r plot-fit-sb-gpd}
plot(fit_sb_gpd, family = c("density", "geweke", "caterpillar"))
```

```{r predict-mean-sb-gpd}
pred_mean_gpd <- predict(fit_sb_gpd, x = x_eval, type = "mean",
                         interval = "credible", nsim_mean = 100)
plot(pred_mean_gpd)
```

```{r predict-quantile-sb-gpd}
pred_q_gpd <- predict(fit_sb_gpd, x = x_eval, type = "quantile",
                      p = 0.5, interval = "credible")
plot(pred_q_gpd)
```

```{r predict-density-sb-gpd}
pred_d_gpd <- predict(fit_sb_gpd, x = x_eval, y = y_eval,
                      type = "density", interval = "credible")
plot(pred_d_gpd)
```

```{r predict-survival-sb-gpd}
pred_surv_gpd <- predict(fit_sb_gpd, x = x_eval, y = y_eval,
                         type = "survival", interval = "credible")
plot(pred_surv_gpd)
```

```{r ate-sb-gpd}
ate_gpd <- ate(fit_sb_gpd, newdata = x_eval,
               interval = "credible", nsim_mean = 100)
print(ate_gpd)
summary(ate_gpd)
```

```{r fig.width=7, fig.height=5}
ate_plots_gpd <- plot(ate_gpd)
ate_plots_gpd$treatment_effect
```

```{r qte-sb-gpd}
qte_gpd <- qte(fit_sb_gpd, probs = c(0.25, 0.5, 0.75),
               newdata = x_eval, interval = "credible")
print(qte_gpd)
summary(qte_gpd)
```

```{r fig.width=7, fig.height=5}
qte_plots_gpd <- plot(qte_gpd)
qte_plots_gpd$treatment_effect
```






