---
<<<<<<< HEAD
<<<<<<< HEAD
title: "20. Troubleshooting"
=======
title: "Legacy: 20. Troubleshooting"
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
title: "Legacy: 20. Troubleshooting"
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
output:
  rmarkdown::html_vignette:
    code_folding: show
vignette: >
<<<<<<< HEAD
<<<<<<< HEAD
  %\VignetteIndexEntry{20. Troubleshooting}
=======
  %\VignetteIndexEntry{Legacy: 20. Troubleshooting}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
  %\VignetteIndexEntry{Legacy: 20. Troubleshooting}
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

<<<<<<< HEAD
<<<<<<< HEAD
```{r legacy-setup, include=FALSE}
source(file.path(dirname(knitr::current_input(dir = TRUE)), "_legacy-setup.R"))
```


=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
<<<<<<< HEAD
<<<<<<< HEAD
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = TRUE,
  cache.path = 'legacy-cache/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

library(nimble)
library(DPmixGPD)

# Precomputed results for fast website builds
PRECOMP_DIR <- file.path("articles", "legacy-precomputed")
if (!dir.exists(PRECOMP_DIR)) dir.create(PRECOMP_DIR, recursive = TRUE, showWarnings = FALSE)

.precomp_path <- function(tag) file.path(PRECOMP_DIR, paste0(tag, ".rds"))

load_or_fit <- function(tag, expr) {
  path <- .precomp_path(tag)
  if (isTRUE(FAST) && file.exists(path)) {
    return(readRDS(path))
  }
  # If FAST but cache missing, run the expression with fast MCMC and save.
  out <- eval(substitute(expr), envir = parent.frame())
  # Save only if it looks like a fit object
  try(saveRDS(out, path), silent = TRUE)
  out
}


quiet_mcmc <- function(expr) {
  nullfile <- if (.Platform$OS.type == 'windows') 'NUL' else '/dev/null'
  out <- NULL
  utils::capture.output(
    out <- eval(substitute(expr), envir = parent.frame()),
    file = nullfile
  )
  out
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (Introduction, Model Spec, MCMC Workflow, Unconditional/Conditional/Causal, Backends, S3 Reference).

=======
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
  message = FALSE,
  warning = FALSE,
  eval = TRUE
)
library(DPmixGPD)
```

> **Legacy note:** This page is preserved for historical context and extended detail. It predates the streamlined official vignettes and may include longer runs or exploratory material.
<<<<<<< HEAD
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090

# Troubleshooting

This vignette collects common errors and fixes encountered when building or running DPmixGPD models.

---

## 1. "Number of columns in 'x' does not match training design matrix"

**Cause**: New data has different columns than the training `X`.

**Fix**: Use the same column names and order as the training matrix.

```r
X_train <- fit$data$X
X_new <- X_train[1:5, , drop = FALSE]
X_new <- X_new[, colnames(X_train), drop = FALSE]

predict(fit, x = X_new, type = "mean")
```

---

## 2. "Cauchy kernels are never paired with GPD tails"

**Cause**: The Cauchy kernel is not allowed with GPD tails.

**Fix**: Use a different kernel or disable GPD.

```r
bundle <- build_nimble_bundle(
  y = y,
  kernel = "cauchy",
  backend = "sb",
  GPD = FALSE
)
```

---

## 3. "design='observational' requires non-empty X"

**Cause**: Observational causal modeling requires covariates for PS estimation.

**Fix**: Provide `X`, or switch to `design = "rct"`.

```r
cb <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  design = "observational",
  PS = "logit"
)
```

---

## 4. "GPD tail not moving" (threshold/shape stuck)

**Cause**: Very short chains or overly restrictive priors.

**Fix**: Increase iterations or relax tail priors.

```r
bundle <- build_nimble_bundle(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = TRUE,
  components = 6,
<<<<<<< HEAD
<<<<<<< HEAD
  mcmc = mcmc
=======
  mcmc = list(niter = 1500, nburnin = 300, thin = 1, nchains = 2)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
  mcmc = list(niter = 1500, nburnin = 300, thin = 1, nchains = 2)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
)
```

---

## 5. "future.globals.maxSize exceeded"

**Cause**: Parallel predictions export large functions/objects.

**Fix**: Set a higher limit or use `ncores = 1`.

```r
options(future.globals.maxSize = 4 * 1024^3)

pred <- predict(fit, x = X_new, y = y_grid, type = "density", ncores = 2)
```

---

## 6. MCMC is slow

**Cause**: Large datasets, many components, or heavy kernels.

**Fix**: Use smaller `components`/`J`, shorter `niter`, or start with CRP.

```r
bundle <- build_nimble_bundle(
  y = y,
  backend = "crp",
  kernel = "normal",
  components = 4,
<<<<<<< HEAD
<<<<<<< HEAD
  mcmc = mcmc
=======
  mcmc = list(niter = 300, nburnin = 80, nchains = 1)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
=======
  mcmc = list(niter = 300, nburnin = 80, nchains = 1)
>>>>>>> 50289162bd36853addda01bb01ee507dfa332090
)
```

---

## 7. Prediction returns constant fitted values in unconditional models

**Cause**: Unconditional fitted values are model-based population summaries.

**Fix**: This is expected. Use conditional models if you need observation-specific fitted values.

```r
fit_vals <- fitted(fit)
head(fit_vals)
```

---

## 8. PS not used in causal model

**Cause**: `PS=FALSE` or `X` missing.

**Fix**: Provide `X` and enable PS.

```r
cb <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  PS = "logit"
)
fit <- run_mcmc_causal(cb)
```

---

## 9. Kernel mismatch when mixing arms

**Cause**: Using kernels with different supports across arms.

**Fix**: Ensure both arms use kernels on the same support (e.g., both positive-support or both real-line).

```r
cb <- build_causal_bundle(
  y = y,
  T = T,
  X = X,
  kernel = c("gamma", "lognormal"),
  backend = c("sb", "crp")
)
```



