---
title: "Umbrella: Custom Models (All-in-One)"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Umbrella: Custom Models (All-in-One)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
devtools::load_all()
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(ggplot2)
library(kableExtra)
library(DPmixGPD)
set.seed(1)

use_fast_mcmc <- TRUE
mcmc_long <- list(niter = 2000, nburnin = 500, thin = 3, nchains = 3, seed = 1:3)
mcmc_fast <- list(niter = 600, nburnin = 150, thin = 2, nchains = 1, seed = 1)
mcmc <- if (use_fast_mcmc) mcmc_fast else mcmc_long
```

# Overview

This vignette collects two end-to-end examples with customized priors and links:

1) **Conditional CRP + Amoroso + GPD** with covariates, custom links, and tail settings.  
2) **Causal model (two arms)** with SB backends: Normal (treated) and Laplace (control),
   with covariate-linked locations and custom priors.

Both examples use `mtcars` and show the main **user-level functions** plus **S3 methods**
(`print`, `summary`, `plot`, `predict`, `fitted`, `residuals`, `params`, `ate`, `qte`).

> Note on link priors: both SB and CRP backends currently require **normal** priors for
> link-mode regression coefficients. The requested **lognormal** prior for the scale
> link is therefore shown below as a **non-executed** spec block.

# Data

```{r}
library(DPmixGPD)

data("mtcars", package = "datasets")
df <- mtcars
y <- df$mpg
X <- df[, c("wt", "hp")]
X <- as.data.frame(X)
```

# Model 1: CRP + Amoroso + GPD (Conditional)

## Specification

- **Backend**: CRP  
- **Kernel**: Amoroso  
- **Covariates**: `wt`, `hp`  
- **Bulk**:
  - `loc`: identity link, normal prior on coefficients
  - `scale`: identity link, normal prior on coefficients (CRP restriction)
  - `shape1`: fixed at 1
  - `shape2`: default prior
- **Tail**:
  - `threshold`: exp link, normal prior on coefficients
  - `tail_scale`: exp link (default prior)

```{r}
param_specs_amoroso <- list(
  bulk = list(
    loc = list(
      mode = "link",
      link = "identity",
      beta_prior = list(dist = "normal", args = list(mean = 0, sd = 2))
    ),
    scale = list(
      mode = "link",
      link = "identity",
      beta_prior = list(dist = "normal", args = list(mean = 0, sd = 2))
    ),
    shape1 = list(mode = "fixed", value = 1)
  ),
  gpd = list(
    threshold = list(
      mode = "link",
      link = "exp",
      beta_prior = list(dist = "normal", args = list(mean = 0, sd = 0.2))
    ),
    tail_scale = list(
      mode = "link",
      link = "exp"
    )
  )
)
```

```{r, eval=FALSE}
# Requested variant (not runnable): link-mode beta priors must be normal
param_specs_amoroso_requested <- list(
  bulk = list(
    loc = list(
      mode = "link",
      link = "identity",
      beta_prior = list(dist = "normal", args = list(mean = 0, sd = 2))
    ),
    scale = list(
      mode = "link",
      link = "identity",
      beta_prior = list(dist = "lognormal", args = list(meanlog = 0, sdlog = 1))
    ),
    shape1 = list(mode = "fixed", value = 1)
  ),
  gpd = list(
    threshold = list(
      mode = "link",
      link = "exp",
      beta_prior = list(dist = "normal", args = list(mean = 0, sd = 0.2))
    ),
    tail_scale = list(
      mode = "link",
      link = "exp"
    )
  )
)
```

## Build + Run (CRP note)

NIMBLE's CRP sampler cannot cluster **deterministic** nodes created by link-mode
parameters. As a result, CRP + covariate links are not runnable with the current
backend. We still record the **requested CRP specification** below (for reference),
and then run an **SB-equivalent** model that supports the same customization and
lets this vignette compile end-to-end.

```{r, eval=FALSE}
# Requested CRP specification (not runnable due to CRP link-mode restriction)
bundle_amoroso_crp <- build_nimble_bundle(
  y = y,
  X = X,
  backend = "crp",
  kernel = "amoroso",
  GPD = TRUE,
  components = 5,
  param_specs = param_specs_amoroso,
  mcmc = mcmc
)
```

```{r, results='hide'}
# Runnable SB equivalent (supports link-mode parameters)
bundle_amoroso <- build_nimble_bundle(
  y = y,
  X = X,
  backend = "sb",
  kernel = "amoroso",
  GPD = TRUE,
  components = 5,
  param_specs = param_specs_amoroso,
  mcmc = mcmc
)

fit_amoroso <- run_mcmc_bundle_manual(bundle_amoroso, show_progress = FALSE)
```

## User-Level Functions

```{r}
print(bundle_amoroso)
summary(bundle_amoroso)
```

```{r}
print(fit_amoroso)
summary(fit_amoroso)
```

## S3 Methods (Fit)

```{r}
tryCatch(
  plot(fit_amoroso, family = "trace"),
  error = function(e) {
    message("Plotting 'fit_amoroso' with family = 'trace' failed: ", conditionMessage(e))
  }
)
```

```{r}
pred_mean <- predict(fit_amoroso, x = X, type = "mean", cred.level = 0.90, interval = "credible")
pred_q90 <- predict(fit_amoroso, x = X, type = "quantile", index = 0.90, cred.level = 0.90, interval = "credible")
head(pred_mean$fit)
head(pred_q90$fit)
```

```{r}
fvals <- fitted(fit_amoroso, type = "mean", level = 0.90)
head(fvals)
summary(fvals$residuals)
```

```{r}
raw_resid <- residuals(fit_amoroso, type = "raw")
head(raw_resid)
```

```{r}
p <- params(fit_amoroso)
names(p)
```

# Model 2: Causal (SB Normal vs SB Laplace)

## Specification

- **Backend**: SB (both arms)  
- **Treated arm (T=1)**: Normal kernel  
  - Location: identity link, Gaussian prior  
  - SD: Inverse-Gamma prior  
- **Control arm (T=0)**: Laplace kernel  
  - Location: identity link, Gaussian prior  
  - Scale: Gamma prior  

```{r}
X_causal <- df[, c("wt", "hp", "qsec", "cyl")]
X_causal <- as.data.frame(X_causal)
T_ind <- df$am

param_specs_causal <- list(
  trt = list(
    bulk = list(
      mean = list(
        mode = "link",
        link = "identity",
        beta_prior = list(dist = "normal", args = list(mean = 0, sd = 2))
      ),
      sd = list(
        mode = "dist",
        dist = "invgamma",
        args = list(shape = 2, scale = 1)
      )
    )
  ),
  con = list(
    bulk = list(
      location = list(
        mode = "link",
        link = "identity",
        beta_prior = list(dist = "normal", args = list(mean = 0, sd = 2))
      ),
      scale = list(
        mode = "dist",
        dist = "gamma",
        args = list(shape = 2, rate = 1)
      )
    )
  )
)
```

## Build + Run

```{r, results='hide'}
causal_bundle <- build_causal_bundle(
  y = y,
  X = X_causal,
  T = T_ind,
  backend = "sb",
  kernel = c("normal", "laplace"),
  GPD = FALSE,
  components = 5,
  param_specs = param_specs_causal,
  PS = "logit",
  design = "observational",
  mcmc_outcome = mcmc,
  mcmc_ps = mcmc
)

causal_fit <- run_mcmc_causal(causal_bundle, show_progress = FALSE)
```

## User-Level Functions

```{r}
print(causal_bundle)
summary(causal_bundle)
```

```{r}
print(causal_fit)
summary(causal_fit)
```

## S3 Methods (Causal)

```{r}
try(plot(causal_fit, family = "trace"), silent = TRUE)
```

```{r}
pred_causal_mean <- predict(causal_fit, x = X_causal, type = "mean", interval = "credible")
head(pred_causal_mean)
```

```{r}
ate_result <- ate(causal_fit, interval = "credible", nsim_mean = 50)
print(ate_result)
summary(ate_result)
```

```{r}
qte_result <- qte(causal_fit, probs = c(0.25, 0.5, 0.75), interval = "credible")
# Plotting may fail on some graphics devices (e.g., non-interactive or CI environments),
# so errors are intentionally suppressed to ensure the vignette renders successfully.
try({
  plot(ate_result, type = "effect")
  plot(qte_result, type = "effect")
}, silent = TRUE)
```


```{r sessioninfo, include=FALSE}
sessionInfo()
```
