---
title: "Backends, Kernels, and Workflow Map"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = '../examples/legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
DO_PLOTS <- !isTRUE(FAST)
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, workers = 2)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_chains = TRUE, workers = 2, z_update_every = 2)

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(CausalMixGPD)

# Ensure package is fully loaded for data access

# Precomputed results for fast website builds
source("_legacy-setup.R")
```


> **Website workflow note.**
> This page reflects the current exported API and recommended wrapper-first usage.
> Last updated: **2026-02-19**.
>
> For the full package narrative, see the main package vignettes (basic, unconditional, conditional, and causal).

[Workflow roadmap](roadmap.html){.btn .btn-outline-light}


# Big picture

CausalMixGPD has two orthogonal dials you turn when building models:

- **Backend** (how the mixture weights / clustering are represented)
  - **CRP**: Chinese Restaurant Process representation.
  - **SB**: stick-breaking truncation with a fixed number of components.

- **Kernel / family** (what distribution models the bulk of the data)
  - Examples: normal, lognormal, gamma, inverse-Gaussian, Laplace, Cauchy, Amoroso, etc.

Optionally, you can also turn on:

- **GPD = TRUE/FALSE** to splice a Generalized Pareto tail beyond a threshold.

> For the math behind the bulk--tail “spliced” construction (threshold exceedances, DPM bulk, and spliced quantiles), see [Theory: GPD tails + DPM bulk + splicing](../advanced/theory/01-gpd-dp-setup.qmd).

# What changes between CRP and SB?

Both backends target the same posterior over densities. The difference is representation:

- **CRP** learns a random number of occupied clusters within a finite `components` cap.
- **SB** uses the same finite `components` cap and learns stick-breaking weights.

Practical rule of thumb:

- **CRP** is convenient when you want adaptive complexity while still using a finite `components` cap.
- **SB** is convenient when you want predictable memory/time and easy vectorization.

# What does the workflow look like?

CausalMixGPD uses a consistent build -> run -> summarize loop:

1. **Build a bundle** using `bundle()` (or causal builders if you are doing TE work).
2. **Fit directly** using `dpmix()` (or `dpmgpd()` when `GPD = TRUE`).
3. **Inspect and summarize** using `print()`, `summary()`, `plot()`.
4. **Predict** using `predict()`. For conditional (covariate) models, `fitted()` is also available.

```{r, cache = FALSE}
# Build
bundle <- bundle(
  x = rnorm(50),
  backend = "crp",
  kernel = "normal",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)

# Run
fit <- load_or_fit("quickstart-backends-and-workflow-fit", dpmix(bundle))

# Summarize
print(fit)
summary(fit)
if (DO_PLOTS) plot(fit)
```

# Kernel support quick check

Use `kernel_support_table()` and the kernel registry helpers to confirm what is available.

```{r}
kernel_support_table()
```

## Manual runners and simulation helpers

Most users should stick to `dpmix()` / `dpmgpd()` (and their causal / clustering counterparts). For debugging or reproducibility workflows, the package also exports:

- `run_mcmc_bundle_manual()` / `run_mcmc_causal()` for an explicit **build → compile → run** pipeline (useful when you want to inspect intermediate NIMBLE objects).
- `sim_bulk_tail()`, `sim_causal_qte()`, and `sim_survival_tail()` for generating toy datasets that match the package’s bulk/tail and causal examples.

# Where to go next

- **Available distributions**: see the [Kernels hub](../kernels.html) and kernel catalog pages.
- **Basic build/compile/run**: see **Start basic-model-compile-run**.
- **Unconditional / Conditional / Causal**: continue through ex01+.

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [basic-model-compile-run](basic-model-compile-run.html)
- Next: [ex01-unconditional-dpm-crp](../examples/ex01-unconditional-dpm-crp.html)
- Workflow index: [Roadmap](roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [Start Here](start-here.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->





