---
title: "Basic Workflow: Model Specification, Bundle, & MCMC"
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = '../examples/legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
max_workers <- local({
  n_phys <- suppressWarnings(tryCatch(parallel::detectCores(logical = FALSE), error = function(e) NA_integer_))
  n_log <- suppressWarnings(tryCatch(parallel::detectCores(logical = TRUE), error = function(e) NA_integer_))
  n <- if (is.finite(n_phys) && !is.na(n_phys) && n_phys >= 1L) n_phys else n_log
  if (!is.finite(n) || is.na(n) || n < 1L) n <- 1L
  as.integer(min(10L, n))
})
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, workers = max_workers)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_chains = TRUE, workers = max_workers, z_update_every = 2)

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Load package (with all_quiet to suppress startup messages)
library(CausalMixGPD)

# Ensure package is fully loaded for data access

# Precomputed results for fast website builds
source("_legacy-setup.R")
max_workers <- min(max_workers, CMGPD_MAX_WORKERS)
```


> **Website workflow note.**
> This page reflects the current exported API and recommended wrapper-first usage.
> Last updated: **2026-02-19**.
>
> For the full package narrative, see the main package vignettes (basic, unconditional, conditional, and causal).

[Workflow roadmap](roadmap.html){.btn .btn-outline-light}


# Workflow Overview

CausalMixGPD supports a wrapper-first workflow for most users:

1. **Fit directly** with `dpmix()` or `dpmgpd()`.
2. **Inspect/predict** with `summary()`, `plot()`, and `predict()`.

The lower-level bundle route (`bundle()` then `mcmc()`) remains available for advanced control.

---

## Phase 1: Bundle (NIMBLE Code Generation & Compilation)

**Purpose**: Generate NIMBLE model code, compile sampler, prepare for MCMC execution.

> For the underlying model math (GPD threshold exceedances, DPM bulk regression, and the spliced quantile construction), see [Theory: GPD tails + DPM bulk + splicing](../advanced/theory/01-gpd-dp-setup.qmd).

### Building Directly

```{r bundle-direct, cache = FALSE}
# Load packaged data
data("nc_pos200_k3")
y <- nc_pos200_k3$y

# Direct call
bundle_direct <- bundle(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)
```

### Inspecting Bundle Contents

```{r bundle-inspect}
# Bundle is an S3 object with structure
bundle_class <- class(bundle_direct)
bundle_names <- names(bundle_direct)
bundle_mcmc <- bundle_direct$mcmc_settings

bundle_class
bundle_names
bundle_mcmc
```

---

## Phase 2: MCMC Execution

**Purpose**: Run posterior sampling from the compiled bundle.

### Basic MCMC Run

```{r mcmc-basic, cache = FALSE}
# Run MCMC
fit <- load_or_fit("quickstart-basic-model-compile-run-fit", dpmix(bundle_direct, mcmc = list(show_progress = FALSE)))
invisible(fit)
```

### Accessing Posterior Samples

```{r mcmc-posterior}
summary(fit)

# Posterior mean parameters in original form
params_fit <- params(fit)
params_fit
```

### Diagnostic Plots

```{r mcmc-diagnostics}
# Trace plots
plot(fit, params = "alpha|beta", family = c("traceplot", "running", "autocorrelation"))
```

---

## Complete Workflow: End-to-End Example

```{r workflow-complete, cache = FALSE}
# Load packaged data
data("nc_pos200_k3")
y_data <- nc_pos200_k3$y

# PHASE 1: Bundle
bundle_final <- bundle(
  y = y_data,
  kernel = "gamma",
  backend = "crp",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)

# PHASE 2: MCMC
fit_final <- load_or_fit("quickstart-basic-model-compile-run-fit_final", dpmix(bundle_final, mcmc = list(show_progress = FALSE)))
summary(fit_final)
```

---

## Backend Comparison: CRP vs Stick-Breaking

### CRP Backend

```{r backend-crp, cache = FALSE}
# Chinese Restaurant Process
bundle_crp <- bundle(
  y = y_data,
  kernel = "gamma",
  backend = "crp",
  components = 5,
  mcmc = mcmc
)

fit_crp <- load_or_fit("quickstart-basic-model-compile-run-fit_crp", dpmix(bundle_crp, mcmc = list(show_progress = FALSE)))
invisible(fit_crp)
```

### Stick-Breaking Backend

```{r backend-sb, cache = FALSE}
# Stick-Breaking Process
bundle_sb <- bundle(
  y = y_data,
  kernel = "gamma",
  backend = "sb",
  components = 5,
  mcmc = mcmc
)

fit_sb <- load_or_fit("quickstart-basic-model-compile-run-fit_sb", dpmix(bundle_sb, mcmc = list(show_progress = FALSE)))
invisible(fit_sb)
```

---

## Kernel Selection Guide

```{r kernel-guide}
kernels_available <- c("gamma", "lognormal", "normal", "laplace", "invgauss", "amoroso", "cauchy")

kernels_available
```

---

## GPD Tail Augmentation

### Unconditional with GPD

```{r gpd-example, eval=TRUE, cache = FALSE}
# Data with tail behavior
data("nc_pos_tail200_k4")
y_tail <- nc_pos_tail200_k4$y

# Build with GPD
bundle_gpd <- bundle(
  y = y_tail,
  kernel = "gamma",
  backend = "sb",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit_gpd <- load_or_fit("quickstart-basic-model-compile-run-fit_gpd", dpmgpd(bundle_gpd, mcmc = list(show_progress = FALSE)))
summary(fit_gpd)
```

---

## Summary of Key Functions

| Phase | Function | Input | Output |
|-------|----------|-------|--------|
| **1. Fit (bulk only)** | `dpmix()` | y, X (optional), kernel, backend, components, mcmc | `mixgpd_fit` |
| **2. Fit (bulk + tail)** | `dpmgpd()` | y, X (optional), kernel, backend, components, mcmc | `mixgpd_fit` |

---

## Common Parameter Settings

```{r param-reference}
recommended_mcmc <- data.frame(
  profile = c("Quick test", "Standard", "Production"),
  niter = c(500, 1000, 1000),
  nburnin = c(100, 250, 250),
  nchains = c(1, 2, 3)
)

recommended_mcmc
```

---

## Next Steps

- Move to **ex01-ex02** for unconditional models (CRP vs SB backends)
- Move to **ex03-ex04** for tail modeling with GPD
- Move to **ex05-ex08** for conditional models with covariates
- Move to **ex09-ex14** for causal inference workflows

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [start-here](start-here.html)
- Next: [backends-and-workflow](backends-and-workflow.html)
- Workflow index: [Roadmap](roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [Start Here](start-here.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->


