---
title: "Introduction to DPmixGPD"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache=TRUE,
  cache.path = '../examples/legacy-cache-v2/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2))

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy

# Ensure package is fully loaded for data access
if (!require("devtools", quietly = TRUE)) {
  library(devtools)
}
pkg_root <- rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
  devtools::load_all(path = pkg_root, quiet = TRUE)

# Precomputed results for fast website builds
source("_legacy-setup.R")
if (!exists("safe_plot", mode = "function")) {
  safe_plot <- function(x, ...) {
    obj <- tryCatch(plot(x, ...), error = function(e) e)
    if (inherits(obj, "ggplot")) {
      try(print(obj), silent = TRUE)
    } else if (is.list(obj)) {
      for (elt in obj) if (inherits(elt, "ggplot")) try(print(elt), silent = TRUE)
    }
    invisible(obj)
  }
}
```


> **Legacy vignette (for the website / historical notes).**
> These files may not match the current exported API one-to-one.
> Last verified: **2026-01-18**.
>
> For the up-to-date workflow, see the main package vignettes (basic, unconditional, conditional, and causal)..

[Workflow roadmap](../roadmap.html){.btn .btn-outline-light}


# DPmixGPD: Bayesian Nonparametric Mixture Models with GPD Tails

## Overview

**DPmixGPD** is an R package for Bayesian nonparametric Dirichlet Process mixture models with flexible kernels and optional Generalized Pareto Distribution (GPD) tail augmentation.

### Key Features

- **Flexible mixture kernels**: normal, gamma, lognormal, laplace, invgauss, amoroso, cauchy
- **Two MCMC backends**: Stick-Breaking (SB) and Chinese Restaurant Process (CRP)
- **Optional GPD tail**: Enhance tail probability estimation
- **Conditional modeling**: Add covariates via parameter linking
- **Causal workflows**: Propensity score adjustment with multiple PS strategies
- **Comprehensive prediction**: density, survival, quantile, sample, mean

---

## Model Structure

### Single-Outcome Model

**Bulk**: Dirichlet Process mixture
$$Y_i \sim \sum_{k=1}^{\infty} w_k f_k(y_i; \theta_k)$$

**Tail (optional)**: GPD beyond threshold
$$P(Y > u | Y > u) \sim \text{GPD}(\text{scale}, \text{shape})$$

### Causal Model

Two outcome arms with optional propensity score (PS) adjustment:
- **Control** ($T=0$): $Y_i^{(0)} \sim \text{DPM}_{\text{con}}(X_i, PS_i)$
- **Treated** ($T=1$): $Y_i^{(1)} \sim \text{DPM}_{\text{trt}}(X_i, PS_i)$

---

## MCMC Backends

### Stick-Breaking (SB)
- Fixed number of mixture components `components`
- Exact truncation at the chosen `components`
- Faster per-iteration computation
- Use when computational cost is primary concern

### Chinese Restaurant Process (CRP)
- Finite `components` cap for sampling
- Adaptively discovers number of occupied clusters within that cap
- More flexible cluster allocation
- Default for exploratory analysis

---

## Propensity Score (PS) Models

**Three PS strategies** (causal workflows only):

1. **`PS="logit"`** (default): Bayesian logistic regression
2. **`PS="probit"`**: Bayesian probit regression
3. **`PS="naive"`**: Gaussian naive Bayes classifier
4. **`PS=FALSE`**: No PS; outcomes depend on $X$ only

When PS is estimated, scores are injected into outcome model covariates for confounding adjustment.

---

## Workflow Overview

### Direct Model Building

```r
bundle <- DPmixGPD::build_nimble_bundle(
  y = y,
  kernel = "gamma",
  backend = "crp",
  GPD = TRUE,
  components = 5,
  mcmc = mcmc
)
fit <- run_mcmc_bundle_manual(bundle)
```

---

## Quick Start

```{r intro-quickstart, cache=TRUE}
# Load packaged dataset
data("nc_pos200_k3", package = "DPmixGPD")
y <- nc_pos200_k3$y

# Build unconditional model: bulk-only, SB backend
bundle <- DPmixGPD::build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "gamma",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)

# Print bundle
print(bundle)
```

---

## Outline of This Vignette Series

1. Introduction
2. Available Distributions
3. Basic Model Workflow
4. ex01 Unconditional DPM CRP
5. ex02 Unconditional DPM SB
6. ex03 Unconditional DPMGPD CRP
7. ex04 Unconditional DPMGPD SB
8. ex05 Conditional DPM CRP
9. ex06 Conditional DPM SB
10. ex07 Conditional DPMGPD CRP
11. ex08 Conditional DPMGPD SB
12. ex09 Causal no X CRP
13. ex10 Causal X no PS SB
14. ex11 Causal same backend CRP
15. ex12 Causal same backend SB
16. ex13 Causal different backends CRP
17. ex14 Causal different backends SB

---

## Dependencies

```{r intro-deps}
packageVersion("DPmixGPD")
packageVersion("nimble")
packageVersion("ggplot2")
```

Each vignette demonstrates complete workflows with reproducible examples.

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: [start-here](start-here.html)
- Next: [basic-model-compile-run](basic-model-compile-run.html)
- Workflow index: [Roadmap](../roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->




















