---
title: "Start Here"
execute:
  eval: false
---

```{r setup, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = NA,
  fig.width = 8,
  fig.height = 5,
  message = FALSE,
  warning = FALSE,
  cache = FALSE,
  cache.path = '../examples/legacy-cache-v3/'
)

# Reproducibility
set.seed(1)

# Fast-by-default rendering. Flip to FALSE for a fuller run.
FAST <- if (nzchar(Sys.getenv("LEGACY_FAST"))) as.logical(Sys.getenv("LEGACY_FAST")) else TRUE
mcmc_fast  <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1, timing = TRUE, workers = 2)
mcmc_heavy <- list(niter = 4000, nburnin = 1000, thin = 5, nchains = 2, seed = c(1, 2), timing = TRUE, parallel_chains = TRUE, workers = 2, z_update_every = 2)

# Use `mcmc = mcmc` in bundle builders unless a vignette overrides it.
mcmc <- if (isTRUE(FAST)) mcmc_fast else mcmc_heavy


# Ensure package is fully loaded for data access

# Precomputed results for fast website builds
source("_legacy-setup.R")
```


> **Website workflow note.**
> This page reflects the current exported API and recommended wrapper-first usage.
> Last updated: **2026-02-19**.
>
> For the full package narrative, see the main package vignettes (basic, unconditional, conditional, and causal).

[Workflow roadmap](roadmap.html){.btn .btn-outline-light}


## Quick start

This vignette gives a minimal, fully working workflow for an **unconditional** model and a **conditional** model. Everything uses short MCMC runs so the vignette renders quickly.

---

## Unconditional Model (CRP, bulk-only)

```{r start-uncond-data}
data("nc_pos200_k3")
y <- nc_pos200_k3$y
```

```{r start-uncond-bundle, cache = FALSE}
bundle_uncond <- bundle(
  y = y,
  backend = "crp",
  kernel = "gamma",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)
```

```{r start-uncond-fit, cache = FALSE}
fit_uncond <- load_or_fit("quickstart-start-here-fit_uncond", quiet_mcmc(dpmix(bundle_uncond, mcmc = list(show_progress = FALSE))))
summary(fit_uncond)
```

```{r start-uncond-predict, cache = FALSE}
pred_q <- predict(fit_uncond, type = "quantile", p = c(0.5, 0.9), interval = "credible")
head(pred_q$fit)
plot(pred_q)
```

---

## Conditional Model (SB, bulk-only)

```{r start-cond-data}
data("nc_posX100_p3_k2")
yc <- nc_posX100_p3_k2$y
X <- as.matrix(nc_posX100_p3_k2$X)
```

```{r start-cond-bundle, cache = FALSE}
bundle_cond <- bundle(
  y = yc,
  X = X,
  backend = "sb",
  kernel = "lognormal",
  GPD = FALSE,
  components = 5,
  mcmc = mcmc
)
```

```{r start-cond-fit, cache = FALSE}
fit_cond <- load_or_fit("quickstart-start-here-fit_cond", quiet_mcmc(dpmix(bundle_cond, mcmc = list(show_progress = FALSE))))
summary(fit_cond)
```

```{r start-cond-predict, cache = FALSE}
x_new <- X[1:20, , drop = FALSE]
pred_mean <- predict(fit_cond, x = x_new, type = "mean", interval = "credible", nsim_mean = 200)
head(pred_mean$fit)
plot(pred_mean)
```

---

## Useful S3 Methods

```{r start-s3}
params(fit_uncond)
plot(fit_uncond, family = c("traceplot", "running"))
```

---

## Next Steps

- `start/roadmap`: workflow map and routing
- `examples/ex01`: full three-phase workflow (spec -> bundle -> MCMC)
- `examples/ex01-ex08`: unconditional/conditional models with and without GPD
- `examples/ex09-ex14`: causal workflows

---

<!-- workflow-nav:start -->
## Workflow Navigation

- Previous: none (start of workflow series)
- Next: [basic-model-compile-run](basic-model-compile-run.html)
- Workflow index: [Roadmap](roadmap.html)
- Practical entry: [Examples](../examples/index.html)
<!-- workflow-nav:end -->

<!-- cmgpd-footer:start -->
## Prereqs
- Required packages and data for this page are listed in the setup chunks above.

## Outputs
- This page renders model fits, diagnostics, and summary artifacts generated by package APIs.

## Interpretation
- Canonical concept page: [Start Here](start-here.qmd)
- Treat this page as an application/example view and use the canonical page for core definitions.

## Next
- Continue to the linked canonical concept page, then return for implementation-specific details.
<!-- cmgpd-footer:end -->





