---
title: "DPmixGPD: Causal treatment effects (ATE/QTE) via two-arm outcome models"
bibliography: dpmixgpd.bib
link-citations: true
format:
  html:
    toc: true
execute:
  cache: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r}
library(DPmixGPD)
set.seed(1)
```

DPmixGPD includes a causal workflow for estimating treatment effects on the *entire* outcome distribution. Rather than restricting attention to a mean difference, the causal module targets:

- mean-scale effects (ATE),
- quantile-scale effects (QTE), and
- truncated/restricted mean effects (available through `ate(..., type="rmean")`).

The causal workflow reuses the same outcome modeling machinery as the unconditional and conditional vignettes: a Dirichlet process mixture (bulk) optionally spliced with a generalized Pareto tail (GPD). The mathematical model for the DPM prior, the GPD density/CDF, and the bulk–tail splice is developed once in the vignette **“DPmixGPD: Model specification and posterior computation”**. This vignette focuses on causal identification, how DPmixGPD builds a two-arm model, and how to compute ATE/QTE from posterior draws.

# Causal setup and identification

## Potential outcomes and estimands

Let \(T\in\{0,1\}\) denote a binary treatment indicator and \(Y(0), Y(1)\) the potential outcomes. The observed outcome is
\[
Y = T\,Y(1) + (1-T)\,Y(0).
\]
For a covariate vector \(\boldsymbol{x}\), define the conditional CDFs and quantiles
\[
F_a(y\mid \boldsymbol{x}) = \Pr\{Y(a)\le y \mid \boldsymbol{X}=\boldsymbol{x}\}, \qquad
Q_a(\tau\mid \boldsymbol{x}) = F_a^{-1}(\tau\mid \boldsymbol{x}), \quad \tau\in(0,1),
\]
and the conditional quantile treatment effect
\[
\mathrm{QTE}(\tau\mid \boldsymbol{x})
=
Q_1(\tau\mid \boldsymbol{x}) - Q_0(\tau\mid \boldsymbol{x}).
\]
The conditional mean effect (ATE at \(\boldsymbol{x}\)) is
\[
\mathrm{ATE}(\boldsymbol{x})
=
\mathbb{E}\{Y(1)\mid \boldsymbol{x}\} - \mathbb{E}\{Y(0)\mid \boldsymbol{x}\}.
\]

DPmixGPD’s `qte()` and `ate()` functions compute these treated-minus-control functionals from a fitted causal object. If you supply `newdata`, you obtain effects evaluated at those covariate profiles; otherwise DPmixGPD defaults to the training design \(X\) when present.

## Assumptions

DPmixGPD follows the standard “selection on observables” identification strategy (Rubin causal model) [@rubin1974]. The identifying assumptions are:

1. **Consistency / SUTVA**: the observed outcome equals the potential outcome under the received treatment; no interference between units.
2. **Unconfoundedness**:
\[
\{Y(0),Y(1)\} \perp T \mid \boldsymbol{X}.
\]
3. **Overlap** (positivity):
\[
0 < e(\boldsymbol{x}) < 1 \quad \text{for covariate values of interest,}
\]
where \(e(\boldsymbol{x})=\Pr(T=1\mid \boldsymbol{X}=\boldsymbol{x})\) is the propensity score.

Under these assumptions,
\[
F_a(y\mid \boldsymbol{x}) = \Pr(Y\le y \mid T=a,\boldsymbol{X}=\boldsymbol{x}),
\]
so causal effects can be estimated by modeling the *observed* conditional outcome distribution within each treatment arm.

# DPmixGPD causal strategy: two-arm outcome models

DPmixGPD builds:

- a control-arm outcome model for \(Y\mid T=0\),
- a treated-arm outcome model for \(Y\mid T=1\),
- and (optionally) a propensity score (PS) model for \(T\mid \boldsymbol{X}\).

Each outcome arm is a full DPmixGPD model (bulk DPM with optional GPD tail). You can choose different kernels/backends/tail settings per arm by passing length-2 vectors for `kernel`, `backend`, `GPD`, and `components`.

## Why include a propensity score module?

Rosenbaum and Rubin showed the propensity score is a balancing score: adjusting for \(e(\boldsymbol{x})\) is sufficient to remove bias due to observed covariates [@rosenbaum1983]. DPmixGPD offers a lightweight PS model and can pass a PS-derived covariate into both outcome arms as an additional predictor. This is optional:

- set `PS = FALSE` to skip propensity score estimation and use only \(X\) in the outcome models;
- set `PS = "logit"` / `"probit"` / `"naive"` to estimate PS and include it in the outcome arms.

Important details:

- `ps_scale = "prob"` uses the posterior summary \( \hat e(\boldsymbol{x}) \) directly.
- `ps_scale = "logit"` uses \(\mathrm{logit}(\hat e(\boldsymbol{x}))\) (often numerically friendlier as a covariate).
- `ps_summary = "mean"` or `"median"` chooses how the posterior for \(e(\boldsymbol{x})\) is summarized before being passed into the outcome arms.
- `ps_clamp` clamps \(\hat e(\boldsymbol{x})\) to \((\varepsilon, 1-\varepsilon)\) to avoid infinities under the logit transform.

# Building and fitting a causal model in code

We illustrate a typical workflow on a small synthetic example. For real analyses, increase MCMC iterations and check diagnostics as in any Bayesian mixture model.

## Simulate data with confounding (for demonstration)

```{r}
set.seed(2)
N <- 250
X <- cbind(x1 = rnorm(N), x2 = rbinom(N, 1, 0.5))

# Treatment assignment via a propensity score
ps_true <- plogis(-0.2 + 0.9 * X[, 1] - 0.6 * X[, 2])
T <- rbinom(N, 1, ps_true)

# Heavy-tailed positive outcome: treatment shifts the bulk and changes tail risk
base <- exp(0.2 + 0.4 * X[, 1] + 0.2 * X[, 2] + rt(N, df = 3) / 3)
y <- base * exp(0.25 * T)
```

## Build a causal bundle

The causal bundle constructor is `build_causal_bundle()`. It creates:

- a propensity score bundle (if `PS` is not `FALSE` and \(X\) is present),
- a control outcome bundle,
- a treated outcome bundle.

```{r, eval=TRUE}
cb <- build_causal_bundle(
  y = y,
  X = X,
  T = T,
  backend = "sb",
  kernel = "lognormal",
  GPD = TRUE,
  components = 6,
  PS = "logit",            # "logit", "probit", "naive", or FALSE
  ps_scale = "logit",      # "logit" or "prob"
  ps_summary = "mean",     # "mean" or "median"
  mcmc_outcome = list(niter = 500, nburnin = 150, thin = 2, nchains = 1, seed = 1),
  mcmc_ps      = list(niter = 400, nburnin = 120, thin = 2, nchains = 1, seed = 2)
)

cb
```

### Arm-specific configuration (optional but supported)

You can use different kernels/backends/tail settings across arms:

```{r, eval=TRUE}
cb2 <- build_causal_bundle(
  y = y, X = X, T = T,
  kernel = c("lognormal", "gamma"),      # treated, control
  backend = c("sb", "sb"),
  GPD = c(TRUE, TRUE),
  components = c(6, 6),
  PS = "probit"
)
```

### Arm-specific priors and link specifications via `param_specs`

`param_specs` works the same way as in `build_nimble_bundle()`, but you may pass either:

- one list (used for both arms), or
- a list with `con` and `trt` entries to override each arm separately.

For example, to let the bulk location depend on \(X\) in both arms (conditional outcome distributions), you can use link mode (see the conditional vignette for full details):

```{r, eval=TRUE}
param_specs <- list(
  bulk = list(
    meanlog = list(
      mode = "link",
      link = "identity",
      beta_prior = list(dist = "normal", args = list(mean = 0, sd = 2))
    ),
    sdlog = list(
      mode = "dist",
      dist = "invgamma",
      args = list(shape = 2, scale = 1)
    )
  ),
  gpd = list(
    tail_scale = list(
      mode = "dist",
      dist = "gamma",
      args = list(shape = 2, rate = 1)
    ),
    tail_shape = list(
      mode = "dist",
      dist = "normal",
      args = list(mean = 0, sd = 0.25)
    )
  )
)

cb3 <- build_causal_bundle(
  y = y, X = X, T = T,
  backend = "sb",
  kernel = "lognormal",
  GPD = TRUE,
  components = 6,
  PS = "logit",
  param_specs = param_specs
)
```

# Run MCMC

```{r, eval=TRUE}
fit <- run_mcmc_causal(cb, show_progress = FALSE)
fit
```

The result is a `dpmixgpd_causal_fit` containing:

- `ps_fit` (if enabled),
- `outcome_fit$con` and `outcome_fit$trt` (two `mixgpd_fit` objects),
- and metadata needed for downstream ATE/QTE computations.

# Estimating QTE and ATE

## Quantile treatment effects (QTE)

DPmixGPD computes QTE by:

1. predicting \(Q_1(\tau\mid \boldsymbol{x})\) from the treated outcome model,
2. predicting \(Q_0(\tau\mid \boldsymbol{x})\) from the control outcome model,
3. subtracting draw-by-draw across posterior samples to obtain a posterior for the difference.

This is aligned with classical quantile treatment effect definitions [@firpo2007] and the broader quantile regression literature [@koenker1978].

```{r, eval=TRUE}
q <- qte(
  fit,
  probs = c(0.25, 0.5, 0.75, 0.9, 0.95),
  interval = "credible",
  level = 0.95
)

q$fit           # posterior mean QTE at each prob
q$lower; q$upper
```

### Conditional QTE at selected covariate profiles

Pass a covariate matrix in `newdata`. The output is then per-row of `newdata`.

```{r, eval=TRUE}
Xnew <- rbind(
  c(0, 0),
  c(1, 0),
  c(0, 1)
)

q_new <- qte(
  fit,
  probs = c(0.5, 0.9, 0.95),
  newdata = Xnew,
  interval = "credible",
  level = 0.90
)
q_new$fit
```

## Average treatment effects (ATE)

`ate()` computes treated-minus-control differences in posterior means. Because posterior means for mixtures with tails are usually computed via posterior predictive simulation, `ate()` uses `nsim_mean` posterior predictive draws per posterior iteration.

```{r, eval=TRUE}
a <- ate(
  fit,
  interval = "credible",
  level = 0.95,
  nsim_mean = 80
)
a$fit
a$lower; a$upper
```

### Restricted mean difference

DPmixGPD also supports a truncated/restricted mean target through `ate(..., type="rmean", cutoff = ...)`, which computes the mean of \(\min(Y, c)\) under each arm and then differences them. This can be a robust alternative when extreme tails dominate the mean.

```{r, eval=TRUE}
a_r <- ate(
  fit,
  type = "rmean",
  cutoff = as.numeric(stats::quantile(y, 0.90)),
  interval = "credible",
  level = 0.95,
  nsim_mean = 80
)
a_r$fit
```

# Diagnostics and practical guidance

1. **Check overlap**. If treated and control units have little overlap in propensity score, causal estimands at covariate values outside overlap become extrapolation problems regardless of outcome model flexibility. Rosenbaum–Rubin diagnostics (PS histograms by group) remain useful here [@rosenbaum1983].

2. **Tail identifiability matters**. If you enable `GPD=TRUE` but only a small number of exceedances are effectively informed by the splice, the tail parameters can be weakly identified and extreme QTEs (e.g., \(\tau=0.99\)) may have wide intervals.

3. **Start simpler**. If this is your first causal DPmixGPD run, consider:
   - bulk-only (`GPD=FALSE`),
   - linking only one bulk location parameter to \(X\),
   - then incrementally adding tail modeling and more link-mode parameters.

4. **Backend choice**. For causal models that use link-mode parameters (conditional outcome distributions), `backend="sb"` is the recommended starting point.

# References



