---
title: "Conditional DP Mixture Models"
bibliography: dpmixgpd.bib
format:
  html:
    toc: true
execute:
  cache: true
---


DPmixGPD supports conditional (covariate-dependent) density estimation by allowing one or more *kernel parameters*—and, optionally, selected *tail parameters*—to vary with a design matrix \(X\). The goal is to estimate an entire conditional distribution \(F(y \mid \boldsymbol{x})\), not just a mean curve.

This vignette focuses on how covariates enter the model, how to control that behavior via `param_specs`, and how to generate conditional predictions. For the bulk–tail spliced likelihood, GPD density/CDF definitions, and the overall “bulk + GPD tail” construction, see the *Model fundamentals* vignette in this package (we will refer to it rather than repeating the full splice derivation here).

### Conditional mixture model

Let \(y_i \in \mathbb{R}\) be an outcome and \(\boldsymbol{x}_i \in \mathbb{R}^p\) a covariate vector. Under a truncated stick-breaking (SB) Dirichlet process mixture, DPmixGPD uses
\[
  y_i \mid z_i=j,\boldsymbol{x}_i \sim k\!\left(y_i \mid \boldsymbol{\theta}_j(\boldsymbol{x}_i)\right), \qquad
  z_i \sim \mathrm{Categorical}(w_1,\dots,w_J),
\]
with stick-breaking weights
\[
  v_j \sim \mathrm{Beta}(1,\alpha),\quad
  w_1=v_1,\quad
  w_j=v_j\prod_{\ell<j}(1-v_\ell)\ (j\ge 2),\quad
  \alpha \sim \text{(prior)}.
\]
This is the same DPM idea as in classic DP mixtures, but applied to conditional distributions; see, e.g., [@escobar1995; @muller1996]. The key difference from the unconditional vignette is that \(\boldsymbol{\theta}_j(\boldsymbol{x})\) may depend on \(\boldsymbol{x}\).

DPmixGPD implements covariate dependence through *per-parameter modes*:

- `mode = "dist"`: parameter is random (drawn from a prior) but does **not** depend on \(X\).
- `mode = "fixed"`: parameter is fixed at a constant value.
- `mode = "link"`: parameter depends on \(X\) through a GLM-style link, with component-specific regression coefficients.

This is set separately for each bulk parameter (kernel parameter), and for selected tail parameters when `GPD = TRUE`.

### A practical note about the design matrix \(X\)

DPmixGPD treats the matrix you pass as `X` as the design matrix used in linear predictors. It does **not** automatically add an intercept column. In practice you will almost always want `model.matrix()` so an intercept is included and factor coding is handled correctly.

A recommended workflow is:

1. Build \(X\) using `model.matrix()` (includes intercept by default).
2. Standardize/center continuous covariates (especially if you do not include an intercept).
3. Keep the same column names and ordering for prediction. `predict()` will validate them.

```{r}
library(DPmixGPD)

# Example covariates
set.seed(1)
N <- 200
dat <- data.frame(
  x1 = rnorm(N),
  x2 = rbinom(N, 1, 0.5)
)

# Recommended: model.matrix() gives an intercept and consistent column names
X <- model.matrix(~ x1 + x2, dat)

# (Optional) scale continuous columns but keep intercept untouched
X[, "x1"] <- scale(X[, "x1"])
```

### Which parameters can depend on \(X\)?

Kernel-specific parameter names are stored in the kernel registry. You can inspect them:

```{r}
init_kernel_registry()
reg <- get_kernel_registry()

# Names of available kernels
names(reg)

# Bulk parameters for a specific kernel
reg$normal$bulk_params
reg$gamma$bulk_params

# Defaults for covariate behavior (when X is present)
reg$normal$defaults_X
reg$gamma$defaults_X
```

The registry is the authoritative source for (i) valid kernel names, (ii) bulk parameter names, and (iii) each kernel’s default covariate behavior.

### Link-mode parameters: general form

If a bulk parameter \(\vartheta\) is assigned `mode = "link"`, DPmixGPD uses a linear predictor and link function:
\[
  \eta_{\vartheta,ij} = \boldsymbol{x}_i^\top \boldsymbol{\beta}_{\vartheta,j},\qquad
  \vartheta_{ij} = g_\vartheta(\eta_{\vartheta,ij}).
\]
Here \(\boldsymbol{\beta}_{\vartheta,j}\) is component-specific (indexed by mixture component \(j\)), so different mixture components can represent different conditional regimes.

Supported link functions include:

- `"identity"`: \(\vartheta = \eta\)
- `"exp"`: \(\vartheta = \exp(\eta)\) (useful for positive parameters)
- `"softplus"`: \(\vartheta = \log(1+\exp(\eta))\) (also enforces positivity)
- `"power"`: \(\vartheta = \eta^{\kappa}\) (requires `link_power = kappa`; use with care if \(\eta\) can be negative)

The best default for strictly-positive parameters is typically `"exp"` or `"softplus"`.

### Tail parameters under GPD

When `GPD = TRUE`, DPmixGPD uses a GPD tail above a threshold (see the *Model fundamentals* vignette for the splice and GPD density). For conditional variants, the following tail parameters can depend on \(X\):

- `gpd$threshold`: `fixed`, `dist`, or `link`
- `gpd$tail_scale`: `fixed`, `dist`, or `link`
- `gpd$tail_shape`: `fixed` or `dist` (no `link` mode)

Two distinct “link” behaviors exist for the threshold:

- Deterministic link: \(u_i = g(\boldsymbol{x}_i^\top \beta_u)\)
- Stochastic link via `link_dist = list(dist="lognormal", ...)`: \(u_i \sim \mathrm{LogNormal}(\mu_i,\sigma_u)\) with \(\mu_i=\boldsymbol{x}_i^\top\beta_u\)

This mirrors common nonstationary threshold modeling ideas in extremes [@davison1990; @chavezdemoulin2005], but placed inside the full Bayesian mixture framework [@behrens2004; @donascimento2012].

### Controlling covariate dependence with `param_specs`

`build_nimble_bundle()` accepts `param_specs` to override defaults. The structure is:

```r
param_specs <- list(
  bulk = list(
    <bulk_param_1> = list(mode = "dist"/"fixed"/"link", ...),
    <bulk_param_2> = list(mode = "dist"/"fixed"/"link", ...),
    ...
  ),
  gpd = list(
    threshold  = list(mode = "dist"/"fixed"/"link", ...),
    tail_scale = list(mode = "dist"/"fixed"/"link", ...),
    tail_shape = list(mode = "dist"/"fixed", ...),
    # optionally: sdlog_u if using lognormal threshold link_dist
    sdlog_u    = list(mode = "dist", dist = "invgamma", args = list(shape = 2, scale = 1))
  )
)
```

In `mode = "link"` you may specify:

- `link`: one of `"identity"`, `"exp"`, `"softplus"`, `"power"`
- `link_power`: numeric, only for `"power"`
- `beta_prior`: a list describing priors for regression coefficients, e.g.
  `list(dist="normal", args=list(mean=0, sd=2))`

In `mode = "dist"` you specify:

- `dist`: e.g., `"normal"`, `"gamma"`, `"invgamma"`, etc. (used internally in NIMBLE code generation)
- `args`: named list of distribution parameters

### Backend note: SB vs CRP with covariates

DPmixGPD supports two backends:

- `backend = "sb"`: truncated stick-breaking mixture (default for covariate-dependent parameters)
- `backend = "crp"`: finite CRP representation

For conditional models, the SB backend is recommended. Internally, the CRP backend avoids default link-mode bulk parameters because deterministic, cluster-indexed covariate nodes can interfere with NIMBLE’s CRP samplers. If you want covariate links on bulk parameters, you should generally use `backend = "sb"`.

### Worked example: conditional normal kernel with a GPD tail

We simulate data where the conditional mean depends on \(X\), fit a conditional model, then generate conditional predictions.

```{r, results='hide'}
# Simulate a simple signal + heavy-tailed noise
set.seed(2)
beta <- c(-0.3, 0.8, -0.6)   # coefficients for (Intercept, x1, x2)
mu <- drop(X %*% beta)
y <- mu + rt(N, df = 3)     # heavy-tailed noise

mcmc <- list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)

bundle <- build_nimble_bundle(
  y = y,
  X = X,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE, quiet = TRUE)
```

A quick summary of the fitted object:

```{r}
print(fit)
```

### Conditional prediction

For conditional models, `predict()` uses `x` for the prediction design matrix \(X_{\text{pred}}\). For `type = "density"` or `"survival"`, you must also provide `y` as the grid of outcome values where the function is evaluated.

```{r}
# Prediction points: two covariate profiles
newdat <- data.frame(x1 = c(-1, 1), x2 = c(0, 1))
Xpred <- model.matrix(~ x1 + x2, newdat)
Xpred[, "x1"] <- scale(Xpred[, "x1"])  # apply same scaling convention

ygrid <- seq(quantile(y, 0.01), quantile(y, 0.99), length.out = 120)

dens_pred <- predict(fit, x = Xpred, y = ygrid, type = "density", cred.level = 0.95)
q_pred <- predict(fit, x = Xpred, type = "quantile", index = c(0.5, 0.9, 0.95), cred.level = 0.95)
```

The returned objects contain point estimates and uncertainty summaries. For example:

```{r}
# Quantile predictions for each covariate profile
q_pred$fit
```

You can also plot prediction objects directly:

```{r, eval=TRUE}
plot(dens_pred)
plot(q_pred)
```

(Plots are enabled in this vignette build; for faster local iteration you can temporarily set this chunk to `eval=FALSE`.)

### What to tune in practice

1. Decide which parameters should depend on \(X\). A good start is to link a location parameter (e.g., `mean` or `meanlog`) and keep scale/shape parameters in `dist` mode, then add complexity if diagnostics warrant it.

2. Use links consistent with support: `"exp"`/`"softplus"` for strictly positive parameters.

3. Prefer `backend = "sb"` for covariate links. Use `backend = "crp"` mainly for unconditional models or when you intentionally keep bulk parameters in `dist` mode.

4. Increase `components`, `niter`, and `nburnin` for real analyses. The small values in this vignette are chosen only to keep examples quick.

### References




