---
title: "MCMC Workflow"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{MCMC Workflow}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
library(ggplot2)
library(kableExtra)
library(DPmixGPD)
set.seed(1)
FAST <- TRUE
mcmc <- if (FAST) {
  list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
} else {
  list(niter = 6000, nburnin = 2000, thin = 5, nchains = 2, seed = c(1, 2))
}
```

## Overview

This vignette demonstrates the MCMC workflow: build, run, inspect, and extract.

## Theory (brief)

The posterior is explored with Markov chain Monte Carlo. Diagnostics summarize
mixing and convergence behavior, such as traceplots, effective sample size, and
Geweke statistics comparing early vs late chain segments. These checks help
ensure that posterior summaries are reliable.

## Model Building and Sampling

```{r, results='hide'}
library(DPmixGPD)

data("faithful", package = "datasets")
y <- faithful$eruptions
bundle <- build_nimble_bundle(
  y = y,
  backend = "sb",
  kernel = "normal",
  GPD = TRUE,
  components = 6,
  mcmc = mcmc
)

fit <- run_mcmc_bundle_manual(bundle, show_progress = FALSE)
```

## Model Summary

```{r}
print(fit)
summary(fit)
```

## Diagnostic Plots

```{r}
if (requireNamespace("ggmcmc", quietly = TRUE) && requireNamespace("coda", quietly = TRUE)) {
  smp <- fit$mcmc$samples
  params <- if (!is.null(smp)) {
    cn <- colnames(as.matrix(smp))
    cn[1:min(6, length(cn))]
  } else {
    NULL
  }
  plot(fit, family = "geweke", params = params)
} else {
  message("Plotting requires 'ggmcmc' and 'coda' packages.")
}
```

## Posterior Sample Extraction

```{r}
if (!is.null(fit$mcmc$samples)) {
  s <- fit$mcmc$samples
  if (requireNamespace("coda", quietly = TRUE)) {
    mat <- as.matrix(s)
    dim(mat)
    colnames(mat)[1:min(20, ncol(mat))]
  } else {
    message("Sample extraction requires 'coda' package.")
  }
}
```

## Re-running with Different Settings

```{r, eval=FALSE}
# Rebuild with modified MCMC settings
# bundle2 <- update_mcmc(bundle, niter = 8000, nburnin = 2000)
# fit2 <- run_mcmc_bundle_manual(bundle2)
```

```{r sessioninfo, include=FALSE}
sessionInfo()
```






