---
title: "DPmixGPD: Unconditional bulk–tail density estimation"
bibliography: dpmixgpd.bib
link-citations: true
format:
  html:
    toc: true
execute:
  cache: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center"
)
```

```{r}
library(DPmixGPD)
set.seed(1)
```

This vignette shows how to fit the *unconditional* DPmixGPD model, i.e. without covariates (\(X=\mathrm{NULL}\)). The unconditional case is the simplest entry point for two reasons:

1. It isolates the central design choice: which kernel family is an appropriate *bulk* model for your outcome support and shape.
2. It makes clear how the bulk (DPM) and tail (GPD) are estimated *jointly* rather than via a two-stage procedure.

All mathematical definitions of the Dirichlet process mixture prior, the GPD tail density/CDF, and the spliced bulk–tail density are developed in the vignette **“DPmixGPD: Model specification and posterior computation”**. This vignette treats those definitions as *fixed* and focuses on: model configuration, prior control, and posterior summaries in the unconditional setting.

# What “unconditional” means in DPmixGPD

With no covariates, the bulk model is an ordinary Dirichlet process mixture:
\[
f_{\mathrm{DPM}}(y) \;=\; \sum_{j=1}^{\infty} w_j \, k(y\mid \Theta_j),
\]
with DP stick-breaking weights \(w_j\) [@sethuraman1994] and blocked Gibbs truncation at \(J=\texttt{components}\) [@ishwaran2001]. In practice, DPmixGPD uses a finite representation for both SB and CRP backends, controlled by the single argument `components`.

If `GPD=TRUE`, DPmixGPD fits the (upper) tail using a generalized Pareto exceedance model and couples it to the bulk via a proper bulk–tail splice (see the “Model specification…” vignette for the full density/CDF). This joint construction is in the same spirit as Bayesian semiparametric bulk–tail models that avoid fixing a hard threshold and instead estimate bulk and tail together [@frigessi2002; @doNascimento2012; @behrens2004].

# Choosing a kernel (bulk family)

A good kernel choice is driven by *support* first and *shape* second:

- If your outcome is supported on \(\mathbb{R}\): Normal / Laplace / Cauchy kernels are natural.
- If your outcome is supported on \(\mathbb{R}_+\): Gamma / Lognormal / Inverse-Gaussian / Amoroso kernels are typically appropriate.
- If your data are heavy-tailed already in the bulk, a more flexible positive-support kernel (e.g., Lognormal, Amoroso) may absorb structure that a lighter kernel would push into the tail, which can affect tail stability.

DPmixGPD ships a registry describing which kernels support which features.

```{r}
kernel_support_table()
```

# SB vs CRP backends

DPmixGPD supports `backend="sb"` (stick-breaking) and `backend="crp"` (Chinese restaurant process). Both correspond to the same DP prior but expose different computational representations [@neal2000].

Practical guidance in the unconditional case:

- Start with `backend="sb"` if you want a stable finite representation with an explicit truncation level \(J\).
- Use `backend="crp"` if you prefer a partition-driven representation (often convenient conceptually for clustering), noting that the finite NIMBLE model still uses `components` as a maximum representable number of clusters.

# Enabling/disabling the GPD tail

The decision `GPD=TRUE/FALSE` is not cosmetic; it changes the likelihood and the posterior targets.

- `GPD=FALSE` fits a pure DPM density on the full support. This is appropriate if you mainly want flexible density estimation without explicit extreme extrapolation.
- `GPD=TRUE` fits the joint bulk–tail model where exceedances beyond a threshold are modeled by a GPD and the total density remains proper by splicing. This is appropriate if you want stable *extreme quantiles* or *tail probabilities* beyond the observed bulk.

Even in the unconditional case, the tail inference is coupled to the bulk via the survival mass at the threshold \(S_{\mathrm{DPM}}(u)=1-F_{\mathrm{DPM}}(u)\). This is exactly why the model tends to be more stable than “fit the bulk, then fit a tail” pipelines when the exceedance sample size is small.

# Model building in code

## Minimal build: bulk only

```{r, eval=TRUE}
y <- abs(rnorm(120)) + 0.1

bundle <- build_nimble_bundle(
  y = y,
  X = NULL,
  backend = "sb",
  kernel = "lognormal",
  GPD = FALSE,
  components = 6,
  mcmc = list(niter = 400, nburnin = 100, thin = 2, nchains = 1, seed = 1)
)

bundle
summary(bundle)
```

## Minimal build: bulk + GPD tail

```{r, eval=TRUE}
bundle_gpd <- build_nimble_bundle(
  y = y,
  X = NULL,
  backend = "sb",
  kernel = "lognormal",
  GPD = TRUE,
  components = 6,
  mcmc = list(niter = 500, nburnin = 150, thin = 2, nchains = 1, seed = 1)
)
```

## Run posterior sampling

```{r, eval=TRUE}
fit <- run_mcmc_bundle_manual(bundle_gpd, show_progress = FALSE)
fit
summary(fit)
```

The returned object is a `mixgpd_fit`, which stores posterior draws, the compiled spec, and the settings needed for downstream summaries and plotting.

# Prior control via `param_specs`

DPmixGPD uses a single structured object `param_specs` to control whether each parameter is:

- `mode="fixed"`: treated as constant,
- `mode="dist"`: assigned a prior and sampled,
- `mode="link"`: made covariate-dependent via a link function (only relevant when `X` is provided; see the conditional vignette).

The expected structure is:

```{r, eval=TRUE}
param_specs <- list(
  bulk = list(
    # kernel-specific parameters go here
  ),
  gpd = list(
    threshold  = list(mode = "fixed", value = 1.0),
    tail_scale = list(mode = "dist", dist = "gamma", args = list(shape = 2, rate = 1)),
    tail_shape = list(mode = "dist", dist = "normal", args = list(mean = 0, sd = 0.25))
  ),
  concentration = list(mode = "dist", dist = "gamma", args = list(shape = 2, rate = 1))
)
```

## Example: treat tail parameters as random, but fix the threshold

Fixing \(u\) can be useful for sensitivity analysis. A common pragmatic choice is to set \(u\) to a high empirical quantile and let \((\sigma,\xi)\) be learned from exceedances. (When \(u\) is random, DPmixGPD will propagate threshold uncertainty automatically.)

```{r, eval=TRUE}
u0 <- as.numeric(stats::quantile(y, 0.90))

param_specs <- list(
  gpd = list(
    threshold  = list(mode = "fixed", value = u0),
    tail_scale = list(mode = "dist",  dist = "gamma",  args = list(shape = 2, rate = 1)),
    tail_shape = list(mode = "dist",  dist = "normal", args = list(mean = 0, sd = 0.25))
  ),
  concentration = list(
    mode = "dist", dist = "gamma", args = list(shape = 2, rate = 1)
  )
)

bundle_gpd <- build_nimble_bundle(
  y = y,
  X = NULL,
  backend = "sb",
  kernel = "lognormal",
  GPD = TRUE,
  components = 6,
  param_specs = param_specs
)
```

## Example: fully stochastic threshold, with a weakly informative prior

DPmixGPD supports stochastic thresholds under `gpd$threshold` when `GPD=TRUE`. In the conditional model the threshold can be linked to covariates; in the unconditional model it is scalar.

```{r, eval=TRUE}
param_specs2 <- list(
  gpd = list(
    threshold  = list(mode = "dist", dist = "lognormal", args = list(meanlog = log(median(y)), sdlog = 0.4)),
    tail_scale = list(mode = "dist", dist = "gamma",     args = list(shape = 2, rate = 1)),
    tail_shape = list(mode = "dist", dist = "normal",    args = list(mean = 0, sd = 0.25))
  )
)
```

# Posterior summaries: density, CDF, and quantiles

The core inferential objects are functionals of the full spliced distribution \(F(y)\) (defined in the “Model specification…” vignette). Given posterior draws \(\{\Theta^{(s)},\Phi^{(s)}\}_{s=1}^S\), DPmixGPD computes summaries by:

1. evaluating the functional under each draw, and
2. aggregating across draws (posterior mean, median, and credible intervals).

## Quantiles under the splice (conceptual formula)

Let \(u^{(s)}\) be the threshold in draw \(s\), and let \(p_u^{(s)} = F_{\mathrm{DPM}}^{(s)}(u^{(s)})\). For a target level \(\tau\):

- If \(\tau \le p_u^{(s)}\), then the quantile is in the bulk: \(Q^{(s)}(\tau)=Q_{\mathrm{DPM}}^{(s)}(\tau)\).
- If \(\tau > p_u^{(s)}\), then it is in the tail. Writing \(q = (\tau - p_u^{(s)})/(1-p_u^{(s)})\), the tail quantile is
\[
Q^{(s)}(\tau) = u^{(s)} + Q_{\mathrm{GPD}}^{(s)}(q),
\]
where \(Q_{\mathrm{GPD}}^{(s)}(\cdot)\) is the GPD quantile function for exceedances.

DPmixGPD implements these calculations using the posterior samples, so you do not need to code them manually unless you are building custom functionals.

## Practical API

```{r, eval=TRUE}
# Posterior predictive density/survival at a grid:
grid <- seq(min(y), max(y) * 1.5, length.out = 200)

dhat <- predict(fit, y = grid, type = "density")
shat <- predict(fit, y = grid, type = "survival")

# Posterior quantiles:
qs <- predict(fit, type = "quantile", p = c(0.90, 0.95, 0.99))

# Posterior predictive mean:
mhat <- predict(fit, type = "mean", cred.level = 0.90)
```

The exact `type` names and available options are documented in the function reference for `predict.mixgpd_fit`. Use `?predict.mixgpd_fit` after installation.

# Diagnostics (what to check)

Posterior computation for mixture models is sensitive to:

- truncation level `components` (SB) or maximum clusters (CRP),
- mixing of allocations \(\{z_i\}\),
- tail identifiability when the exceedance sample is small.

In practice, you should:

1. run multiple chains if feasible,
2. inspect trace plots for key tail parameters \((u,\sigma,\xi)\) when `GPD=TRUE`,
3. check stability of extreme quantiles (e.g., 0.95/0.99) across chains and across small perturbations of priors.

DPmixGPD prints a compact summary and supports standard MCMC diagnostics tooling.

```{r, eval=TRUE}
summary(fit)
plot(fit)
```

# Low-level density utilities (optional)

DPmixGPD exports both CamelCase and lower-case density/CDF/quantile/random-generation helpers for kernels and their mix/splice variants. As a practical rule:

- the CamelCase forms (e.g., `dNormMixGpd`) are intended for NIMBLE model code and internal use,
- the lower-case forms (e.g., `dnormmixgpd`) are convenient for direct use in R scripts.

This separation makes it easier to validate pieces of the model outside MCMC (e.g., checking a GPD tail density shape) without rebuilding a full bundle.

# References



