---
title: "Customization: Build NIMBLE Models from Scratch"
output:
  rmarkdown::html_vignette:
    highlight: null
vignette: >
  %\VignetteIndexEntry{Customization: Build NIMBLE Models from Scratch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = NA,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  fig.height = 4.5,
  fig.align = "center",
  cache = TRUE,
  cache.path = "legacy-cache/",
  cache.lazy = FALSE
)

source("_load_pkg.R")
library(nimble)
library(ggplot2)
library(kableExtra)

set.seed(1)

use_fast_mcmc <- TRUE
mcmc_long <- list(niter = 2000, nburnin = 500, thin = 3, nchains = 3, seed = 1:3)
mcmc_fast <- list(niter = 600, nburnin = 150, thin = 2, nchains = 1, seed = 1)
mcmc <- if (use_fast_mcmc) mcmc_fast else mcmc_long

# Avoid compiling NIMBLE models during pkgdown builds.
is_pkgdown <- any(grepl("pkgdown|callr", Sys.getenv(), ignore.case = TRUE)) ||
  exists(".__PKGDOWN__", envir = .GlobalEnv, inherits = FALSE) ||
  any(grepl("pkgdown", commandArgs(), ignore.case = TRUE))
if (is_pkgdown) {
  knitr::opts_chunk$set(eval = FALSE)
}
```

# Overview

This vignette shows how to **build NIMBLE models from scratch** using the exported
`d* / p* / q* / r*` nimbleFunctions in **DPmixGPD**, then walks through the full
pipeline **(data -> constants -> inits -> code -> compile -> run -> extract)**.

We cover three workflows:

1) **SB case**: manual stick-breaking Normal mixture using `dNormMix`.
2) **CRP + GPD case**: build a CRP model with a GPD tail using DPmixGPD codegen,
   then run the manual NIMBLE steps explicitly.
3) **Causal model**: build a causal model and use S3 methods for inference.

> We use fast MCMC settings to keep this vignette lightweight. Increase the
> iterations if you want more stable posterior summaries.

# Data

```{r}
data("mtcars", package = "datasets")

df <- mtcars
y <- df$mpg
X <- df[, c("wt", "hp")]
X <- as.data.frame(X)
```

# SB Case: Manual Stick-Breaking Model with `dNormMix`

This section shows a **from-scratch** SB model that uses the exported
`dNormMix` nimbleFunction directly as a likelihood. We register it as a
custom NIMBLE distribution and then build the model step by step.

## Note on custom likelihoods

To keep the model fully NIMBLE-compatible without registering a custom
distribution, we use a **zeros-trick** likelihood based on the exported
`dNormMix` nimbleFunction.

## Constants, data, and inits

```{r}
N <- length(y)
K <- 3

constants_sb <- list(N = N, K = K, C = 1000)
data_sb <- list(y = y, zeros = rep(0, N))

inits_sb <- function() {
  list(
    alpha = 1,
    v = c(rep(0.5, K - 1), 1),
    mu = rnorm(K, mean(y), 1),
    sigma = rep(sd(y), K)
  )
}
```

## NIMBLE code

```{r}
code_sb <- nimble::nimbleCode({
  alpha ~ dgamma(1, 1)

  for (k in 1:(K - 1)) {
    v[k] ~ dbeta(1, alpha)
  }
  v[K] <- 1

  w[1] <- v[1]
  for (k in 2:K) {
    w[k] <- v[k] * prod(1 - v[1:(k - 1)])
  }

  for (k in 1:K) {
    mu[k] ~ dnorm(0, sd = 5)
    sigma[k] ~ dunif(0.1, 5)
  }

  for (i in 1:N) {
    loglik[i] <- dNormMix(y[i], w[1:K], mu[1:K], sigma[1:K], log = 1)
    zeros[i] ~ dpois(-loglik[i] + C)
  }
})
```

## Compile and run (manual)

```{r, cache=TRUE, results='hide'}
Rmodel_sb <- nimble::nimbleModel(
  code = code_sb,
  constants = constants_sb,
  data = data_sb,
  inits = inits_sb()
)

conf_sb <- nimble::configureMCMC(
  Rmodel_sb,
  monitors = c("alpha", "v", "w", "mu", "sigma")
)

Rmcmc_sb <- nimble::buildMCMC(conf_sb)
Cmodel_sb <- nimble::compileNimble(Rmodel_sb, showCompilerOutput = FALSE)
Cmcmc_sb <- nimble::compileNimble(Rmcmc_sb, project = Rmodel_sb, showCompilerOutput = FALSE)

samples_sb <- nimble::runMCMC(
  Cmcmc_sb,
  niter = mcmc$niter,
  nburnin = mcmc$nburnin,
  thin = mcmc$thin,
  nchains = mcmc$nchains,
  setSeed = mcmc$seed
)
```

## Sample extraction

```{r}
head(samples_sb[, 1:min(6, ncol(samples_sb))])
summary(samples_sb[, c("alpha", "mu[1]", "mu[2]", "mu[3]")])
```

## Optional: S3 methods via the DPmixGPD bundle

For comparison, we build a **bundle** for the same SB Normal mixture and use
DPmixGPD's S3 methods for posterior inference.

```{r, cache=TRUE, results='hide'}
bundle_sb <- build_nimble_bundle(
  y = y,
  X = NULL,
  backend = "sb",
  kernel = "normal",
  GPD = FALSE,
  components = K,
  mcmc = mcmc
)

fit_sb <- run_mcmc_bundle_manual(bundle_sb, show_progress = FALSE)
```

```{r, results='asis'}
print(bundle_sb)
summary(bundle_sb)
```

```{r, results='asis'}
print(fit_sb)
summary(fit_sb)
```

```{r}
if (interactive()) plot(fit_sb, family = "traceplot")
```

```{r}
pred_sb <- predict(fit_sb, type = "mean", interval = "credible")
head(pred_sb$fit)
```

# SB + GPD Case: Manual Pipeline with Codegen

This example uses **SB + Amoroso + GPD**, but we still walk through the
manual NIMBLE steps (constants, inits, code, compile, run). The generated
NIMBLE code calls DPmixGPD's GPD kernels under the hood.

## Specification

```{r}
param_specs_crp <- list(
  bulk = list(
    loc = list(
      mode = "dist",
      dist = "normal",
      args = list(mean = 0, sd = 2)
    ),
    scale = list(
      mode = "dist",
      dist = "gamma",
      args = list(shape = 2, rate = 1)
    ),
    shape1 = list(mode = "fixed", value = 1)
  ),
  gpd = list(
    threshold = list(
      mode = "dist",
      dist = "gamma",
      args = list(shape = 2, rate = 1)
    ),
    tail_scale = list(
      mode = "dist",
      dist = "gamma",
      args = list(shape = 2, rate = 1)
    )
  )
)
```

## Build bundle and extract pieces

```{r}
X_crp <- NULL
bundle_crp <- build_nimble_bundle(
  y = y,
  X = X_crp,
  backend = "sb",
  kernel = "amoroso",
  GPD = TRUE,
  components = 5,
  param_specs = param_specs_crp,
  mcmc = mcmc
)

code_crp <- bundle_crp$code$nimble
constants_crp <- bundle_crp$constants
data_crp <- bundle_crp$data
inits_crp <- if (is.function(bundle_crp$inits_fun)) bundle_crp$inits_fun else bundle_crp$inits
```

## Compile and run (manual)

```{r, cache=TRUE, results='hide'}
Rmodel_crp <- nimble::nimbleModel(
  code = code_crp,
  constants = constants_crp,
  data = data_crp,
  inits = if (is.function(inits_crp)) inits_crp() else inits_crp
)

conf_crp <- nimble::configureMCMC(
  Rmodel_crp,
  monitors = bundle_crp$monitors
)

if (!is.null(conf_crp$samplerConfs) && length(conf_crp$samplerConfs) > 0) {
  for (i in seq_along(conf_crp$samplerConfs)) {
    ctl <- conf_crp$samplerConfs[[i]]$control
    if (is.null(ctl)) ctl <- list()
    if (is.null(ctl$checkConjugacy)) ctl$checkConjugacy <- FALSE
    conf_crp$samplerConfs[[i]]$control <- ctl
  }
}

Rmcmc_crp <- nimble::buildMCMC(conf_crp)
Cmodel_crp <- nimble::compileNimble(Rmodel_crp, showCompilerOutput = FALSE)
Cmcmc_crp <- nimble::compileNimble(Rmcmc_crp, project = Rmodel_crp, showCompilerOutput = FALSE)

samples_crp <- nimble::runMCMC(
  Cmcmc_crp,
  niter = mcmc$niter,
  nburnin = mcmc$nburnin,
  thin = mcmc$thin,
  nchains = mcmc$nchains,
  setSeed = mcmc$seed
)
```

## Sample extraction

```{r}
head(samples_crp[, 1:min(6, ncol(samples_crp))])
```

## S3 methods for posterior inference

```{r, cache=TRUE, results='hide'}
fit_crp <- run_mcmc_bundle_manual(bundle_crp, show_progress = FALSE)
```

```{r, results='asis'}
print(fit_crp)
summary(fit_crp)
```

```{r}
if (interactive()) plot(fit_crp, family = "traceplot")
```

```{r}
pred_crp_mean <- predict(fit_crp, type = "mean", interval = "credible")
pred_crp_q90 <- predict(fit_crp, type = "quantile", index = 0.90, interval = "credible")
head(pred_crp_mean$fit)
head(pred_crp_q90$fit)
```

# Causal Model: From Bundle to S3 Inference

We build a causal model using SB backends for treated and control arms. This
uses the same fast MCMC settings and then demonstrates `ate()` and `qte()`.

```{r}
X_causal <- df[, c("wt", "hp", "qsec", "cyl")]
X_causal <- as.data.frame(X_causal)
T_ind <- df$am

param_specs_causal <- list(
  trt = list(
    bulk = list(
      mean = list(
        mode = "link",
        link = "identity",
        beta_prior = list(dist = "normal", args = list(mean = 0, sd = 2))
      ),
      sd = list(
        mode = "dist",
        dist = "invgamma",
        args = list(shape = 2, scale = 1)
      )
    )
  ),
  con = list(
    bulk = list(
      location = list(
        mode = "link",
        link = "identity",
        beta_prior = list(dist = "normal", args = list(mean = 0, sd = 2))
      ),
      scale = list(
        mode = "dist",
        dist = "gamma",
        args = list(shape = 2, rate = 1)
      )
    )
  )
)
```

```{r, cache=TRUE, results='hide'}
causal_bundle <- build_causal_bundle(
  y = y,
  X = X_causal,
  T = T_ind,
  backend = "sb",
  kernel = c("normal", "laplace"),
  GPD = FALSE,
  components = 5,
  param_specs = param_specs_causal,
  PS = "logit",
  mcmc_outcome = mcmc,
  mcmc_ps = mcmc
)

causal_fit <- run_mcmc_causal(causal_bundle, show_progress = FALSE)
```

```{r, results='asis'}
print(causal_bundle)
summary(causal_bundle)
```

```{r, results='asis'}
print(causal_fit)
summary(causal_fit)
```

```{r}
if (interactive()) plot(causal_fit, family = "traceplot")
```

```{r}
pred_causal_mean <- predict(causal_fit, x = X_causal, type = "mean", interval = "credible")
head(pred_causal_mean)
```

```{r, results='asis'}
ate_result <- ate(causal_fit, interval = "credible", nsim_mean = 50)
qte_result <- qte(causal_fit, probs = c(0.25, 0.5, 0.75), interval = "credible")

print(ate_result)
summary(ate_result)

if (interactive()) plot(ate_result, type = "effect")
if (interactive()) plot(qte_result, type = "effect")
```

```{r sessioninfo, include=FALSE}
sessionInfo()
```


